### 2019-09-16 paper-bedpca.tex
init

%% LyX 1.3 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english, 12pt]{article}
\usepackage{times}
%\usepackage{algorithm2e}
\usepackage{url}
\usepackage{bbm}
\usepackage[T1]{fontenc}
\usepackage[latin1]{inputenc}
\usepackage{geometry}
\geometry{verbose,letterpaper,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm}
\usepackage{rotating}
\usepackage{color}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{setspace}
\usepackage{lineno}
\usepackage{hyperref}
\usepackage{bbm}

%\usepackage{xr}
%\externaldocument{SCT-supp}

%\linenumbers
%\doublespacing
\onehalfspacing
%\usepackage[authoryear]{natbib}
\usepackage{natbib} \bibpunct{(}{)}{;}{author-year}{}{,}

%Pour les rajouts
\usepackage{color}
\definecolor{trustcolor}{rgb}{0,0,1}

\usepackage{dsfont}
\usepackage[warn]{textcomp}
\usepackage{adjustbox}
\usepackage{multirow}
\usepackage{graphicx}
\graphicspath{{../figures/}}
\DeclareMathOperator*{\argmin}{\arg\!\min}

\let\tabbeg\tabular
\let\tabend\endtabular
\renewenvironment{tabular}{\begin{adjustbox}{max width=0.9\textwidth}\tabbeg}{\tabend\end{adjustbox}}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% Bold symbol macro for standard LaTeX users
%\newcommand{\boldsymbol}[1]{\mbox{\boldmath $#1$}}

%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}

\usepackage{babel}
\makeatother

\begin{document}

\title{Principal Component Analysis of Population Genetic Data}
\author{Florian Priv\'e,$^{\text{1,2,}*}$ Michael G.B. Blum,$^{\text{1,}*}$ and Bjarni J. Vilhj\'almsson$^{\text{2}}$}

\date{~ }
\maketitle

\noindent$^{\text{\sf 1}}$Laboratoire TIMC-IMAG, UMR 5525, Univ.\ Grenoble Alpes, CNRS, La Tronche, France, \\
\noindent$^{\text{\sf 2}}$National Centre for Register-based Research (NCRR), Aarhus University, Denmark. \\

\noindent$^\ast$To whom correspondence should be addressed.\\

\noindent Contacts:
\begin{itemize}
\item \url{florian.prive.21@gmail.com}
\item \url{michael.blum@univ-grenoble-alpes.fr}
\item \url{bjv@econ.au.dk}
\end{itemize}

\newpage

\abstract{

}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage

\section{Introduction}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Material and Methods}

\subsection{PCA with missing values}

When there is no missing value, we compute the partial Singular Value Decomposition (SVD) $U \Delta V^T$ of the scaled genotype matrix $\tilde{G}_{i,j} = \frac{G_{i,j} - 2 f_j}{\sqrt{2 f_j (1 - f_j)}}$ where $G_{i,j}$ is the allele count (genotype) of individual $i$ and variant $j$, and $f_j$ is the allele frequency of variant $j$ ($2 f_j$ is the mean allele count of variant $j$). Then, $U \Delta$ are the first $K$ PC scores and $V$ are the first $K$ PC loadings, where $K$ is the number of PCs computed (e.g.\ $K=20$).

When there is some missing values, we effectively compute the partial SVD of the following matrix: $\breve{G}_{i,j} = \frac{G_{i,j} - 2 f_j}{\sqrt{2 f_j (1 - f_j) ~ p_i^{(1)} p_j^{(2)}}}$, where missing values are replaced by the mean (i.e.\ $G_{i,j} - 2 f_j = 0$ if $G_{i,j}$ is missing) and where the proportion of missing values of individual $i$, $p_i^{(1)}$, and variant $j$, $p_j^{(2)}$, are accounting for the shrinkage due to mean imputation [REWORD?].
Note that this decomposition is equivalent to the decomposition presented above when there is no missing value.

\subsection{Projecting PCs of the 1000 Genomes}

We provide a subsetted version of the 1000 genomes (1000G) project data \cite[]{10002015global,meyer2019genotype}.
Variants are restricted to the ones in common with HapMap3 and UK Biobank \cite[]{international2010integrating,bycroft2017genome}. 
Moreover, we apply some quality control filters; we remove variants having a minor allele frequency < 0.01, variants or individuals with more than 10\% missing value, variants with P-value of the Hardy-Weinberg exact test < $10^{-50}$, and non-autosomal variants. 
To remove related individuals with first-degree relationship or more, we apply KING cutoff of 0.177 to the data \cite[]{manichaikul2010robust,chang2015second}.
This results in 2493 individuals of the 1000G project (phase 3) in PLINK bed/bim/fam format. 
Resulting PLINK files and R code to generate these files are available at \url{https://doi.org/10.6084/m9.figshare.9208979.v2}. 
To easily download these data, we provide a function called \texttt{download\_1000G} in R package bigsnpr.

There are [X] steps to project PCs of a reference dataset (e.g.\ 1000G) to a target genotype data:
\begin{enumerate}
\item matching the variants of each dataset, including removing ambiguous alleles [A/T] and [C/G], and matching strand and direction of the alleles;
\item computing PCA of the reference dataset using the matched variants only;
\item projecting computed PCs to the target data;
\item possibly adjusting projected PCs for shrinkage;
\item[5.] [ANYTHING ELSE?].
\end{enumerate}
We implement these steps in function \texttt{bed\_projectPCA} of package bigsnpr.

For step 4, we implement the shrinkage-bias adjustment proposed by \cite{dey2019asymptotic}.
This requires computing all eigenvalues of the reference dataset, which is relatively fast to compute for the 2493 individuals of the 1000G. However, it can be computionally expensive if the reference data used includes many more individuals [BETWEEN O($N^2$) AND O($N^3$); MENTION IT?].

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Results}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Discussion}

\subsection{Conclusion}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\newpage

\section*{Funding}

\section*{Acknowledgements}

This research has been conducted using the UK Biobank Resource under Application Number 25589.

~\\ \textit{Conflict of Interest:} none declared.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\newpage

\bibliographystyle{natbib}
\bibliography{refs}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\section*{Supplementary Materials}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\clearpage

\vspace*{1em}

\begin{figure}[!htpb]
%\centerline{\includegraphics[width=0.8\textwidth]{AUC-simu2.pdf}}
\caption{}
\label{fig:AUC-simu2}
\end{figure}

\vspace*{1em}

% latex table generated in R 3.5.2 by xtable 1.8-4 package
% Wed Jun 19 15:19:38 2019
\begin{table}[!htpb]
\caption{AUC values on the test set for simulations with less well imputed variants (mean [95\% CI] from $10^4$ bootstrap samples). These results are plotted in figure \ref{fig:AUC-simu2}.\label{tab:AUC-simu2}}
\vspace*{0.5em}
\centering
\begin{tabular}{|l|c|c|c|c|}
\hline
Scenario & stdCT & maxCT & SCT & lassosum \\ 
\hline
100  & 77.4 [76.0-78.8] & 83.9 [83.4-84.4] & 83.1 [82.6-83.6] & 80.1 [79.5-80.8] \\ 
10K  & 69.4 [68.4-70.5] & 73.0 [72.5-73.4] & 72.9 [72.5-73.3] & 71.2 [70.6-71.7] \\ 
1M   & 64.0 [63.6-64.4] & 64.0 [63.6-64.4] & 62.7 [62.3-63.0] & 64.1 [63.3-64.8] \\ 
2chr & 70.0 [68.8-71.2] & 74.4 [73.6-75.2] & 78.5 [77.9-79.1] & 73.2 [72.5-73.8] \\ 
err  & 67.0 [66.0-68.1] & 68.6 [67.7-69.5] & 69.5 [68.9-70.1] & 65.6 [64.9-66.3] \\ 
HLA  & 74.8 [72.9-76.3] & 75.3 [73.5-76.9] & 76.4 [74.5-78.0] & 75.8 [74.2-77.2] \\ 
\hline
\end{tabular}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2019-09-26 paper-bedpca.tex
save

@@
 
-When there is some missing values, we effectively compute the partial SVD of the following matrix: $\breve{G}_{i,j} = \frac{G_{i,j} - 2 f_j}{\sqrt{2 f_j (1 - f_j) ~ p_i^{(1)} p_j^{(2)}}}$, where missing values are replaced by the mean (i.e.\ $G_{i,j} - 2 f_j = 0$ if $G_{i,j}$ is missing) and where the proportion of missing values of individual $i$, $p_i^{(1)}$, and variant $j$, $p_j^{(2)}$, are accounting for the shrinkage due to mean imputation [REWORD?].
+When there is some missing values, we effectively compute the partial SVD of the following matrix: $\breve{G}_{i,j} = \frac{G_{i,j} - 2 f_j}{\sqrt{2 f_j (1 - f_j) ~ p_i^{(1)} p_j^{(2)}}}$ [VERIFY THAT THIS IS CENTERED], where missing values are replaced by the mean (i.e.\ $G_{i,j} - 2 f_j = 0$ if $G_{i,j}$ is missing) and where the proportion of missing values of individual $i$, $p_i^{(1)}$, and variant $j$, $p_j^{(2)}$, are accounting for the shrinkage due to mean imputation [REWORD?].
 Note that this decomposition is equivalent to the decomposition presented above when there is no missing value.
 
-\subsection{Projecting PCs of the 1000 Genomes}
+\subsection{Outlier detection in PCA}
+
+We propose [BOF] different outlier detection techniques for PCA.
+
+For detecting outlier variants in PCA that are due to long-range Linkage Disequilibrium (LD) regions, we use the same procedure as described by \cite{prive2017efficient}: we compute pcadapt statistics \cite[]{luu2017pcadapt} that summarises the contribution of each variant in all PC loadings.
+This captures both population-differentiating variants as well as long-range LD regions.
+We then smooth the statistics to mostly capture consecutive outliers that corresponds to long-range LD regions.
+
+For detecting outlier samples in PCA, we use the Local Outlier Factor (LOF) on PCs \cite[]{breunig2000lof}.
+To make it robust to the choice of the number of nearest neighbours (kNN) and to the number of PCs used, we take the maximum value of the statistic computed with kNN $\in$ \{4, 10, 30\} and varying the number of PCs from 2 to the number of available PCs.
+We make use of the fast K nearest neighbours implementation of R package nabor to implement the LOF statistic efficiently \cite[]{elseberg2012comparison}.
+
+For detecting samples that have a different ancestry from most of the samples in the data, we compute the pairwise orthogonalized Gnanadesikan-Kettenrin robust Mahalanobis distance of PCs \cite[]{gnanadesikan1972robust,maronna2002robust}, and do not include individuals whose log-distance is larger than some threshold determined based on visual inspection [DO IT AUTOMATIC?].
+
+Finally, to choose the threshold of outlierness based on the previously described statistics, we use a modified version of Tukey's rule, a standard rule for detecting outliers \cite[]{tukey77}. 
+The standard upper limit defined by Tukey's rule is $q_{75\%}(x) + 1.5 \cdot IQR(x)$, where $x$ is the vector of computed statistics and $IQR(x) = q_{75\%}(x) - q_{25\%}(x)$ is the interquartile range.
+However, there are two pitfalls when using Tukey's rule.
+First, Tukey's rule assumes a normally distributed sample.
+For example, when the data is skewed, it does not work that well.
+We account for skewness using the medcouple \cite[]{hubert2008adjusted}.
+Standard Tukey's rule also uses a fixed coefficient (1.5) that does not account for multiple testing, which means that for large samples, you will almost always get some outliers if using 1.5.
+To solve these two issues, we implement \texttt{tukey\_mc\_up} in R package bigutilsr and use it here, which accounts both for skewness and multiple testing by default.
+
+\subsection{Projecting PCs from a reference dataset}
 
@@
 
-There are [X] steps to project PCs of a reference dataset (e.g.\ 1000G) to a target genotype data:
+There are 4 steps to project PCs of a reference dataset (e.g.\ 1000G) to a target genotype data:
 \begin{enumerate}
@@
 \item projecting computed PCs to the target data;
-\item possibly adjusting projected PCs for shrinkage;
-\item[5.] [ANYTHING ELSE?].
+\item adjusting projected PCs for shrinkage.
 \end{enumerate}
@@
 
-For step 4, we implement the shrinkage-bias adjustment proposed by \cite{dey2019asymptotic}.
-This requires computing all eigenvalues of the reference dataset, which is relatively fast to compute for the 2493 individuals of the 1000G.
+For step 4, we implement a shrinkage-bias adjustment method that combines the two methods proposed by \cite{zhang2019fast}.
+First, for only a subset of target individuals, we compute the Online Augmentation, Decomposition, and Procrustes (OADP) transformation (using $K''=K'=K$).
+For these individuals, we regress the new PC scores resulting from OADP by the projected PC scores without adjustement; this gives us shrinkage-adjustment factors that we use to correct projected PC scores for all target individuals.
+
+%This requires computing all eigenvalues of the reference dataset, which is relatively fast to compute for the 2493 individuals of the 1000G.
 However, it can be computionally expensive if the reference data used includes many more individuals [BETWEEN O($N^2$) AND O($N^3$); MENTION IT?].
@@
 \section{Discussion}
+
+[LIMITATION: RUN MULTIPLE TIMES / FAIL TO SPEED UP]
 

### 2019-09-27 paper-bedpca.tex
add intro

@@
 \title{Principal Component Analysis of Population Genetic Data}
-\author{Florian Priv\'e,$^{\text{1,2,}*}$ Michael G.B. Blum,$^{\text{1,}*}$ and Bjarni J. Vilhj\'almsson$^{\text{2}}$}
+\author{Florian Priv\'e,$^{\text{1,2,}*}$ Keurcien Luu, Clara Albi\~nana Climent,$^{\text{2}}$ Michael G.B. Blum,$^{\text{1,}*}$ and Bjarni J. Vilhj\'almsson$^{\text{2}}$}
 
@@
 \item \url{florian.prive.21@gmail.com}
+\item \url{keurcien.luu@gmail.com}
+\item \url{clara@econ.au.dk}
 \item \url{michael.blum@univ-grenoble-alpes.fr}
@@
 \section{Introduction}
+
+Principal Component Analysis (PCA) has been widely used in genetics has been used for many years and in many contexts such as e.g.\ to adjust for population structure in Genome-Wide Association Studies (GWAS) by adding PCs as covariates \cite[]{price2006principal} and to detect loci under selection based on PCA \cite[]{galinsky2016fast,luu2017pcadapt}.
+Much work has been devoted to developing scalable algorithms to compute PCA on data as large as the UK Biobank \cite[]{bycroft2017genome}, which is now possible thanks to software such as FastPCA (fast mode), FlashPCA2, PLINK 2.0 (approx mode), bigstatsr/bigsnpr and TeraPCA \cite[]{galinsky2016fast,abraham2017flashpca2,chang2015second,prive2017efficient,bose2019terapca}.
+
+However, some pitfalls related to PCA of genotype data has been documented and we recall them here.
+This includes some possible lack of precision of computed PCs of software such as FastPCA and PLINK 2.0, which is not an issue for FlashPCA2 and bigstatsr/bigsnpr \cite[]{abraham2017flashpca2,prive2017efficient}.
+This also includes capturing Linkage Disequilibrium (LD) structure in PCA instead of population structure only \cite[]{price2008long,abdellaoui2013population,prive2017efficient}.
+This leads to e.g.\ an over-correction of GWAS variants within long-range LD regions [REWORD/PRECISE].
+Finally, another issue may arise when projecting PCs of a reference dataset to another study dataset: projected PCs are shrunk towards 0 in the new dataset \cite[]{lee2010convergence,wang2015improved,zhang2019fast}.
+This shrinkage makes it dangerous to use the projected PCs for analysis such as PC regression, ancestry detection and GWAS [REWORD/PRECISE].
+
+In this paper, we derive a new implementation of PCA that can be used directly on bed/bim/fam PLINK files with some missing values, which is available in R package bigsnpr v1.0.0.
+Then, we recall some of the pitfalls of PCA computation on genotype data and provide efficient solutions to these issues, such as accouting for LD structure in PCA, detecting outliers [NOT INTRODUCED BEFORE] and projecting PCs to another study sample without shrinkage bias.
 

### 2019-09-28 paper-bedpca.tex
simple NA

@@
 
-When there is no missing value, we compute the partial Singular Value Decomposition (SVD) $U \Delta V^T$ of the scaled genotype matrix $\tilde{G}_{i,j} = \frac{G_{i,j} - 2 f_j}{\sqrt{2 f_j (1 - f_j)}}$ where $G_{i,j}$ is the allele count (genotype) of individual $i$ and variant $j$, and $f_j$ is the allele frequency of variant $j$ ($2 f_j$ is the mean allele count of variant $j$).
+When there is no missing value, we compute the partial Singular Value Decomposition (SVD) $U \Delta V^T$ of the scaled genotype matrix $\tilde{G}_{i,j} = \frac{G_{i,j} - 2 \hat{f_j}}{\sqrt{2 \hat{f_j} (1 - \hat{f_j})}}$ where $G_{i,j}$ is the allele count (genotype) of individual $i$ and variant $j$, and $\hat{f_j}$ is the estimated allele frequency of variant $j$ ($2 \hat{f_j}$ is the mean allele count of variant $j$).
 Then, $U \Delta$ are the first $K$ PC scores and $V$ are the first $K$ PC loadings, where $K$ is the number of PCs computed (e.g.\ $K=20$).
 
-When there is some missing values, we effectively compute the partial SVD of the following matrix: $\breve{G}_{i,j} = \frac{G_{i,j} - 2 f_j}{\sqrt{2 f_j (1 - f_j) ~ p_i^{(1)} p_j^{(2)}}}$ [VERIFY THAT THIS IS CENTERED], where missing values are replaced by the mean (i.e.\ $G_{i,j} - 2 f_j = 0$ if $G_{i,j}$ is missing) and where the proportion of missing values of individual $i$, $p_i^{(1)}$, and variant $j$, $p_j^{(2)}$, are accounting for the shrinkage due to mean imputation [REWORD?].
-Note that this decomposition is equivalent to the decomposition presented above when there is no missing value.
+When there is some missing values, we compute the partial SVD similarly, except that missing values are replaced by the variant means (i.e.\ $G_{i,j} - 2 \hat{f_j} = 0$ when $G_{i,j}$ is missing) and  the $\hat{f_j}$ for each variant are estimated using only non missing values of each variant.
+Note that this decomposition is equivalent to the decomposition presented above after imputation by the variant means [SURE?].
 
@@
 We account for skewness using the medcouple \cite[]{hubert2008adjusted}.
-Standard Tukey's rule also uses a fixed coefficient (1.5) that does not account for multiple testing, which means that for large samples, you will almost always get some outliers if using 1.5.
+Standard Tukey's rule also uses a fixed coefficient (1.5) that does not account for multiple testing, which means that for large samples, there are always some outliers detected when using 1.5.
 To solve these two issues, we implement \texttt{tukey\_mc\_up} in R package bigutilsr and use it here, which accounts both for skewness and multiple testing by default.
@@
 Moreover, we apply some quality control filters; we remove variants having a minor allele frequency < 0.01, variants or individuals with more than 10\% missing value, variants with P-value of the Hardy-Weinberg exact test < $10^{-50}$, and non-autosomal variants.
-To remove related individuals with first-degree relationship or more, we apply KING cutoff of 0.177 to the data \cite[]{manichaikul2010robust,chang2015second}.
+To remove related individuals with first-degree relationship or more, we apply KING cutoff of 0.177 [BE MORE STRINGENT?] to the data using PLINK 2.0 \cite[]{manichaikul2010robust,chang2015second}.
 This results in 2493 individuals of the 1000G project (phase 3) in PLINK bed/bim/fam format. 
@@
 For step 4, we implement a shrinkage-bias adjustment method that combines the two methods proposed by \cite{zhang2019fast}.
-First, for only a subset of target individuals, we compute the Online Augmentation, Decomposition, and Procrustes (OADP) transformation (using $K''=K'=K$).
+First, for only a subset of target individuals, we compute the Online Augmentation, Decomposition, and Procrustes (OADP) transformation (using $K''=K'=K$ [REMOVE?]).
 For these individuals, we regress the new PC scores resulting from OADP by the projected PC scores without adjustement; this gives us shrinkage-adjustment factors that we use to correct projected PC scores for all target individuals.

### 2019-10-01 paper-bedpca.tex
save

@@
 
-For detecting outlier variants in PCA that are due to long-range Linkage Disequilibrium (LD) regions, we use the same procedure as described by \cite{prive2017efficient}: we compute pcadapt statistics \cite[]{luu2017pcadapt} that summarises the contribution of each variant in all PC loadings.
+For detecting outlier variants in PCA that are due to long-range Linkage Disequilibrium (LD) regions, we use the same procedure as described by \cite{prive2017efficient}: we compute pcadapt statistics \cite[]{luu2017pcadapt} that summarises the contribution of each variant in all PC loadings computed.
 This captures both population-differentiating variants as well as long-range LD regions.
-We then smooth the statistics to mostly capture consecutive outliers that corresponds to long-range LD regions.
+We then smooth the statistics [BE MORE PRECISE?] to mostly capture consecutive outliers that corresponds to long-range LD regions.
 
 For detecting outlier samples in PCA, we use the Local Outlier Factor (LOF) on PCs \cite[]{breunig2000lof}.
-To make it robust to the choice of the number of nearest neighbours (kNN) and to the number of PCs used, we take the maximum value of the statistic computed with kNN $\in$ \{4, 10, 30\} and varying the number of PCs from 2 to the number of available PCs.
-We make use of the fast K nearest neighbours implementation of R package nabor to implement the LOF statistic efficiently \cite[]{elseberg2012comparison}.
+To make it robust to the choice of the number of nearest neighbours (kNN), we take the maximum value of the statistic computed with kNN $\in$ \{4, 5, 6, 7, 8\}.
+We do not use larger values for `kNN' because it would remove small ancestry groups (e.g.\ composed of 10-15 people only).
+We make use of the fast K nearest neighbours implementation of R package nabor to implement the LOF statistic efficiently in our new R package bigutilsr \cite[]{elseberg2012comparison}.
 
@@
 Moreover, we apply some quality control filters; we remove variants having a minor allele frequency < 0.01, variants or individuals with more than 10\% missing value, variants with P-value of the Hardy-Weinberg exact test < $10^{-50}$, and non-autosomal variants.
-To remove related individuals with first-degree relationship or more, we apply KING cutoff of 0.177 [BE MORE STRINGENT?] to the data using PLINK 2.0 \cite[]{manichaikul2010robust,chang2015second}.
-This results in 2493 individuals of the 1000G project (phase 3) in PLINK bed/bim/fam format. 
-Resulting PLINK files and R code to generate these files are available at \url{https://doi.org/10.6084/m9.figshare.9208979.v2}. 
+To remove related individuals with second-degree relationship or more, we apply KING cutoff of 0.0884 to the data using PLINK 2.0 \cite[]{manichaikul2010robust,chang2015second}.
+This results in 2490 individuals of the 1000G project (phase 3) in PLINK bed/bim/fam format. 
+Resulting PLINK files and R code to generate these files are available at \url{https://doi.org/10.6084/m9.figshare.9208979.v3}. 
 To easily download these data, we provide a function called \texttt{download\_1000G} in R package bigsnpr.

### 2019-10-01 paper-bedpca.tex
add opti OADP

@@
 
-\subsection{Projecting PCs from a reference dataset}
+\subsection{Subsetted version of 1000G data [RENAME?]}
 
@@
 
-There are 4 steps to project PCs of a reference dataset (e.g.\ 1000G) to a target genotype data:
+\subsection{Projecting PCs from a reference dataset}
+
+To project PCs of a reference dataset (e.g.\ 1000G) to a target genotype data, we implement the following 3 steps in function \texttt{bed\_projectPCA} of package bigsnpr:
 \begin{enumerate}
@@
 \item computing PCA of the reference dataset using the matched variants only;
-\item projecting computed PCs to the target data;
-\item adjusting projected PCs for shrinkage.
+\item projecting computed PCs to the target data using an optimised implementation of the Online Augmentation, Decomposition, and Procrustes (OADP) transformation \cite[]{zhang2019fast} described in Supplementary Materials.
 \end{enumerate}
-We implement these steps in function \texttt{bed\_projectPCA} of package bigsnpr.
 
-For step 4, we implement a shrinkage-bias adjustment method that combines the two methods proposed by \cite{zhang2019fast}.
+%For step 4, we implement a shrinkage-bias adjustment method that combines the two methods proposed by \cite{zhang2019fast}.
 First, for only a subset of target individuals, we compute the Online Augmentation, Decomposition, and Procrustes (OADP) transformation (using $K''=K'=K$ [REMOVE?]).
@@
 [LIMITATION: RUN MULTIPLE TIMES / FAIL TO SPEED UP]
+
+\noindent[DISCUSS ABOUT $K''=K'=K$? FASTER, EASIER, DO NOT WANT MORE DIM?]
 
@@
 \section*{Supplementary Materials}
+
+\subsection*{Optimised OADP transformation}
+
+We implement an optimised version of the Online Augmentation, Decomposition, and Procrustes (OADP) transformation when using $K''=K'=K$ \cite[]{zhang2019fast}.
+We assume that the $K$-partial Singular Value Decomposition (SVD) of the reference matrix $X$ (of size $n \times p$) has been computed as $U \Delta V^T$.
+There are several steps to perform OADP transformation for each sample $y$ (of size $1 \times p$) of target matrix $Y$ (of size $m \times p$):
+\begin{enumerate}
+\item Calculate $l = y \cdot V$ (of size $1 \times K$), where $V$ are the K PC loadings.
+And $g = y \cdot h^T$ (of size $1 \times 1$), where $h = (y - l \cdot V^T) / ||y - l \cdot V^T||_2$.
+Actually, $||y - l \cdot V^T||_2^2 = y \cdot y^T - 2 \cdot y \cdot V \cdot l^T  + l \cdot V^T \cdot V \cdot l^T = y \cdot y^T - l \cdot l^T$ and $y \cdot (y - l \cdot V^T)^T = y \cdot y^T - y \cdot V \cdot l^T = y \cdot y^T - l \cdot l^T$.
+Then $g = \sqrt{y \cdot y^T - l \cdot l^T}$.
+\item Calculate $Q^T Q$ where $$Q = \begin{bmatrix} \Delta & l^T \\ 0 & g \end{bmatrix}$$ so that $$Q^TQ = \begin{bmatrix} \Delta^2 & \Delta \cdot l^T \\ l \cdot \Delta & g^2 + l \cdot l^T \end{bmatrix} = \begin{bmatrix} \Delta^2 & \Delta \cdot l^T \\ l \cdot \Delta & y \cdot y^T \end{bmatrix}.$$ Note that we do not actually need to compute $g$, and that we can update the last row and column of $Q^T Q$ instead of computing it from an updated version of $Q$.
+\item Get the eigen decomposition $Q^T Q = V' {\Delta'}^2 {V'}^T$ (truncated to $K$ components out of the $K+1$).
+\item Calculate $$\begin{bmatrix} \widetilde{U} \\ \widetilde{u}  \end{bmatrix} = \begin{bmatrix} U & 0 \\ 0 & 1 \end{bmatrix} V' = \begin{bmatrix} U V'[1{:}K,~] \\ V'[K{+}1,~] \end{bmatrix} $$
+\item Find the Procrustes transformation from $\widetilde{U}$ to $U$.
+As $\widetilde{U}$ and $U$ have both their columns centered already, the Procrustes transformation $\rho ~\widetilde{U} A$, where $\rho$ is a scaling coefficient and $A$ is an orthonomal projection matrix that minimise the Frobenius norm of $(\rho \widetilde{U} A - U)$, is given by $A = V'' {U''}^T$ and $\rho = \frac{\text{trace}(\Delta'')}{\text{trace}(\widetilde{U}^T \widetilde{U})}$ where $U'' \Delta'' {V''}^T$ is the full SVD of $U^T \widetilde{U} = V'[1{:}K,~]$ \cite[]{wang2015improved}.
+Then, $\rho = \frac{\text{trace}(\Delta'')}{\text{trace}(\widetilde{U}^T \widetilde{U})} = \frac{\text{trace}(\Delta'')}{\text{trace}({V'[1{:}K,~]}^T V'[1{:}K,~])} = \frac{\text{trace}(\Delta'')}{\text{trace}(\Delta''^2)}$.
+\item Apply the previous transformation to $\widetilde{u}$ to get the projection of $y$ in the reference PCA space (i.e.\ $\rho \widetilde{u} A$),.
+\item Finally, multipy by $\Delta$.
+\end{enumerate}
+
+[CAN COMPUTE SVD OF $V'[1{:}K,~]$ FROM LOW-RANK UPDATE?]
 

### 2019-10-06 paper-bedpca.tex
add figures projection 1000G

@@
 \usepackage{graphicx}
-\graphicspath{{../figures/}}
+\graphicspath{{figures/}}
 \DeclareMathOperator*{\argmin}{\arg\!\min}
@@
 
-\subsection{Outlier detection in PCA}
+\subsection{Outlier detection in PCA [SPLIT IN MULTIPLE SECTIONS]}
 
@@
 
-\subsection{Subsetted version of 1000G data [RENAME?]}
+\subsection{Subsetted version of 1000G data [RENAME?]\label{1000G}}
 
@@
 
-\subsection{Projecting PCs from a reference dataset}
+\subsection{Projecting PCs from a reference dataset\label{proj}}
 
@@
 \item computing PCA of the reference dataset using the matched variants only;
-\item projecting computed PCs to the target data using an optimised implementation of the Online Augmentation, Decomposition, and Procrustes (OADP) transformation \cite[]{zhang2019fast} described in Supplementary Materials.
+\item projecting computed PCs to the target data using an optimised implementation of the Online Augmentation, Decomposition, and Procrustes (OADP) transformation \cite[]{zhang2019fast}, which we describe in Supplementary Materials.
 \end{enumerate}
@@
 \section{Results}
+
+\subsection{Projecting PCs from a reference dataset}
+
+We use 60\% of individuals in the 1000G data (section \ref{1000G}) to compute the PCA ($K=20$ PCs).
+Then, we project the remaining 40\% individuals using two methods: 1/ simply multiplying the genotypes of these individuals by the previously computed loadings, and 2/ the OADP projection (section \ref{proj}).
+When simply projecting using loadings, there is no apparent shrinkage for PC1 and PC2, a weak [REWORD?] shrinkage for PC3 and PC4, and a large shrinkage for PC5 to PC8 (Figure \ref{fig:proj1000G}).
+In contrast, there is no apparent shrinkage when projecting new individuals with OADP (Figure \ref{fig:proj1000G}).
+Simple projection is affected even more by this shrinkage for PC9 to PC20, while OADP appears free of this issue (Figure \ref{fig:proj1000G-2}).
+
+\begin{figure}[!htpb]
+\centerline{\includegraphics[width=0.8\textwidth]{proj1000G-PC1-8.pdf}}
+\caption{Principal Component (PC) scores 1 to 8 of the 1000 Genomes project.
+Black points are the 60\% individuals used for computing PCA.
+Red points are the 40\% remaining individuals, projected by simply multiplying their genotypes by the corresponding PC loadings.
+Blue points are the 40\% remaining individuals, projected using the Online Augmentation, Decomposition, and Procrustes (OADP) transformation.
+\label{fig:proj1000G}}
+\end{figure}
 
@@
 
+\renewcommand{\thefigure}{S\arabic{figure}}
+\setcounter{figure}{0}
+\renewcommand{\thetable}{S\arabic{table}}
+\setcounter{table}{0}
+
 \subsection*{Optimised OADP transformation}
@@
 
-[CAN COMPUTE SVD OF $V'[1{:}K,~]$ FROM LOW-RANK UPDATE?]
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{figure}[!htpb]
-%\centerline{\includegraphics[width=0.8\textwidth]{AUC-simu2.pdf}}
-\caption{}
-\label{fig:AUC-simu2}
+\centerline{\includegraphics[width=0.8\textwidth]{proj1000G-PC9-20.pdf}}
+\caption{Principal Component (PC) scores 9 to 20 of the 1000 Genomes project.
+Black points are the 60\% individuals used for computing PCA.
+Red points are the 40\% remaining individuals, projected by simply multiplying their genotypes by the corresponding PC loadings.
+Blue points are the 40\% remaining individuals, projected using the Online Augmentation, Decomposition, and Procrustes (OADP) transformation.
+\label{fig:proj1000G-2}}
 \end{figure}
@@
 \begin{table}[!htpb]
-\caption{AUC values on the test set for simulations with less well imputed variants (mean [95\% CI] from $10^4$ bootstrap samples).
-These results are plotted in figure \ref{fig:AUC-simu2}.\label{tab:AUC-simu2}}
+\caption{\label{tab:AUC-simu2}}
 \vspace*{0.5em}

### 2019-10-06 code/project-1000G.R

@@ -1,0 +1,56 @@
+library(bigsnpr)
+bedfile.ref <- download_1000G("tmp-data")
+bed.ref <- bed(bedfile.ref)
+
+set.seed(1)
+train <- sample(nrow(bed.ref), 0.6 * nrow(bed.ref))
+test <- bed_projectPCA(
+  bed.new = bed.ref,
+  bed.ref = bed.ref,
+  ind.row.new = rows_along(bed.ref)[-train],
+  ind.row.ref = train,
+  k = 20,
+  kNN = 6,
+  ncores = nb_cores()
+)
+
+PC.ref <- predict(test$obj.svd.ref)
+proj1 <- test$simple_proj
+proj2 <- test$OADP_proj
+
+library(ggplot2)
+plot_grid(plotlist = lapply(1:4, function(k) {
+  k1 <- 2 * k - 1
+  k2 <- 2 * k
+  plot_grid(
+    qplot(PC.ref[, k1], PC.ref[, k2], size = I(2)) +
+      geom_point(aes(proj1[, k1], proj1[, k2]), color = "red") +
+      theme_bigstatsr() +
+      labs(x = paste0("PC", k1), y = paste0("PC", k2)),
+    qplot(PC.ref[, k1], PC.ref[, k2], size = I(2)) +
+      geom_point(aes(proj2[, k1], proj2[, k2]), color = "blue") +
+      theme_bigstatsr() +
+      labs(x = paste0("PC", k1), y = paste0("PC", k2)),
+    scale = 0.95
+  )
+}))
+
+# ggsave("figures/proj1000G-PC1-8.pdf", width = 15, height = 7.5)
+
+plot_grid(plotlist = lapply(5:10, function(k) {
+  k1 <- 2 * k - 1
+  k2 <- 2 * k
+  plot_grid(
+    qplot(PC.ref[, k1], PC.ref[, k2], size = I(2)) +
+      geom_point(aes(proj1[, k1], proj1[, k2]), color = "red") +
+      theme_bigstatsr() +
+      labs(x = paste0("PC", k1), y = paste0("PC", k2)),
+    qplot(PC.ref[, k1], PC.ref[, k2], size = I(2)) +
+      geom_point(aes(proj2[, k1], proj2[, k2]), color = "blue") +
+      theme_bigstatsr() +
+      labs(x = paste0("PC", k1), y = paste0("PC", k2)),
+    scale = 0.95
+  )
+}), nrow = 3)
+
[…]

### 2019-10-14 paper-bedpca.tex
save

@@
 
-\title{Principal Component Analysis of Population Genetic Data}
-\author{Florian Priv\'e,$^{\text{1,2,}*}$ Keurcien Luu, Clara Albi\~nana Climent,$^{\text{2}}$ Michael G.B. Blum,$^{\text{1,}*}$ and Bjarni J. Vilhj\'almsson$^{\text{2}}$}
+\title{Efficient toolkit implementing best practices for principal component analysis of population genetic data}
+\author{Florian Priv\'e,$^{\text{1,}*}$ John J. McGrath,$^{\text{1,2,3}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,}*}$}
 
@@
 
-\noindent$^{\text{\sf 1}}$Laboratoire TIMC-IMAG, UMR 5525, Univ.\ Grenoble Alpes, CNRS, La Tronche, France, \\
-\noindent$^{\text{\sf 2}}$National Centre for Register-based Research (NCRR), Aarhus University, Denmark. \\
+\noindent$^{\text{\sf 1}}$National Centre for Register-Based Research, Aarhus University, Aarhus, 8210, Denmark. \\
+\noindent$^{\text{\sf 2}}$Queensland Brain Institute, University of Queensland, St. Lucia, 4072, Queensland, Australia. \\
+\noindent$^{\text{\sf 3}}$Queensland Centre for Mental Health Research, The Park Centre for Mental Health, Wacol, 4076, Queensland, Australia.\\
 
@@
 \item \url{florian.prive.21@gmail.com}
-\item \url{keurcien.luu@gmail.com}
-\item \url{clara@econ.au.dk}
-\item \url{michael.blum@univ-grenoble-alpes.fr}
+\item \url{john_mcgrath@qcmhr.uq.edu.au}
 \item \url{bjv@econ.au.dk}
@@
 \abstract{
+Principal Component Analysis (PCA) of genetic data is routinely used to infer ancestry and control for population structure in various genetic analyses.
+However, conducting PCA in large populations is complicated, and there are several potential pitfalls.
+These pitfalls include (1) capturing Linkage Disequilibrium (LD) structure instead of population structure, (2) projected PCs that suffer from shrinkage bias when projecting PCA from a reference dataset to another independent dataset, and (3) detecting sample outliers.
+In this work, we explore these pitfalls of using PCA, and present efficient solutions to these pitfalls.
+Following applications to diverse datasets, we make recommendations for best practices and implement the proposed solutions in an efficient and user-friendly toolkit for performing these analyses in the R packages bigsnpr and bigutilsr.
 
+For example, we show that PC18 to PC40 in the UK Biobank are capturing LD structure.
+Using our automatic algorithm for removing long-range LD region, we are able to recover only 17 PCs that capture population structure only.
+Therefore, we recommend using only 17 PCs from the UK Biobank.
+Another example: when projecting the PCA computed on the 1000 Genomes project, although PC1 to PC4 suffer from only moderate shrinkage (1.00-1.08), PC5 for example suffers from a shrinkage factor of 1.51 and PC10 from a shrinkage factor of 3.16.
+We provide a fast way to project new individuals that is mostly free of this shrinkage bias.
+We also show how to restrict to individuals that are genetically homogeneous based on PCA.
+
+Overall, we believe this work would be of interest for anyone using PCA in their analyses of genetic data.
 }
@@
 
-To project PCs of a reference dataset (e.g.\ 1000G) to a target genotype data, we implement the following 3 steps in function \texttt{bed\_projectPCA} of package bigsnpr:
-\begin{enumerate}
-\item matching the variants of each dataset, including removing ambiguous alleles [A/T] and [C/G], and matching strand and direction of the alleles;
-\item computing PCA of the reference dataset using the matched variants only;
-\item projecting computed PCs to the target data using an optimised implementation of the Online Augmentation, Decomposition, and Procrustes (OADP) transformation \cite[]{zhang2019fast}, which we describe in Supplementary Materials.
-\end{enumerate}
-
-%For step 4, we implement a shrinkage-bias adjustment method that combines the two methods proposed by \cite{zhang2019fast}.
-First, for only a subset of target individuals, we compute the Online Augmentation, Decomposition, and Procrustes (OADP) transformation (using $K''=K'=K$ [REMOVE?]).
-For these individuals, we regress the new PC scores resulting from OADP by the projected PC scores without adjustement; this gives us shrinkage-adjustment factors that we use to correct projected PC scores for all target individuals.
-
-%This requires computing all eigenvalues of the reference dataset, which is relatively fast to compute for the 2493 individuals of the 1000G.
-However, it can be computionally expensive if the reference data used includes many more individuals [BETWEEN O($N^2$) AND O($N^3$); MENTION IT?].
+To project PCs of a reference dataset (e.g.\ 1000G) to a target genotype dataset, we implement the following 3 steps in function \texttt{bed\_projectPCA} of package bigsnpr: 1) matching the variants of each dataset, including removing ambiguous alleles [A/T] and [C/G], and matching strand and direction of the alleles; 2) computing PCA of the reference dataset using the matched variants only; 3) projecting computed PCs to the target data using an optimised implementation (see Supplementary Materials) of the Online Augmentation, Decomposition, and Procrustes (OADP) transformation \cite[]{zhang2019fast}.
+Note that you can also use function \texttt{bed\_projectPCA} to project individuals from the reference dataset that were not used during the PCA computation (e.g.\ related people).
 
@@
 \item Calculate $Q^T Q$ where $$Q = \begin{bmatrix} \Delta & l^T \\ 0 & g \end{bmatrix}$$ so that $$Q^TQ = \begin{bmatrix} \Delta^2 & \Delta \cdot l^T \\ l \cdot \Delta & g^2 + l \cdot l^T \end{bmatrix} = \begin{bmatrix} \Delta^2 & \Delta \cdot l^T \\ l \cdot \Delta & y \cdot y^T \end{bmatrix}.$$ Note that we do not actually need to compute $g$, and that we can update the last row and column of $Q^T Q$ instead of computing it from an updated version of $Q$.
-\item Get the eigen decomposition $Q^T Q = V' {\Delta'}^2 {V'}^T$ (truncated to $K$ components out of the $K+1$).
-\item Calculate $$\begin{bmatrix} \widetilde{U} \\ \widetilde{u}  \end{bmatrix} = \begin{bmatrix} U & 0 \\ 0 & 1 \end{bmatrix} V' = \begin{bmatrix} U V'[1{:}K,~] \\ V'[K{+}1,~] \end{bmatrix} $$
-\item Find the Procrustes transformation from $\widetilde{U}$ to $U$.
-As $\widetilde{U}$ and $U$ have both their columns centered already, the Procrustes transformation $\rho ~\widetilde{U} A$, where $\rho$ is a scaling coefficient and $A$ is an orthonomal projection matrix that minimise the Frobenius norm of $(\rho \widetilde{U} A - U)$, is given by $A = V'' {U''}^T$ and $\rho = \frac{\text{trace}(\Delta'')}{\text{trace}(\widetilde{U}^T \widetilde{U})}$ where $U'' \Delta'' {V''}^T$ is the full SVD of $U^T \widetilde{U} = V'[1{:}K,~]$ \cite[]{wang2015improved}.
-Then, $\rho = \frac{\text{trace}(\Delta'')}{\text{trace}(\widetilde{U}^T \widetilde{U})} = \frac{\text{trace}(\Delta'')}{\text{trace}({V'[1{:}K,~]}^T V'[1{:}K,~])} = \frac{\text{trace}(\Delta'')}{\text{trace}(\Delta''^2)}$.
-\item Apply the previous transformation to $\widetilde{u}$ to get the projection of $y$ in the reference PCA space (i.e.\ $\rho \widetilde{u} A$),.
-\item Finally, multipy by $\Delta$.
+\item Get the eigen decomposition $Q^T Q = V' {\Delta'}^2 {V'}^T$ (truncated to $K$ components out of the $K+1$). Let us denote $V_2 = V' {\Delta'}$.
+\item Calculate $$\begin{bmatrix} \widetilde{U} \\ \widetilde{u}  \end{bmatrix} = \begin{bmatrix} U & 0 \\ 0 & 1 \end{bmatrix} V_2 = \begin{bmatrix} U V_2[1{:}K,~] \\ V_2[K{+}1,~] \end{bmatrix} $$
+\item Find the Procrustes transformation from $\widetilde{U}$ to $U \Delta$.
+As $\widetilde{U}$ and $U \Delta$ have both their columns centered already (since $U$ does), the Procrustes transformation $\rho \widetilde{U} A$, where $\rho$ is a scaling coefficient and $A$ is an orthonomal projection matrix that minimise the Frobenius norm of $(\rho \widetilde{U} A - U)$, is given by $A = V'' {U''}^T$ and $\rho = \frac{\text{trace}(\Delta'')}{\text{trace}(\widetilde{U}^T \widetilde{U})}$ where $U'' \Delta'' {V''}^T$ is the full SVD of $(U \Delta)^T \widetilde{U}$ \cite[]{wang2015improved}.
+As $U^T U = I$, we note that $(U \Delta)^T \widetilde{U} = \Delta V_2[1{:}K,~]$ and that $\rho = \frac{\text{trace}(\Delta'')}{\text{trace}({V_2[1{:}K,~]}^T V_2[1{:}K,~])}$, therefore we do not need to explicitly compute $\widetilde{U}$.
+\item Apply the previous transformation to $\widetilde{u}$ to get the projection of $y$ in the reference PCA space (i.e.\ $\rho \widetilde{u} A$).
 \end{enumerate}
@@
 \begin{figure}[!htpb]
-\centerline{\includegraphics[width=0.8\textwidth]{proj1000G-PC9-20.pdf}}
+\centerline{\includegraphics[width=0.9\textwidth]{proj1000G-PC9-20.pdf}}
 \caption{Principal Component (PC) scores 9 to 20 of the 1000 Genomes project.
+Black points are the 60\% individuals used for computing PCA.
+Green points are the 40\% remaining individuals, projected by simply multiplying their genotypes by the corresponding PC loadings, corrected using theoritical asymptotic shrinkage factors.
+Blue points are the 40\% remaining individuals, projected using the Online Augmentation, Decomposition, and Procrustes (OADP) transformation.
+\label{fig:proj1000G-2}}
+\end{figure}
+
+\vspace*{1em}
+
+\begin{figure}[!htpb]
+\centerline{\includegraphics[width=0.9\textwidth]{proj1000G-PC5-12.pdf}}
+\caption{Principal Component (PC) scores 5 to 12 of the 1000 Genomes project.
 Black points are the 60\% individuals used for computing PCA.
@@
 Blue points are the 40\% remaining individuals, projected using the Online Augmentation, Decomposition, and Procrustes (OADP) transformation.
-\label{fig:proj1000G-2}}
+\label{fig:proj1000G-3}}
 \end{figure}

### 2019-10-16 paper-bedpca.tex
all proj 1000G

@@
 
+[SEPARATE STEPS?]
+
 To project PCs of a reference dataset (e.g.\ 1000G) to a target genotype dataset, we implement the following 3 steps in function \texttt{bed\_projectPCA} of package bigsnpr: 1) matching the variants of each dataset, including removing ambiguous alleles [A/T] and [C/G], and matching strand and direction of the alleles; 2) computing PCA of the reference dataset using the matched variants only; 3) projecting computed PCs to the target data using an optimised implementation (see Supplementary Materials) of the Online Augmentation, Decomposition, and Procrustes (OADP) transformation \cite[]{zhang2019fast}.
-Note that you can also use function \texttt{bed\_projectPCA} to project individuals from the reference dataset that were not used during the PCA computation (e.g.\ related people).
 
@@
 We use 60\% of individuals in the 1000G data (section \ref{1000G}) to compute the PCA ($K=20$ PCs).
-Then, we project the remaining 40\% individuals using two methods: 1/ simply multiplying the genotypes of these individuals by the previously computed loadings, and 2/ the OADP projection (section \ref{proj}).
+Then, we project the remaining 40\% individuals using three methods: 1/ simply multiplying the genotypes of these individuals by the previously computed loadings, 2/ correcting the simple projections using asymptotic shrinkage factors as determined by R package hdpca v1.0.0 \cite[]{dey2019asymptotic} and 3/ the OADP projection (section \ref{proj}).
 When simply projecting using loadings, there is no apparent shrinkage for PC1 and PC2, a weak [REWORD?] shrinkage for PC3 and PC4, and a large shrinkage for PC5 to PC8 (Figure \ref{fig:proj1000G}).
@@
 Simple projection is affected even more by this shrinkage for PC9 to PC20, while OADP appears free of this issue (Figure \ref{fig:proj1000G-2}).
+When correcting projected PC scores with asymptotic shrinkage factors, bias is smaller than when simply projecting loadings, yet, it is apparent for PC7-8 and PC11-12 (Figure \ref{fig:proj1000G-3}).
+Finally, to assess if OADP could be used to project individuals that are related to some individuals that were used to compute PCA, we projected these individuals (as if we were projecting their monozygotic twins) using OADP.
+Projections of related individuals using OADP shows some reverse [REWORD?] shrinkage (Figure \ref{fig:proj1000G-4}).
 
@@
 [LIMITATION: RUN MULTIPLE TIMES / FAIL TO SPEED UP]
+
+\noindent[LIMITATION 2: FAIL TO PROJECT RELATED INDIVIDUALS WITH OADP.]
+
+\noindent[LIMITATION 3: FAIL TO AUTOMATICALLY DETECT OUTLIER SAMPLES.]
 
@@
 Black points are the 60\% individuals used for computing PCA.
-Green points are the 40\% remaining individuals, projected by simply multiplying their genotypes by the corresponding PC loadings, corrected using theoritical asymptotic shrinkage factors.
+Red points are the 40\% remaining individuals, projected by simply multiplying their genotypes by the corresponding PC loadings.
 Blue points are the 40\% remaining individuals, projected using the Online Augmentation, Decomposition, and Procrustes (OADP) transformation.
@@
 Black points are the 60\% individuals used for computing PCA.
-Red points are the 40\% remaining individuals, projected by simply multiplying their genotypes by the corresponding PC loadings.
+Green points are the 40\% remaining individuals, projected by multiplying their genotypes by the corresponding PC loadings, further corrected using theoritical asymptotic shrinkage factors.
 Blue points are the 40\% remaining individuals, projected using the Online Augmentation, Decomposition, and Procrustes (OADP) transformation.
 \label{fig:proj1000G-3}}
+\end{figure}
+
+\begin{figure}[!htpb]
+\centerline{\includegraphics[width=0.9\textwidth]{proj1000G-related.pdf}}
+\caption{Principal Component (PC) scores 5 to 12 of the 1000 Genomes project.
+Black points are the 60\% individuals used for computing PCA.
+Blue points are the same 60\% individuals, projected using the Online Augmentation, Decomposition, and Procrustes (OADP) transformation.
+\label{fig:proj1000G-4}}
 \end{figure}

### 2019-10-16 code/project-1000G.R

@@ -1,24 +1,31 @@
 library(bigsnpr)
-bedfile.ref <- download_1000G("tmp-data")
-bed.ref <- bed(bedfile.ref)
+library(ggplot2)
 
-set.seed(1)
-train <- sample(nrow(bed.ref), 0.6 * nrow(bed.ref))
-test <- bed_projectPCA(
-  bed.new = bed.ref,
-  bed.ref = bed.ref,
-  ind.row.new = rows_along(bed.ref)[-train],
-  ind.row.ref = train,
-  k = 20,
-  kNN = 6,
-  ncores = nb_cores()
-)
+bedfile.ref <- download_1000G("tmp-data")
+obj.bed <- bed(bedfile.ref)
+
+set.seed(1); train <- sample(nrow(obj.bed), 0.6 * nrow(obj.bed))
+
+# obj.svd <- bed_autoSVD2(obj.bed, ind.row = train, k = 20, ncores = nb_cores())
+# saveRDS(obj.svd, "tmp-data/SVD_1000G.rds")
+obj.svd <- readRDS("tmp-data/SVD_1000G.rds")
+
+test <- bed_projectSelfPCA(obj.svd, obj.bed,
+                           ind.row = rows_along(obj.bed)[-train],
+                           ncores = nb_cores())
 
 PC.ref <- predict(test$obj.svd.ref)
 proj1 <- test$simple_proj
 proj2 <- test$OADP_proj
 
-library(ggplot2)
+# shrinkage coefficients
+shrinkage <- unname(sapply(1:20, function(k) {
+  MASS::rlm(proj2[, k] ~ proj1[, k] + 0)$coef
+}))
+round(shrinkage, 2)
+#  [1] 1.01 1.02 1.07 1.09 1.51 1.68 1.94 1.40 2.88 3.17 2.90 2.92 3.23
+# [14] 5.13 5.25 5.04 4.58 5.69 6.26 6.32
+
 plot_grid(plotlist = lapply(1:4, function(k) {
   k1 <- 2 * k - 1
   k2 <- 2 * k
@@ -54,3 +61,48 @@ plot_grid(plotlist = lapply(5:10, function(k) {
 }), nrow = 3)
 
 # ggsave("figures/proj1000G-PC9-20.pdf", width = 13.5, height = 8)
+
+
+ind.keep <- attr(obj.svd, "subset")
+K <- bed_tcrossprodSelf(obj.bed, ind.row = train, ind.col = ind.keep)
[…]

### 2019-10-16 code/project-UKBB-to-1000G.R
proj UKBB

@@ -1,0 +1,43 @@
+library(bigsnpr)
+library(ggplot2)
+
+obj.bed <- bed("data/ukbb_488282.bed")
+bed.ref <- bed(download_1000G("data"))
+
+system.time(
+  test <- bed_projectPCA(bed.ref, obj.bed, k = 20, ncores = nb_cores())
+) # 20 min (including 12 min for projection)
+
+PC.ref <- predict(test$obj.svd.ref)
+proj1 <- test$simple_proj
+proj2 <- test$OADP_proj
+
+# shrinkage coefficients
+shrinkage <- unname(sapply(1:20, function(k) {
+  MASS::rlm(proj2[, k] ~ proj1[, k] + 0)$coef
+}))
+round(shrinkage, 2)
+#  [1] 1.01 1.02 1.06 1.08 1.35 1.82 2.32 2.35 2.78 2.84 2.98 3.47 4.37
+# [14] 4.64 4.94 5.30 5.74 6.49 6.70 6.74
+
+ind <- sample(nrow(proj1), 10e3)
+
+plot_grid(plotlist = lapply(1:6, function(k) {
+  k1 <- 2 * k - 1
+  k2 <- 2 * k
+  plot_grid(
+    qplot(PC.ref[, k1], PC.ref[, k2], size = I(2)) +
+      geom_point(aes(proj1[ind, k1], proj1[ind, k2]), color = "red") +
+      theme_bigstatsr(0.7) +
+      labs(x = paste0("PC", k1), y = paste0("PC", k2)),
+    qplot(PC.ref[, k1], PC.ref[, k2], size = I(2)) +
+      geom_point(aes(proj2[ind, k1], proj2[ind, k2]), color = "blue") +
+      theme_bigstatsr(0.7) +
+      labs(x = paste0("PC", k1), y = paste0("PC", k2)),
+    scale = 0.95
+  )
+}), nrow = 3)
+
+# ggsave("figures/proj1000G-UKBB.pdf", width = 10.5, height = 6.5)
+
+

### 2019-10-16 code/project-UKBB-to-1000G.R
20K points until PC16

@@ -20,24 +20,24 @@ round(shrinkage, 2)
 #  [1] 1.01 1.02 1.06 1.08 1.35 1.82 2.32 2.35 2.78 2.84 2.98 3.47 4.37
 # [14] 4.64 4.94 5.30 5.74 6.49 6.70 6.74
 
-ind <- sample(nrow(proj1), 10e3)
+ind <- sample(nrow(proj1), 20e3)
 
-plot_grid(plotlist = lapply(1:6, function(k) {
+plot_grid(plotlist = lapply(1:8, function(k) {
   k1 <- 2 * k - 1
   k2 <- 2 * k
   plot_grid(
     qplot(PC.ref[, k1], PC.ref[, k2], size = I(2)) +
       geom_point(aes(proj1[ind, k1], proj1[ind, k2]), color = "red") +
-      theme_bigstatsr(0.7) +
+      theme_bigstatsr(0.5) +
       labs(x = paste0("PC", k1), y = paste0("PC", k2)),
     qplot(PC.ref[, k1], PC.ref[, k2], size = I(2)) +
       geom_point(aes(proj2[ind, k1], proj2[ind, k2]), color = "blue") +
-      theme_bigstatsr(0.7) +
+      theme_bigstatsr(0.5) +
       labs(x = paste0("PC", k1), y = paste0("PC", k2)),
     scale = 0.95
   )
-}), nrow = 3)
+}), nrow = 4)
 
-# ggsave("figures/proj1000G-UKBB.pdf", width = 10.5, height = 6.5)
+# ggsave("figures/proj1000G-UKBB.png", width = 11, height = 7)
 
 

### 2019-10-17 paper-bedpca.tex

@@
 
+\subsection{Robust Mahalanobis distance\label{maha}}
+
+In this paper, for many applications, we use the pairwise orthogonalized Gnanadesikan-Kettenrin robust Mahalanobis distance as implemented in function \texttt{covRob} of R package robust with parameter \texttt{estim = "pairwiseGK"} \cite[]{gnanadesikan1972robust,yohai1988high,maronna2002robust,todorov2009object}.
+We reexport this function in R package bigutilsr for convenience.
+
 \subsection{Outlier detection in PCA [SPLIT IN MULTIPLE SECTIONS]}
@@
 
-[SEPARATE STEPS?]
+%[SEPARATE STEPS?]
 
 To project PCs of a reference dataset (e.g.\ 1000G) to a target genotype dataset, we implement the following 3 steps in function \texttt{bed\_projectPCA} of package bigsnpr: 1) matching the variants of each dataset, including removing ambiguous alleles [A/T] and [C/G], and matching strand and direction of the alleles; 2) computing PCA of the reference dataset using the matched variants only; 3) projecting computed PCs to the target data using an optimised implementation (see Supplementary Materials) of the Online Augmentation, Decomposition, and Procrustes (OADP) transformation \cite[]{zhang2019fast}.
+
+\subsection{Ancestry estimation}
+
+We use the 1000 Genomes (1000G) Project data (section \ref{1000G}) to infer the ancestry of individuals from a study dataset following the following steps: 1) we compute the robust center and covariance parameters of the Mahalanobis distance (section \ref{maha}) for each of the 26 populations of the 10000G data; 2) we project individuals from the study dataset onto the PCA space computed using the 1000G data (section \ref{proj}); 3) we compute the Mahalanobis distance for each projected individual of the study data and each population of the 1000G data (using the center and covariance parameters computed previously); 4) we derive p-values from those distances, as they are approximately chi-squared distributed with the number of PCs used for computing the distances as degrees of freedom; 5) we affect each individual to the largest corresponding p-value (smallest distance), but discard individuals whose maximum p-value is less than 0.05 (i.e.
+individuals far from every population of the 1000G data).
 
@@
 Simple projection is affected even more by this shrinkage for PC9 to PC20, while OADP appears free of this issue (Figure \ref{fig:proj1000G-2}).
+We show the same results when projecting the full UK Biobank data onto PCA computed using 1000G data  (Figure \ref{fig:proj1000G-5}).
 When correcting projected PC scores with asymptotic shrinkage factors, bias is smaller than when simply projecting loadings, yet, it is apparent for PC7-8 and PC11-12 (Figure \ref{fig:proj1000G-3}).
@@
 
+\begin{figure}[!htpb]
+\centerline{\includegraphics[width=0.9\textwidth]{proj1000G-UKBB.png}}
+\caption{Principal Component (PC) scores 1 to 16 of the 1000 Genomes project and projected individuals from the UK Biobank.
+Black points are PC scores of 1000G individuals used for computing PCA.
+Blue points are the 488,371 individuals from the UK Biobank, projected using the Online Augmentation, Decomposition, and Procrustes (OADP) transformation.
+Note that only 20,000 random individuals are represented in this plot.
+\label{fig:proj1000G-5}}
+\end{figure}
+
 \vspace*{1em}

### 2019-10-17 code/project-UKBB-to-1000G.R
ancestry estimation UKBB

@@ -40,4 +40,114 @@ plot_grid(plotlist = lapply(1:8, function(k) {
 
 # ggsave("figures/proj1000G-UKBB.png", width = 11, height = 7)
 
+#### Ancestry estimation ####
 
+fam2 <- bigreadr::fread2("data/1000G_phase3_common_norel.fam2")
+
+plot_grid(plotlist = lapply(5:10, function(k) {
+  k1 <- 2 * k - 1
+  k2 <- 2 * k
+  qplot(PC.ref[, k1], PC.ref[, k2], size = I(2), color = fam2$Population) +
+    theme_bigstatsr(0.7) +
+    labs(x = paste0("PC", k1), y = paste0("PC", k2))
+}))
+
+unAsIs <- function(X) {
+  class(X) <- setdiff(class(X), "AsIs")
+  X
+}
+
+library(dplyr)
+seq_PC <- c(1:14, 16)
+all_covRob <- cbind(fam2, PC = I(PC.ref)) %>%
+  group_by(`Super Population`, `Population Description`, Population) %>%
+  summarize(maha = list(bigutilsr::covRob(unAsIs(PC[, seq_PC]), estim = "pairwiseGK")))
+POP <- paste(all_covRob$`Super Population`, all_covRob$Population, sep = "_")
+
+all_dist <- sapply(all_covRob$maha, function(maha) {
+  mahalanobis(proj2[, seq_PC], center = maha$center, cov = maha$cov)
+})
+colnames(all_dist) <- POP
+
+round(all_prob <- pchisq(all_dist, df = length(seq_PC), lower.tail = FALSE), 3)
+round(corr <- 100 * cor(all_prob), 1)
+ind <- which(corr > 10 & lower.tri(corr), arr.ind = TRUE)
+data.frame(pop1 = POP[ind[, 1]], pop2 = POP[ind[, 2]], cor = corr[ind])
+# 1  ACB  ASW 26.57052
+# 2  YRI  ESN 55.01614
+# 3  KHV  CDX 39.74580
+# 4  CHS  CHB 65.06130
+# 5  CEU  GBR 65.59252
+# 6  STU  ITU 70.99639
+
+choose_pop <- apply(all_prob, 1, which.max)
+pval_max <- all_prob[cbind(rows_along(all_prob), choose_pop)]
+choose_pop2 <- ifelse(pval_max < 0.05, NA, POP[choose_pop])
+
+ggplot() +
[…]

### 2019-10-17 paper-bedpca.tex
save UKBB ancestry

@@
 
-\subsection{Outlier detection in PCA [SPLIT IN MULTIPLE SECTIONS]}
+\subsection{Detecting LD structure in PCA}
 
-We propose [BOF] different outlier detection techniques for PCA.
-
-For detecting outlier variants in PCA that are due to long-range Linkage Disequilibrium (LD) regions, we use the same procedure as described by \cite{prive2017efficient}: we compute pcadapt statistics \cite[]{luu2017pcadapt} that summarises the contribution of each variant in all PC loadings computed.
+For detecting outlier variants in PCA that are due to long-range Linkage Disequilibrium (LD) regions, we use a slightly modified version of the procedure described by \cite{prive2017efficient}. 
+To summarise the contribution of each variant in all PC loadings, we compute the robust Mahalanobis distances (Section \ref{maha}) of PC loadings.  
 This captures both population-differentiating variants as well as long-range LD regions. 
-We then smooth the statistics [BE MORE PRECISE?] to mostly capture consecutive outliers that corresponds to long-range LD regions.
-
-For detecting outlier samples in PCA, we use the Local Outlier Factor (LOF) on PCs \cite[]{breunig2000lof}.
-To make it robust to the choice of the number of nearest neighbours (kNN), we take the maximum value of the statistic computed with kNN $\in$ \{4, 5, 6, 7, 8\}.
-We do not use larger values for `kNN' because it would remove small ancestry groups (e.g.\ composed of 10-15 people only).
-We make use of the fast K nearest neighbours implementation of R package nabor to implement the LOF statistic efficiently in our new R package bigutilsr \cite[]{elseberg2012comparison}.
-
-For detecting samples that have a different ancestry from most of the samples in the data, we compute the pairwise orthogonalized Gnanadesikan-Kettenrin robust Mahalanobis distance of PCs \cite[]{gnanadesikan1972robust,maronna2002robust}, and do not include individuals whose log-distance is larger than some threshold determined based on visual inspection [DO IT AUTOMATIC?].
+Then, to capture consecutive outliers that corresponds to long-range LD regions, we apply a Gaussian smoothing to these statistics.
 
@@
 To solve these two issues, we implement \texttt{tukey\_mc\_up} in R package bigutilsr and use it here, which accounts both for skewness and multiple testing by default.
+
+\subsection{Detecting outlier samples in PCA}
+
+We propose [BOF] different outlier detection techniques for PCA.
+
+For detecting outlier samples in PCA, we use the Local Outlier Factor (LOF) on PCs \cite[]{breunig2000lof}.
+To make it robust to the choice of the number of nearest neighbours (kNN), we take the maximum value of the statistic computed with kNN $\in$ \{4, 5, 6, 7, 8\}.
+We do not use larger values for `kNN' because it would remove small ancestry groups (e.g.\ composed of 10-15 people only).
+We make use of the fast K nearest neighbours implementation of R package nabor to implement the LOF statistic efficiently in our new R package bigutilsr \cite[]{elseberg2012comparison}.
+
+For detecting samples that have a different ancestry from most of the samples in the data, we compute the pairwise orthogonalized Gnanadesikan-Kettenrin robust Mahalanobis distance of PCs \cite[]{gnanadesikan1972robust,maronna2002robust}, and do not include individuals whose log-distance is larger than some threshold determined based on visual inspection [DO IT AUTOMATIC?].
 
@@
 
-\subsection{Ancestry estimation}
+\subsection{Ancestry estimation\label{ancestry}}
 
-We use the 1000 Genomes (1000G) Project data (section \ref{1000G}) to infer the ancestry of individuals from a study dataset following the following steps: 1) we compute the robust center and covariance parameters of the Mahalanobis distance (section \ref{maha}) for each of the 26 populations of the 10000G data; 2) we project individuals from the study dataset onto the PCA space computed using the 1000G data (section \ref{proj}); 3) we compute the Mahalanobis distance for each projected individual of the study data and each population of the 1000G data (using the center and covariance parameters computed previously); 4) we derive p-values from those distances, as they are approximately chi-squared distributed with the number of PCs used for computing the distances as degrees of freedom; 5) we affect each individual to the largest corresponding p-value (smallest distance), but discard individuals whose maximum p-value is less than 0.05 (i.e.
-individuals far from every population of the 1000G data).
+We use the 1000 Genomes (1000G) Project data (section \ref{1000G}) to infer the ancestry of individuals from a study dataset following the following steps: 1) we compute the robust center and covariance parameters of the Mahalanobis distance (section \ref{maha}) for each of the 26 populations of the 10000G data; 2) we project individuals from the study dataset onto the PCA space computed using the 1000G data (section \ref{proj}); 3) we compute the Mahalanobis distance for each projected individual of the study data and each population of the 1000G data (using the center and covariance parameters computed previously); 4) we derive p-values from those distances, as they are approximately chi-squared distributed with the number of PCs used for computing the distances as degrees of freedom; 5) we affect each individual to the largest corresponding p-value (smallest distance), but discard individuals whose maximum p-value is less than 0.05 (i.e.\ individuals far from every population of the 1000G data).
 
@@
 
+\subsection{Ancestry estimation}
+
+[REDO WITH OUTLIER REMOVAL]
+
+We performed ancestry estimation of the individuals in the UK Biobank using the 1000G data (Section \ref{ancestry}), discarding individuals with unknown ancestry or mixed ancestry. 
+We did not infer ancestry for 15.5\% of UK Biobank that had a maximum p-value smaller than 0.05 (Figure \ref{fig:ancestry-pval}). 
+More precisely, among ``White'', ``British'' and ``Irish'' ancestries, this represented respectively 20.4\%, 10.5\% and 19.8\%, while this represented between 48.5\% (``Chinese'') and 81.4\% (``Bangladeshi'').
+In other words, most of the UK Biobank individual who self reports as ``Bangladeshi'' could not be attributed to any 1000G population, although 1000G data includes some ``Bengali in Bangladesh'' population.
+Only 10 people out of 401,048 were wrongly classified in super population of the 1000G; e.g.\ one Chinese was classified as European by our method (Table \ref{tab:ancestry-pred}). 
+Our method is able to accurately differentiate between sub-continental populations such as differentiating between Pakistani, Bangladeshi and Chinese people (Table \ref{tab:ancestry-pred}).
+
+% latex table generated in R 3.6.0 by xtable 1.8-4 package
+% Thu Oct 17 13:58:41 2019
+\begin{table}[ht]
+\centering
+\caption{Self-reported ancestry (top) of UKBB individuals and their matching to 1000G populations (left) by our method.
+See the description of 1000G populations at \url{https://www.internationalgenome.org/category/population/}.}
+\label{tab:ancestry-pred}
+\begin{tabular}{|l|c|c|c|c|c|c|c|c|c|c|c|c|}
+  \hline
+ & British & Irish & White & Other White & Indian & Pakistani & Bangladeshi & Chinese & Other Asian & Caribbean & African & Other Black \\ 
+  \hline
+AFR\_ACB &  &  &  &  &  &  &  &  &  & 1635 & 510 & 27 \\ 
+  AFR\_ASW &  &  &  &  &  &  &  &  &  & 488 & 23 & 14 \\ 
+  AFR\_ESN &  &  &  &  &  &  &  &  &  &  & 103 &  \\ 
+  AFR\_GWD &  &  &  &  &  &  &  &  &  &  & 15 &  \\ 
+  AFR\_LWK &  &  &  &  &  &  &  &  &  &  & 38 &  \\ 
+  AFR\_MSL &  &  &  &  &  &  &  &  &  &  & 25 &  \\ 
+  AFR\_YRI &  &  &  &  &  &  &  &  &  & 4 & 433 & 4 \\ 
+   \hline
+AMR\_CLM & 1 &  &  & 53 &  &  &  &  &  &  &  &  \\ 
+  AMR\_MXL &  &  & 1 & 3 &  &  &  &  &  &  &  &  \\ 
+  AMR\_PEL &  &  &  & 1 &  &  &  &  &  &  &  &  \\ 
+  AMR\_PUR &  &  &  &  &  &  &  &  &  &  &  &  \\ 
+   \hline
+EAS\_CDX &  &  &  &  &  &  &  & 1 &  &  &  &  \\ 
+  EAS\_CHB &  &  & 1 &  &  &  &  & 408 & 11 &  &  &  \\ 
+  EAS\_CHS &  &  &  &  &  &  &  & 350 & 6 &  &  &  \\ 
+  EAS\_JPT &  &  &  &  &  &  &  & 2 & 24 &  &  &  \\ 
+  EAS\_KHV &  &  &  &  &  &  &  & 3 & 16 &  &  &  \\ 
+   \hline
+EUR\_CEU & 293299 & 5483 & 327 & 4486 &  &  &  & 1 &  &  &  &  \\ 
+  EUR\_FIN & 2 &  &  & 116 &  &  &  &  &  &  &  &  \\ 
+  EUR\_GBR & 91862 & 4881 & 99 & 765 &  &  &  &  &  &  & 1 &  \\ 
+  EUR\_IBS & 32 &  & 4 & 481 &  &  &  &  &  &  &  &  \\ 
+  EUR\_TSI & 20 &  & 1 & 166 &  &  &  &  &  &  &  &  \\ 
+   \hline
+SAS\_BEB &  &  &  &  & 12 &  & 50 &  &  &  &  &  \\ 
+  SAS\_GIH &  &  &  &  & 302 &  &  &  &  &  &  &  \\ 
+  SAS\_ITU & 1 &  &  &  & 307 & 3 &  &  & 251 & 3 &  &  \\ 
+  SAS\_PJL &  &  &  &  & 1452 & 712 &  &  & 82 & 1 &  &  \\ 
+  SAS\_STU &  &  &  &  & 42 &  &  &  & 172 &  &  &  \\ 
+   \hline
+\end{tabular}
+\end{table}
+
+\noindent[OKAY TO CLASSIFY SOME OTHER WHITE AS CENTRAL/SOUTH AMERICANS?]
+
+\noindent[INCLUDE RESULTS FOR SIMPLE PROJ? -> HIST PVAL BAD -> 38\% NOT MATCHED, ESP PAKISTANI]
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-[LIMITATION: RUN MULTIPLE TIMES / FAIL TO SPEED UP]
+\noindent[LIMITATION: RUN MULTIPLE TIMES / FAIL TO SPEED UP]
 
@@
 
+\subsection*{Supplementary figures}
+
 \vspace*{1em}
@@
 
-% latex table generated in R 3.5.2 by xtable 1.8-4 package
-% Wed Jun 19 15:19:38 2019
-\begin{table}[!htpb]
-\caption{\label{tab:AUC-simu2}}
-\vspace*{0.5em}
-\centering
-\begin{tabular}{|l|c|c|c|c|}
-\hline
-Scenario & stdCT & maxCT & SCT & lassosum \\ 
-\hline
-100  & 77.4 [76.0-78.8] & 83.9 [83.4-84.4] & 83.1 [82.6-83.6] & 80.1 [79.5-80.8] \\ 
-10K  & 69.4 [68.4-70.5] & 73.0 [72.5-73.4] & 72.9 [72.5-73.3] & 71.2 [70.6-71.7] \\ 
-1M   & 64.0 [63.6-64.4] & 64.0 [63.6-64.4] & 62.7 [62.3-63.0] & 64.1 [63.3-64.8] \\ 
-2chr & 70.0 [68.8-71.2] & 74.4 [73.6-75.2] & 78.5 [77.9-79.1] & 73.2 [72.5-73.8] \\ 
-err  & 67.0 [66.0-68.1] & 68.6 [67.7-69.5] & 69.5 [68.9-70.1] & 65.6 [64.9-66.3] \\ 
-HLA  & 74.8 [72.9-76.3] & 75.3 [73.5-76.9] & 76.4 [74.5-78.0] & 75.8 [74.2-77.2] \\ 
-\hline
-\end{tabular}
-\end{table}
+\begin{figure}[!htpb]
+\centerline{\includegraphics[width=0.8\textwidth]{hist-pval-max.pdf}}
+\caption{Maximum p-values based on robust Mahalabonis distances of UK Biobank individuals from each of the 26 1000G populations.
+\label{fig:ancestry-pval}}
+\end{figure}
 

### 2019-10-17 paper-bedpca.tex
save

@@
 
+% latex table generated in R 3.5.2 by xtable 1.8-4 package
+% Thu Oct 17 16:12:09 2019
+\begin{table}[ht]
+\centering
+\caption{Reported ancestry (left) of POPRES individuals and their matching to 1000G populations (top) by our method.
+See the description of 1000G populations at \url{https://www.internationalgenome.org/category/population/}.}
+\label{tab:ancestry-pred-popres}
+\begin{tabular}{|l|c|c|c|c|c|c|}
+  \hline
+ & EUR\_CEU & EUR\_FIN & EUR\_GBR & EUR\_IBS & EUR\_TSI & NA \\ 
+  \hline
+Anglo-Irish Isles & 109 &  & 108 &  &  & 49 \\ 
+  Belgium & 25 &  & 4 &  &  & 14 \\ 
+  Central Europe & 5 &  &  &  & 1 & 49 \\ 
+  Eastern Europe & 1 &  &  &  &  & 29 \\ 
+  France & 13 &  &  & 14 &  & 62 \\ 
+  Germany & 31 &  & 3 &  &  & 37 \\ 
+  Italy &  &  &  & 13 & 19 & 187 \\ 
+  Netherlands & 12 &  & 1 &  &  & 4 \\ 
+  Scandinavia & 3 & 1 &  &  &  & 11 \\ 
+  SE Europe &  &  &  &  & 6 & 88 \\ 
+  SW Europe &  &  &  & 211 &  & 53 \\ 
+  Switzerland & 29 &  & 4 & 7 &  & 182 \\ 
+   \hline
+\end{tabular}
+\end{table}
+
 \end{document}

### 2019-10-17 code/project-POPRES-to-1000G.R

@@ -1,0 +1,113 @@
+library(bigsnpr)
+library(ggplot2)
+
+obj.bed <- bed("~/Bureau/POPRES_data/POPRES_allchr_QC.bed")
+bed.ref <- bed(download_1000G("tmp-data"))
+
+system.time(
+  test <- bed_projectPCA(bed.ref, obj.bed, k = 20, ncores = nb_cores(),
+                         build.new = "hg18", liftOver = "tmp-data/liftOver")
+)
+
+PC.ref <- predict(test$obj.svd.ref)
+proj1 <- test$simple_proj
+proj2 <- test$OADP_proj
+
+# shrinkage coefficients
+shrinkage <- unname(sapply(1:20, function(k) {
+  MASS::rlm(proj2[, k] ~ proj1[, k] + 0)$coef
+}))
+round(shrinkage, 2)
+#  [1] 1.01 1.01 1.04 1.05 1.31 1.33 1.50 1.69 1.91 2.05 2.25 2.29 2.31
+# [14] 2.34 2.49 2.82 2.90 3.03 3.42 3.67
+
+plot_grid(plotlist = lapply(1:8, function(k) {
+  k1 <- 2 * k - 1
+  k2 <- 2 * k
+  plot_grid(
+    qplot(PC.ref[, k1], PC.ref[, k2], size = I(2)) +
+      geom_point(aes(proj1[, k1], proj1[, k2]), color = "red") +
+      theme_bigstatsr(0.5) +
+      labs(x = paste0("PC", k1), y = paste0("PC", k2)),
+    qplot(PC.ref[, k1], PC.ref[, k2], size = I(2)) +
+      geom_point(aes(proj2[, k1], proj2[, k2]), color = "blue") +
+      theme_bigstatsr(0.5) +
+      labs(x = paste0("PC", k1), y = paste0("PC", k2)),
+    scale = 0.95
+  )
+}), nrow = 4)
+
+# ggsave("figures/proj1000G-Celiac.png", width = 11, height = 7)
+
+#### Ancestry estimation ####
+
+fam2 <- bigreadr::fread2("tmp-data/1000G_phase3_common_norel.fam2")
+
+plot_grid(plotlist = lapply(5:10, function(k) {
+  k1 <- 2 * k - 1
+  k2 <- 2 * k
+  qplot(PC.ref[, k1], PC.ref[, k2], size = I(2), color = fam2$Population) +
+    theme_bigstatsr(0.7) +
[…]

### 2019-10-17 code/project-celiac-to-1000G.R

@@ -1,0 +1,118 @@
+library(bigsnpr)
+library(ggplot2)
+
+obj.bed <- bed("~/Bureau/Dubois2010_data/FinnuncorrNLITUK1UK3hap300_QC.bed")
+bed.ref <- bed(download_1000G("tmp-data"))
+
+system.time(
+  test <- bed_projectPCA(bed.ref, obj.bed, k = 20, ncores = nb_cores(),
+                         build.new = "hg18", liftOver = "tmp-data/liftOver")
+)
+
+PC.ref <- predict(test$obj.svd.ref)
+proj1 <- test$simple_proj
+proj2 <- test$OADP_proj
+
+# shrinkage coefficients
+shrinkage <- unname(sapply(1:20, function(k) {
+  MASS::rlm(proj2[, k] ~ proj1[, k] + 0)$coef
+}))
+round(shrinkage, 2)
+#  [1] 1.01 1.01 1.04 1.05 1.30 1.33 1.51 1.72 1.93 2.08 2.30 2.31 2.36
+# [14] 2.39 2.54 2.87 2.97 3.25 3.65 3.91
+
+plot_grid(plotlist = lapply(1:8, function(k) {
+  k1 <- 2 * k - 1
+  k2 <- 2 * k
+  plot_grid(
+    qplot(PC.ref[, k1], PC.ref[, k2], size = I(2)) +
+      geom_point(aes(proj1[, k1], proj1[, k2]), color = "red") +
+      theme_bigstatsr(0.5) +
+      labs(x = paste0("PC", k1), y = paste0("PC", k2)),
+    qplot(PC.ref[, k1], PC.ref[, k2], size = I(2)) +
+      geom_point(aes(proj2[, k1], proj2[, k2]), color = "blue") +
+      theme_bigstatsr(0.5) +
+      labs(x = paste0("PC", k1), y = paste0("PC", k2)),
+    scale = 0.95
+  )
+}), nrow = 4)
+
+# ggsave("figures/proj1000G-Celiac.png", width = 11, height = 7)
+
+#### Ancestry estimation ####
+
+fam2 <- bigreadr::fread2("tmp-data/1000G_phase3_common_norel.fam2")
+
+plot_grid(plotlist = lapply(5:10, function(k) {
+  k1 <- 2 * k - 1
+  k2 <- 2 * k
+  qplot(PC.ref[, k1], PC.ref[, k2], size = I(2), color = fam2$Population) +
[…]

### 2019-10-17 code/project-me-to-POPRES.R

@@ -1,0 +1,88 @@
+library(bigsnpr)
+library(ggplot2)
+
+plink <- download_plink("tmp-data")
+mygenome <- "~/Bureau/thesis/backingfiles/genome_Florian_Prive_v5_Full_20190212023418.txt"
+
+system(glue::glue(
+  "{plink} --23file {mygenome}",
+  " --autosome --geno 0 --snps-only",
+  " --a2-allele {sub_bed(download_1000G('tmp-data'), '.bim')} 5 2",
+  " --make-bed --out tmp-data/mygenome"
+))
+
+obj.bed <- bed("tmp-data/mygenome.bed")
+bed.ref <- bed("~/Bureau/POPRES_data/POPRES_allchr_QC.bed")
+
+system.time(
+  test <- bed_projectPCA(bed.ref, obj.bed, k = 10, ncores = nb_cores(),
+                         build.ref = "hg18", liftOver = "tmp-data/liftOver",
+                         match.min.prop = 0, roll.size = 20)
+) # using 18,764 variants only
+
+PC.ref <- predict(test$obj.svd.ref)
+proj2 <- test$OADP_proj
+
+
+#### Ancestry estimation ####
+
+pop <- bed.ref$fam$family.ID
+pop2 <- dplyr::case_when(
+  pop %in% c("Portugal", "Spain") ~ "SW Europe",
+  pop %in% c("?orway", "Sweden", "Finland", "Denmark") ~ "Scandinavia",
+  pop %in% c("Hungary", "Slovakia", "Czech", "Austria", "Slovenia",
+             "Croatia") ~ "Central Europe",
+  pop %in% c("Russian", "Latvia", "Ukraine", "Poland") ~ "Eastern Europe",
+  pop %in% c("Greece", "Turkey", "Serbia", "Cyprus", "Albania", "Kosovo",
+             "Bosnia", "Macedonia,",  "Romania", "Bulgaria") ~ "SE Europe",
+  pop %in% c("Swiss-German", "Swiss-French", "Swiss-Italian") ~ "Switzerland",
+  pop == "?etherlands" ~ "Netherlands",
+  pop %in% c("United", "Scotland", "Ireland") ~ "Anglo-Irish Isles",
+  TRUE ~ pop
+)
+
+plot_grid(plotlist = lapply(1:2, function(k) {
+  k1 <- 2 * k - 1
+  k2 <- 2 * k
[…]

### 2019-10-18 paper-bedpca.tex
save

--- 2019-10-17 paper-bedpca.tex
@@
 
+% latex table generated in R 3.6.0 by xtable 1.8-4 package
+% Sat Oct 19 00:18:24 2019
+\begin{table}[ht]
+\centering
+\caption{Ancestry (left) of POPRES individuals and their matching to 1000G populations (top) by our method.
+See the description of 1000G populations at \url{https://www.internationalgenome.org/category/population/}.}
+\label{tab:ancestry-pred-popres}
+\begin{tabular}{|l|c|c|c|c|c|}
+  \hline
+ & EUR\_CEU & EUR\_GBR & EUR\_IBS & EUR\_TSI & NA \\ 
+  \hline
+Anglo-Irish Isles & 159 & 57 &  &  & 50 \\ 
+  Belgium & 28 & 3 &  &  & 12 \\ 
+  Central Europe & 3 &  &  & 1 & 51 \\ 
+  Eastern Europe & 2 &  &  &  & 28 \\ 
+  France & 14 &  & 13 &  & 62 \\ 
+  Germany & 37 &  &  &  & 34 \\ 
+  Italy &  &  & 12 & 19 & 188 \\ 
+  Netherlands & 10 & 2 &  &  & 5 \\ 
+  Scandinavia & 5 &  &  &  & 10 \\ 
+  SE Europe &  &  &  & 5 & 89 \\ 
+  SW Europe &  &  & 193 &  & 71 \\ 
+  Switzerland & 33 & 1 & 6 &  & 182 \\ 
+   \hline
+\end{tabular}
+\end{table}
+
+% latex table generated in R 3.6.0 by xtable 1.8-4 package
+% Sat Oct 19 00:28:24 2019
+\begin{table}[ht]
+\centering
+\caption{Ancestry (left) of Celiac individuals and their matching to 1000G populations (top) by our method.
+See the description of 1000G populations at \url{https://www.internationalgenome.org/category/population/}.}
+\label{tab:ancestry-pred-celiac}
+\begin{tabular}{|l|c|c|c|c|c|c|c|}
+  \hline
+ & AMR\_MXL & EUR\_CEU & EUR\_FIN & EUR\_GBR & EUR\_IBS & EUR\_TSI & NA \\ 
+  \hline
+Finland &  & 4 & 1543 & 1 &  & 1 & 922 \\ 
+  Italy &  &  &  &  & 15 & 229 & 795 \\ 
+  Netherlands &  & 802 &  & 382 & 1 &  & 463 \\ 
+  UK & 1 & 2865 & 1 & 2532 & 4 & 5 & 1346 \\ 
+   \hline
+\end{tabular}
+\end{table}
+
 \end{document}

### 2019-10-18 code/project-POPRES-to-1000G.R

@@ -1,6 +1,12 @@
 library(bigsnpr)
 library(ggplot2)
 
+snp_plinkQC(
+  plink.path = download_plink("tmp-data"),
+  prefix.in = "~/Bureau/POPRES_data/POPRES_allchr",
+  geno = 0.01, mind = 0.01,
+  autosome.only = TRUE
+)
 obj.bed <- bed("~/Bureau/POPRES_data/POPRES_allchr_QC.bed")
 bed.ref <- bed(download_1000G("tmp-data"))
 
@@ -13,13 +19,6 @@ PC.ref <- predict(test$obj.svd.ref)
 proj1 <- test$simple_proj
 proj2 <- test$OADP_proj
 
-# shrinkage coefficients
-shrinkage <- unname(sapply(1:20, function(k) {
-  MASS::rlm(proj2[, k] ~ proj1[, k] + 0)$coef
-}))
-round(shrinkage, 2)
-#  [1] 1.01 1.01 1.04 1.05 1.31 1.33 1.50 1.69 1.91 2.05 2.25 2.29 2.31
-# [14] 2.34 2.49 2.82 2.90 3.03 3.42 3.67
 
 plot_grid(plotlist = lapply(1:8, function(k) {
   k1 <- 2 * k - 1
@@ -51,28 +50,20 @@ plot_grid(plotlist = lapply(5:10, function(k) {
     labs(x = paste0("PC", k1), y = paste0("PC", k2))
 }))
 
-unAsIs <- function(X) {
-  class(X) <- setdiff(class(X), "AsIs")
-  X
-}
 
-library(dplyr)
-seq_PC <- c(1:19)
-all_covRob <- cbind(fam2, PC = I(PC.ref)) %>%
-  group_by(`Super Population`, `Population Description`, Population) %>%
-  summarize(maha = list(bigutilsr::covRob(unAsIs(PC[, seq_PC]), estim = "pairwiseGK")))
-POP <- paste(all_covRob$`Super Population`, all_covRob$Population, sep = "_")
+seq_PC <- c(1:18, 20)
+pop <- paste(fam2$`Super Population`, fam2$Population, sep = "_")
+pop_PCs <- vctrs::vec_split(PC.ref, pop)
 
-all_dist <- sapply(all_covRob$maha, function(maha) {
-  mahalanobis(proj2[, seq_PC], center = maha$center, cov = maha$cov)
+all_pval <- sapply(pop_PCs$val, function(PC) {
[…]

### 2019-10-18 code/project-celiac-to-1000G.R

@@ -1,7 +1,14 @@
 library(bigsnpr)
 library(ggplot2)
 
-obj.bed <- bed("~/Bureau/Dubois2010_data/FinnuncorrNLITUK1UK3hap300_QC.bed")
+snp_plinkQC(
+  plink.path = download_plink("tmp-data"),
+  prefix.in = "~/Bureau/Dubois2010_data/FinnuncorrNLITUK3hap550",
+  geno = 0.01, mind = 0.01,
+  autosome.only = TRUE
+)
+
+obj.bed <- bed("~/Bureau/Dubois2010_data/FinnuncorrNLITUK3hap550_QC.bed")
 bed.ref <- bed(download_1000G("tmp-data"))
 
 system.time(
@@ -12,14 +19,6 @@ system.time(
 PC.ref <- predict(test$obj.svd.ref)
 proj1 <- test$simple_proj
 proj2 <- test$OADP_proj
-
-# shrinkage coefficients
-shrinkage <- unname(sapply(1:20, function(k) {
-  MASS::rlm(proj2[, k] ~ proj1[, k] + 0)$coef
-}))
-round(shrinkage, 2)
-#  [1] 1.01 1.01 1.04 1.05 1.30 1.33 1.51 1.72 1.93 2.08 2.30 2.31 2.36
-# [14] 2.39 2.54 2.87 2.97 3.25 3.65 3.91
 
 plot_grid(plotlist = lapply(1:8, function(k) {
   k1 <- 2 * k - 1
@@ -51,27 +50,20 @@ plot_grid(plotlist = lapply(5:10, function(k) {
     labs(x = paste0("PC", k1), y = paste0("PC", k2))
 }))
 
-unAsIs <- function(X) {
-  class(X) <- setdiff(class(X), "AsIs")
-  X
-}
 
-library(dplyr)
-seq_PC <- c(1:19)
-all_covRob <- cbind(fam2, PC = I(PC.ref)) %>%
-  group_by(`Super Population`, `Population Description`, Population) %>%
-  summarize(maha = list(bigutilsr::covRob(unAsIs(PC[, seq_PC]), estim = "pairwiseGK")))
-POP <- paste(all_covRob$`Super Population`, all_covRob$Population, sep = "_")
+seq_PC <- 1:19
+pop <- paste(fam2$`Super Population`, fam2$Population, sep = "_")
+pop_PCs <- vctrs::vec_split(PC.ref, pop)
 
[…]

### 2019-10-19 paper-bedpca.tex
add Celiac & POPRES

@@
 
-\subsection{Robust Mahalanobis distance\label{maha}}
+\subsection{Robust Mahalanobis distance \label{maha}}
 
@@
 
-\subsection{Projecting PCs from a reference dataset\label{proj}}
+\subsection{Projecting PCs from a reference dataset \label{proj}}
 
@@
 
-\subsection{Ancestry estimation\label{ancestry}}
+\subsection{Ancestry matching \label{ancestry}}
 
@@
 
-\subsection{Ancestry estimation}
+\subsection{Ancestry matching}
 
-[REDO WITH OUTLIER REMOVAL]
+[REDO WITH OUTLIER REMOVAL?]
 
@@
 More precisely, among ``White'', ``British'' and ``Irish'' ancestries, this represented respectively 20.4\%, 10.5\% and 19.8\%, while this represented between 48.5\% (``Chinese'') and 81.4\% (``Bangladeshi'').
-In other words, most of the UK Biobank individual who self reports as ``Bangladeshi'' could not be attributed to any 1000G population, although 1000G data includes some ``Bengali in Bangladesh'' population.
-Only 10 people out of 401,048 were wrongly classified in super population of the 1000G; e.g.\ one Chinese was classified as European by our method (Table \ref{tab:ancestry-pred}). 
-Our method is able to accurately differentiate between sub-continental populations such as differentiating between Pakistani, Bangladeshi and Chinese people (Table \ref{tab:ancestry-pred}).
+In other words, most of the UK Biobank individual who self-reports as ``Bangladeshi'' could not be attributed to any 1000G population, although 1000G data includes several ``Bengali in Bangladesh'' (SAS\_BEB).
+Only 10 people out of 401,048 were wrongly classified in ``super'' population of the 1000G; e.g.\ one Chinese person was classified as European by our method (Table \ref{tab:ancestry-pred}). 
+Moreover, our method is able to accurately differentiate between sub-continental populations such as differentiating between Pakistani, Bangladeshi and Chinese people (Table \ref{tab:ancestry-pred}).
 
 % latex table generated in R 3.6.0 by xtable 1.8-4 package
-% Thu Oct 17 13:58:41 2019
+% Sat Oct 19 09:45:04 2019
 \begin{table}[ht]
@@
   \hline
-AFR\_ACB &  &  &  &  &  &  &  &  &  & 1635 & 510 & 27 \\ 
-  AFR\_ASW &  &  &  &  &  &  &  &  &  & 488 & 23 & 14 \\ 
+AFR\_ACB &  &  &  &  &  &  &  &  &  & 1687 & 580 & 28 \\ 
+  AFR\_ASW &  &  &  &  &  &  &  &  &  & 365 & 12 & 12 \\ 
   AFR\_ESN &  &  &  &  &  &  &  &  &  &  & 103 &  \\ 
-  AFR\_GWD &  &  &  &  &  &  &  &  &  &  & 15 &  \\ 
-  AFR\_LWK &  &  &  &  &  &  &  &  &  &  & 38 &  \\ 
-  AFR\_MSL &  &  &  &  &  &  &  &  &  &  & 25 &  \\ 
-  AFR\_YRI &  &  &  &  &  &  &  &  &  & 4 & 433 & 4 \\ 
+  AFR\_GWD &  &  &  &  &  &  &  &  &  &  & 12 &  \\ 
+  AFR\_LWK &  &  &  &  &  &  &  &  &  &  & 33 &  \\ 
+  AFR\_MSL &  &  &  &  &  &  &  &  &  &  & 23 &  \\ 
+  AFR\_YRI &  &  &  &  &  &  &  &  &  & 4 & 403 & 3 \\ 
    \hline
-AMR\_CLM & 1 &  &  & 53 &  &  &  &  &  &  &  &  \\ 
+AMR\_CLM & 1 &  &  & 47 &  &  &  &  &  &  &  &  \\ 
   AMR\_MXL &  &  & 1 & 3 &  &  &  &  &  &  &  &  \\ 
   AMR\_PEL &  &  &  & 1 &  &  &  &  &  &  &  &  \\ 
-  AMR\_PUR &  &  &  &  &  &  &  &  &  &  &  &  \\ 
    \hline
 EAS\_CDX &  &  &  &  &  &  &  & 1 &  &  &  &  \\ 
-  EAS\_CHB &  &  & 1 &  &  &  &  & 408 & 11 &  &  &  \\ 
-  EAS\_CHS &  &  &  &  &  &  &  & 350 & 6 &  &  &  \\ 
-  EAS\_JPT &  &  &  &  &  &  &  & 2 & 24 &  &  &  \\ 
-  EAS\_KHV &  &  &  &  &  &  &  & 3 & 16 &  &  &  \\ 
+  EAS\_CHB & 1 &  &  &  &  &  &  & 369 & 10 &  &  &  \\ 
+  EAS\_CHS &  &  &  &  &  &  &  & 398 & 8 &  &  &  \\ 
+  EAS\_JPT &  &  &  &  &  &  &  & 2 & 23 &  &  &  \\ 
+  EAS\_KHV &  &  &  &  &  &  &  & 3 & 14 &  &  &  \\ 
    \hline
-EUR\_CEU & 293299 & 5483 & 327 & 4486 &  &  &  & 1 &  &  &  &  \\ 
-  EUR\_FIN & 2 &  &  & 116 &  &  &  &  &  &  &  &  \\ 
-  EUR\_GBR & 91862 & 4881 & 99 & 765 &  &  &  &  &  &  & 1 &  \\ 
-  EUR\_IBS & 32 &  & 4 & 481 &  &  &  &  &  &  &  &  \\ 
-  EUR\_TSI & 20 &  & 1 & 166 &  &  &  &  &  &  &  &  \\ 
+EUR\_CEU & 323808 & 6547 & 348 & 4904 &  &  &  & 1 &  &  &  &  \\ 
+  EUR\_FIN & 3 &  &  & 115 &  &  &  &  &  &  &  &  \\ 
+  EUR\_GBR & 62059 & 3682 & 79 & 482 &  &  &  &  &  &  & 1 &  \\ 
+  EUR\_IBS & 35 &  & 5 & 494 &  &  &  &  &  &  &  &  \\ 
+  EUR\_TSI & 20 &  & 1 & 156 &  &  &  &  &  &  &  &  \\ 
    \hline
-SAS\_BEB &  &  &  &  & 12 &  & 50 &  &  &  &  &  \\ 
-  SAS\_GIH &  &  &  &  & 302 &  &  &  &  &  &  &  \\ 
-  SAS\_ITU & 1 &  &  &  & 307 & 3 &  &  & 251 & 3 &  &  \\ 
-  SAS\_PJL &  &  &  &  & 1452 & 712 &  &  & 82 & 1 &  &  \\ 
-  SAS\_STU &  &  &  &  & 42 &  &  &  & 172 &  &  &  \\ 
+SAS\_BEB &  &  &  &  & 10 &  & 41 &  &  &  &  &  \\ 
+  SAS\_GIH &  &  &  &  & 292 &  &  &  &  &  &  &  \\ 
+  SAS\_ITU & 1 &  &  &  & 270 & 1 &  &  & 224 & 2 &  &  \\ 
+  SAS\_PJL &  &  &  &  & 1406 & 641 &  &  & 82 & 2 &  &  \\ 
+  SAS\_STU &  &  &  &  & 39 &  &  &  & 198 &  &  &  \\ 
    \hline
@@
 \noindent[INCLUDE RESULTS FOR SIMPLE PROJ? -> HIST PVAL BAD -> 38\% NOT MATCHED, ESP PAKISTANI]
+
+We also projected individuals using the OADP projection and then applied our ancestry detection technique to other datasets. 
+When applied to the European individuals of the POPRES data \cite[]{nelson2008population}, we show that 56.5\% of individuals could not be matched to one of the 26 populations of the 1000G data (Table \ref{tab:ancestry-pred-popres}).
+This is particularly dramatic for all East-Europeans, whose only 11 people could be matched out of 179 (6.1\%).
+Moreover, even though 1000G data include several ``Toscani in Italia'' (EUR\_TSI), we were able to match 14.2\% of Italians only.
+Nevertheless, for individuals that we could match, they were all identified as European ancestry.
+We could also identify accurately sub-regions of Europe; e.g.\ all Spanish and Portugese individuals that we could match (73.1\%) were identified as ``Iberian Population in Spain'' (EUR\_IBS, Table \ref{tab:ancestry-pred-popres}).
+Finally, we also projected and matched people from a case-control cohort for celiac disease composed of four European countries: Italy, the Netherlands, the UK and Finland \cite[]{dubois2010multiple}.
+Among UK individuals, 5397 (79.9\%) were matched to either ``Utah Residents with Northern and Western European Ancestry'' (EUR\_CEU) or ``British in England and Scotland'' (EUR\_GBR), 10 (0.2\%) to other European populations, 1 to ``Mexican Ancestry from Los Angeles USA'' (AMR\_MXL) and 1346 (19.9\%) could not be matched (Table \ref{tab:ancestry-pred-celiac}).
+Among Finns, 1543 (62.4\%) were matched to ``Finnish in Finland'' (EUR\_FIN), 6 (0.3\%) to other European populations, and 922 (37.3\%) could not be matched.
+Among Italians, 229 (22\%) were matched to EUR\_TSI, 15 (1.5\%) to EUR\_IBS and 795 (76.5\%) could not be matched (Table \ref{tab:ancestry-pred-celiac}).
 

### 2019-10-19 code/project-POPRES-to-1000G.R

@@ -36,7 +36,6 @@ plot_grid(plotlist = lapply(1:8, function(k) {
   )
 }), nrow = 4)
 
-# ggsave("figures/proj1000G-Celiac.png", width = 11, height = 7)
 
 #### Ancestry estimation ####
 
@@ -72,7 +71,6 @@ ggplot() +
   theme_bigstatsr() +
   labs(x = "Maximum p-value testing population belonging")
 
-# ggsave("figures/hist-pval-max.pdf", width = 8.5, height = 5.5)
 
 pop <- obj.bed$fam$family.ID
 pop2 <- dplyr::case_when(

### 2019-10-19 code/project-celiac-to-1000G.R

@@ -38,6 +38,7 @@ plot_grid(plotlist = lapply(1:8, function(k) {
 
 # ggsave("figures/proj1000G-Celiac.png", width = 11, height = 7)
 
+
 #### Ancestry estimation ####
 
 fam2 <- bigreadr::fread2("tmp-data/1000G_phase3_common_norel.fam2")
@@ -71,8 +72,6 @@ ggplot() +
   geom_vline(xintercept = 0.05, color = "red") +
   theme_bigstatsr() +
   labs(x = "Maximum p-value testing population belonging")
-
-# ggsave("figures/hist-pval-max.pdf", width = 8.5, height = 5.5)
 
 
 # Get population from external files

### 2019-10-19 paper-bedpca.tex
save

@@
 We did not infer ancestry for 15.5\% of UK Biobank that had a maximum p-value smaller than 0.05 (Figure \ref{fig:ancestry-pval}). 
-More precisely, among ``White'', ``British'' and ``Irish'' ancestries, this represented respectively 20.4\%, 10.5\% and 19.8\%, while this represented between 48.5\% (``Chinese'') and 81.4\% (``Bangladeshi'').
+More precisely, among ``White'', ``British'' and ``Irish'' ancestries, this represented respectively 20.4\%, 10.5\% and 19.8\%, while this represented between 48.5\% (``Chinese'') and 81.4\% (``Bangladeshi'') for other populations.
 In other words, most of the UK Biobank individual who self-reports as ``Bangladeshi'' could not be attributed to any 1000G population, although 1000G data includes several ``Bengali in Bangladesh'' (SAS\_BEB).
@@
 % latex table generated in R 3.6.0 by xtable 1.8-4 package
-% Sat Oct 19 09:45:04 2019
+% Sat Oct 19 22:53:58 2019
 \begin{table}[ht]
@@
 \label{tab:ancestry-pred}
-\begin{tabular}{|l|c|c|c|c|c|c|c|c|c|c|c|c|}
+\begin{tabular}{|l|c|c|c|c|c|c|c|c|c|c|c|c|c|}
   \hline
- & British & Irish & White & Other White & Indian & Pakistani & Bangladeshi & Chinese & Other Asian & Caribbean & African & Other Black \\ 
+ & British & Irish & White & Other White & Indian & Pakistani & Bangladeshi & Chinese & Other Asian & Caribbean & African & Other Black & NA \\ 
   \hline
-AFR\_ACB &  &  &  &  &  &  &  &  &  & 1687 & 580 & 28 \\ 
-  AFR\_ASW &  &  &  &  &  &  &  &  &  & 365 & 12 & 12 \\ 
-  AFR\_ESN &  &  &  &  &  &  &  &  &  &  & 103 &  \\ 
-  AFR\_GWD &  &  &  &  &  &  &  &  &  &  & 12 &  \\ 
-  AFR\_LWK &  &  &  &  &  &  &  &  &  &  & 33 &  \\ 
-  AFR\_MSL &  &  &  &  &  &  &  &  &  &  & 23 &  \\ 
-  AFR\_YRI &  &  &  &  &  &  &  &  &  & 4 & 403 & 3 \\ 
+AFR\_ACB &  &  &  &  &  &  &  &  &  & 1687 & 580 & 28 & 252 \\ 
+  AFR\_ASW &  &  &  &  &  &  &  &  &  & 365 & 12 & 12 & 41 \\ 
+  AFR\_ESN &  &  &  &  &  &  &  &  &  &  & 103 &  & 16 \\ 
+  AFR\_GWD &  &  &  &  &  &  &  &  &  &  & 12 &  & 5 \\ 
+  AFR\_LWK &  &  &  &  &  &  &  &  &  &  & 33 &  & 6 \\ 
+  AFR\_MSL &  &  &  &  &  &  &  &  &  &  & 23 &  & 2 \\ 
+  AFR\_YRI &  &  &  &  &  &  &  &  &  & 4 & 403 & 3 & 66 \\ 
    \hline
-AMR\_CLM & 1 &  &  & 47 &  &  &  &  &  &  &  &  \\ 
-  AMR\_MXL &  &  & 1 & 3 &  &  &  &  &  &  &  &  \\ 
-  AMR\_PEL &  &  &  & 1 &  &  &  &  &  &  &  &  \\ 
+AMR\_CLM & 1 &  &  & 47 &  &  &  &  &  &  &  &  & 127 \\ 
+  AMR\_MXL &  &  & 1 & 3 &  &  &  &  &  &  &  &  & 21 \\ 
+  AMR\_PEL &  &  &  & 1 &  &  &  &  &  &  &  &  & 14 \\ 
    \hline
-EAS\_CDX &  &  &  &  &  &  &  & 1 &  &  &  &  \\ 
-  EAS\_CHB & 1 &  &  &  &  &  &  & 369 & 10 &  &  &  \\ 
-  EAS\_CHS &  &  &  &  &  &  &  & 398 & 8 &  &  &  \\ 
-  EAS\_JPT &  &  &  &  &  &  &  & 2 & 23 &  &  &  \\ 
-  EAS\_KHV &  &  &  &  &  &  &  & 3 & 14 &  &  &  \\ 
+EAS\_CDX &  &  &  &  &  &  &  & 1 &  &  &  &  &  \\ 
+  EAS\_CHB & 1 &  &  &  &  &  &  & 369 & 10 &  &  &  & 9 \\ 
+  EAS\_CHS &  &  &  &  &  &  &  & 398 & 8 &  &  &  & 18 \\ 
+  EAS\_JPT &  &  &  &  &  &  &  & 2 & 23 &  &  &  & 114 \\ 
+  EAS\_KHV &  &  &  &  &  &  &  & 3 & 14 &  &  &  & 19 \\ 
    \hline
-EUR\_CEU & 323808 & 6547 & 348 & 4904 &  &  &  & 1 &  &  &  &  \\ 
-  EUR\_FIN & 3 &  &  & 115 &  &  &  &  &  &  &  &  \\ 
-  EUR\_GBR & 62059 & 3682 & 79 & 482 &  &  &  &  &  &  & 1 &  \\ 
-  EUR\_IBS & 35 &  & 5 & 494 &  &  &  &  &  &  &  &  \\ 
-  EUR\_TSI & 20 &  & 1 & 156 &  &  &  &  &  &  &  &  \\ 
+EUR\_CEU & 323808 & 6547 & 348 & 4904 &  &  &  & 1 &  &  &  &  & 1358 \\ 
+  EUR\_FIN & 3 &  &  & 115 &  &  &  &  &  &  &  &  & 1 \\ 
+  EUR\_GBR & 62059 & 3682 & 79 & 482 &  &  &  &  &  &  & 1 &  & 268 \\ 
+  EUR\_IBS & 35 &  & 5 & 494 &  &  &  &  &  &  &  &  & 10 \\ 
+  EUR\_TSI & 20 &  & 1 & 156 &  &  &  &  &  &  &  &  & 3 \\ 
    \hline
-SAS\_BEB &  &  &  &  & 10 &  & 41 &  &  &  &  &  \\ 
-  SAS\_GIH &  &  &  &  & 292 &  &  &  &  &  &  &  \\ 
-  SAS\_ITU & 1 &  &  &  & 270 & 1 &  &  & 224 & 2 &  &  \\ 
-  SAS\_PJL &  &  &  &  & 1406 & 641 &  &  & 82 & 2 &  &  \\ 
-  SAS\_STU &  &  &  &  & 39 &  &  &  & 198 &  &  &  \\ 
+SAS\_BEB &  &  &  &  & 10 &  & 41 &  &  &  &  &  & 4 \\ 
+  SAS\_GIH &  &  &  &  & 292 &  &  &  &  &  &  &  & 3 \\ 
+  SAS\_ITU & 1 &  &  &  & 270 & 1 &  &  & 224 & 2 &  &  & 69 \\ 
+  SAS\_PJL &  &  &  &  & 1406 & 641 &  &  & 82 & 2 &  &  & 147 \\ 
+  SAS\_STU &  &  &  &  & 39 &  &  &  & 198 &  &  &  & 47 \\ 
+   \hline
+NA & 45101 & 2526 & 111 & 9614 & 3699 & 1106 & 180 & 730 & 1188 & 2237 & 2037 & 75 & 7051 \\ 
    \hline

### 2019-10-19 code/project-me-to-POPRES.R

@@ -1,88 +1,0 @@
-library(bigsnpr)
-library(ggplot2)
-
-plink <- download_plink("tmp-data")
-mygenome <- "~/Bureau/thesis/backingfiles/genome_Florian_Prive_v5_Full_20190212023418.txt"
-
-system(glue::glue(
-  "{plink} --23file {mygenome}",
-  " --autosome --geno 0 --snps-only",
-  " --a2-allele {sub_bed(download_1000G('tmp-data'), '.bim')} 5 2",
-  " --make-bed --out tmp-data/mygenome"
-))
-
-obj.bed <- bed("tmp-data/mygenome.bed")
-bed.ref <- bed("~/Bureau/POPRES_data/POPRES_allchr_QC.bed")
-
-system.time(
-  test <- bed_projectPCA(bed.ref, obj.bed, k = 10, ncores = nb_cores(),
-                         build.ref = "hg18", liftOver = "tmp-data/liftOver",
-                         match.min.prop = 0, roll.size = 20)
-) # using 18,764 variants only
-
-PC.ref <- predict(test$obj.svd.ref)
-proj2 <- test$OADP_proj
-
-
-#### Ancestry estimation ####
-
-pop <- bed.ref$fam$family.ID
-pop2 <- dplyr::case_when(
-  pop %in% c("Portugal", "Spain") ~ "SW Europe",
-  pop %in% c("?orway", "Sweden", "Finland", "Denmark") ~ "Scandinavia",
-  pop %in% c("Hungary", "Slovakia", "Czech", "Austria", "Slovenia",
-             "Croatia") ~ "Central Europe",
-  pop %in% c("Russian", "Latvia", "Ukraine", "Poland") ~ "Eastern Europe",
-  pop %in% c("Greece", "Turkey", "Serbia", "Cyprus", "Albania", "Kosovo",
-             "Bosnia", "Macedonia,",  "Romania", "Bulgaria") ~ "SE Europe",
-  pop %in% c("Swiss-German", "Swiss-French", "Swiss-Italian") ~ "Switzerland",
-  pop == "?etherlands" ~ "Netherlands",
-  pop %in% c("United", "Scotland", "Ireland") ~ "Anglo-Irish Isles",
-  TRUE ~ pop
-)
-
-plot_grid(plotlist = lapply(1:2, function(k) {
-  k1 <- 2 * k - 1
-  k2 <- 2 * k
[…]

### 2019-10-19 code/project-us-to-POPRES.R

@@ -1,0 +1,102 @@
+library(bigsnpr)
+library(dplyr)
+library(ggplot2)
+
+bedfiles <-
+  sapply(c("~/Bureau/thesis/backingfiles/genome_Florian_Prive_v5_Full_20190212023418.txt",
+           "tmp-data/genome_B_V_v5_Full_20191017230843.txt"),
+         function(genome) {
+           name <- sub("\\.txt$", "", basename(genome))
+           genome <- bigreadr::fread2(genome)
+           bim <- bigreadr::fread2(sub_bed(download_1000G('tmp-data'), '.bim'))
+           genome2 <- genome %>%
+             tidyr::separate(genotype, c("g1", "g2"), sep = 1) %>%
+             mutate(chromosome = as.integer(chromosome)) %>%
+             left_join(bim, by = c(chromosome = "V1", position = "V4")) %>%
+             mutate_at(c("g1", "g2"), ~ {
+               ifelse(. %in% c("A", "T", "C", "G"), . == V5, NA)
+             })
+           big_snp <- snp_fake(1, nrow(genome2))
+           big_snp$map[c(1:2, 4:6)] <- genome2[c(2, 1, 3, 8:9)]
+           big_snp$fam$sample.ID <- name
+           g <- genome2$g1 + genome2$g2
+           big_snp$genotypes[] <- ifelse(is.na(g), 3, g)
+           snp_writeBed(big_snp, tempfile(fileext = ".bed"))
+         })
+
+write(sapply(bedfiles, sub_bed), tmp <- tempfile(), ncolumns = 1)
+plink <- download_plink("tmp-data")
+
+system(glue::glue(
+  "{plink} --merge-list {tmp}",
+  " --autosome --geno 0",
+  " --make-bed --out tmp-data/our_genome"
+))
+
+obj.bed <- bed("tmp-data/our_genome.bed")
+bed.ref <- bed("~/Bureau/POPRES_data/POPRES_allchr_QC.bed")
+
+system.time(
+  test <- bed_projectPCA(bed.ref, obj.bed, k = 10, ncores = nb_cores(),
+                         build.ref = "hg18", liftOver = "tmp-data/liftOver",
+                         match.min.prop = 0, roll.size = 20)
+) # using 18,764 variants only
+
+PC.ref <- predict(test$obj.svd.ref)
+proj2 <- test$OADP_proj
[…]

### 2019-10-21 code/download-ukbb.R
save PCA UKBB

@@ -1,0 +1,61 @@
+wd <- setwd("data/ukbb_bed")
+for (chr in 1:22) {
+  # bed
+  system(paste0("./ukbgene cal -c", chr))
+  # fam
+  system(paste0("./ukbgene cal -c", chr, " -m"))
+}
+# bim
+download.file("https://biobank.ctsu.ox.ac.uk/crystal/crystal/auxdata/ukb_snp_bim.tar", "ukb_snp_bim.tar")
+untar("ukb_snp_bim.tar")
+
+write(sapply(1:22, function(chr) {
+  c(paste0("ukb_cal_chr", chr, "_v2.bed"),
+    paste0("ukb_snp_chr", chr, "_v2.bim"),
+    paste0("ukb25589_cal_chr", chr, "_v2_s488282.fam"))
+}), tmp <- tempfile(), ncolumns = 3)
+
+library(bigsnpr)
+snp_plinkQC(
+  plink.path = download_plink("."),
+  prefix.in = tmp,
+  file.type = "--merge-list",
+  prefix.out = "ukbb_488282",
+  geno = 0.01,
+  autosome.only = TRUE
+)
+# PLINK v1.90b6.10 64-bit (17 Jun 2019)          www.cog-genomics.org/plink/1.9/
+# (C) 2005-2019 Shaun Purcell, Christopher Chang   GNU General Public License v3
+# Logging to ukbb_488282.log.
+# Options in effect:
+#   --autosome
+#   --geno 0.01
+#   --hwe 1e-50
+#   --maf 0.01
+#   --make-bed
+#   --merge-list /tmp/RtmpHoVtFZ/file32bc56c9faf9b
+#   --mind 0.1
+#   --out ukbb_488282
+#
+# 513440 MB RAM detected; reserving 256720 MB for main workspace.
+# Performing single-pass merge (488377 people, 784256 variants).
+# Merged fileset written to ukbb_488282-merge.bed + ukbb_488282-merge.bim +
+#   ukbb_488282-merge.fam .
+# 784256 variants loaded from .bim file.
+# 488377 people (223474 males, 264808 females, 95 ambiguous) loaded from .fam.
+# Ambiguous sex IDs written to ukbb_488282.nosex .
+# 6 people removed due to missing genotype data (--mind).
+# IDs written to ukbb_488282.irem .
+# Using 1 thread (no multithreaded calculations invoked).
+# Before main variant filters, 488371 founders and 0 nonfounders present.
+# Calculating allele frequencies... done.
[…]

### 2019-10-21 paper-bedpca.tex
save

@@
 Another example: when projecting the PCA computed on the 1000 Genomes project, although PC1 to PC4 suffer from only moderate shrinkage (1.00-1.08), PC5 for example suffers from a shrinkage factor of 1.51 and PC10 from a shrinkage factor of 3.16.
-We provide a fast way to project new individuals that is mostly free of this shrinkage bias.
+We provide a fast way to project new individuals and that is not affected by this shrinkage bias.
 We also show how to restrict to individuals that are genetically homogeneous based on PCA.
@@
 In this paper, for many applications, we use the pairwise orthogonalized Gnanadesikan-Kettenrin robust Mahalanobis distance as implemented in function \texttt{covRob} of R package robust with parameter \texttt{estim = "pairwiseGK"} \cite[]{gnanadesikan1972robust,yohai1988high,maronna2002robust,todorov2009object}.
-We reexport this function in R package bigutilsr for convenience.
+We reexport function \texttt{covRob} in R package bigutilsr for convenience.
 
@@
 For detecting outlier variants in PCA that are due to long-range Linkage Disequilibrium (LD) regions, we use a slightly modified version of the procedure described by \cite{prive2017efficient}.
-To summarise the contribution of each variant in all PC loadings, we compute the robust Mahalanobis distances (Section \ref{maha}) of PC loadings.  
-This captures both population-differentiating variants as well as long-range LD regions. 
-Then, to capture consecutive outliers that corresponds to long-range LD regions, we apply a Gaussian smoothing to these statistics.
+We first apply a first round of clumping, prioritising variants by higher minor allele count, at e.g.\ $r^2$ > 0.2.
+Then, we compute $K$ PC scores and loadings.
+To summarise the contribution of each variant in all $K$ PC loadings, we compute the robust Mahalanobis distances (Section \ref{maha}) of these PC loadings.  
+This captures both population-differentiating variants as well as long-range LD regions; to capture consecutive outliers that corresponds to long-range LD regions, we apply a Gaussian smoothing to these statistics.
 
@@
 To solve these two issues, we implement \texttt{tukey\_mc\_up} in R package bigutilsr and use it here, which accounts both for skewness and multiple testing by default.
+
+We remove the detected outlier variants, compute the PC scores and loadings again, and iterate until there is no detected outlier variant anymore.
+This procedure is implemented in function \texttt{bed\_autoSVD} or R package bigsnpr.
 
@@
 
+\noindent[TODO SECTION ON SAMPLE OUTLIER REMOVAL]
+
+\noindent[TALK ABOUT EFFECT OF MISSING VALUES ON PCA?]
+
+\noindent[TALK ABOUT REMOVING OF LRLDR?]
+
+\subsection{Performance}
+
+To show that we provide very performant implementations of the different methods presented in this paper, we apply them to the UK Biobank. We use 20 physical cores for most of the computations.
+It takes 22 minutes to perform a first phase of clumping on 488,371 individuals genotyped over 504,139 variants, which reduces the number of variants to 261,307.
+It then takes 34 minutes to compute the first 20 PCs using these 261,307 variants.
+If performing the automatic procedure for LD detection, it takes 5 hours to perform the initial clumping step, 6 computation of PCs and 5 outlier variant detection (i.e.\ 5 iterations of outlier detection and one final computation of PCs).
+
+It takes 8 minutes to match the 1000G data to the UKBB data and compute 20 PCs of the 1000G data using the automatic outlier detection technique.
+It takes 12 minutes more to do the OADP projection of all 488,371 individuals of the UKBB data to the PCA computed using the 1000G data.
+Once projections are available, et takes only a couple minutes to perform the ancestry estimation.
+
+[TALK ABOUT KNN]
+
 \subsection{Projecting PCs from a reference dataset}
@@
 % latex table generated in R 3.6.0 by xtable 1.8-4 package
-% Sat Oct 19 22:53:58 2019
+% Mon Oct 21 15:14:34 2019
 \begin{table}[ht]
@@
   \hline
-AFR\_ACB &  &  &  &  &  &  &  &  &  & 1687 & 580 & 28 & 252 \\ 
-  AFR\_ASW &  &  &  &  &  &  &  &  &  & 365 & 12 & 12 & 41 \\ 
-  AFR\_ESN &  &  &  &  &  &  &  &  &  &  & 103 &  & 16 \\ 
-  AFR\_GWD &  &  &  &  &  &  &  &  &  &  & 12 &  & 5 \\ 
-  AFR\_LWK &  &  &  &  &  &  &  &  &  &  & 33 &  & 6 \\ 
-  AFR\_MSL &  &  &  &  &  &  &  &  &  &  & 23 &  & 2 \\ 
-  AFR\_YRI &  &  &  &  &  &  &  &  &  & 4 & 403 & 3 & 66 \\ 
+AFR\_ACB &  &  &  &  &  &  &  &  &  & 1832 & 580 & 31 & 259 \\ 
+  AFR\_ASW &  &  &  &  &  &  &  &  &  & 315 & 13 & 10 & 39 \\ 
+  AFR\_ESN &  &  &  &  &  &  &  &  &  &  & 155 & 1 & 29 \\ 
+  AFR\_GWD &  &  &  &  &  &  &  &  &  &  & 20 &  & 4 \\ 
+  AFR\_LWK &  &  &  &  &  &  &  &  &  &  & 34 &  & 4 \\ 
+  AFR\_MSL &  &  &  &  &  &  &  &  &  &  & 26 &  &  \\ 
+  AFR\_YRI &  &  &  &  &  &  &  &  &  & 5 & 469 & 3 & 81 \\ 
    \hline
-AMR\_CLM & 1 &  &  & 47 &  &  &  &  &  &  &  &  & 127 \\ 
-  AMR\_MXL &  &  & 1 & 3 &  &  &  &  &  &  &  &  & 21 \\ 
-  AMR\_PEL &  &  &  & 1 &  &  &  &  &  &  &  &  & 14 \\ 
+AMR\_CLM & 1 &  &  & 50 &  &  &  &  &  &  &  &  & 142 \\ 
+  AMR\_MXL &  &  & 1 & 3 &  &  &  &  &  &  &  &  & 20 \\ 
+  AMR\_PEL &  &  &  & 2 &  &  &  &  &  &  &  &  & 19 \\ 
    \hline
-EAS\_CDX &  &  &  &  &  &  &  & 1 &  &  &  &  &  \\ 
-  EAS\_CHB & 1 &  &  &  &  &  &  & 369 & 10 &  &  &  & 9 \\ 
-  EAS\_CHS &  &  &  &  &  &  &  & 398 & 8 &  &  &  & 18 \\ 
-  EAS\_JPT &  &  &  &  &  &  &  & 2 & 23 &  &  &  & 114 \\ 
-  EAS\_KHV &  &  &  &  &  &  &  & 3 & 14 &  &  &  & 19 \\ 
+EAS\_CDX &  &  &  &  &  &  &  & 2 &  &  &  &  & 1 \\ 
+  EAS\_CHB & 1 &  &  &  &  &  &  & 445 & 10 &  &  &  & 7 \\ 
+  EAS\_CHS &  &  &  &  &  &  &  & 380 & 9 &  &  &  & 21 \\ 
+  EAS\_JPT &  &  &  &  &  &  &  & 3 & 25 &  &  &  & 122 \\ 
+  EAS\_KHV &  &  &  &  &  &  &  & 4 & 16 &  &  &  & 22 \\ 
    \hline
-EUR\_CEU & 323808 & 6547 & 348 & 4904 &  &  &  & 1 &  &  &  &  & 1358 \\ 
-  EUR\_FIN & 3 &  &  & 115 &  &  &  &  &  &  &  &  & 1 \\ 
-  EUR\_GBR & 62059 & 3682 & 79 & 482 &  &  &  &  &  &  & 1 &  & 268 \\ 
-  EUR\_IBS & 35 &  & 5 & 494 &  &  &  &  &  &  &  &  & 10 \\ 
-  EUR\_TSI & 20 &  & 1 & 156 &  &  &  &  &  &  &  &  & 3 \\ 
+EUR\_CEU & 325725 & 6675 & 350 & 4855 &  &  &  & 1 &  &  &  &  & 1363 \\ 
+  EUR\_FIN & 2 &  &  & 118 &  &  &  &  &  &  &  &  & 1 \\ 
+  EUR\_GBR & 57497 & 3486 & 79 & 454 &  &  &  &  &  &  & 1 &  & 244 \\ 
+  EUR\_IBS & 28 &  & 4 & 452 &  &  &  &  &  &  &  &  & 8 \\ 
+  EUR\_TSI & 25 &  & 1 & 177 &  &  &  &  &  &  &  &  & 5 \\ 
    \hline
-SAS\_BEB &  &  &  &  & 10 &  & 41 &  &  &  &  &  & 4 \\ 
-  SAS\_GIH &  &  &  &  & 292 &  &  &  &  &  &  &  & 3 \\ 
-  SAS\_ITU & 1 &  &  &  & 270 & 1 &  &  & 224 & 2 &  &  & 69 \\ 
-  SAS\_PJL &  &  &  &  & 1406 & 641 &  &  & 82 & 2 &  &  & 147 \\ 
-  SAS\_STU &  &  &  &  & 39 &  &  &  & 198 &  &  &  & 47 \\ 
+SAS\_BEB &  &  &  &  & 11 &  & 57 &  &  &  &  &  & 7 \\ 
+  SAS\_GIH &  &  &  &  & 334 &  &  &  & 1 &  &  &  & 4 \\ 
+  SAS\_ITU & 1 &  &  &  & 314 & 1 &  &  & 264 & 3 &  &  & 82 \\ 
+  SAS\_PJL & 1 &  &  &  & 1749 & 767 &  &  & 104 & 2 &  &  & 177 \\ 
+  SAS\_STU &  &  &  &  & 38 &  &  &  & 185 &  &  &  & 40 \\ 
    \hline
-NA & 45101 & 2526 & 111 & 9614 & 3699 & 1106 & 180 & 730 & 1188 & 2237 & 2037 & 75 & 7051 \\ 
+NA & 47748 & 2594 & 110 & 9705 & 3270 & 980 & 164 & 669 & 1133 & 2140 & 1906 & 73 & 6970 \\ 
    \hline
@@
 
-\noindent[LIMITATION: RUN MULTIPLE TIMES / FAIL TO SPEED UP]
+\noindent[LIMITATION 1: RUN MULTIPLE TIMES / FAIL TO SPEED UP]
 
@@
 \noindent[LIMITATION 3: FAIL TO AUTOMATICALLY DETECT OUTLIER SAMPLES.]
+
+\noindent[DISCUSS LACK OF CERTAIN POPULATIONS IN 1000G]
 

### 2019-10-21 paper-bedpca.tex

@@
 We performed ancestry estimation of the individuals in the UK Biobank using the 1000G data (Section \ref{ancestry}), discarding individuals with unknown ancestry or mixed ancestry.
-We did not infer ancestry for 15.5\% of UK Biobank that had a maximum p-value smaller than 0.05 (Figure \ref{fig:ancestry-pval}). 
-More precisely, among ``White'', ``British'' and ``Irish'' ancestries, this represented respectively 20.4\%, 10.5\% and 19.8\%, while this represented between 48.5\% (``Chinese'') and 81.4\% (``Bangladeshi'') for other populations.
-In other words, most of the UK Biobank individual who self-reports as ``Bangladeshi'' could not be attributed to any 1000G population, although 1000G data includes several ``Bengali in Bangladesh'' (SAS\_BEB).
+We used from PC1 to PC14 and PC16 based on visual inspection of PC scores (Figure \ref{fig:PC-1000G-UKBB}).
+We did not infer ancestry for 15.9\% of UK Biobank that had a maximum p-value smaller than 0.05 (Figure \ref{fig:ancestry-pval}). 
+More precisely, among ``White'', ``British'' and ``Irish'' ancestries, this represented respectively 20.2\%, 11.1\% and 20.3\%, while this represented between 44.5\% (``Chinese'') and 74.3\% (``Bangladeshi'') for other populations.
+In other words, 3 out of 4 of the UK Biobank individual who self-reports as ``Bangladeshi'' could not be attributed to any 1000G population, although 1000G data includes several ``Bengali in Bangladesh'' (SAS\_BEB).
 Only 10 people out of 401,048 were wrongly classified in ``super'' population of the 1000G; e.g.\ one Chinese person was classified as European by our method (Table \ref{tab:ancestry-pred}). 
@@
 \begin{figure}[!htpb]
+\centerline{\includegraphics[width=0.8\textwidth]{PC-1000G-UKBB.pdf}}
+\caption{Principal Component (PC) scores 1 to 20 of the 1000 Genomes (1000G) project using variants matched with the UK Biobank. Colors represent populations of the 1000G data.
+\label{fig:PC-1000G-UKBB}}
+\end{figure}
+
+\vspace*{1em}
+
+\begin{figure}[!htpb]
 \centerline{\includegraphics[width=0.8\textwidth]{hist-pval-max.pdf}}

### 2019-10-21 paper-bedpca.tex
save PCA UKBB

@@
 
+[REDO]
+
 We propose [BOF] different outlier detection techniques for PCA.
@@
 [TALK ABOUT KNN]
+
+\subsection{PCA of the UK Biobank}
+
+Principal Components (PC) loadings reported by the UK Biobank clearly show that PC19 to PC40 capture LD structure, which is also the case for PC16 and PC18, although less pronounced (see peaks in figure \ref{fig:UKBB-loadings40}).
+Moreover, our automatic procedure removing long-range LD regions did not converge after 5 iterations, meaning that it kept detecting long-range LD regions at each iteration.
+Therefore, we were able to capture only 16 PCs that shows population stratification and not LD structure (Figures \ref{fig:UKBB-screeplot}, \ref{fig:UKBB-scores} and \ref{fig:UKBB-loadings}).
+
+\noindent[SHOW INITIAL PC SCORES OF UKBB?]
+
+\noindent[COMPUTE MAHA ON LOADINGS TO SHOW THE PEAKS SUMMARIZED IN ALL 16-18 PCS?]
 
@@
 
+\begin{figure}[!htpb]
+\centerline{\includegraphics[width=0.9\textwidth]{UKBB-loadings1-40.pdf}}
+\caption{Principal Component (PC) loadings 1 to 40 reported by the UK Biobank.
+\label{fig:UKBB-loadings40}}
+\end{figure}
+
+\begin{figure}[!htpb]
+\centerline{\includegraphics[width=0.8\textwidth]{UKBB-screeplot.pdf}}
+\caption{Scree plot: plot of singular values computed from the UK Biobank.
+\label{fig:UKBB-screeplot}}
+\end{figure}
+
+\begin{figure}[!htpb]
+\centerline{\includegraphics[width=0.9\textwidth]{UKBB-PC1-20.png}}
+\caption{Principal Component (PC) scores 1 to 20 computed from the UK Biobank.
+\label{fig:UKBB-scores}}
+\end{figure}
+
+\begin{figure}[!htpb]
+\centerline{\includegraphics[width=0.9\textwidth]{UKBB-loadings.pdf}}
+\caption{Principal Component (PC) loadings 1 to 20 computed from the UK Biobank.
+\label{fig:UKBB-loadings}}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \vspace*{1em}
@@
 \end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2019-10-21 paper-bedpca.tex
save

@@
 
+[FIGURE TO ILLUSTRATE THIS?]
+
 Finally, to choose the threshold of outlierness based on the previously described statistics, we use a modified version of Tukey's rule, a standard rule for detecting outliers \cite[]{tukey77}. 
@@
 
+[SHOULD DESCRIBE QC FOR OTHER DATASETS TOO?]
+
 \subsection{Projecting PCs from a reference dataset \label{proj}}
 
-%[SEPARATE STEPS?]
+[SCHEME TO ILLUSTRATE OADP?]
 
@@
 To show that we provide very performant implementations of the different methods presented in this paper, we apply them to the UK Biobank. We use 20 physical cores for most of the computations.
-It takes 22 minutes to perform a first phase of clumping on 488,371 individuals genotyped over 504,139 variants, which reduces the number of variants to 261,307.
+It takes 22 minutes to perform a first phase of clumping on 406,545 unrelated individuals genotyped over 504,139 variants, which reduces the number of variants to 261,307.
 It then takes 34 minutes to compute the first 20 PCs using these 261,307 variants.
@@
 Once projections are available, et takes only a couple minutes to perform the ancestry estimation.
+
+Finally, it took only 6 minutes to compute the 30-Nearest Neighbours of PC scores for all UK Biobank individuals.
 
@@
 \noindent[COMPUTE MAHA ON LOADINGS TO SHOW THE PEAKS SUMMARIZED IN ALL 16-18 PCS?]
+
+\noindent[SHOW THAT PROCEDURE IS OKAY FOR SMALLER DATASETS?]
 

### 2019-10-21 pca-toolkit.Rmd
save

---
title: "PCA Toolkit"
output:
  xaringan::moon_reader:
    df_print: paged
    seal: false
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center', dev = "svg", out.width = "80%",
                      echo = FALSE)
options(width = 80)
```

class: title-slide center middle inverse

<br>

# Efficient toolkit <br> implementing best practices <br> for principal component analysis <br> of population genetic data

<br>

## Florian Privé

<br>

<br>

**Slides:** `https://privefl.github.io/paper4-bedpca/pca-toolkit.html`

---

## Motivation

<br>

Perform PCA analyses in iPSYCH appropriately and quickly.

<br>

--

<br>

## Disclaimer

<br>

Still unfinished work.

---

## Range of analyses

<br>

- Compute PCA on PLINK .bed files (with missing values)

- Perform pruning to reduce effect of Linkage Disequilibrium (LD)

- Further removing of long-range LD regions

- Detection of outlier samples?

- Detection of homogeneous samples -> Danes

- Projection of new individuals onto reference PCA space

- Ancestry matching

---

## Pruning and PCA

<br>

Using 20 physical cores,

- takes 22 minutes to perform a first phase of clumping on 406,545 unrelated individuals genotyped over 504,139 variants, which reduces the number of variants to 261,307.

- then takes 34 minutes to compute the first 20 PCs using these 261,307 variants.

--

<br>

Method:

- Directly memory-map the bed file (access it as if it were a matrix in memory)

- Same PCA algorithm based on random projections, already used in PCA of R package bigstatsr (Privé et al. 2018) and in FlashPCA2 (Abraham et al. 2017).

---

## Problem of LD: PC reported by UK Biobank

```{r, out.width="98%"}
knitr::include_graphics("figures/UKBB-loadings1-40.svg")
```

---

## Problem of LD: PC computed after LD removal

Method: Robust Mahalanobis distance on PC loadings + Gaussian Smoothing = removing long-range LD outliers.

```{r, out.width="98%"}
knitr::include_graphics("figures/UKBB-loadings.svg")
```

---

## Remaining structure: 16 PCs

```{r, out.width="85%"}
knitr::include_graphics("figures/UKBB-PC1-20.png")
```

---

## Detection of outlier samples in 1000G

Method: compare density around point with density around its K-Nearest Neighbours.

```{r}
knitr::include_graphics("figures/outliers-1000G.svg")
```

---

## Restricting to homogeneous individuals

Method: Robust Mahalanobis distance on PC scores (approx. $\chi^2(K)$).

```{r, out.width="85%"}
knitr::include_graphics("figures/homogeneous.svg")
```

---

## Projection of new individuals into reference PCA space

**Issue:** shrinkage bias when projecting new individuals (when p > n).

PCA on 60% of 1000G (**black**) / Projecting on remaining 40% (**red**):

```{r, out.width="95%"}
knitr::include_graphics("figures/proj1000G-PC1-8.svg")
```

Shrinkage factors: `1.01 1.02 1.07 1.09 1.51 1.68 1.94 1.40 2.88 3.17 2.90 2.92 3.23  5.13 5.25 5.04 4.58 5.69 6.26 6.32`.

---

## Projection of new individuals into reference PCA space

**Solution:** Online Augmentation, Decomposition, and Procrustes (OADP) projection (Zhang et al. 2019). We provide faster implementation.

PCA on 60% of 1000G (**black**) / OADP projection on remaining 40% (**blue**):

```{r, out.width="95%"}
knitr::include_graphics("figures/proj1000G-PC1-8.svg")
```

---

## Projection of new individuals into reference PCA space

**Limitation:** Projection on related individuals does not work.

PCA on 60% of 1000G (**black**) / OADP projection on same 60% (**blue**):

```{r, out.width="75%"}
knitr::include_graphics("figures/proj1000G-related.svg")
```

---

## Ancestry matching

Method: Robust Mahalanobis distances by reference population.

```{r}

```

---

class: inverse, center, middle

# Thanks!

<br>

<br>

`r icon::fa("twitter")` [privefl](https://twitter.com/privefl) &nbsp;&nbsp;&nbsp;&nbsp; `r icon::fa("github")` [privefl](https://github.com/privefl) &nbsp;&nbsp;&nbsp;&nbsp; `r icon::fa("stack-overflow")` [F. Privé](https://stackoverflow.com/users/6103040/f-priv%c3%a9)

.footnote[Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan).]

### 2019-10-21 code/outlier-samples.R

@@ -1,0 +1,50 @@
+#### Detection of individual outliers ####
+
+library(bigsnpr)
+
+obj.bed <- bed(download_1000G("tmp-data"))
+obj.svd <- bed_autoSVD(obj.bed, k = 20, ncores = nb_cores())
+
+prob <- bigutilsr::prob_dist(obj.svd$u)
+hist(S <- prob$dist.self / sqrt(prob$dist.nn))
+
+library(ggplot2)
+plot_grid(plotlist = lapply(7:10, function(k) {
+  plot(obj.svd, type = "scores", scores = 2 * k - 1:0, coeff = 0.6) +
+    aes(color = S) +
+    scale_colour_viridis_c()
+}), scale = 0.95)
+
+# ggsave("figures/outliers-1000G.pdf", width = 7, height = 5)
+
+#### Restricting to homogeneous individuals ####
+
+library(bigsnpr)
+
+obj.bed <- bed("~/Bureau/Dubois2010_data/FinnuncorrNLITUK3hap550_QC.bed")
+obj.svd <- bed_autoSVD(obj.bed, k = 10, ncores = nb_cores())
+
+plot(obj.svd, type = "scores", scores = 1:10, coeff = 0.7)
+
+dist <- bigutilsr::covRob(obj.svd$u, estim = "pairwiseGK")$dist
+pval <- pchisq(dist, df = 10, lower.tail = FALSE)
+
+# Get population from external files
+pop.files <- list.files(path = "~/Bureau/Dubois2010_data/",
+                        pattern = "cluster_*", full.names = TRUE)
+pop <- snp_getSampleInfos(obj.bed, pop.files)[[1]]
+pop_celiac <- c("Netherlands", "Italy", "UK", "UK", "Finland")[pop]
+
+library(ggplot2)
+plot_grid(plotlist = lapply(1:2, function(k) {
+  plot_grid(
+    plot(obj.svd, type = "scores", scores = 2 * k - 1:0, coeff = 0.7) +
+      aes(color = pop_celiac),
+    plot(obj.svd, type = "scores", scores = 2 * k - 1:0, coeff = 0.7) +
+      aes(color = pval < 0.001) +
+      scale_colour_viridis_d(),
+    ncol = 1
+  )
+}), scale = 0.95)
+
+# ggsave("figures/homogeneous.pdf", width = 7, height = 5)

### 2019-10-21 pdf2png.R

@@ -1,0 +1,8 @@
+library(magick)
+
+"figures/proj1000G-related.pdf" %>% {
+  svg <- sub("\\.pdf$", ".svg", .)
+  img <- image_read_pdf(.)
+  image_write(img, svg)
+  print(svg)
+}

### 2019-10-21 pca-toolkit.Rmd
save

@@
 
-```{r}
+```{r, out.width="70%"}
+knitr::include_graphics("figures/us.svg")
+```
 
+---
+
+## Ancestry matching
+
+Matching POPRES populations to 1000G populations    
+(do not match maximum p-values < 0.05):
+
+```{r, out.width="90%"}
+knitr::include_graphics("figures/ancestry-popres.png")
+```
+
+---
+
+## Ancestry matching
+
+Matching UKBB populations to 1000G populations (20 minutes)  
+(do not match maximum p-values < 0.05):
+
+```{r, out.width="98%"}
+knitr::include_graphics("figures/ancestry-UKBB.png")
 ```

### 2019-10-21 pdf2png.R

@@ -1,6 +1,6 @@
 library(magick)
 
-"figures/proj1000G-related.pdf" %>% {
+"figures/us.pdf" %>% {
   svg <- sub("\\.pdf$", ".svg", .)
   img <- image_read_pdf(.)
   image_write(img, svg)

### 2019-10-22 pca-toolkit.Rmd
best
save

@@
 
-Perform PCA analyses in iPSYCH appropriately and quickly.
+Perform PCA analyses in iPSYCH quickly and following best practices.
 
@@
 
-- Perform pruning to reduce effect of Linkage Disequilibrium (LD)
+- Perform pruning or clumping to reduce effect of Linkage Disequilibrium (LD)
 
@@
 
+Method:
+
+- Directly memory-map the bed file (access it as if it were a matrix in memory)
+
+- Same PCA algorithm based on random projections, already used in PCA of R package bigstatsr (Privé et al. 2018) and in FlashPCA2 (Abraham et al. 2017).
+
+--
+
+<br>
+
 Using 20 physical cores,
@@
 - then takes 34 minutes to compute the first 20 PCs using these 261,307 variants.
-
---
-
-<br>
-
-Method:
-
-- Directly memory-map the bed file (access it as if it were a matrix in memory)
-
-- Same PCA algorithm based on random projections, already used in PCA of R package bigstatsr (Privé et al. 2018) and in FlashPCA2 (Abraham et al. 2017).
 
@@
 
-Method: Robust Mahalanobis distances by reference population.
+Method: Robust Mahalanobis distance for each reference population    
+\+ minimum distance (or maximum p-value).
 
-```{r, out.width="70%"}
+```{r, out.width="65%"}
 knitr::include_graphics("figures/us.svg")

### 2019-10-22 code/project-1000G-to-UKBB.R
save

@@ -1,0 +1,68 @@
+library(bigsnpr)
+library(bigreadr)
+rel <- fread2("data/ukb25589_rel_s488346.dat")
+fam <- fread2("data/ukbb_bed/ukbb_488282.fam")
+ind.row <- which(!fam$V2 %in% rel$ID2)
+bed.ref <- bed("data/ukbb_bed/ukbb_488282.bed")
+obj.bed <- bed(download_1000G("data"))
+
+system.time(
+  test <- bed_projectPCA(bed.ref, obj.bed, ind.row.ref = ind.row,
+                         k = 20, ncores = nb_cores())
+) # 4.4 H
+length(attr(test$obj.svd.ref, "subset")) # 171977
+
+PC.ref <- predict(test$obj.svd.ref)
+proj1 <- test$simple_proj
+proj2 <- test$OADP_proj
+
+# shrinkage coefficients
+shrinkage <- unname(sapply(1:20, function(k) {
+  MASS::rlm(proj2[, k] ~ proj1[, k] + 0)$coef
+}))
+round(shrinkage, 2)
+#  [1] 1.00 1.00 1.00 1.01 1.01 1.02 1.03 1.03 1.04 1.04 1.04 1.05 1.05
+# [14] 1.06 1.07 1.07 1.08 1.08 1.08 1.08
+
+
+library(ggplot2)
+ind <- sample(nrow(PC.ref), 20e3)
+
+plot_grid(plotlist = lapply(1:8, function(k) {
+  k1 <- 2 * k - 1
+  k2 <- 2 * k
+  plot_grid(
+    qplot(PC.ref[ind, k1], PC.ref[ind, k2], size = I(2)) +
+      geom_point(aes(proj1[, k1], proj1[, k2]), color = "red") +
+      theme_bigstatsr(0.5) +
+      labs(x = paste0("PC", k1), y = paste0("PC", k2)),
+    qplot(PC.ref[ind, k1], PC.ref[ind, k2], size = I(2)) +
+      geom_point(aes(proj2[, k1], proj2[, k2]), color = "blue") +
+      theme_bigstatsr(0.5) +
+      labs(x = paste0("PC", k1), y = paste0("PC", k2)),
+    scale = 0.95
+  )
+}), nrow = 4)
+
+# ggsave("figures/proj1000G-UKBB2.png", width = 10, height = 7)
+
+library(bigreadr)
[…]

### 2019-10-22 paper-bedpca.tex
save

@@
 
-[TALK ABOUT KNN]
-
 \subsection{PCA of the UK Biobank}
@@
 
+\noindent[SHOW ABLE TO RECONSTRUCT PCA? MAYBE NOT PC17]
+
 \subsection{Projecting PCs from a reference dataset}
 
-We use 60\% of individuals in the 1000G data (section \ref{1000G}) to compute the PCA ($K=20$ PCs).
+We use 60\% of individuals in the 1000G data (section \ref{1000G}) to compute the PCA ($K=20$ PCs) [EXPLICIT SAMPLE SIZE].
 Then, we project the remaining 40\% individuals using three methods: 1/ simply multiplying the genotypes of these individuals by the previously computed loadings, 2/ correcting the simple projections using asymptotic shrinkage factors as determined by R package hdpca v1.0.0 \cite[]{dey2019asymptotic} and 3/ the OADP projection (section \ref{proj}).
 When simply projecting using loadings, there is no apparent shrinkage for PC1 and PC2, a weak [REWORD?] shrinkage for PC3 and PC4, and a large shrinkage for PC5 to PC8 (Figure \ref{fig:proj1000G}).
+[USE VALUES?]
 In contrast, there is no apparent shrinkage when projecting new individuals with OADP (Figure \ref{fig:proj1000G}).
@@
 \end{figure}
+
+When computing the PCA on the UK Biobank using 406,545 unrelated individuals and 171,977 variants, and projecting the 1000G data onto this reference PCA space, shrinkage is much smaller ($\leq 1.08$ for all 20 first PCs, Figure \ref{fig:proj1000G-6}).
 
@@
 
+\begin{figure}[!htpb]
+\centerline{\includegraphics[width=0.8\textwidth]{proj1000G-UKBB2.png}}
+\caption{Principal Component (PC) scores 1 to 16 from the UK Biobank and projected individuals of the 1000 Genomes (1000G) project.
+Black points are the UK Biobank individuals used for computing PCA.
+Red points are the individuals from 1000G, projected by simply multiplying their genotypes by the corresponding PC loadings.
+Blue points are the individuals from 1000G, projected using the Online Augmentation, Decomposition, and Procrustes (OADP) transformation.
+\label{fig:proj1000G-6}}
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2019-10-24 code/project-us-to-POPRES.R
add esben

@@ -4,12 +4,18 @@ library(ggplot2)
 
 bedfiles <-
   sapply(c("tmp-data/genome_Florian_Prive_v5_Full_20190212023418.txt",
-           "tmp-data/genome_B_V_v5_Full_20191017230843.txt"),
+           "tmp-data/genome_B_V_v5_Full_20191017230843.txt",
+           "tmp-data/genome_Esben_Agerbo_v3_Full_20190131054436.txt"),
          function(genome) {
-           name <- sub("\\.txt$", "", basename(genome))
+           name <- gsubfn::strapply(
+             basename(genome), "^genome_(.+_.+)_v.+_Full_[0-9]+\\.txt$")[[1]]
            genome <- bigreadr::fread2(genome)
            bim <- bigreadr::fread2(sub_bed(download_1000G('tmp-data'), '.bim'))
            genome2 <- genome %>%
+             filter(chromosome %in% 1:22 &
+                      `# rsid` != "rs4001921",
+                      !vctrs::vec_duplicate_detect(genome[2:3]) &
+                      !vctrs::vec_duplicate_detect(genome[[1]])) %>%
              tidyr::separate(genotype, c("g1", "g2"), sep = 1) %>%
              mutate(chromosome = as.integer(chromosome)) %>%
              left_join(bim, by = c(chromosome = "V1", position = "V4")) %>%
@@ -81,13 +87,21 @@ all_pval <- sapply(pop_PCs$val, function(PC) {
 })
 
 colnames(all_pval) <- pop_PCs$key
-round(100 * all_pval, 2)
-#      Central Europe SE Europe Eastern Europe Anglo-Irish Isles Scandinavia
-# [1,]           0.05         0              0             45.63       85.82
-# [2,]           1.24         0              0              0.22        0.18
-#      Italy Germany SW Europe Switzerland Belgium France Netherlands
-# [1,]  0.00    1.64         0        0.00    0.01   0.06        0.00
-# [2,]  0.01    9.63         0       28.37   30.90  24.60        0.79
+rownames(all_pval) <- obj.bed$fam$sample.ID
+t(round(100 * all_pval, 2))
+#                     B_V Esben_Agerbo Florian_Prive
+# Central Europe     0.10         0.01          4.28
+# SE Europe          0.00         0.00          0.00
[…]

### 2019-10-24 code/project-us-to-POPRES.R
rm file

@@ -1,146 +1,0 @@
-library(bigsnpr)
-library(dplyr)
-library(ggplot2)
-
-bedfiles <-
-  sapply(c("tmp-data/genome_Florian_Prive_v5_Full_20190212023418.txt",
-           "tmp-data/genome_B_V_v5_Full_20191017230843.txt",
-           "tmp-data/genome_Esben_Agerbo_v3_Full_20190131054436.txt"),
-         function(genome) {
-           name <- gsubfn::strapply(
-             basename(genome), "^genome_(.+_.+)_v.+_Full_[0-9]+\\.txt$")[[1]]
-           genome <- bigreadr::fread2(genome)
-           bim <- bigreadr::fread2(sub_bed(download_1000G('tmp-data'), '.bim'))
-           genome2 <- genome %>%
-             filter(chromosome %in% 1:22 &
-                      `# rsid` != "rs4001921",
-                      !vctrs::vec_duplicate_detect(genome[2:3]) &
-                      !vctrs::vec_duplicate_detect(genome[[1]])) %>%
-             tidyr::separate(genotype, c("g1", "g2"), sep = 1) %>%
-             mutate(chromosome = as.integer(chromosome)) %>%
-             left_join(bim, by = c(chromosome = "V1", position = "V4")) %>%
-             mutate_at(c("g1", "g2"), ~ {
-               ifelse(. %in% c("A", "T", "C", "G"), . == V5, NA)
-             })
-           big_snp <- snp_fake(1, nrow(genome2))
-           big_snp$map[c(1:2, 4:6)] <- genome2[c(2, 1, 3, 8:9)]
-           big_snp$fam$sample.ID <- name
-           g <- genome2$g1 + genome2$g2
-           big_snp$genotypes[] <- ifelse(is.na(g), 3, g)
-           snp_writeBed(big_snp, tempfile(fileext = ".bed"))
-         })
-
-write(sapply(bedfiles, sub_bed), tmp <- tempfile(), ncolumns = 1)
-plink <- download_plink("tmp-data")
-
-system(glue::glue(
-  "{plink} --merge-list {tmp}",
-  " --autosome --geno 0",
-  " --make-bed --out tmp-data/our_genome"
-))
-
-obj.bed <- bed("tmp-data/our_genome.bed")
-bed.ref <- bed("~/Bureau/POPRES_data/POPRES_allchr_QC.bed")
[…]

### 2019-10-24 paper-bedpca.tex
add method overview

@@
 \usepackage{geometry}
-\geometry{verbose,letterpaper,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm}
+\geometry{verbose,letterpaper,tmargin=2cm,bmargin=2cm,lmargin=1.5cm,rmargin=1.5cm}
 \usepackage{rotating}
@@
 \usepackage{bbm}
+\usepackage{makecell}
+
+\renewcommand{\arraystretch}{2}
 
@@
 \title{Efficient toolkit implementing best practices for principal component analysis of population genetic data}
-\author{Florian Priv\'e,$^{\text{1,}*}$ John J. McGrath,$^{\text{1,2,3}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,}*}$}
+\author{Florian Priv\'e,$^{\text{1,}*}$ Keurcien Luu, Michael G.B. Blum, John J. McGrath,$^{\text{1,2,3}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,}*}$}
 
@@
 \section{Material and Methods}
+
+\begin{table}[h]
+\centering
+\caption{Overview of existing methods. [TODO: ADD CITATIONS]} 
+\label{tab:method-overview}
+\begin{tabular}{|c|c|c|c|}
+\hline
+Analysis & Method / Software & Pros & Cons \\
+\hline
+\multirow{3}{*}{PCA} & bigstatsr / bigsnpr & \makecell{Fast and accurate + handle dosages \\ + thinning options directly included} & \makecell{Own format without missing values \\ (fast functions are available \\ for converting and imputing)} \\
+ & FlashPCA2 & Fast and accurate & Not parallelised? \\
+ & \makecell{PLINK 2.0 \\ (reimplementation of FastPCA)} & Very fast (AVX2 only?) & Not always 100\% accurate \\
+\hline
+\makecell{Detection of \\ oultier samples} & \makecell{"6 SDs from the mean" \\ (EIGENSOFT)} & Simple & Assume Gaussian distritbution \\
+\hline
+\makecell{Detection of \\ homogeneous \\ samples} & Aberrant & Robust & Only two statistics at once? \\
+\hline
+\multirow{5}{*}{\makecell{Projection of \\ new individuals \\ onto reference \\ PCA space}} & Simple projection (multiplication by loadings) & Simple & Shrinkage biased \\
+  & Bias-adjusted projection & Independent of new samples & \makecell{Assume same shrinkage for all individuals \\ + model-based \\ + need all eigenvalues of reference} \\
+  & \makecell{Augmentation, Decomposition, \\ and Procrustes transformation (ADP)} & Accurate & Slow \\
+  & Online ADP (OADP) & Much faster than OADP & Does not work for related individuals \\ 
+\hline
+\end{tabular}
+\end{table}
 

### 2019-10-25 code/pca-UKBB.R
save UKBB related

@@ -6,8 +6,9 @@ ind.row <- which(!fam$V2 %in% rel$ID2)
 obj.bed <- bed("data/ukbb_bed/ukbb_488282.bed")
 
 system.time(
-  obj.svd <- bed_autoSVD2(obj.bed, ind.row = ind.row, k = 20, ncores = nb_cores())
+  obj.svd <- bed_autoSVD(obj.bed, ind.row = ind.row, k = 20, ncores = nb_cores())
 ) # 5h
+saveRDS(obj.svd, "tmp-results/SVD_UKBB.rds")
 # Phase of clumping (on MAC) at r^2 > 0.2.. keep 261307 variants.
 # Discarding 0 variant with MAC < 10.
 #
@@ -40,7 +41,6 @@ system.time(
 # Computing SVD..
 # Maximum number of iterations reached.
 
-
 system.time(
   ind.keep2 <- bed_clumping(obj.bed, ind.row = ind.row, ncores = nb_cores())
 ) # 22 min
@@ -61,8 +61,44 @@ plot(obj.svd, type = "loadings", loadings = 1:20, coeff = 0.4)
 plot(obj.svd, type = "scores", scores = 1:20, coeff = 0.4, ncol = 5)
 # ggsave("figures/UKBB-PC1-20.png", width = 9, height = 7)
 
-
+# Benchmark KNN
 PCs <- predict(obj.svd)
 system.time(
   knn <- bigutilsr::knn_parallel(PCs, k = 31, ncores = nb_cores())
 ) # 6 min
+
+
+#### Reported by the UK Biobank ####
+
+snp_qc <- bigreadr::fread2("https://biobank.ctsu.ox.ac.uk/crystal/crystal/auxdata/ukb_snp_qc.txt")
+str(loadings <- snp_qc[grep("_loading$", names(snp_qc))])
+
+plot_grid(plotlist = lapply(1:40, function(i) {
+  l_i <- loadings[[i]]
+  ind <- which(!is.na(l_i))
+  ggplot() +
+    geom_hex(aes(ind, l_i[ind])) +
+    theme_bigstatsr(0.4) +
+    labs(title = paste0("Loadings of PC", i), x = "Column index", y = NULL) +
+    scale_fill_viridis_c()
+}), align = "hv", ncol = 5, scale = 0.95)
+
+# ggsave("figures/UKBB-loadings1-40.pdf", width = 15, height = 15)
+
[…]

### 2019-10-25 code/project-1000G-to-UKBB.R

@@ -3,6 +3,7 @@ library(bigreadr)
 rel <- fread2("data/ukb25589_rel_s488346.dat")
 fam <- fread2("data/ukbb_bed/ukbb_488282.fam")
 ind.row <- which(!fam$V2 %in% rel$ID2)
+length(ind.row)  # 406545
 bed.ref <- bed("data/ukbb_bed/ukbb_488282.bed")
 obj.bed <- bed(download_1000G("data"))
 

### 2019-10-25 code/project-1000G.R
rerun with R 3.6

@@ -6,7 +6,7 @@ obj.bed <- bed(bedfile.ref)
 
 set.seed(1); train <- sample(nrow(obj.bed), 0.6 * nrow(obj.bed))
 
-# obj.svd <- bed_autoSVD2(obj.bed, ind.row = train, k = 20, ncores = nb_cores())
+# obj.svd <- bed_autoSVD(obj.bed, ind.row = train, k = 20, ncores = nb_cores())
 # saveRDS(obj.svd, "tmp-data/SVD_1000G.rds")
 obj.svd <- readRDS("tmp-data/SVD_1000G.rds")
 
@@ -23,8 +23,8 @@ shrinkage <- unname(sapply(1:20, function(k) {
   MASS::rlm(proj2[, k] ~ proj1[, k] + 0)$coef
 }))
 round(shrinkage, 2)
-#  [1] 1.01 1.02 1.07 1.09 1.51 1.68 1.94 1.40 2.88 3.17 2.90 2.92 3.23
-# [14] 5.13 5.25 5.04 4.58 5.69 6.26 6.32
+#  [1] 1.01 1.02 1.06 1.09 1.50 1.69 1.98 1.39 2.79 3.14
+# [11] 3.64 3.18 2.47 3.88 5.31 5.84 3.45 6.55 3.68 6.70
 
 plot_grid(plotlist = lapply(1:4, function(k) {
   k1 <- 2 * k - 1
@@ -70,8 +70,8 @@ stopifnot(all.equal(sqrt(eigval[1:20]), obj.svd$d))
 proj3 <- hdpca::pc_adjust(eigval, p = length(ind.keep), n = length(train),
                           test.scores = proj1, n.spikes.max = 50)
 round(shrinkage2 <- (proj3 / proj1)[1, ], 2)
-#  [1] 1.01 1.02 1.07 1.10 1.43 1.54 1.73 1.82 2.51 2.78 2.88 2.91 3.11
-# [14] 1.00 1.00 1.00 1.00 1.00 1.00 1.00
+#  [1] 1.01 1.02 1.07 1.10 1.43 1.54 1.74 1.79 2.47 2.77
+# [11] 2.84 3.15 1.00 1.00 1.00 1.00 1.00 1.00 1.00 1.00
 
 plot_grid(plotlist = lapply(3:6, function(k) {
   k1 <- 2 * k - 1
[…]

### 2019-10-29 paper-bedpca.tex
save

--- 2019-10-22 paper-bedpca.tex
@@
 \title{Efficient toolkit implementing best practices for principal component analysis of population genetic data}
-\author{Florian Priv\'e,$^{\text{1,}*}$ John J. McGrath,$^{\text{1,2,3}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,}*}$}
+\author{Florian Priv\'e,$^{\text{1,}*}$ Keurcien Luu, Michael G.B. Blum,$^{\text{4}}$ John J. McGrath,$^{\text{1,2,3}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,}*}$}
 
@@
 \noindent$^{\text{\sf 2}}$Queensland Brain Institute, University of Queensland, St. Lucia, 4072, Queensland, Australia. \\
-\noindent$^{\text{\sf 3}}$Queensland Centre for Mental Health Research, The Park Centre for Mental Health, Wacol, 4076, Queensland, Australia.\\
-
+\noindent$^{\text{\sf 3}}$Queensland Centre for Mental Health Research, The Park Centre for Mental Health, Wacol, 4076, Queensland, Australia, \\
+\noindent$^{\text{\sf 4}}$OWKIN France, Paris, 75010, France, \\
 \noindent$^\ast$To whom correspondence should be addressed.\\
@@
 \item \url{florian.prive.21@gmail.com}
-\item \url{john_mcgrath@qcmhr.uq.edu.au}
 \item \url{bjv@econ.au.dk}
@@
 Principal Component Analysis (PCA) has been widely used in genetics has been used for many years and in many contexts such as e.g.\ to adjust for population structure in Genome-Wide Association Studies (GWAS) by adding PCs as covariates \cite[]{price2006principal} and to detect loci under selection based on PCA \cite[]{galinsky2016fast,luu2017pcadapt}.
-Much work has been devoted to developing scalable algorithms to compute PCA on data as large as the UK Biobank \cite[]{bycroft2017genome}, which is now possible thanks to software such as FastPCA (fast mode), FlashPCA2, PLINK 2.0 (approx mode), bigstatsr/bigsnpr and TeraPCA \cite[]{galinsky2016fast,abraham2017flashpca2,chang2015second,prive2017efficient,bose2019terapca}.
+Recently, the advent of large population-scale genetic datasets, such as the UK biobank data, has prompted research on developing scalable algorithms to compute PCA on data as large as the UK Biobank \cite[]{bycroft2017genome}.
+This is now possible thanks to software such as FastPCA (fast mode), FlashPCA2, PLINK 2.0 (approx mode), bigstatsr/bigsnpr and TeraPCA \cite[]{galinsky2016fast,abraham2017flashpca2,chang2015second,prive2017efficient,bose2019terapca}.
 
-However, some pitfalls related to PCA of genotype data has been documented and we recall them here.
-This includes some possible lack of precision of computed PCs of software such as FastPCA and PLINK 2.0, which is not an issue for FlashPCA2 and bigstatsr/bigsnpr \cite[]{abraham2017flashpca2,prive2017efficient}.
-This also includes capturing Linkage Disequilibrium (LD) structure in PCA instead of population structure only \cite[]{price2008long,abdellaoui2013population,prive2017efficient}.
-This leads to e.g.\ an over-correction of GWAS variants within long-range LD regions [REWORD/PRECISE].
-Finally, another issue may arise when projecting PCs of a reference dataset to another study dataset: projected PCs are shrunk towards 0 in the new dataset \cite[]{lee2010convergence,wang2015improved,zhang2019fast}.
-This shrinkage makes it dangerous to use the projected PCs for analysis such as PC regression, ancestry detection and GWAS [REWORD/PRECISE].
+However, some pitfalls related to PCA of genotype data have been documented and not all of the currently available software addresses all of these. 
+In the following, we recall common pitfalls and explain when they are relevant.
+First, efficient methods for PCA use approximations, which can results in some lack of precision of computed PCs.
+This has been shown to be a possible issue for software such as FastPCA and PLINK 2.0, but not for FlashPCA2 and bigstatsr/bigsnpr \cite[]{abraham2017flashpca2,prive2017efficient}.
+Second, some of the PCs may capture LD structure rather than population structure \cite[]{price2008long,abdellaoui2013population,prive2017efficient}.
+Including PCs that capture LD as covariates in genetic analyses can lead to reduced power, e.g.\ for detecting genetic association within these LD regions in GWAS.
+Third, another issue may arise when projecting PCs of a reference dataset to another study dataset: projected PCs are shrunk towards 0 in the new dataset \cite[]{lee2010convergence,wang2015improved,zhang2019fast}.
+This shrinkage makes it dangerous to use the projected PCs for analyses such as PC regression, ancestry detection and correction for ancestry.
+Finally, PC scores may capture outliers that are due to family structure, population structure or other reasons; it might be benificial to detect and remove these individuals to maximise the population structure captured by the PCA (in the case of removing a few outliers) or to restrict analyses to genetically homogeneous samples.
 
-In this paper, we derive a new implementation of PCA that can be used directly on bed/bim/fam PLINK files with some missing values, which is available in R package bigsnpr v1.0.0.
-Then, we recall some of the pitfalls of PCA computation on genotype data and provide efficient solutions to these issues, such as accouting for LD structure in PCA, detecting outliers [NOT INTRODUCED BEFORE] and projecting PCs to another study sample without shrinkage bias.
+In this paper, we derive a new implementation of PCA that can be used directly on bed/bim/fam PLINK files with some missing values, which we make available in R package bigsnpr v1.0.0.
+Then, we recall some of the pitfalls of PCA computation on genotype data and provide efficient solutions to these issues.
+These includes accouting for LD structure in PCA, detecting outliers and projecting PCs to another study sample without shrinkage bias.
+
+[SHOULD ALSO TALK ABOUT PCADAPT AND LD]
 
@@
 \section{Material and Methods}
+
+\begin{table}[h]
+\centering
+\caption{Overview of existing methods. [TODO: ADD CITATIONS + RESOLVE QUESTIONS]} 
+\label{tab:method-overview}
+\begin{tabular}{|c|c|c|c|}
+\hline
+Analysis & Method / Software & Pros & Cons \\
+\hline
+\multirow{3}{*}{PCA} & bigstatsr / bigsnpr \cite[]{prive2017efficient} & \makecell{Fast and accurate + handle dosages \\ + thinning options directly included} & \makecell{Own format without missing values \\ (fast functions are available \\ for converting and imputing)} \\
+ & FlashPCA2 \cite[]{abraham2017flashpca2} & Fast and accurate & Sequential implementation \\
+ & \makecell{PLINK 2.0 (reimplementation of FastPCA) \\ \cite[]{galinsky2016fast,chang2015second}} & Very fast (AVX2 only?) & \makecell{Not always 100\% accurate \\ \cite[]{abraham2017flashpca2,prive2017efficient}} \\
+\hline
+\makecell{Detection of \\ oultier samples} & \makecell{"6 SDs from the mean" \\ (EIGENSOFT, \cite{patterson2006population})} & Simple & Assume Gaussian distritbution \\
+\hline
+\makecell{Detection of \\ homogeneous \\ samples} & Aberrant \cite[]{bellenguez2011robust} & Robust & Only two statistics at once? \\
+\hline
+\multirow{5}{*}{\makecell{Projection of \\ new individuals \\ onto reference \\ PCA space}} & Simple projection (multiplication by loadings) & Simple & Shrinkage biased \\
+  & Bias-adjusted projection \cite[]{dey2019asymptotic} & Independent of new samples & \makecell{Assume same shrinkage for all individuals \\ + model-based \\ + need all eigenvalues of reference} \\
+  & \makecell{Augmentation, Decomposition, \\ and Procrustes transformation \\ (ADP, \cite{wang2015improved})} & Accurate & Slow \\
+  & Online ADP (OADP, \cite{zhang2019fast}) & Much faster than ADP & Does not work for related individuals \\ 
+\hline
+\end{tabular}
+\end{table}
 
@@
 We remove the detected outlier variants, compute the PC scores and loadings again, and iterate until there is no detected outlier variant anymore.
-This procedure is implemented in function \texttt{bed\_autoSVD} or R package bigsnpr.
+This procedure is implemented in function \texttt{bed\_autoSVD} of R package bigsnpr.
 
-\subsection{Detecting outlier samples in PCA}
+\subsection{Detecting outlier samples in PCA \label{outlier-sample}}
 
-[REDO]
+For detecting outlier samples in PCA, we use a modified version of the Probabilistic Local Outlier Factor (PLOF) statistic on PCs \cite[]{kriegel2009loop}.
+Using K nearest neighbours (KNN), this consists in comparing the distance from an point $j$ to its KNNs ($pd_j = \sqrt{\frac{1}{K} \sum_{k=1}^K d^2_{j \leftrightarrow j_k}}$, where $j_k$ is the $k$-th NN of $j$) with the distances from its KNNs to their KNNs ($\frac{1}{K} \sum_{k=1}^K pd_{j_k}$).
+The idea is that an outlier should be far from all other points, but this point is even more outlier if its KNNs are in a very dense cluster.
+In practice, we use $pd_j / \sqrt{\frac{1}{K} \sum_{k=1}^K pd_{j_k}}$ (note the squared root) as it otherwise detect any point next to a very dense cluster as an outlier.
+We implement this statistic in function \texttt{prob\_dist} of our new R package bigutilsr.
+To make it fast, we use and parallelise (\texttt{knn\_parallel}) the fast K nearest neighbours implementation of R package nabor \cite[]{elseberg2012comparison}.
+Automatic detection of outliers is very hard, then we recommend users to choose a threshold of outlierness for this statistic based on visual inspection.
 
-We propose [BOF] different outlier detection techniques for PCA.
+For detecting samples that have a different ancestry from most of the samples in the data, we compute the pairwise orthogonalized Gnanadesikan-Kettenrin robust Mahalanobis distance on PC scores (Section \ref{maha}).
+We then restrict to individuals whose log-distance (alternatively p-value) is smaller (larger) than some threshold determined based on visual inspection.
 
-For detecting outlier samples in PCA, we use the Local Outlier Factor (LOF) on PCs \cite[]{breunig2000lof}.
-To make it robust to the choice of the number of nearest neighbours (kNN), we take the maximum value of the statistic computed with kNN $\in$ \{4, 5, 6, 7, 8\}.
-We do not use larger values for `kNN' because it would remove small ancestry groups (e.g.\ composed of 10-15 people only).
-We make use of the fast K nearest neighbours implementation of R package nabor to implement the LOF statistic efficiently in our new R package bigutilsr \cite[]{elseberg2012comparison}.
+\subsection{Projecting PCs from a reference dataset \label{proj}}
 
-For detecting samples that have a different ancestry from most of the samples in the data, we compute the pairwise orthogonalized Gnanadesikan-Kettenrin robust Mahalanobis distance of PCs \cite[]{gnanadesikan1972robust,maronna2002robust}, and do not include individuals whose log-distance is larger than some threshold determined based on visual inspection [DO IT AUTOMATIC?].
+[SCHEME TO ILLUSTRATE OADP?]
 
-\subsection{Subsetted version of 1000G data [RENAME?]\label{1000G}}
+To project PCs of a reference dataset (e.g.\ 1000G) to a target genotype dataset, we implement the following 3 steps in function \texttt{bed\_projectPCA} of package bigsnpr: 1) matching the variants of each dataset, including removing ambiguous alleles [A/T] and [C/G], and matching strand and direction of the alleles; 2) computing PCA of the reference dataset using the matched variants only; 3) projecting computed PCs to the target data using an optimised implementation (see Supplementary Materials) of the Online Augmentation, Decomposition, and Procrustes (OADP) transformation \cite[]{zhang2019fast}.
+
+\subsection{Data\label{1000G}}
 
@@
 
-[SHOULD DESCRIBE QC FOR OTHER DATASETS TOO?]
+[REMOVE QC NA FOR 1000G?]
 
-\subsection{Projecting PCs from a reference dataset \label{proj}}
+[DESCRIBE QC FOR OTHER DATASETS TOO]
 
-[SCHEME TO ILLUSTRATE OADP?]
+\subsection{Reproducibility}
 
-To project PCs of a reference dataset (e.g.\ 1000G) to a target genotype dataset, we implement the following 3 steps in function \texttt{bed\_projectPCA} of package bigsnpr: 1) matching the variants of each dataset, including removing ambiguous alleles [A/T] and [C/G], and matching strand and direction of the alleles; 2) computing PCA of the reference dataset using the matched variants only; 3) projecting computed PCs to the target data using an optimised implementation (see Supplementary Materials) of the Online Augmentation, Decomposition, and Procrustes (OADP) transformation \cite[]{zhang2019fast}.
-
-\subsection{Ancestry matching \label{ancestry}}
-
-We use the 1000 Genomes (1000G) Project data (section \ref{1000G}) to infer the ancestry of individuals from a study dataset following the following steps: 1) we compute the robust center and covariance parameters of the Mahalanobis distance (section \ref{maha}) for each of the 26 populations of the 10000G data; 2) we project individuals from the study dataset onto the PCA space computed using the 1000G data (section \ref{proj}); 3) we compute the Mahalanobis distance for each projected individual of the study data and each population of the 1000G data (using the center and covariance parameters computed previously); 4) we derive p-values from those distances, as they are approximately chi-squared distributed with the number of PCs used for computing the distances as degrees of freedom; 5) we affect each individual to the largest corresponding p-value (smallest distance), but discard individuals whose maximum p-value is less than 0.05 (i.e.\ individuals far from every population of the 1000G data).
+[ALL CODE..]
 
@@
 
-To show that we provide very performant implementations of the different methods presented in this paper, we apply them to the UK Biobank. We use 20 physical cores for most of the computations.
+To show that we provide very fast implementations of the different methods presented in this paper, we apply them to the UK Biobank. We use 20 physical cores for most of the computations.
 It takes 22 minutes to perform a first phase of clumping on 406,545 unrelated individuals genotyped over 504,139 variants, which reduces the number of variants to 261,307.
@@
 
-It takes 8 minutes to match the 1000G data to the UKBB data and compute 20 PCs of the 1000G data using the automatic outlier detection technique.
+It takes 8 minutes to match the 1000G data to the UKBB data and compute 20 PCs of the 1000G data using the automatic LD detection technique.
 It takes 12 minutes more to do the OADP projection of all 488,371 individuals of the UKBB data to the PCA computed using the 1000G data.
-Once projections are available, et takes only a couple minutes to perform the ancestry estimation.
 
-Finally, it took only 6 minutes to compute the 30-Nearest Neighbours of PC scores for all UK Biobank individuals.
-
-\subsection{PCA of the UK Biobank}
-
-Principal Components (PC) loadings reported by the UK Biobank clearly show that PC19 to PC40 capture LD structure, which is also the case for PC16 and PC18, although less pronounced (see peaks in figure \ref{fig:UKBB-loadings40}).
-Moreover, our automatic procedure removing long-range LD regions did not converge after 5 iterations, meaning that it kept detecting long-range LD regions at each iteration.
-Therefore, we were able to capture only 16 PCs that shows population stratification and not LD structure (Figures \ref{fig:UKBB-screeplot}, \ref{fig:UKBB-scores} and \ref{fig:UKBB-loadings}).
-
-\noindent[SHOW INITIAL PC SCORES OF UKBB?]
-
-\noindent[COMPUTE MAHA ON LOADINGS TO SHOW THE PEAKS SUMMARIZED IN ALL 16-18 PCS?]
-
-\noindent[SHOW THAT PROCEDURE IS OKAY FOR SMALLER DATASETS?]
-
-\noindent[SHOW ABLE TO RECONSTRUCT PCA? MAYBE NOT PC17]
+Finally, it takes only 6 minutes to compute the 30-Nearest Neighbours of PC scores for all UK Biobank individuals, which is the main computational aspect of computing statistics for detection individual outlier samples (Section \ref{outlier-sample}).
 
@@
 We use 60\% of individuals in the 1000G data (section \ref{1000G}) to compute the PCA ($K=20$ PCs) [EXPLICIT SAMPLE SIZE].
-Then, we project the remaining 40\% individuals using three methods: 1/ simply multiplying the genotypes of these individuals by the previously computed loadings, 2/ correcting the simple projections using asymptotic shrinkage factors as determined by R package hdpca v1.0.0 \cite[]{dey2019asymptotic} and 3/ the OADP projection (section \ref{proj}).
+Then, we project the remaining 40\% individuals using three methods: 1/ simply multiplying the genotypes of these individuals by the previously computed loadings, 2/ correcting the simple projections using asymptotic shrinkage factors as determined by R package hdpca v1.1.3 \cite[]{dey2019asymptotic} and 3/ the OADP projection (section \ref{proj}).
 When simply projecting using loadings, there is no apparent shrinkage for PC1 and PC2, a weak [REWORD?] shrinkage for PC3 and PC4, and a large shrinkage for PC5 to PC8 (Figure \ref{fig:proj1000G}).
@@
 
-\subsection{Ancestry matching}
+\subsection{PCA of the UK Biobank}
 
-[REDO WITH OUTLIER REMOVAL?]
+Principal Components (PC) loadings reported by the UK Biobank clearly show that PC19 to PC40 capture LD structure, which is also the case for PC16 and PC18, although less pronounced (see peaks in figure \ref{fig:UKBB-loadings40}).
+Moreover, our automatic procedure removing long-range LD regions did not converge after 5 iterations, meaning that it kept detecting long-range LD regions at each iteration.
+Therefore, we were able to capture only 16 PCs that shows population stratification and not LD structure (Figures \ref{fig:UKBB-screeplot}, \ref{fig:UKBB-scores} and \ref{fig:UKBB-loadings}).
 
-We performed ancestry estimation of the individuals in the UK Biobank using the 1000G data (Section \ref{ancestry}), discarding individuals with unknown ancestry or mixed ancestry.
-We used from PC1 to PC14 and PC16 based on visual inspection of PC scores (Figure \ref{fig:PC-1000G-UKBB}).
-We did not infer ancestry for 15.9\% of UK Biobank that had a maximum p-value smaller than 0.05 (Figure \ref{fig:ancestry-pval}). 
-More precisely, among ``White'', ``British'' and ``Irish'' ancestries, this represented respectively 20.2\%, 11.1\% and 20.3\%, while this represented between 44.5\% (``Chinese'') and 74.3\% (``Bangladeshi'') for other populations.
-In other words, 3 out of 4 of the UK Biobank individual who self-reports as ``Bangladeshi'' could not be attributed to any 1000G population, although 1000G data includes several ``Bengali in Bangladesh'' (SAS\_BEB).
-Only 10 people out of 401,048 were wrongly classified in ``super'' population of the 1000G; e.g.\ one Chinese person was classified as European by our method (Table \ref{tab:ancestry-pred}). 
-Moreover, our method is able to accurately differentiate between sub-continental populations such as differentiating between Pakistani, Bangladeshi and Chinese people (Table \ref{tab:ancestry-pred}).
+Then, we computed again PCA after restricting the individuals included in the computations: we randomly subsampled UKBB data to use only 10,000 British individuals (out of 431,029) and to 5000 Irish individuals (out of 12,755), while keeping all individuals with other self-reported individuals.
+We further removed all individuals for all pairs of related individuals reported by the UKBB. 
+This resulted in 48,942 individuals that we used to compute 50 PCs in less than 3 hours using function \texttt{bed\_autoSVD}.
 
-% latex table generated in R 3.6.0 by xtable 1.8-4 package
-% Mon Oct 21 15:14:34 2019
-\begin{table}[ht]
-\centering
-\caption{Self-reported ancestry (top) of UKBB individuals and their matching to 1000G populations (left) by our method.
-See the description of 1000G populations at \url{https://www.internationalgenome.org/category/population/}.}
-\label{tab:ancestry-pred}
-\begin{tabular}{|l|c|c|c|c|c|c|c|c|c|c|c|c|c|}
-  \hline
- & British & Irish & White & Other White & Indian & Pakistani & Bangladeshi & Chinese & Other Asian & Caribbean & African & Other Black & NA \\ 
-  \hline
-AFR\_ACB &  &  &  &  &  &  &  &  &  & 1832 & 580 & 31 & 259 \\ 
-  AFR\_ASW &  &  &  &  &  &  &  &  &  & 315 & 13 & 10 & 39 \\ 
-  AFR\_ESN &  &  &  &  &  &  &  &  &  &  & 155 & 1 & 29 \\ 
-  AFR\_GWD &  &  &  &  &  &  &  &  &  &  & 20 &  & 4 \\ 
-  AFR\_LWK &  &  &  &  &  &  &  &  &  &  & 34 &  & 4 \\ 
-  AFR\_MSL &  &  &  &  &  &  &  &  &  &  & 26 &  &  \\ 
-  AFR\_YRI &  &  &  &  &  &  &  &  &  & 5 & 469 & 3 & 81 \\ 
-   \hline
-AMR\_CLM & 1 &  &  & 50 &  &  &  &  &  &  &  &  & 142 \\ 
-  AMR\_MXL &  &  & 1 & 3 &  &  &  &  &  &  &  &  & 20 \\ 
-  AMR\_PEL &  &  &  & 2 &  &  &  &  &  &  &  &  & 19 \\ 
-   \hline
-EAS\_CDX &  &  &  &  &  &  &  & 2 &  &  &  &  & 1 \\ 
-  EAS\_CHB & 1 &  &  &  &  &  &  & 445 & 10 &  &  &  & 7 \\ 
-  EAS\_CHS &  &  &  &  &  &  &  & 380 & 9 &  &  &  & 21 \\ 
-  EAS\_JPT &  &  &  &  &  &  &  & 3 & 25 &  &  &  & 122 \\ 
-  EAS\_KHV &  &  &  &  &  &  &  & 4 & 16 &  &  &  & 22 \\ 
-   \hline
-EUR\_CEU & 325725 & 6675 & 350 & 4855 &  &  &  & 1 &  &  &  &  & 1363 \\ 
-  EUR\_FIN & 2 &  &  & 118 &  &  &  &  &  &  &  &  & 1 \\ 
-  EUR\_GBR & 57497 & 3486 & 79 & 454 &  &  &  &  &  &  & 1 &  & 244 \\ 
-  EUR\_IBS & 28 &  & 4 & 452 &  &  &  &  &  &  &  &  & 8 \\ 
-  EUR\_TSI & 25 &  & 1 & 177 &  &  &  &  &  &  &  &  & 5 \\ 
-   \hline
-SAS\_BEB &  &  &  &  & 11 &  & 57 &  &  &  &  &  & 7 \\ 
-  SAS\_GIH &  &  &  &  & 334 &  &  &  & 1 &  &  &  & 4 \\ 
-  SAS\_ITU & 1 &  &  &  & 314 & 1 &  &  & 264 & 3 &  &  & 82 \\ 
-  SAS\_PJL & 1 &  &  &  & 1749 & 767 &  &  & 104 & 2 &  &  & 177 \\ 
-  SAS\_STU &  &  &  &  & 38 &  &  &  & 185 &  &  &  & 40 \\ 
-   \hline
-NA & 47748 & 2594 & 110 & 9705 & 3270 & 980 & 164 & 669 & 1133 & 2140 & 1906 & 73 & 6970 \\ 
-   \hline
-\end{tabular}
-\end{table}
+\noindent[SHOW INITIAL PC SCORES OF UKBB?]
 
-\noindent[OKAY TO CLASSIFY SOME OTHER WHITE AS CENTRAL/SOUTH AMERICANS?]
+\noindent[COMPUTE MAHA ON LOADINGS TO SHOW THE PEAKS SUMMARIZED IN ALL 16-18 PCS?]
 
-\noindent[INCLUDE RESULTS FOR SIMPLE PROJ? -> HIST PVAL BAD -> 38\% NOT MATCHED, ESP PAKISTANI]
+\noindent[SHOW THAT PROCEDURE IS OKAY FOR SMALLER DATASETS?]
 
-We also projected individuals using the OADP projection and then applied our ancestry detection technique to other datasets. 
-When applied to the European individuals of the POPRES data \cite[]{nelson2008population}, we show that 56.5\% of individuals could not be matched to one of the 26 populations of the 1000G data (Table \ref{tab:ancestry-pred-popres}).
-This is particularly dramatic for all East-Europeans, whose only 11 people could be matched out of 179 (6.1\%).
-Moreover, even though 1000G data include several ``Toscani in Italia'' (EUR\_TSI), we were able to match 14.2\% of Italians only.
-Nevertheless, for individuals that we could match, they were all identified as European ancestry.
-We could also identify accurately sub-regions of Europe; e.g.\ all Spanish and Portugese individuals that we could match (73.1\%) were identified as ``Iberian Population in Spain'' (EUR\_IBS, Table \ref{tab:ancestry-pred-popres}).
-Finally, we also projected and matched people from a case-control cohort for celiac disease composed of four European countries: Italy, the Netherlands, the UK and Finland \cite[]{dubois2010multiple}.
-Among UK individuals, 5397 (79.9\%) were matched to either ``Utah Residents with Northern and Western European Ancestry'' (EUR\_CEU) or ``British in England and Scotland'' (EUR\_GBR), 10 (0.2\%) to other European populations, 1 to ``Mexican Ancestry from Los Angeles USA'' (AMR\_MXL) and 1346 (19.9\%) could not be matched (Table \ref{tab:ancestry-pred-celiac}).
-Among Finns, 1543 (62.4\%) were matched to ``Finnish in Finland'' (EUR\_FIN), 6 (0.3\%) to other European populations, and 922 (37.3\%) could not be matched.
-Among Italians, 229 (22\%) were matched to EUR\_TSI, 15 (1.5\%) to EUR\_IBS and 795 (76.5\%) could not be matched (Table \ref{tab:ancestry-pred-celiac}).
+\noindent[SHOW ABLE TO RECONSTRUCT PCA? MAYBE NOT PC17]
 
@@
 
-\noindent[LIMITATION 2: FAIL TO PROJECT RELATED INDIVIDUALS WITH OADP.]
-
-\noindent[LIMITATION 3: FAIL TO AUTOMATICALLY DETECT OUTLIER SAMPLES.]
-
-\noindent[DISCUSS LACK OF CERTAIN POPULATIONS IN 1000G]
-
-\noindent[DISCUSS ABOUT $K''=K'=K$? FASTER, EASIER, DO NOT WANT MORE DIM?]
+\noindent[LIMITATION 2: FAIL TO PROJECT RELATED INDIVIDUALS WITH OADP -> RESTRICT MORE.]
 
@@
 This research has been conducted using the UK Biobank Resource under Application Number 25589.
+
+\section{Declaraction of Interests}
+
+As competing interest, you can write
+Michael Blum is an employee at OWKIN.
 

### 2019-10-29 code/pca-UKBB-restricted.R
add restricted PCA

@@ -1,0 +1,79 @@
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+rel <- fread2("data/ukb25589_rel_s488346.dat")
+fam <- fread2("data/ukbb_bed/ukbb_488282.fam")
+is_rel <- (fam$V2 %in% rel$ID1 | fam$V2 %in% rel$ID2)
+obj.bed <- bed("data/ukbb_bed/ukbb_488282.bed")
+
+## Self-reported ancestry (https://biobank.ctsu.ox.ac.uk/crystal/coding.cgi?id=1001)
+code_ancestry <- fread2("data/coding1001.tsv")
+csv <- "data/ukb22544.csv"
+df0 <- fread2(csv, select = c("eid", "21000-0.0", "22006-0.0"),
+              col.names = c("eid", "pop", "is_caucasian")) %>%
+  mutate(
+    pop  = factor(pop, levels = code_ancestry$coding,
+                  labels = code_ancestry$meaning),
+    is_caucasian = as.logical(is_caucasian)
+  )
+pop_UKBB <- df0$pop[match(obj.bed$fam$sample.ID, df0$eid)]
+table(pop_UKBB)
+
+ind_used <- c(sample(which(pop_UKBB == "British"), 10e3),
+              sample(which(pop_UKBB == "Irish"), 5e3),
+              which(!pop_UKBB %in% c("British", "Irish")))
+ind.train <- intersect(ind_used, which(!is_rel))
+table(pop_UKBB[ind.train])
+
+system.time(
+  obj.svd <- bed_autoSVD(obj.bed, ind.row = ind.train,
+                         k = 50, ncores = nb_cores())
+) # 2.9H
+
+plot(obj.svd)
+# ggsave("figures/UKBB-screeplot-restricted.pdf", width = 7, height = 5)
+
+plot_grid(plotlist = lapply(14:25, function(k) {
+  plot(obj.svd, type = "scores", scores = 2 * k - 1:0, coeff = 0.4) +
+    aes(color = pop_UKBB[ind.train]) +
+    theme(legend.position = "none")
+}))
+# ggsave("figures/UKBB-scores-restricted.png", width = 11, height = 6)
+
+
+system.time(
+  proj <- bed_projectSelfPCA(obj.svd, obj.bed,
+                             ind.row = rows_along(obj.bed)[-ind.train],
[…]

### 2019-10-29 code/pca-UKBB-restricted.R
add shrinkage

@@ -51,6 +51,15 @@ system.time(
 proj1 <- proj$simple_proj
 proj2 <- proj$OADP_proj
 
+# shrinkage coefficients
+shrinkage <- unname(sapply(1:50, function(k) {
+  MASS::rlm(proj2[, k] ~ proj1[, k] + 0)$coef
+}))
+round(shrinkage, 2)
+#  [1] 1.00 1.00 1.01 1.01 1.04 1.05 1.07 1.07 1.11 1.12 1.13 1.20 1.23 1.25 1.30 1.31 1.35
+# [18] 1.37 1.40 1.43 1.44 1.50 1.52 1.54 1.56 1.56 1.59 1.65 1.66 1.70 1.73 1.74 1.74 1.77
+# [35] 1.77 1.78 1.78 1.79 1.79 1.79 1.80 1.80 1.80 1.80 1.80 1.80 1.80 1.80 1.80 1.80
+
 
 PC.ref <- predict(obj.svd)
 ind <- rows_along(PC.ref) # sample(nrow(PC.ref), 10e3)

### 2019-10-29 paper-bedpca.tex
save

@@
 \abstract{
+
+[REDO LATER]
+
 Principal Component Analysis (PCA) of genetic data is routinely used to infer ancestry and control for population structure in various genetic analyses.
@@
 
-[SHOULD ALSO TALK ABOUT PCADAPT AND LD]
+[SHOULD ALSO TALK ABOUT PCADAPT -> CONFOUNDED BY LD + SHOULD COMPUTE MAHA ON UNLINKED]
 
@@
 
-\noindent[TODO SECTION ON SAMPLE OUTLIER REMOVAL]
+\noindent[TODO SECTION ON SAMPLE OUTLIER REMOVAL / HOMOGENEOUS SAMPLES]
 
@@
 
+When computing the PCA on the UK Biobank using 406,545 unrelated individuals and 171,977 variants, and projecting the 1000G data onto this reference PCA space, shrinkage is much smaller ($\leq 1.08$ for all 20 first PCs, Figure \ref{fig:proj1000G-6}).
+
 \begin{figure}[!htpb]
@@
 
-When computing the PCA on the UK Biobank using 406,545 unrelated individuals and 171,977 variants, and projecting the 1000G data onto this reference PCA space, shrinkage is much smaller ($\leq 1.08$ for all 20 first PCs, Figure \ref{fig:proj1000G-6}).
-
 \subsection{PCA of the UK Biobank}
@@
 
-Then, we computed again PCA after restricting the individuals included in the computations: we randomly subsampled UKBB data to use only 10,000 British individuals (out of 431,029) and to 5000 Irish individuals (out of 12,755), while keeping all individuals with other self-reported individuals.
-We further removed all individuals for all pairs of related individuals reported by the UKBB. 
-This resulted in 48,942 individuals that we used to compute 50 PCs in less than 3 hours using function \texttt{bed\_autoSVD}.
+Then, we computed again PCA after restricting the individuals included in the computations: we randomly subsampled UKBB data to use only 10,000 British individuals (out of 431,029) and 5000 Irish individuals (out of 12,755), while keeping all individuals with other self-reported individuals.
+We further removed all pairs of related individuals reported by the UKBB (both individuals in each pair). 
+This resulted in 48,942 individuals that we used to compute 50 PCs, which took less than 3 hours using function \texttt{bed\_autoSVD} (that converged after 4 iterations of automatic LD removal). 
+We show that we are able to capture at least 40 PCs that display visual population structure (Figures \ref{fig:UKBB-scores2} and \ref{fig:UKBB-screeplot2}).
+We then projected all 439,429 remaining individuals from UKBB onto this PCA space in 41 minutes.
+Note that these individuals are not supposed to be related to any of the 48,942 individuals used for training PCA because we removed both individuals from each pair of related individuals in UKBB.
+Projection of new individuals show again clear shrinkage when using simple projection (between 1.00 for PC1 and 1.80 for PC50), but no visible bias when using OADP projection (Figure \ref{fig:projUKBB-related}).
 
@@
 
-\section*{Funding}
-
 \section*{Acknowledgements}
 
+Authors thank Rounak Dey and Seunggeun Lee for helpful discussions about PCA projection.
 This research has been conducted using the UK Biobank Resource under Application Number 25589.
 
-\section{Declaraction of Interests}
+F.P., J.J.M.\ and B.J.V.\ acknowledge Niels Bohr Professorship from the Danish National Research Foundation to Prof.\ John J.\ McGrath, and the Lundbeck Foundation Initiative for Integrative Psychiatric Research, iPSYCH (R248-2017-2003).
 
-As competing interest, you can write
-Michael Blum is an employee at OWKIN.
+\section*{Declaraction of Interests}
 
-~\\ \textit{Conflict of Interest:} none declared.
+Michael Blum is an employee of OWKIN France.
+The other authors declare no competing interests.
 
@@
 
-\begin{figure}[!htpb]
-\centerline{\includegraphics[width=0.9\textwidth]{UKBB-loadings1-40.pdf}}
-\caption{Principal Component (PC) loadings 1 to 40 reported by the UK Biobank.
-\label{fig:UKBB-loadings40}}
-\end{figure}
-
-\begin{figure}[!htpb]
-\centerline{\includegraphics[width=0.8\textwidth]{UKBB-screeplot.pdf}}
-\caption{Scree plot: plot of singular values computed from the UK Biobank.
-\label{fig:UKBB-screeplot}}
-\end{figure}
-
-\begin{figure}[!htpb]
-\centerline{\includegraphics[width=0.9\textwidth]{UKBB-PC1-20.png}}
-\caption{Principal Component (PC) scores 1 to 20 computed from the UK Biobank.
-\label{fig:UKBB-scores}}
-\end{figure}
-
-\begin{figure}[!htpb]
-\centerline{\includegraphics[width=0.9\textwidth]{UKBB-loadings.pdf}}
-\caption{Principal Component (PC) loadings 1 to 20 computed from the UK Biobank.
-\label{fig:UKBB-loadings}}
-\end{figure}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Blue points are the 488,371 individuals from the UK Biobank, projected using the Online Augmentation, Decomposition, and Procrustes (OADP) transformation. 
-Note that only 20,000 random individuals are represented in this plot.
+Note that only 20,000 random projected individuals are represented in this plot.
 \label{fig:proj1000G-5}}
@@
 
-\vspace*{1em}
+\begin{figure}[!htpb]
+\centerline{\includegraphics[width=0.9\textwidth]{UKBB-loadings1-40.pdf}}
+\caption{Principal Component (PC) loadings 1 to 40 reported by the UK Biobank.
+\label{fig:UKBB-loadings40}}
+\end{figure}
 
 \begin{figure}[!htpb]
-\centerline{\includegraphics[width=0.8\textwidth]{PC-1000G-UKBB.pdf}}
-\caption{Principal Component (PC) scores 1 to 20 of the 1000 Genomes (1000G) project using variants matched with the UK Biobank. Colors represent populations of the 1000G data.
-\label{fig:PC-1000G-UKBB}}
+\centerline{\includegraphics[width=0.8\textwidth]{UKBB-screeplot.pdf}}
+\caption{Scree plot: plot of singular values computed from the UK Biobank.
+\label{fig:UKBB-screeplot}}
 \end{figure}
 
-\vspace*{1em}
+\begin{figure}[!htpb]
+\centerline{\includegraphics[width=0.9\textwidth]{UKBB-PC1-20.png}}
+\caption{Principal Component (PC) scores 1 to 20 computed from the UK Biobank.
+\label{fig:UKBB-scores}}
+\end{figure}
 
 \begin{figure}[!htpb]
-\centerline{\includegraphics[width=0.8\textwidth]{hist-pval-max.pdf}}
-\caption{Maximum p-values based on robust Mahalabonis distances of UK Biobank individuals from each of the 26 1000G populations.
-\label{fig:ancestry-pval}}
+\centerline{\includegraphics[width=0.9\textwidth]{UKBB-loadings.pdf}}
+\caption{Principal Component (PC) loadings 1 to 20 computed from the UK Biobank.
+\label{fig:UKBB-loadings}}
 \end{figure}
@@
 
-% latex table generated in R 3.6.0 by xtable 1.8-4 package
-% Sat Oct 19 00:18:24 2019
-\begin{table}[ht]
-\centering
-\caption{Ancestry (left) of POPRES individuals and their matching to 1000G populations (top) by our method.
-See the description of 1000G populations at \url{https://www.internationalgenome.org/category/population/}.}
-\label{tab:ancestry-pred-popres}
-\begin{tabular}{|l|c|c|c|c|c|}
-  \hline
- & EUR\_CEU & EUR\_GBR & EUR\_IBS & EUR\_TSI & NA \\ 
-  \hline
-Anglo-Irish Isles & 159 & 57 &  &  & 50 \\ 
-  Belgium & 28 & 3 &  &  & 12 \\ 
-  Central Europe & 3 &  &  & 1 & 51 \\ 
-  Eastern Europe & 2 &  &  &  & 28 \\ 
-  France & 14 &  & 13 &  & 62 \\ 
-  Germany & 37 &  &  &  & 34 \\ 
-  Italy &  &  & 12 & 19 & 188 \\ 
-  Netherlands & 10 & 2 &  &  & 5 \\ 
-  Scandinavia & 5 &  &  &  & 10 \\ 
-  SE Europe &  &  &  & 5 & 89 \\ 
-  SW Europe &  &  & 193 &  & 71 \\ 
-  Switzerland & 33 & 1 & 6 &  & 182 \\ 
-   \hline
-\end{tabular}
-\end{table}
+\begin{figure}[!htpb]
+\centerline{\includegraphics[width=0.8\textwidth]{UKBB-screeplot-restricted.pdf}}
+\caption{Scree plot: plot of singular values computed from the UK Biobank using 48,942 individuals of diverse ancestries.
+\label{fig:UKBB-screeplot2}}
+\end{figure}
 
-% latex table generated in R 3.6.0 by xtable 1.8-4 package
-% Sat Oct 19 00:28:24 2019
-\begin{table}[ht]
-\centering
-\caption{Ancestry (left) of Celiac individuals and their matching to 1000G populations (top) by our method.
-See the description of 1000G populations at \url{https://www.internationalgenome.org/category/population/}.}
-\label{tab:ancestry-pred-celiac}
-\begin{tabular}{|l|c|c|c|c|c|c|c|}
-  \hline
- & AMR\_MXL & EUR\_CEU & EUR\_FIN & EUR\_GBR & EUR\_IBS & EUR\_TSI & NA \\ 
-  \hline
-Finland &  & 4 & 1543 & 1 &  & 1 & 922 \\ 
-  Italy &  &  &  &  & 15 & 229 & 795 \\ 
-  Netherlands &  & 802 &  & 382 & 1 &  & 463 \\ 
-  UK & 1 & 2865 & 1 & 2532 & 4 & 5 & 1346 \\ 
-   \hline
-\end{tabular}
-\end{table}
+\begin{figure}[!htpb]
+\centerline{\includegraphics[width=0.9\textwidth]{UKBB-scores-restricted.png}}
+\caption{Principal Component (PC) scores 27 to 50 computed from the UK Biobank using 48,942 individuals of diverse ancestries. Different colors represent different self-reported ancestries.
+\label{fig:UKBB-scores2}}
+\end{figure}
+
+\begin{figure}[!htpb]
+\centerline{\includegraphics[width=0.9\textwidth]{projUKBB-related.png}}
+\caption{Principal Component (PC) scores 27 to 50 of the UK Biobank.
+Black points are the 48,942 individuals of diverse ancestries used for computing PCA.
+Red points are the remaining UKBB individuals, projected by simply multiplying their genotypes by the corresponding PC loadings.
+Blue points are the remaining UKBB individuals, projected using the Online Augmentation, Decomposition, and Procrustes (OADP) transformation.
+Note that only 20,000 random projected individuals are represented in this plot.
+\label{fig:projUKBB-related}}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2019-11-01 pca-toolkit.Rmd
add method overview
add restricted PCs

--- 2019-10-21 pca-toolkit.Rmd
@@
 
+## Overview of existing methods
+
+<br>
+
+```{r, out.width="98%"}
+knitr::include_graphics("figures/method-overview.png")
+```
+
+---
+
 ## Pruning and PCA
+
+<br>
+
+Method:
+
+- Directly memory-map the bed file (access it as if it were a matrix in memory)
+
+- Same PCA algorithm based on random projections, already used in PCA of R package bigstatsr (Privé et al. 2018) and in FlashPCA2 (Abraham et al. 2017).
+
+--
 
@@
 knitr::include_graphics("figures/UKBB-PC1-20.png")
+```
+
+---
+
+## Computing PCs using a subset of individuals
+
+When restricting British to 10K and Irish to 5K: capturing at least 40 PCs!
+
+<br>
+
+```{r, out.width="95%"}
+knitr::include_graphics("figures/UKBB-scores-restricted.png")
 ```
@@
 
-## Ancestry matching
-
-Method: Robust Mahalanobis distances by reference population.
-
-```{r}
-
-```
-
----
-
 class: inverse, center, middle

### 2019-11-01 pdf2png.R
proper SVGs

@@ -3,6 +3,6 @@ library(magick)
 "figures/us.pdf" %>% {
   svg <- sub("\\.pdf$", ".svg", .)
   img <- image_read_pdf(.)
-  image_write(img, svg)
+  image_write(img, svg, format = "SVG")
   print(svg)
 }

### 2019-11-04 code/project-1000G.R
save

@@ -6,9 +6,7 @@ obj.bed <- bed(bedfile.ref)
 
 set.seed(1); train <- sample(nrow(obj.bed), 0.6 * nrow(obj.bed))
 
-# obj.svd <- bed_autoSVD(obj.bed, ind.row = train, k = 20, ncores = nb_cores())
-# saveRDS(obj.svd, "tmp-data/SVD_1000G.rds")
-obj.svd <- readRDS("tmp-data/SVD_1000G.rds")
+obj.svd <- bed_autoSVD(obj.bed, ind.row = train, k = 20, ncores = nb_cores())
 
 test <- bed_projectSelfPCA(obj.svd, obj.bed,
                            ind.row = rows_along(obj.bed)[-train],

### 2019-11-05 code/pca-UKBB-restricted.R
rerun analyses with newest version

@@ -47,7 +47,7 @@ system.time(
   proj <- bed_projectSelfPCA(obj.svd, obj.bed,
                              ind.row = rows_along(obj.bed)[-ind.train],
                              ncores = nb_cores())
-) # 41 min
+) # 21 min
 proj1 <- proj$simple_proj
 proj2 <- proj$OADP_proj
 

### 2019-11-05 code/pca-UKBB.R

@@ -8,8 +8,7 @@ obj.bed <- bed("data/ukbb_bed/ukbb_488282.bed")
 system.time(
   obj.svd <- bed_autoSVD(obj.bed, ind.row = ind.row, k = 20, ncores = nb_cores())
 ) # 5h
-saveRDS(obj.svd, "tmp-results/SVD_UKBB.rds")
-# Phase of clumping (on MAC) at r^2 > 0.2.. keep 261307 variants.
+# Phase of clumping (on MAC) at r^2 > 0.2.. keep 261306 variants.
 # Discarding 0 variant with MAC < 10.
 #
 # Iteration 1:
@@ -19,22 +18,22 @@ saveRDS(obj.svd, "tmp-results/SVD_UKBB.rds")
 #
 # Iteration 2:
 # Computing SVD..
-# 4550 outlier variants detected..
+# 4549 outlier variants detected..
 # 11 long-range LD regions detected..
 #
 # Iteration 3:
 # Computing SVD..
-# 3882 outlier variants detected..
+# 3881 outlier variants detected..
 # 11 long-range LD regions detected..
 #
 # Iteration 4:
 # Computing SVD..
-# 3832 outlier variants detected..
+# 3839 outlier variants detected..
 # 15 long-range LD regions detected..
 #
 # Iteration 5:
 # Computing SVD..
-# 5500 outlier variants detected..
+# 5488 outlier variants detected..
 # 17 long-range LD regions detected..
 #
 # Iteration 6:
@@ -44,7 +43,7 @@ saveRDS(obj.svd, "tmp-results/SVD_UKBB.rds")
 system.time(
   ind.keep2 <- bed_clumping(obj.bed, ind.row = ind.row, ncores = nb_cores())
 ) # 22 min
-length(ind.keep2)  # 261307
+length(ind.keep2)  # 261,306
 
 system.time(
   obj.svd2 <- bed_randomSVD(obj.bed, ind.row = ind.row, ind.col = ind.keep2,

### 2019-11-05 code/project-1000G-to-UKBB.R

@@ -11,7 +11,7 @@ system.time(
   test <- bed_projectPCA(bed.ref, obj.bed, ind.row.ref = ind.row,
                          k = 20, ncores = nb_cores())
 ) # 4.4 H
-length(attr(test$obj.svd.ref, "subset")) # 171977
+length(attr(test$obj.svd.ref, "subset")) # 171,987
 
 PC.ref <- predict(test$obj.svd.ref)
 proj1 <- test$simple_proj
@@ -46,24 +46,3 @@ plot_grid(plotlist = lapply(1:8, function(k) {
 }), nrow = 4)
 
 # ggsave("figures/proj1000G-UKBB2.png", width = 10, height = 7)
-
-library(bigreadr)
-## Self-reported ancestry (https://biobank.ctsu.ox.ac.uk/crystal/coding.cgi?id=1001)
-code_ancestry <- fread2("data/coding1001.tsv")
-csv <- "data/ukb22544.csv"
-library(dplyr)
-df0 <- fread2(csv, select = c("eid", "21000-0.0", "22006-0.0"),
-              col.names = c("eid", "pop", "is_caucasian")) %>%
-  mutate(
-    pop  = factor(pop, levels = code_ancestry$coding,
-                  labels = code_ancestry$meaning),
-    is_caucasian = as.logical(is_caucasian)
-  )
-
-eid <- bed.ref$fam$sample.ID[ind.row[PC.ref[, 6] > -50 & PC.ref[, 5] < -10]]
-pop_UKBB <- df0$pop[match(eid, df0$eid)]
-table(pop_UKBB)
-
-
-fam2 <- bigreadr::fread2("data/1000G_phase3_common_norel.fam2")
-fam2$Population[proj2[, 6] > -50 & proj2[, 5] < -10]

### 2019-11-05 code/project-POPRES-to-1000G.R

@@ -1,124 +1,0 @@
-library(bigsnpr)
-library(ggplot2)
-
-snp_plinkQC(
-  plink.path = download_plink("tmp-data"),
-  prefix.in = "~/Bureau/POPRES_data/POPRES_allchr",
-  geno = 0.01, mind = 0.01,
-  autosome.only = TRUE
-)
-obj.bed <- bed("~/Bureau/POPRES_data/POPRES_allchr_QC.bed")
-bed.ref <- bed(download_1000G("tmp-data"))
-
-system.time(
-  test <- bed_projectPCA(bed.ref, obj.bed, k = 20, ncores = nb_cores(),
-                         build.new = "hg18", liftOver = "tmp-data/liftOver")
-)
-
-PC.ref <- predict(test$obj.svd.ref)
-proj1 <- test$simple_proj
-proj2 <- test$OADP_proj
-
-
-plot_grid(plotlist = lapply(1:8, function(k) {
-  k1 <- 2 * k - 1
-  k2 <- 2 * k
-  plot_grid(
-    qplot(PC.ref[, k1], PC.ref[, k2], size = I(2)) +
-      geom_point(aes(proj1[, k1], proj1[, k2]), color = "red") +
-      theme_bigstatsr(0.5) +
-      labs(x = paste0("PC", k1), y = paste0("PC", k2)),
-    qplot(PC.ref[, k1], PC.ref[, k2], size = I(2)) +
-      geom_point(aes(proj2[, k1], proj2[, k2]), color = "blue") +
-      theme_bigstatsr(0.5) +
-      labs(x = paste0("PC", k1), y = paste0("PC", k2)),
-    scale = 0.95
-  )
-}), nrow = 4)
-
-
-#### Ancestry estimation ####
-
-fam2 <- bigreadr::fread2("tmp-data/1000G_phase3_common_norel.fam2")
-
-plot_grid(plotlist = lapply(5:10, function(k) {
-  k1 <- 2 * k - 1
-  k2 <- 2 * k
-  qplot(PC.ref[, k1], PC.ref[, k2], size = I(2), color = fam2$Population) +
-    theme_bigstatsr(0.7) +
-    labs(x = paste0("PC", k1), y = paste0("PC", k2))
-}))
-
-
-seq_PC <- c(1:18, 20)
-pop <- paste(fam2$`Super Population`, fam2$Population, sep = "_")
-pop_PCs <- vctrs::vec_split(PC.ref, pop)
-
[…]

### 2019-11-05 code/project-UKBB-to-1000G.R

@@ -39,101 +39,3 @@ plot_grid(plotlist = lapply(1:8, function(k) {
 }), nrow = 4)
 
 # ggsave("figures/proj1000G-UKBB.png", width = 11, height = 7)
-
-#### Ancestry estimation ####
-
-fam2 <- bigreadr::fread2("data/1000G_phase3_common_norel.fam2")
-
-plot(test$obj.svd.ref)
-plot_grid(plotlist = lapply(5:10, function(k) {
-  k1 <- 2 * k - 1
-  k2 <- 2 * k
-  qplot(PC.ref[, k1], PC.ref[, k2], size = I(2), color = fam2$Population) +
-    theme_bigstatsr(0.7) +
-    labs(x = paste0("PC", k1), y = paste0("PC", k2)) +
-    theme(legend.position = "none")
-}))
-
-# ggsave("figures/PC-1000G-UKBB.pdf", width = 10, height = 6)
-
-seq_PC <- c(1:14, 16)
-pop <- paste(fam2$`Super Population`, fam2$Population, sep = "_")
-pop_PCs <- vctrs::vec_split(PC.ref, pop)
-
-all_pval <- sapply(pop_PCs$val, function(PC) {
-  maha <- bigutilsr::covRob(PC[, seq_PC], estim = "pairwiseGK")
-  dist <- mahalanobis(proj2[, seq_PC], center = maha$center, cov = maha$cov)
-  pval <- pchisq(dist, df = length(seq_PC), lower.tail = FALSE)
-})
-
-choose_pop <- apply(all_pval, 1, which.max)
-pval_max <- all_pval[cbind(rows_along(all_pval), choose_pop)]
-choose_pop2 <- ifelse(pval_max < 0.05, NA, pop_PCs$key[choose_pop])
-
-ggplot() +
-  geom_histogram(aes(pval_max), breaks = seq(0, 1, by = 0.05),
-                 color = "#FFFFFF", fill = "#000000", alpha = 0.5) +
-  geom_vline(xintercept = 0.05, color = "red") +
-  theme_bigstatsr() +
-  labs(x = "Maximum p-value testing population belonging")
-
-# ggsave("figures/hist-pval-max.pdf", width = 8.5, height = 5.5)
-
-library(bigreadr)
-## Self-reported ancestry (https://biobank.ctsu.ox.ac.uk/crystal/coding.cgi?id=1001)
-code_ancestry <- fread2("data/coding1001.tsv")
[…]

### 2019-11-05 code/project-celiac-to-1000G.R

@@ -1,109 +1,0 @@
-library(bigsnpr)
-library(ggplot2)
-
-snp_plinkQC(
-  plink.path = download_plink("tmp-data"),
-  prefix.in = "~/Bureau/Dubois2010_data/FinnuncorrNLITUK3hap550",
-  geno = 0.01, mind = 0.01,
-  autosome.only = TRUE
-)
-
-obj.bed <- bed("~/Bureau/Dubois2010_data/FinnuncorrNLITUK3hap550_QC.bed")
-bed.ref <- bed(download_1000G("tmp-data"))
-
-system.time(
-  test <- bed_projectPCA(bed.ref, obj.bed, k = 20, ncores = nb_cores(),
-                         build.new = "hg18", liftOver = "tmp-data/liftOver")
-)
-
-PC.ref <- predict(test$obj.svd.ref)
-proj1 <- test$simple_proj
-proj2 <- test$OADP_proj
-
-plot_grid(plotlist = lapply(1:8, function(k) {
-  k1 <- 2 * k - 1
-  k2 <- 2 * k
-  plot_grid(
-    qplot(PC.ref[, k1], PC.ref[, k2], size = I(2)) +
-      geom_point(aes(proj1[, k1], proj1[, k2]), color = "red") +
-      theme_bigstatsr(0.5) +
-      labs(x = paste0("PC", k1), y = paste0("PC", k2)),
-    qplot(PC.ref[, k1], PC.ref[, k2], size = I(2)) +
-      geom_point(aes(proj2[, k1], proj2[, k2]), color = "blue") +
-      theme_bigstatsr(0.5) +
-      labs(x = paste0("PC", k1), y = paste0("PC", k2)),
-    scale = 0.95
-  )
-}), nrow = 4)
-
-# ggsave("figures/proj1000G-Celiac.png", width = 11, height = 7)
-
-
-#### Ancestry estimation ####
-
-fam2 <- bigreadr::fread2("tmp-data/1000G_phase3_common_norel.fam2")
-
-plot_grid(plotlist = lapply(5:10, function(k) {
-  k1 <- 2 * k - 1
-  k2 <- 2 * k
-  qplot(PC.ref[, k1], PC.ref[, k2], size = I(2), color = fam2$Population) +
-    theme_bigstatsr(0.7) +
-    labs(x = paste0("PC", k1), y = paste0("PC", k2))
-}))
-
-
-seq_PC <- 1:19
[…]

### 2019-11-06 code/pca-UKBB.R
save

@@ -1,5 +1,8 @@
 library(bigsnpr)
 library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
 rel <- fread2("data/ukb25589_rel_s488346.dat")
 fam <- fread2("data/ukbb_bed/ukbb_488282.fam")
 ind.row <- which(!fam$V2 %in% rel$ID2)
@@ -40,6 +43,8 @@ system.time(
 # Computing SVD..
 # Maximum number of iterations reached.
 
+# saveRDS(obj.svd, "tmp-results/SVD_UKBB.rds")
+
 system.time(
   ind.keep2 <- bed_clumping(obj.bed, ind.row = ind.row, ncores = nb_cores())
 ) # 22 min
@@ -57,8 +62,26 @@ plot(obj.svd)
 plot(obj.svd, type = "loadings", loadings = 1:20, coeff = 0.4)
 # ggsave("figures/UKBB-loadings.pdf", width = 13, height = 7)
 
-plot(obj.svd, type = "scores", scores = 1:20, coeff = 0.4, ncol = 5)
-# ggsave("figures/UKBB-PC1-20.png", width = 9, height = 7)
+## Self-reported ancestry (https://biobank.ctsu.ox.ac.uk/crystal/coding.cgi?id=1001)
+code_ancestry <- fread2("data/coding1001.tsv")
+csv <- "data/ukb22544.csv"
+df0 <- fread2(csv, select = c("eid", "21000-0.0", "22006-0.0"),
+              col.names = c("eid", "pop", "is_caucasian")) %>%
+  mutate(
+    pop  = factor(pop, levels = code_ancestry$coding,
+                  labels = code_ancestry$meaning),
+    is_caucasian = as.logical(is_caucasian)
+  )
+pop_UKBB <- df0$pop[match(obj.bed$fam$sample.ID, df0$eid)]
+table(pop_UKBB)
+
+plot_grid(plotlist = lapply(1:10, function(k) {
+  plot(obj.svd, type = "scores", scores = 2 * k - 1:0, coeff = 0.5) +
+    aes(color = pop_UKBB[ind.row]) +
+    theme(legend.position = "none")
+}), ncol = 5)
+# ggsave("figures/UKBB-PC1-20.png", width = 11, height = 8)
+
 
 # Benchmark KNN
 PCs <- predict(obj.svd)

### 2019-11-07 paper-bedpca.tex
save

@@
 
-\renewcommand{\arraystretch}{2}
+%\renewcommand{\arraystretch}{1.8}
 
@@
 
-Principal Component Analysis (PCA) has been widely used in genetics has been used for many years and in many contexts such as e.g.\ to adjust for population structure in Genome-Wide Association Studies (GWAS) by adding PCs as covariates \cite[]{price2006principal} and to detect loci under selection based on PCA \cite[]{galinsky2016fast,luu2017pcadapt}.
+Principal Component Analysis (PCA) has been widely used in genetics for many years and in many contexts.
+For instance, adding PCs as covariates has been used to adjust for population structure in Genome-Wide Association Studies (GWAS) \cite[]{price2006principal}.
+PCA has also been used to detect loci under selection \cite[]{galinsky2016fast,luu2017pcadapt}.
 Recently, the advent of large population-scale genetic datasets, such as the UK biobank data, has prompted research on developing scalable algorithms to compute PCA on data as large as the UK Biobank \cite[]{bycroft2017genome}.
-This is now possible thanks to software such as FastPCA (fast mode), FlashPCA2, PLINK 2.0 (approx mode), bigstatsr/bigsnpr and TeraPCA \cite[]{galinsky2016fast,abraham2017flashpca2,chang2015second,prive2017efficient,bose2019terapca}.
+This is now possible thanks to software such as FastPCA (fast mode of EIGENSOFT), FlashPCA2, PLINK 2.0 (approx mode), bigstatsr/bigsnpr and TeraPCA \cite[]{galinsky2016fast,abraham2017flashpca2,chang2015second,prive2017efficient,bose2019terapca}.
 
-However, some pitfalls related to PCA of genotype data have been documented and not all of the currently available software addresses all of these. 
+However, some pitfalls related to PCA of genotype data have been documented and not all of the currently available software address all of these. 
 In the following, we recall common pitfalls and explain when they are relevant.
@@
 Third, another issue may arise when projecting PCs of a reference dataset to another study dataset: projected PCs are shrunk towards 0 in the new dataset \cite[]{lee2010convergence,wang2015improved,zhang2019fast}.
-This shrinkage makes it dangerous to use the projected PCs for analyses such as PC regression, ancestry detection and correction for ancestry.
+This shrinkage makes it potentially dangerous to use the projected PCs for analyses such as PC regression, ancestry detection and correction for ancestry.
 Finally, PC scores may capture outliers that are due to family structure, population structure or other reasons; it might be benificial to detect and remove these individuals to maximise the population structure captured by the PCA (in the case of removing a few outliers) or to restrict analyses to genetically homogeneous samples.
 
-In this paper, we derive a new implementation of PCA that can be used directly on bed/bim/fam PLINK files with some missing values, which we make available in R package bigsnpr v1.0.0.
-Then, we recall some of the pitfalls of PCA computation on genotype data and provide efficient solutions to these issues.
+In this paper, we derive a new implementation of PCA that can be used directly on bed/bim/fam PLINK files with some missing values, which we make available in a new release of R package bigsnpr (v1.0.0).
+Then, we recall some of the pitfalls of PCA computation on genotype data and provide efficient solutions to these issues, which we implement in R packages bigsnpr and bigutilsr.
 These includes accouting for LD structure in PCA, detecting outliers and projecting PCs to another study sample without shrinkage bias.
-
-[SHOULD ALSO TALK ABOUT PCADAPT -> CONFOUNDED BY LD + SHOULD COMPUTE MAHA ON UNLINKED]
 
@@
 
+\begingroup
+\renewcommand*{\arraystretch}{2}
 \begin{table}[h]
@@
 \hline
-\makecell{Detection of \\ homogeneous \\ samples} & Aberrant \cite[]{bellenguez2011robust} & Robust & Only two statistics at once? \\
+\makecell{Detection of \\ homogeneous \\ samples} & Aberrant \cite[]{bellenguez2011robust} & Robust & Only two statistics at once \\
 \hline
@@
 \end{table}
+\endgroup
 
@@
 
+When computing these distances on $K$ approximately independent variables such as the first $K$ PC scores, the computed squared distances follow a $\chi^2(K)$ distribution.
+
 \subsection{Detecting LD structure in PCA}
 
-For detecting outlier variants in PCA that are due to long-range Linkage Disequilibrium (LD) regions, we use a slightly modified version of the procedure described by \cite{prive2017efficient}.
+For detecting outlier variants in PCA that are due to long-range Linkage Disequilibrium (LD) regions, we use a similar procedure as described by \cite{prive2017efficient}.
 We first apply a first round of clumping, prioritising variants by higher minor allele count, at e.g.\ $r^2$ > 0.2.
@@
 This captures both population-differentiating variants as well as long-range LD regions; to capture consecutive outliers that corresponds to long-range LD regions, we apply a Gaussian smoothing to these statistics.
-
-[FIGURE TO ILLUSTRATE THIS?]
 
@@
 The idea is that an outlier should be far from all other points, but this point is even more outlier if its KNNs are in a very dense cluster.
-In practice, we use $pd_j / \sqrt{\frac{1}{K} \sum_{k=1}^K pd_{j_k}}$ (note the squared root) as it otherwise detect any point next to a very dense cluster as an outlier.
-We implement this statistic in function \texttt{prob\_dist} of our new R package bigutilsr.
+In practice, we use $pd_j / \sqrt{\frac{1}{K} \sum_{k=1}^K pd_{j_k}}$ (note the square root) as it otherwise detect any point next to a very dense cluster as an outlier.
+We implement this statistic in function \texttt{prob\_dist} of R package bigutilsr.
 To make it fast, we use and parallelise (\texttt{knn\_parallel}) the fast K nearest neighbours implementation of R package nabor \cite[]{elseberg2012comparison}.
-Automatic detection of outliers is very hard, then we recommend users to choose a threshold of outlierness for this statistic based on visual inspection.
+Automatic outlier detection is very hard, therefore we recommend users to choose a threshold of outlierness for this statistic based on visual inspection (histogram of these statistics and PC scores colored by these statistics).
 
-For detecting samples that have a different ancestry from most of the samples in the data, we compute the pairwise orthogonalized Gnanadesikan-Kettenrin robust Mahalanobis distance on PC scores (Section \ref{maha}).
+For detecting samples that have a different ancestry from most of the samples in the data, we compute the pairwise orthogonalized Gnanadesikan-Kettenrin robust Mahalanobis distances on PC scores (Section \ref{maha}).
 We then restrict to individuals whose log-distance (alternatively p-value) is smaller (larger) than some threshold determined based on visual inspection.
@@
 To project PCs of a reference dataset (e.g.\ 1000G) to a target genotype dataset, we implement the following 3 steps in function \texttt{bed\_projectPCA} of package bigsnpr: 1) matching the variants of each dataset, including removing ambiguous alleles [A/T] and [C/G], and matching strand and direction of the alleles; 2) computing PCA of the reference dataset using the matched variants only; 3) projecting computed PCs to the target data using an optimised implementation (see Supplementary Materials) of the Online Augmentation, Decomposition, and Procrustes (OADP) transformation \cite[]{zhang2019fast}.
+To project individuals from the same dataset as the ones used for computing PCA, you can use function \texttt{bed\_projectSelfPCA}. 
+Note that the new individuals to be projected should not be related to the ones used for computing PCA.
 
@@
 We provide a subsetted version of the 1000 genomes (1000G) project data \cite[]{10002015global,meyer2019genotype}.
-Variants are restricted to the ones in common with HapMap3 and UK Biobank \cite[]{international2010integrating,bycroft2017genome}. 
-Moreover, we apply some quality control filters; we remove variants having a minor allele frequency < 0.01, variants or individuals with more than 10\% missing value, variants with P-value of the Hardy-Weinberg exact test < $10^{-50}$, and non-autosomal variants.
+Variants are restricted to the ones in common with HapMap3 or UK Biobank \cite[]{international2010integrating,bycroft2017genome}. 
+Moreover, we apply some quality control filters; we remove variants having a minor allele frequency < 0.01, variants with P-value of the Hardy-Weinberg exact test < $10^{-50}$, and non-autosomal variants.
 To remove related individuals with second-degree relationship or more, we apply KING cutoff of 0.0884 to the data using PLINK 2.0 \cite[]{manichaikul2010robust,chang2015second}.
-This results in 2490 individuals of the 1000G project (phase 3) in PLINK bed/bim/fam format. 
+This results in 2490 individuals and 1,664,852 variants of the 1000G project (phase 3) in PLINK bed/bim/fam format. 
 Resulting PLINK files and R code to generate these files are available at \url{https://doi.org/10.6084/m9.figshare.9208979.v3}. 
-To easily download these data, we provide a function called \texttt{download\_1000G} in R package bigsnpr.
-
-[REMOVE QC NA FOR 1000G?]
+To easily download these data, we provide function \texttt{download\_1000G} in R package bigsnpr.
 
@@
 
-\noindent[TALK ABOUT REMOVING OF LRLDR?]
-
 \subsection{Performance}
@@
 It then takes 34 minutes to compute the first 20 PCs using these 261,307 variants.
-If performing the automatic procedure for LD detection, it takes 5 hours to perform the initial clumping step, 6 computation of PCs and 5 outlier variant detection (i.e.\ 5 iterations of outlier detection and one final computation of PCs).
+If performing the automatic procedure for LD detection, it takes 5 hours to perform the initial clumping step, 6 rounds of computation of PCs and 5 rounds of outlier variant detection (i.e.\ 5 iterations of outlier detection and one final computation of PCs).
 
@@
 
-Finally, it takes only 6 minutes to compute the 30-Nearest Neighbours of PC scores for all UK Biobank individuals, which is the main computational aspect of computing statistics for detection individual outlier samples (Section \ref{outlier-sample}).
+Finally, it takes only 6 minutes to compute the 30-Nearest Neighbours of 20 PC scores for 406,545 UK Biobank individuals, which is the main computational aspect of computing statistics for detection individual outlier samples (Section \ref{outlier-sample}).
+
+[SHOULD INCLUDE PLINK 2.0 WITH AND WITHOUT AVX2?]
 
@@
 We show that we are able to capture at least 40 PCs that display visual population structure (Figures \ref{fig:UKBB-scores2} and \ref{fig:UKBB-screeplot2}).
-We then projected all 439,429 remaining individuals from UKBB onto this PCA space in 41 minutes.
-Note that these individuals are not supposed to be related to any of the 48,942 individuals used for training PCA because we removed both individuals from each pair of related individuals in UKBB.
-Projection of new individuals show again clear shrinkage when using simple projection (between 1.00 for PC1 and 1.80 for PC50), but no visible bias when using OADP projection (Figure \ref{fig:projUKBB-related}).
+We then projected all 439,429 remaining individuals from UKBB onto this PCA space in 21 minutes only.
+Note that these individuals should not be related to any of the 48,942 individuals used for training PCA because we removed both individuals from each pair of related individuals in UKBB.
+Projection of new individuals show again a clear shrinkage when using simple projection (between 1.00 for PC1 and 1.80 for PC50), but no visible bias when using OADP projection (Figure \ref{fig:projUKBB-related}).
 
@@
 \noindent[COMPUTE MAHA ON LOADINGS TO SHOW THE PEAKS SUMMARIZED IN ALL 16-18 PCS?]
-
-\noindent[SHOW THAT PROCEDURE IS OKAY FOR SMALLER DATASETS?]
-
-\noindent[SHOW ABLE TO RECONSTRUCT PCA? MAYBE NOT PC17]
 
@@
 \noindent[LIMITATION 2: FAIL TO PROJECT RELATED INDIVIDUALS WITH OADP -> RESTRICT MORE.]
+
+\noindent[ADD DISCU ON OTHER USE + OTHER POSSIBLE PROBLEMS -> METHYL]
 
@@
 \centerline{\includegraphics[width=0.8\textwidth]{UKBB-screeplot.pdf}}
-\caption{Scree plot: plot of singular values computed from the UK Biobank.
+\caption{Scree plot: plot of singular values computed on the UK Biobank.
 \label{fig:UKBB-screeplot}}
@@
 \centerline{\includegraphics[width=0.9\textwidth]{UKBB-PC1-20.png}}
-\caption{Principal Component (PC) scores 1 to 20 computed from the UK Biobank.
+\caption{Principal Component (PC) scores 1 to 20 computed on the UK Biobank.
 \label{fig:UKBB-scores}}
@@
 \centerline{\includegraphics[width=0.9\textwidth]{UKBB-loadings.pdf}}
-\caption{Principal Component (PC) loadings 1 to 20 computed from the UK Biobank.
+\caption{Principal Component (PC) loadings 1 to 20 computed on the UK Biobank.
 \label{fig:UKBB-loadings}}
@@
 \centerline{\includegraphics[width=0.8\textwidth]{UKBB-screeplot-restricted.pdf}}
-\caption{Scree plot: plot of singular values computed from the UK Biobank using 48,942 individuals of diverse ancestries.
+\caption{Scree plot: plot of singular values computed on the UK Biobank using 48,942 individuals of diverse ancestries.
 \label{fig:UKBB-screeplot2}}
@@
 \centerline{\includegraphics[width=0.9\textwidth]{UKBB-scores-restricted.png}}
-\caption{Principal Component (PC) scores 27 to 50 computed from the UK Biobank using 48,942 individuals of diverse ancestries. Different colors represent different self-reported ancestries.
+\caption{Principal Component (PC) scores 27 to 50 computed on the UK Biobank using 48,942 individuals of diverse ancestries. Different colors represent different self-reported ancestries.
 \label{fig:UKBB-scores2}}

### 2019-11-07 code/outlier-samples.R

@@ -1,6 +1,7 @@
 #### Detection of individual outliers ####
 
 library(bigsnpr)
+library(ggplot2)
 
 obj.bed <- bed(download_1000G("tmp-data"))
 obj.svd <- bed_autoSVD(obj.bed, k = 20, ncores = nb_cores())
@@ -8,7 +9,6 @@ obj.svd <- bed_autoSVD(obj.bed, k = 20, ncores = nb_cores())
 prob <- bigutilsr::prob_dist(obj.svd$u)
 hist(S <- prob$dist.self / sqrt(prob$dist.nn))
 
-library(ggplot2)
 plot_grid(plotlist = lapply(7:10, function(k) {
   plot(obj.svd, type = "scores", scores = 2 * k - 1:0, coeff = 0.6) +
     aes(color = S) +
@@ -17,9 +17,8 @@ plot_grid(plotlist = lapply(7:10, function(k) {
 
 # ggsave("figures/outliers-1000G.pdf", width = 7, height = 5)
 
-#### Restricting to homogeneous individuals ####
 
-library(bigsnpr)
+#### Restricting to homogeneous individuals ####
 
 obj.bed <- bed("~/Bureau/Dubois2010_data/FinnuncorrNLITUK3hap550_QC.bed")
 obj.svd <- bed_autoSVD(obj.bed, k = 10, ncores = nb_cores())
@@ -39,7 +38,8 @@ library(ggplot2)
 plot_grid(plotlist = lapply(1:2, function(k) {
   plot_grid(
     plot(obj.svd, type = "scores", scores = 2 * k - 1:0, coeff = 0.7) +
-      aes(color = pop_celiac),
+      aes(color = pop_celiac) +
+      labs(color = "Population"),
     plot(obj.svd, type = "scores", scores = 2 * k - 1:0, coeff = 0.7) +
       aes(color = pval < 0.001) +
       scale_colour_viridis_d(),
@@ -47,4 +47,4 @@ plot_grid(plotlist = lapply(1:2, function(k) {
   )
 }), scale = 0.95)
 
-# ggsave("figures/homogeneous.pdf", width = 7, height = 5)
+# ggsave("figures/homogeneous.pdf", width = 8, height = 6)

### 2019-11-07 code/outlier-samples.R
outlier detection

@@ -6,8 +6,17 @@ library(ggplot2)
 obj.bed <- bed(download_1000G("tmp-data"))
 obj.svd <- bed_autoSVD(obj.bed, k = 20, ncores = nb_cores())
 
-prob <- bigutilsr::prob_dist(obj.svd$u)
-hist(S <- prob$dist.self / sqrt(prob$dist.nn))
+prob <- bigutilsr::prob_dist(obj.svd$u, ncores = nb_cores())
+S <- prob$dist.self / sqrt(prob$dist.nn)
+
+ggplot() +
+  geom_histogram(aes(S), color = "#000000", fill = "#000000", alpha = 0.5) +
+  scale_x_continuous(breaks = 0:5 / 5, limits = c(0, NA)) +
+  scale_y_sqrt(breaks = c(10, 100, 500)) +
+  theme_bigstatsr() +
+  labs(x = "Statistic of outlierness", y = "Frequency (sqrt-scale)")
+
+# ggsave("figures/hist-outliers-1000G.pdf", width = 9, height = 6)
 
 plot_grid(plotlist = lapply(7:10, function(k) {
   plot(obj.svd, type = "scores", scores = 2 * k - 1:0, coeff = 0.6) +
@@ -17,34 +26,39 @@ plot_grid(plotlist = lapply(7:10, function(k) {
 
 # ggsave("figures/outliers-1000G.pdf", width = 7, height = 5)
 
-
-#### Restricting to homogeneous individuals ####
-
-obj.bed <- bed("~/Bureau/Dubois2010_data/FinnuncorrNLITUK3hap550_QC.bed")
-obj.svd <- bed_autoSVD(obj.bed, k = 10, ncores = nb_cores())
-
-plot(obj.svd, type = "scores", scores = 1:10, coeff = 0.7)
-
-dist <- bigutilsr::covRob(obj.svd$u, estim = "pairwiseGK")$dist
-pval <- pchisq(dist, df = 10, lower.tail = FALSE)
-
-# Get population from external files
-pop.files <- list.files(path = "~/Bureau/Dubois2010_data/",
-                        pattern = "cluster_*", full.names = TRUE)
-pop <- snp_getSampleInfos(obj.bed, pop.files)[[1]]
-pop_celiac <- c("Netherlands", "Italy", "UK", "UK", "Finland")[pop]
-
-library(ggplot2)
-plot_grid(plotlist = lapply(1:2, function(k) {
-  plot_grid(
[…]

### 2019-11-08 code/homogeneous-samples.R
figs homogeneous

@@ -1,0 +1,121 @@
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+
+rel <- fread2("data/ukb25589_rel_s488346.dat")
+fam <- fread2("data/ukbb_bed/ukbb_488282.fam")
+ind.row <- which(!fam$V2 %in% rel$ID2)
+obj.bed <- bed("data/ukbb_bed/ukbb_488282.bed")
+
+obj.svd <- readRDS("tmp-results/SVD_UKBB.rds")
+PC <- predict(obj.svd)
+
+dist <- bigutilsr::covRob(PC, estim = "pairwiseGK")$dist
+
+ggplot() +
+  geom_histogram(aes(log(dist)), color = "#000000", fill = "#000000", alpha = 0.5) +
+  scale_x_continuous(breaks = seq(0, 20, by = 2)) +
+  theme_bigstatsr() +
+  labs(x = "log Mahalanobis distance", y = "Frequency")
+
+# ggsave("figures/hist-Maha-dist.pdf", width = 9, height = 6)
+
+plot_grid(plotlist = lapply(1:10, function(k) {
+  k1 <- 2 * k - 1; k2 <- 2 * k
+  qplot(PC[, k1], PC[, k2]) +
+    labs(x = paste0("PC", k1), y = paste0("PC", k2)) +
+    theme_bigstatsr(0.5) +
+    aes(color = dist) +
+    theme(legend.position = "none") +
+    scale_color_viridis_c(trans = "log") +
+    coord_equal()
+}), ncol = 5)
+
+# ggsave("figures/UKBB-Maha-dist.png", width = 11, height = 8)
+
+## Self-reported ancestry (https://biobank.ctsu.ox.ac.uk/crystal/coding.cgi?id=1001)
+code_ancestry <- fread2("data/coding1001.tsv")
+csv <- "data/ukb22544.csv"
+df0 <- fread2(csv, select = c("eid", "21000-0.0", "22006-0.0"),
+              col.names = c("eid", "pop", "is_caucasian")) %>%
+  mutate(
+    pop  = factor(pop, levels = code_ancestry$coding,
+                  labels = code_ancestry$meaning),
+    is_caucasian = as.logical(is_caucasian)
+  )
+pop_UKBB <- df0$pop[match(obj.bed$fam$sample.ID, df0$eid)]
+
[…]

### 2019-11-08 paper-bedpca.tex
save

@@
 
+\noindent[ADD: implement for bed files with missing values (as opposed to a complete analysis with bigstatsr/bigsnpr), because useful on its own (and can be use only genotyped SNPs).]
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \centering
-\caption{Overview of existing methods. [TODO: ADD CITATIONS + RESOLVE QUESTIONS]} 
+\caption{Overview of existing methods.} 
 \label{tab:method-overview}
@@
 \hline
-Analysis & Method / Software & Pros & Cons \\
+Analysis & Method and/or Software & Pros & Cons \\
 \hline
@@
  & FlashPCA2 \cite[]{abraham2017flashpca2} & Fast and accurate & Sequential implementation \\
- & \makecell{PLINK 2.0 (reimplementation of FastPCA) \\ \cite[]{galinsky2016fast,chang2015second}} & Very fast (AVX2 only?) & \makecell{Not always 100\% accurate \\ \cite[]{abraham2017flashpca2,prive2017efficient}} \\
+ & \makecell{PLINK 2.0 (reimplementation of FastPCA) \\ \cite[]{galinsky2016fast,chang2015second}} & Fast & \makecell{Possible lack of accuracy \\ \cite[]{abraham2017flashpca2,prive2017efficient}} \\
 \hline
-\makecell{Detection of \\ oultier samples} & \makecell{"6 SDs from the mean" \\ (EIGENSOFT, \cite{patterson2006population})} & Simple & Assume Gaussian distritbution \\
+\makecell{Detection of \\ oultier samples} & \makecell{``6 SDs from the mean'' \\ in EIGENSOFT \cite[]{patterson2006population}} & Simple & Assume Gaussian distritbution \\
 \hline
-\makecell{Detection of \\ homogeneous \\ samples} & Aberrant \cite[]{bellenguez2011robust} & Robust & Only two statistics at once \\
+\makecell{Detection of \\ homogeneous \\ samples} & R package aberrant \cite[]{bellenguez2011robust} & Robust & Only two statistics at once \\
 \hline
 \multirow{5}{*}{\makecell{Projection of \\ new individuals \\ onto reference \\ PCA space}} & Simple projection (multiplication by loadings) & Simple & Shrinkage biased \\
-  & Bias-adjusted projection \cite[]{dey2019asymptotic} & Independent of new samples & \makecell{Assume same shrinkage for all individuals \\ + model-based \\ + need all eigenvalues of reference} \\
-  & \makecell{Augmentation, Decomposition, \\ and Procrustes transformation \\ (ADP, \cite{wang2015improved})} & Accurate & Slow \\
-  & Online ADP (OADP, \cite{zhang2019fast}) & Much faster than ADP & Does not work for related individuals \\ 
+  & Bias-adjusted projection in R package hdpca \cite[]{dey2019asymptotic} & Independent of new samples & \makecell{Assume same shrinkage for all individuals \\ + model-based \\ + need all eigenvalues of reference} \\
+  & \makecell{Augmentation, Decomposition and Procrustes transformation \\ (ADP) in LASER 2.0 \cite[]{wang2015improved}} & Accurate & Slow (a new PCA for each new sample) \\
+  & Online ADP (OADP) in python package FRAPOSA \cite[]{zhang2019fast} & Much faster than ADP & \makecell{Does not work for related individuals} \\ 
 \hline
@@
 When there is some missing values, we compute the partial SVD similarly, except that missing values are replaced by the variant means (i.e.\ $G_{i,j} - 2 \hat{f_j} = 0$ when $G_{i,j}$ is missing) and  the $\hat{f_j}$ for each variant are estimated using only non missing values of each variant.
-Note that this decomposition is equivalent to the decomposition presented above after imputation by the variant means [SURE?].
+Note that this decomposition is equivalent to the decomposition presented above after imputation by the variant means.
+
+To compute this decomposition easily and efficiently, we implement an accessor that memory-map the bed file to use it directly as if it were a standard matrix.
+Then, we apply the same algorithm for partial SVD that is used in R package bigstatsr \cite[]{prive2017efficient}.
 
@@
 To summarise the contribution of each variant in all $K$ PC loadings, we compute the robust Mahalanobis distances (Section \ref{maha}) of these PC loadings.  
-This captures both population-differentiating variants as well as long-range LD regions; to capture consecutive outliers that corresponds to long-range LD regions, we apply a Gaussian smoothing to these statistics.
+To capture consecutive outliers that corresponds to long-range LD regions, we apply a Gaussian smoothing to these statistics.
 
@@
 The standard upper limit defined by Tukey's rule is $q_{75\%}(x) + 1.5 \cdot IQR(x)$, where $x$ is the vector of computed statistics and $IQR(x) = q_{75\%}(x) - q_{25\%}(x)$ is the interquartile range.
-However, there are two pitfalls when using Tukey's rule.
-First, Tukey's rule assumes a normally distributed sample.
-For example, when the data is skewed, it does not work that well.
-We account for skewness using the medcouple \cite[]{hubert2008adjusted}.
+One assumption of Tukey's rule is that the sample is normally distributed; we account for skewness in the data using the medcouple \cite[]{hubert2008adjusted}. 
 Standard Tukey's rule also uses a fixed coefficient (1.5) that does not account for multiple testing, which means that for large samples, there are always some outliers detected when using 1.5. 
-To solve these two issues, we implement \texttt{tukey\_mc\_up} in R package bigutilsr and use it here, which accounts both for skewness and multiple testing by default.
+To solve these two potential issues, we implement \texttt{tukey\_mc\_up} in R package bigutilsr and use it here, which accounts both for skewness and multiple testing by default.
 
@@
 For detecting outlier samples in PCA, we use a modified version of the Probabilistic Local Outlier Factor (PLOF) statistic on PCs \cite[]{kriegel2009loop}.
-Using K nearest neighbours (KNN), this consists in comparing the distance from an point $j$ to its KNNs ($pd_j = \sqrt{\frac{1}{K} \sum_{k=1}^K d^2_{j \leftrightarrow j_k}}$, where $j_k$ is the $k$-th NN of $j$) with the distances from its KNNs to their KNNs ($\frac{1}{K} \sum_{k=1}^K pd_{j_k}$).
+Using K nearest neighbours (KNN), this consists in comparing the distance from a point $j$ to its KNNs ($pd_j = \sqrt{\frac{1}{K} \sum_{k=1}^K d^2_{j \leftrightarrow j_k}}$, where $j_k$ is the $k$-th NN of $j$) with the distances from its KNNs to their KNNs ($\frac{1}{K} \sum_{k=1}^K pd_{j_k}$).
 The idea is that an outlier should be far from all other points, but this point is even more outlier if its KNNs are in a very dense cluster.
@@
 
-[SCHEME TO ILLUSTRATE OADP?]
+[DIAGRAM TO ILLUSTRATE OADP?]
 
@@
 
+\noindent[TODO: Make a tuto with 1000G]
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\noindent[TODO SECTION ON SAMPLE OUTLIER REMOVAL / HOMOGENEOUS SAMPLES]
+\noindent[TODO SECTION ON SAMPLE OUTLIER REMOVAL / COMPARE WITH EIGENSOFT]
 
-\noindent[TALK ABOUT EFFECT OF MISSING VALUES ON PCA?]
+\noindent[TODO SECTION ON HOMOGENEOUS SAMPLES / COMPARE WITH ABERRANT]
+
+\noindent[SHOW WHAT HAPPENS WHEN NOT REMOVING RELATEDNESS IN PCA?]
+
+\noindent[MORE EMPHASIS ON COMPUTING ALL EIGEN WITH BED\_TCROSSPRODSELF()?]
 
@@
 It then takes 34 minutes to compute the first 20 PCs using these 261,307 variants.
-If performing the automatic procedure for LD detection, it takes 5 hours to perform the initial clumping step, 6 rounds of computation of PCs and 5 rounds of outlier variant detection (i.e.\ 5 iterations of outlier detection and one final computation of PCs).
+When performing the automatic procedure for LD detection, it takes 5 hours to perform the initial clumping step, 6 rounds of computation of PCs and 5 rounds of outlier variant detection (i.e.\ 5 iterations of outlier detection and one final computation of PCs).
 
@@
 
-[SHOULD INCLUDE PLINK 2.0 WITH AND WITHOUT AVX2?]
+\subsection{Outlier sample detection}
+
+To detect a few sample outliers, we compare the rule of ``6 SDs from the mean'' (6SD) to the statistic we propose in section \ref{outlier-sample}.
+Our statistic identifies only isolated samples or isolated pairs that seems to be outliers driving structure of PC17-20 (Figure \ref{fig:outlier-pd}).
+6SD identifies a lot of outliers, of which some are part of a relatively large cluster (Figure \ref{fig:outlier-sd}).
+[WHERE TO PUT FOLLOWING?] In theory, if all PCs are normally distributed, after correcting for multiple testing of 2500 individuals and 20 PCs, 6SD has a small probability of 0.0001 of detecting one outlier or more.
+
+To restrict to homogeneous samples, we compare the use of R package Aberrant as used for UKBB to the use of the robust Mahalanobis distance we propose.
 
@@
 
-\noindent[SHOW INITIAL PC SCORES OF UKBB?]
-
 \noindent[COMPUTE MAHA ON LOADINGS TO SHOW THE PEAKS SUMMARIZED IN ALL 16-18 PCS?]
@@
 
-\noindent[LIMITATION 1: RUN MULTIPLE TIMES / FAIL TO SPEED UP]
+\noindent[LIMITATION 1: RUN MULTIPLE TIMES / FAIL TO SPEED UP / PRIMME + LOW-RANK UPDATE (educated guesses)]
 
@@
 \noindent[ADD DISCU ON OTHER USE + OTHER POSSIBLE PROBLEMS -> METHYL]
+
+\noindent[OK TO PROJECT DIRECTLY UKBB (BECAUSE OF SIZE) -> INCLUDE? DO NOT CARE IF SUBSETTING]
+
+\noindent[PROMOTE VISUAL INSPECTION]
+
+\noindent[To make sure within one population. Do the Maha in one pop and then apply it to everyone]
+
+\noindent[Show pipeline to compute PCA (diagram like first paper)]
+
+\noindent[DISCUSS ABOUT PC16 -> FROM DIFFERENT ASSESSMENT CENTRES -> 36\% correlation with number of NA for all PC16 > 30]
 
@@
 \subsection*{Supplementary figures}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[htbp]
+\centering
+\begin{subfigure}[b]{0.5\textwidth}
+\includegraphics[width=\textwidth]{hist-outliers-1000G.pdf}
+\caption{Distribution of statistics}
+\end{subfigure}
+\\~\\
+\begin{subfigure}[b]{0.45\textwidth}
+\includegraphics[width=\textwidth]{outliers-1000G.pdf}
+\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by statistic of outlierness.\\~\\}
+\end{subfigure}
+~~~~~~
+\begin{subfigure}[b]{0.45\textwidth}
+\includegraphics[width=\textwidth]{be-outlier-1000G.pdf}
+\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by being detected as an outlier. Threshold of outlierness is determined based on histogram (a).}
+\end{subfigure}
+\caption{Outlier detection in the 1000 Genomes (1000G) project, using \texttt{prob\_dist} (Section \ref{outlier-sample}).\label{fig:outlier-pd}}
+\end{figure}
+
+\begin{figure}[htbp]
+\centering
+\begin{subfigure}[b]{0.5\textwidth}
+\includegraphics[width=\textwidth]{hist-outliers2-1000G.pdf}
+\caption{Distribution of statistics and default threshold (6, in red)}
+\end{subfigure}
+\\~\\
+\begin{subfigure}[b]{0.45\textwidth}
+\includegraphics[width=\textwidth]{outliers2-1000G.pdf}
+\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by statistic of outlierness (log-scale).\\~\\}
+\end{subfigure}
+~~~~~~
+\begin{subfigure}[b]{0.45\textwidth}
+\includegraphics[width=\textwidth]{be-outlier2-1000G.pdf}
+\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by being detected as an outlier. Threshold of outlierness is determined based on histogram (a).}
+\end{subfigure}
+\caption{Outlier detection in the 1000 Genomes (1000G) project, using the rule ``6 SDs from the mean'', i.e.\ where S2 is the maximum (for all PCs) number of SDs from the mean (Section \ref{outlier-sample}).\label{fig:outlier-sd}}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[htbp]
+\centering
+\begin{subfigure}[b]{0.45\textwidth}
+\includegraphics[width=\textwidth]{UKBB-White-British.png}
+\caption{PCs 1 to 20 of UKBB, colored by whether it is in the group of ``White British'' reported by the UKBB (blue).}
+\end{subfigure}
+~~~~~~
+\begin{subfigure}[b]{0.5\textwidth}
+\includegraphics[width=\textwidth]{UKBB-Maha-dist.png}
+\caption{PCs 1 to 20 of UKBB, colored by statistic of robust Mahalanobis distances (log-scale).\\}
+\end{subfigure}
+\\~\\
+\begin{subfigure}[b]{0.45\textwidth}
+\includegraphics[width=\textwidth]{hist-Maha-dist.pdf}
+\caption{Distribution of (log squared) distances.\\~\\~\\~\\}
+\end{subfigure}
+~~~~~~
+\begin{subfigure}[b]{0.45\textwidth}
+\includegraphics[width=\textwidth]{UKBB-Maha-outlier.png}
+\caption{PCs 1 to 20 of UKBB, colored by being detected as an outlier.
+Threshold of outlierness is determined based on histogram (c), where the threshold of 5 is chosen for the logarithm of the distances.}
+\end{subfigure}
+\caption{Homogeneous sample detection in the UK Biobank (UKBB), using robust Mahalanobis distances computed on the first 20 Principal Component scores (PCs) of UKBB.).\label{fig:homogeneous}}
+\end{figure}
 
@@
 \caption{Principal Component (PC) scores 1 to 20 computed on the UK Biobank.
+Different colors represent different self-reported ancestries.
 \label{fig:UKBB-scores}}

### 2019-11-09 code/homogeneous-samples.R
redo with UKBB PCs

@@ -3,13 +3,23 @@ library(bigreadr)
 library(dplyr)
 library(ggplot2)
 
-rel <- fread2("data/ukb25589_rel_s488346.dat")
-fam <- fread2("data/ukbb_bed/ukbb_488282.fam")
-ind.row <- which(!fam$V2 %in% rel$ID2)
-obj.bed <- bed("data/ukbb_bed/ukbb_488282.bed")
+## Self-reported ancestry (https://biobank.ctsu.ox.ac.uk/crystal/coding.cgi?id=1001)
+code_ancestry <- fread2("data/coding1001.tsv")
+df0 <- fread2(
+  "data/ukb22544.csv",
+  select = c("eid", "21000-0.0", "22006-0.0", paste0("22009-0.", 1:20)),
+  col.names = c("eid", "pop", "is_caucasian", paste0("PC", 1:20))
+) %>%
+  mutate(
+    pop  = factor(pop, levels = code_ancestry$coding,
+                  labels = code_ancestry$meaning),
+    is_caucasian = as.logical(is_caucasian)
+  )
 
-obj.svd <- readRDS("tmp-results/SVD_UKBB.rds")
-PC <- predict(obj.svd)
+PC <- na.omit(as.matrix(df0[-(1:3)]))
+ind_na <- attr(PC, "na.action")
+pop_UKBB <- df0$pop[-ind_na]
+
 
 dist <- bigutilsr::covRob(PC, estim = "pairwiseGK")$dist
 
@@ -19,9 +29,10 @@ ggplot() +
   theme_bigstatsr() +
   labs(x = "log Mahalanobis distance", y = "Frequency")
 
-# ggsave("figures/hist-Maha-dist.pdf", width = 9, height = 6)
+# ggsave("figures/hist-Maha-dist.pdf", width = 10, height = 6)
 
-plot_grid(plotlist = lapply(1:10, function(k) {
+
+plot_grid(plotlist = lapply(1:9, function(k) {
   k1 <- 2 * k - 1; k2 <- 2 * k
   qplot(PC[, k1], PC[, k2]) +
     labs(x = paste0("PC", k1), y = paste0("PC", k2)) +
@@ -30,64 +41,47 @@ plot_grid(plotlist = lapply(1:10, function(k) {
     theme(legend.position = "none") +
     scale_color_viridis_c(trans = "log") +
     coord_equal()
-}), ncol = 5)
+}), ncol = 3)
 
[…]

### 2019-11-09 paper-bedpca.tex
save

@@
 \noindent$^{\text{\sf 2}}$Queensland Brain Institute, University of Queensland, St. Lucia, 4072, Queensland, Australia. \\
-\noindent$^{\text{\sf 3}}$Queensland Centre for Mental Health Research, The Park Centre for Mental Health, Wacol, 4076, Queensland, Australia, \\
-\noindent$^{\text{\sf 4}}$OWKIN France, Paris, 75010, France, \\
+\noindent$^{\text{\sf 3}}$Queensland Centre for Mental Health Research, The Park Centre for Mental Health, Wacol, 4076, Queensland, Australia. \\
+\noindent$^{\text{\sf 4}}$OWKIN France, Paris, 75010, France. \\
 \noindent$^\ast$To whom correspondence should be addressed.\\
@@
 \abstract{
-
-[REDO LATER]
-
 Principal Component Analysis (PCA) of genetic data is routinely used to infer ancestry and control for population structure in various genetic analyses.
@@
 In this work, we explore these pitfalls of using PCA, and present efficient solutions to these pitfalls.
-Following applications to diverse datasets, we make recommendations for best practices and implement the proposed solutions in an efficient and user-friendly toolkit for performing these analyses in the R packages bigsnpr and bigutilsr.
+Following applications to the UK Biobank and the 1000 Genomes project datasets, we make recommendations for best practices and provide efficient and user-friendly implementations of the proposed solutions in R packages bigsnpr and bigutilsr.
 
-For example, we show that PC18 to PC40 in the UK Biobank are capturing LD structure.
-Using our automatic algorithm for removing long-range LD region, we are able to recover only 17 PCs that capture population structure only.
-Therefore, we recommend using only 17 PCs from the UK Biobank.
+For example, we show that PC19 to PC40 in the UK Biobank are capturing LD structure.
+Using our automatic algorithm for removing long-range LD region, we are able to recover only 16 PCs that capture population structure only.
+Therefore, we recommend using only 16-18 PCs from the UK Biobank.
 Another example: when projecting the PCA computed on the 1000 Genomes project, although PC1 to PC4 suffer from only moderate shrinkage (1.00-1.08), PC5 for example suffers from a shrinkage factor of 1.51 and PC10 from a shrinkage factor of 3.16.
@@
 
-Overall, we believe this work would be of interest for anyone using PCA in their analyses of genetic data.
+Overall, we believe this work would be of interest for anyone using PCA in their analyses of genetic data, as well as for other omics data.
 }
@@
 PCA has also been used to detect loci under selection \cite[]{galinsky2016fast,luu2017pcadapt}.
-Recently, the advent of large population-scale genetic datasets, such as the UK biobank data, has prompted research on developing scalable algorithms to compute PCA on data as large as the UK Biobank \cite[]{bycroft2017genome}.
-This is now possible thanks to software such as FastPCA (fast mode of EIGENSOFT), FlashPCA2, PLINK 2.0 (approx mode), bigstatsr/bigsnpr and TeraPCA \cite[]{galinsky2016fast,abraham2017flashpca2,chang2015second,prive2017efficient,bose2019terapca}.
+Recently, the advent of large population-scale genetic datasets, such as the UK biobank data, has prompted research on developing scalable algorithms to compute PCA on very large data \cite[]{bycroft2017genome}.
+This is now possible to compute PCA on these very large datasets thanks to software such as FastPCA (fast mode of EIGENSOFT), FlashPCA2, PLINK 2.0 (approx mode), bigstatsr/bigsnpr and TeraPCA \cite[]{galinsky2016fast,abraham2017flashpca2,chang2015second,prive2017efficient,bose2019terapca}.
 
@@
 This shrinkage makes it potentially dangerous to use the projected PCs for analyses such as PC regression, ancestry detection and correction for ancestry.
-Finally, PC scores may capture outliers that are due to family structure, population structure or other reasons; it might be benificial to detect and remove these individuals to maximise the population structure captured by the PCA (in the case of removing a few outliers) or to restrict analyses to genetically homogeneous samples.
+Finally, PC scores may capture outliers that are due to family structure, population structure or other reasons; it might be benificial to detect and remove these individuals to maximise the population structure captured by PCA (in the case of removing a few outliers) or to restrict analyses to genetically homogeneous samples (e.g.\ ``White British'' people in the UK Biobank).
+An overview of existing methods with their respective pros and cons is presented in table \ref{tab:method-overview}.
 
-In this paper, we derive a new implementation of PCA that can be used directly on bed/bim/fam PLINK files with some missing values, which we make available in a new release of R package bigsnpr (v1.0.0).
+For this paper, we derive a new implementation of PCA that can be used directly on bed/bim/fam PLINK files with some missing values, which we make available in a new release of R package bigsnpr (v1.0.0).
 Then, we recall some of the pitfalls of PCA computation on genotype data and provide efficient solutions to these issues, which we implement in R packages bigsnpr and bigutilsr.
-These includes accouting for LD structure in PCA, detecting outliers and projecting PCs to another study sample without shrinkage bias.
+These includes accouting for LD structure in PCA, detecting outliers in PC scores and projecting PCs to another study sample without shrinkage bias.
 
-\noindent[ADD: implement for bed files with missing values (as opposed to a complete analysis with bigstatsr/bigsnpr), because useful on its own (and can be use only genotyped SNPs).]
+[WHERE TO PUT THAT WE GET 40 PCs WHEN SUBSETTING?]
 
@@
 For detecting outlier variants in PCA that are due to long-range Linkage Disequilibrium (LD) regions, we use a similar procedure as described by \cite{prive2017efficient}.
-We first apply a first round of clumping, prioritising variants by higher minor allele count, at e.g.\ $r^2$ > 0.2.
+We first apply a first round of clumping at e.g.\ $r^2$ > 0.2, prioritising variants by higher minor allele count.
 Then, we compute $K$ PC scores and loadings.
 To summarise the contribution of each variant in all $K$ PC loadings, we compute the robust Mahalanobis distances (Section \ref{maha}) of these PC loadings.  
-To capture consecutive outliers that corresponds to long-range LD regions, we apply a Gaussian smoothing to these statistics.
+To capture consecutive outliers that corresponds to long-range LD regions, we apply a Gaussian smoothing to these statistics (moving average with a Gaussian filter over a window of 50 variants as radius by default).
 
@@
 One assumption of Tukey's rule is that the sample is normally distributed; we account for skewness in the data using the medcouple \cite[]{hubert2008adjusted}. 
-Standard Tukey's rule also uses a fixed coefficient (1.5) that does not account for multiple testing, which means that for large samples, there are always some outliers detected when using 1.5. 
+Standard Tukey's rule also uses a fixed coefficient (1.5) that does not account for multiple testing, which means that there are always some outliers detected when using 1.5 for large samples. 
 To solve these two potential issues, we implement \texttt{tukey\_mc\_up} in R package bigutilsr and use it here, which accounts both for skewness and multiple testing by default.
@@
 The idea is that an outlier should be far from all other points, but this point is even more outlier if its KNNs are in a very dense cluster.
-In practice, we use $pd_j / \sqrt{\frac{1}{K} \sum_{k=1}^K pd_{j_k}}$ (note the square root) as it otherwise detect any point next to a very dense cluster as an outlier.
+In practice, we use $pd_j / \sqrt{\frac{1}{K} \sum_{k=1}^K pd_{j_k}}$. Note the square root, as it otherwise detect any point next to a very dense cluster as an outlier.
 We implement this statistic in function \texttt{prob\_dist} of R package bigutilsr.
-To make it fast, we use and parallelise (\texttt{knn\_parallel}) the fast K nearest neighbours implementation of R package nabor \cite[]{elseberg2012comparison}.
+To make it fast, we use the fast K nearest neighbours implementation of R package nabor \cite[]{elseberg2012comparison} and paralleliseit in function \texttt{knn\_parallel} of package bigutilsr.
 Automatic outlier detection is very hard, therefore we recommend users to choose a threshold of outlierness for this statistic based on visual inspection (histogram of these statistics and PC scores colored by these statistics).
 
-For detecting samples that have a different ancestry from most of the samples in the data, we compute the pairwise orthogonalized Gnanadesikan-Kettenrin robust Mahalanobis distances on PC scores (Section \ref{maha}).
+For detecting samples that have a different ancestry from most of the samples in the data, i.e.\ for restricting to homogeneous samples, we compute the pairwise orthogonalized Gnanadesikan-Kettenrin robust Mahalanobis distances on PC scores (Section \ref{maha}).
 We then restrict to individuals whose log-distance (alternatively p-value) is smaller (larger) than some threshold determined based on visual inspection.
@@
 \subsection{Projecting PCs from a reference dataset \label{proj}}
-
-[DIAGRAM TO ILLUSTRATE OADP?]
 
@@
 
-[DESCRIBE QC FOR OTHER DATASETS TOO]
-
-\subsection{Reproducibility}
-
-[ALL CODE..]
-
-\noindent[TODO: Make a tuto with 1000G]
+In this paper, we also analyse the genotyped version of the UK Biobank \cite[]{bycroft2017genome}.
+We apply some quality control filters; we remove individuals with more than 10\% missing values, variants with more than 1\% missing values, variants having a minor allele frequency < 0.01, variants with P-value of the Hardy-Weinberg exact test < $10^{-50}$, and non-autosomal variants.
+This results in 488,371 individuals and 504,139 variants.
+When removing related individuals, we use the list of individual pairs reported by the UK Biobank.
 
@@
 
-\noindent[TODO SECTION ON SAMPLE OUTLIER REMOVAL / COMPARE WITH EIGENSOFT]
-
 \noindent[TODO SECTION ON HOMOGENEOUS SAMPLES / COMPARE WITH ABERRANT]
-
-\noindent[SHOW WHAT HAPPENS WHEN NOT REMOVING RELATEDNESS IN PCA?]
-
-\noindent[MORE EMPHASIS ON COMPUTING ALL EIGEN WITH BED\_TCROSSPRODSELF()?]
 
@@
 To detect a few sample outliers, we compare the rule of ``6 SDs from the mean'' (6SD) to the statistic we propose in section \ref{outlier-sample}.
-Our statistic identifies only isolated samples or isolated pairs that seems to be outliers driving structure of PC17-20 (Figure \ref{fig:outlier-pd}).
+Our statistic identifies only isolated samples or isolated pairs that seems to be outliers driving structure of PC17-20 of 1000G (Figure \ref{fig:outlier-pd}).
 6SD identifies a lot of outliers, of which some are part of a relatively large cluster (Figure \ref{fig:outlier-sd}).
@@
 
-To restrict to homogeneous samples, we compare the use of R package Aberrant as used for UKBB to the use of the robust Mahalanobis distance we propose.
+To restrict to homogeneous samples, we compare the use of R package aberrant used to report the ``White British'' subset of UKBB homogeneous samples, to the use of the robust Mahalanobis distance we propose.
+Here, we choose a threshold of 5 and show that this gives a similar subset of individuals than the ``White British'' subset reported by the UK Biobank (Figure \ref{fig:homogeneous}).
+Using this threshold of 5, only 3 out of 10,936 people of Asian ancestry (1 ``Chinese'' and 2 ``Indian'') are kept, and 1 ``African'' out of 7622 poeple with Black background is kept (Table \ref{tab:homogeneous}).
+In contrast, 416,492 out of 431,090 ``British'' (96.6\%) and 12,620 out of 12759 ``Irish'' (98.9\%) are kept.
 
@@
 We use 60\% of individuals in the 1000G data (section \ref{1000G}) to compute the PCA ($K=20$ PCs) [EXPLICIT SAMPLE SIZE].
-Then, we project the remaining 40\% individuals using three methods: 1/ simply multiplying the genotypes of these individuals by the previously computed loadings, 2/ correcting the simple projections using asymptotic shrinkage factors as determined by R package hdpca v1.1.3 \cite[]{dey2019asymptotic} and 3/ the OADP projection (section \ref{proj}).
+Then, we project the remaining 40\% individuals using three methods: 1/ simply multiplying the genotypes of these individuals by the previously computed loadings, 2/ correcting the simple projections using asymptotic shrinkage factors as determined by R package hdpca v1.1.3 \cite[]{dey2019asymptotic}, with all eigenvalues derived from the Genetic Relationship Matrix (GRM) computed with function \texttt{bed\_tcrossprodSelf} of R package bigsnpr, and 3/ the OADP projection (section \ref{proj}).
 When simply projecting using loadings, there is no apparent shrinkage for PC1 and PC2, a weak [REWORD?] shrinkage for PC3 and PC4, and a large shrinkage for PC5 to PC8 (Figure \ref{fig:proj1000G}).
@@
 We show the same results when projecting the full UK Biobank data onto PCA computed using 1000G data  (Figure \ref{fig:proj1000G-5}).
-When correcting projected PC scores with asymptotic shrinkage factors, bias is smaller than when simply projecting loadings, yet, it is apparent for PC7-8 and PC11-12 (Figure \ref{fig:proj1000G-3}).
+When correcting projected PC scores with asymptotic shrinkage factors, bias is smaller than when simply projecting loadings, yet, there is an apparent bias for PC7-8 (Figure \ref{fig:proj1000G-3}).
 Finally, to assess if OADP could be used to project individuals that are related to some individuals that were used to compute PCA, we projected these individuals (as if we were projecting their monozygotic twins) using OADP.
@@
 
-\noindent[COMPUTE MAHA ON LOADINGS TO SHOW THE PEAKS SUMMARIZED IN ALL 16-18 PCS?]
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %\newpage
+
+\section*{Software and code availability}
+
+All code used for the analyses of this paper is available at \url{https://github.com/privefl/paper4-bedpca/tree/master/code}.
+
+Package bigsnpr can be installed from GitHub (see \url{https://github.com/privefl/bigsnpr}).
+
+A tutorial on the steps to perform PCA on 1000G data is presented at \url{} [TODO].
 
@@
 
+\newpage
+
 \begin{figure}[htbp]
@@
 \includegraphics[width=\textwidth]{hist-Maha-dist.pdf}
-\caption{Distribution of (log squared) distances.\\~\\~\\~\\}
+\caption{Distribution of (log squared) distances.\\~\\~\\~\\~\\}
 \end{subfigure}
@@
 
+% latex table generated in R 3.6.0 by xtable 1.8-4 package
+% Sat Nov  9 10:17:25 2019
+\begin{table}[ht]
+\centering
+\caption{Number of UKBB individuals with (log) Mahalanobis distance lower than some threshold (top), grouped by self-reported ancestry (left).} 
+\label{tab:homogeneous}
+\begin{tabular}{|l|r|r|r|r|r|r|r|r|r|r|}
+  \hline
+ & $<$ 3 & $<$ 4 & $<$ 5 & $<$ 6 & $<$ 7 & $<$ 8 & $<$ 9 & $<$ 10 & $<$ 11 & $<$ 12 \\ 
+  \hline
+Prefer not to answer & 484 & 1013 & 1062 & 1099 & 1139 & 1177 & 1279 & 1405 & 1471 & 1583 \\ 
+  Do not know & 36 & 68 & 76 & 84 & 92 & 118 & 155 & 188 & 196 & 204 \\ 
+  White & 186 & 422 & 457 & 483 & 513 & 533 & 543 & 543 & 545 & 546 \\ 
+  Mixed & 2 & 6 & 6 & 7 & 8 & 15 & 26 & 42 & 46 & 46 \\ 
+  Asian or Asian British &  &  &  &  &  & 3 & 20 & 40 & 42 & 42 \\ 
+  Black or Black British & 1 & 2 & 2 & 2 & 2 & 2 & 2 & 4 & 6 & 26 \\ 
+  Chinese &  & 1 & 1 & 1 & 1 & 2 & 5 & 21 & 1423 & 1504 \\ 
+  Other ethnic group & 57 & 230 & 261 & 314 & 469 & 885 & 1939 & 2761 & 3681 & 4356 \\ 
+  British & 191713 & 400516 & 416492 & 424490 & 427769 & 429172 & 431026 & 431082 & 431089 & 431090 \\ 
+  Irish & 1416 & 12039 & 12620 & 12700 & 12734 & 12743 & 12759 & 12759 & 12759 & 12759 \\ 
+  Any other white background & 1468 & 4747 & 6953 & 9341 & 12979 & 14613 & 15741 & 15810 & 15820 & 15820 \\ 
+  White and Black Caribbean & 1 & 4 & 4 & 4 & 9 & 35 & 142 & 537 & 589 & 597 \\ 
+  White and Black African & 1 & 3 & 3 & 4 & 6 & 29 & 99 & 333 & 400 & 402 \\ 
+  White and Asian & 4 & 7 & 13 & 23 & 79 & 350 & 651 & 790 & 802 & 802 \\ 
+  Any other mixed background & 24 & 66 & 87 & 155 & 274 & 391 & 595 & 884 & 990 & 996 \\ 
+  Indian &  & 2 & 2 & 5 & 6 & 29 & 1682 & 5700 & 5716 & 5716 \\ 
+  Pakistani &  &  &  &  & 1 & 13 & 532 & 1747 & 1748 & 1748 \\ 
+  Bangladeshi &  &  &  &  &  &  & 2 & 220 & 221 & 221 \\ 
+  Any other Asian background &  &  &  & 1 & 6 & 66 & 427 & 1364 & 1730 & 1747 \\ 
+  Caribbean &  &  &  &  &  &  & 3 & 113 & 1323 & 4299 \\ 
+  African &  & 1 & 1 & 1 & 1 & 1 & 3 & 58 & 350 & 3205 \\ 
+  Any other Black background &  &  &  &  &  & 1 & 3 & 22 & 49 & 118 \\ 
+   \hline
+  All & 195393 & 419127 & 438040 & 448714 & 456088 & 460178 & 467634 & 476423 & 480996 & 487827 \\ 
+   \hline
+\end{tabular}
+\end{table}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\vspace*{1em}
+\newpage
 
@@
 
+\newpage
+
 \begin{figure}[!htpb]
@@
 \centerline{\includegraphics[width=0.9\textwidth]{UKBB-scores-restricted.png}}
-\caption{Principal Component (PC) scores 27 to 50 computed on the UK Biobank using 48,942 individuals of diverse ancestries. Different colors represent different self-reported ancestries.
+\caption{[SHOULD BE MAIN FIGURE?] Principal Component (PC) scores 27 to 50 computed on the UK Biobank using 48,942 individuals of diverse ancestries.
+Different colors represent different self-reported ancestries.
 \label{fig:UKBB-scores2}}

### 2019-11-09 paper-bedpca.tex
save

@@
 
-In this paper, for many applications, we use the pairwise orthogonalized Gnanadesikan-Kettenrin robust Mahalanobis distance as implemented in function \texttt{covRob} of R package robust with parameter \texttt{estim = "pairwiseGK"} \cite[]{gnanadesikan1972robust,yohai1988high,maronna2002robust,todorov2009object}.
+In this paper, for many applications, we use the pairwise orthogonalised Gnanadesikan-Kettenrin robust Mahalanobis distance as implemented in function \texttt{covRob} of R package robust with parameter \texttt{estim = "pairwiseGK"} \cite[]{gnanadesikan1972robust,yohai1988high,maronna2002robust,todorov2009object}.
 We reexport function \texttt{covRob} in R package bigutilsr for convenience.
@@
 
-For detecting samples that have a different ancestry from most of the samples in the data, i.e.\ for restricting to homogeneous samples, we compute the pairwise orthogonalized Gnanadesikan-Kettenrin robust Mahalanobis distances on PC scores (Section \ref{maha}).
+For detecting samples that have a different ancestry from most of the samples in the data, i.e.\ for restricting to homogeneous samples, we compute the pairwise orthogonalised Gnanadesikan-Kettenrin robust Mahalanobis distances on PC scores (Section \ref{maha}).
 We then restrict to individuals whose log-distance (alternatively p-value) is smaller (larger) than some threshold determined based on visual inspection.
@@
 
-\noindent[TODO SECTION ON HOMOGENEOUS SAMPLES / COMPARE WITH ABERRANT]
-
 \subsection{Performance}
@@
 
-Finally, it takes only 6 minutes to compute the 30-Nearest Neighbours of 20 PC scores for 406,545 UK Biobank individuals, which is the main computational aspect of computing statistics for detection individual outlier samples (Section \ref{outlier-sample}).
+Finally, it takes only 6 minutes to compute the 30-Nearest Neighbours of 20 PC scores for 406,545 UK Biobank individuals, which is the main computational aspect of computing statistics for detecting individual outlier samples (Section \ref{outlier-sample}).
 
@@
 
-To restrict to homogeneous samples, we compare the use of R package aberrant used to report the ``White British'' subset of UKBB homogeneous samples, to the use of the robust Mahalanobis distance we propose.
+When the goal is to restrict to homogeneous samples, we compare the use of R package aberrant used to report the ``White British'' subset of UKBB homogeneous samples, to the use of the robust Mahalanobis distance we propose.
 Here, we choose a threshold of 5 and show that this gives a similar subset of individuals than the ``White British'' subset reported by the UK Biobank (Figure \ref{fig:homogeneous}).
 Using this threshold of 5, only 3 out of 10,936 people of Asian ancestry (1 ``Chinese'' and 2 ``Indian'') are kept, and 1 ``African'' out of 7622 poeple with Black background is kept (Table \ref{tab:homogeneous}).
-In contrast, 416,492 out of 431,090 ``British'' (96.6\%) and 12,620 out of 12759 ``Irish'' (98.9\%) are kept.
+In contrast, 416,492 out of 431,090 ``British'' (96.6\%) and 12,620 out of 12,759 ``Irish'' (98.9\%) are kept.
 
@@
 
-We use 60\% of individuals in the 1000G data (section \ref{1000G}) to compute the PCA ($K=20$ PCs) [EXPLICIT SAMPLE SIZE].
+We use 60\% of individuals in the 1000G data (section \ref{1000G}) to compute the PCA ($K=20$ PCs).
 Then, we project the remaining 40\% individuals using three methods: 1/ simply multiplying the genotypes of these individuals by the previously computed loadings, 2/ correcting the simple projections using asymptotic shrinkage factors as determined by R package hdpca v1.1.3 \cite[]{dey2019asymptotic}, with all eigenvalues derived from the Genetic Relationship Matrix (GRM) computed with function \texttt{bed\_tcrossprodSelf} of R package bigsnpr, and 3/ the OADP projection (section \ref{proj}).
@@
 When correcting projected PC scores with asymptotic shrinkage factors, bias is smaller than when simply projecting loadings, yet, there is an apparent bias for PC7-8 (Figure \ref{fig:proj1000G-3}).
-Finally, to assess if OADP could be used to project individuals that are related to some individuals that were used to compute PCA, we projected these individuals (as if we were projecting their monozygotic twins) using OADP.
+Finally, to assess if OADP could be used to project individuals that are related to some individuals that were used to compute PCA, we projected these 60\% individuals (as if we were projecting their monozygotic twins) using OADP.
 Projections of related individuals using OADP shows some reverse [REWORD?] shrinkage (Figure \ref{fig:proj1000G-4}).
@@
 
-Principal Components (PC) loadings reported by the UK Biobank clearly show that PC19 to PC40 capture LD structure, which is also the case for PC16 and PC18, although less pronounced (see peaks in figure \ref{fig:UKBB-loadings40}).
-Moreover, our automatic procedure removing long-range LD regions did not converge after 5 iterations, meaning that it kept detecting long-range LD regions at each iteration.
+Principal Component loadings reported by the UK Biobank clearly show that PC19 to PC40 capture LD structure, which is also the case for PC16 and PC18, although less pronounced (see peaks in figure \ref{fig:UKBB-loadings40}).
+When applying our automatic procedure to remove long-range LD regions, it did not converge after 5 iterations, meaning that it kept detecting long-range LD regions at each iteration.
 Therefore, we were able to capture only 16 PCs that shows population stratification and not LD structure (Figures \ref{fig:UKBB-screeplot}, \ref{fig:UKBB-scores} and \ref{fig:UKBB-loadings}).
 
-Then, we computed again PCA after restricting the individuals included in the computations: we randomly subsampled UKBB data to use only 10,000 British individuals (out of 431,029) and 5000 Irish individuals (out of 12,755), while keeping all individuals with other self-reported individuals.
+Then, we computed again PCA after restricting the individuals included in the computations: we randomly subsampled UKBB data to use only 10,000 British individuals (out of 431,029) and 5000 Irish individuals (out of 12,755), while keeping all individuals with other or unkonwn self-reported ancestry.
 We further removed all pairs of related individuals reported by the UKBB (both individuals in each pair). 
 This resulted in 48,942 individuals that we used to compute 50 PCs, which took less than 3 hours using function \texttt{bed\_autoSVD} (that converged after 4 iterations of automatic LD removal). 
-We show that we are able to capture at least 40 PCs that display visual population structure (Figures \ref{fig:UKBB-scores2} and \ref{fig:UKBB-screeplot2}).
-We then projected all 439,429 remaining individuals from UKBB onto this PCA space in 21 minutes only.
+We show that we are able to capture more PCs (at least 40) that display visual population structure (Figures \ref{fig:UKBB-scores2} and \ref{fig:UKBB-screeplot2}).
+We then projected all 439,429 remaining individuals from UKBB onto this PCA space in 21 minutes only using our implementation of the OADP projection.
 Note that these individuals should not be related to any of the 48,942 individuals used for training PCA because we removed both individuals from each pair of related individuals in UKBB.

### 2019-11-09 paper-bedpca.tex
add pipeline

@@
 
+\begin{figure}[!htpb]
+\centerline{\includegraphics[width=0.4\textwidth]{PCA-pipeline.pdf}}
+\caption{Pipeline for computing Principal Components (PCs) using R packages bigsnpr and bigutilsr. [ADD VISUAL INSPECTION? QC? IN THE LEGEND?]
+\label{fig:pipeline}}
+\end{figure}
+
+In this work, we compiled different pitfalls that can occur with analyses related to Principal Component Analysis (PCA) of genetic data.
+We then investigated possible solutions to these pitfalls and choose the ones that we thought to be the most relevant both with respect to properties such as accuracy and robustness, but also that are fast and easy to use.
+We then implemented these solutions in open-source R packages that are available for anyone to use.
+
+As a general recommendation for computing PCA, we show some pipeline in figure \ref{fig:pipeline}. [DESCRIBE ALL STEPS?]
+
 \noindent[LIMITATION 1: RUN MULTIPLE TIMES / FAIL TO SPEED UP / PRIMME + LOW-RANK UPDATE (educated guesses)]

### 2019-11-10 paper-bedpca.tex
add discu

--- 2019-11-09 paper-bedpca.tex
@@
 
+\noindent[MOVE FIGURE TO MAIN TEXT? E.G. \ref{fig:UKBB-scores2}]
+
 \subsection{Performance}
@@
 
-\noindent[LIMITATION 1: RUN MULTIPLE TIMES / FAIL TO SPEED UP / PRIMME + LOW-RANK UPDATE (educated guesses)]
+\begin{figure}[!htpb]
+\centerline{\includegraphics[width=0.4\textwidth]{PCA-pipeline.pdf}}
+\caption{Pipeline for computing Principal Components (PCs) using R packages bigsnpr and bigutilsr. [ADD VISUAL INSPECTION? QC? IN THE LEGEND?]
+\label{fig:pipeline}}
+\end{figure}
 
-\noindent[LIMITATION 2: FAIL TO PROJECT RELATED INDIVIDUALS WITH OADP -> RESTRICT MORE.]
+In this work, we have compiled different pitfalls that can occur with analyses related to Principal Component Analysis (PCA) of genetic data.
+Then, we have investigated possible solutions to these pitfalls and chose the ones that we thought to be the most relevant both with respect to properties such as accuracy and robustness, but also that would be fast and easy to use.
+We then implemented these solutions in open-source R packages.
+These functions can be directly used on genotypes stored as PLINK bed/bim/fam files with some missing values. 
+This is in contrast with our first release of packages bigstatsr and bigsnpr that use their own ``bigSNP'' format that can store both genotype calls and (rounded) dosages, but that mostly provide functions that require no missing value.
+Yet, PCA is a useful tool on its own and does not require extensive imputed data, therefore we decided that operating directly on PLINK files with a few missing values would be more practical for the users.
 
-\noindent[ADD DISCU ON OTHER USE + OTHER POSSIBLE PROBLEMS -> METHYL]
+As a general recommendation for computing PCA, we show some pipeline in figure \ref{fig:pipeline}. [START WITH QC?]
+This requires removing all related individuals, for which we provide an R wrapper to PLINK's implementation of KING robust kinship coefficients \cite[]{manichaikul2010robust,chang2015second}. 
+Note that one should remove both individuals in each related pairs so that projected individuals at the end are not related to the ones used for computing PCA, as we showed that relatedness is a problem when using the OADP projection.
+Then, it requires to do several steps of outlier detection, one for outlier variants that capture long-range LD variation (automatic), and one for detecting outlier samples (semi-automatic and visual).
+To fasten these steps, we tried hard to come up with a solution for not recomputing PCA from scratch when removing a few samples or variants.
+Using educated guesses based on low-rank approximations seemed to be a promising approach but did not fasten computation by much, so we did not pursue this idea \cite[]{brand2003fast,wu2017primme}.
+Once PCA is done, one should look at the loadings to make sure that there is no apparent LD structure left. 
+One should also look at PC scores to identify which PCs capture population structure.
+As in many applications, we believe that a compromise should be made between signal and noise.
+Therefore, we recommend to use only PCs that show population structure (e.g.\ first 16 PCs in figure \ref{fig:UKBB-scores}) [NOTE SURE ABOUT PC15-16..].
 
-\noindent[OK TO PROJECT DIRECTLY UKBB (BECAUSE OF SIZE) -> INCLUDE? DO NOT CARE IF SUBSETTING]
+We also provide some recommendations when the dataset is composed mainly of one population (e.g.\ British people in the UK Biobank).
+We show that it is useful to actually subset these individuals so there is not too much imbalance between the number of people from different ancestries. 
+Indeed, when subsetting British and Irish people, we are able to capture a lot more PCs that show population structure.
+This makes designs such as the 1000 Genomes project, that gathered around 100 people for each of 26 populations, highly relevant for capturing population structure.
+In contrast, when the goal is to have an homogeneous sample, we show that using the Mahalanobis distance on PC scores can efficiently achieve this goal, which we used in previous analyses \cite[]{prive2019efficient}.
+When the homogeneous sample is not predominant in the dataset, one solution is to compute the center and covariance of the robust Mahalanobis distance using only the population of interest, and then computing the distances for all individuals using these parameters.
 
-\noindent[PROMOTE VISUAL INSPECTION]
-
-\noindent[To make sure within one population. Do the Maha in one pop and then apply it to everyone]
-
-\noindent[Show pipeline to compute PCA (diagram like first paper)]
+Finally, we have showed some pitfalls when computing PCA on genotype data, yet we think most of the results in this paper are not inherent to genotype data, and can be transferred to e.g.\ other omics data as well.
+For example, outlier and homogeneous sample detection can be used on PCs of anything.
+Moreover, projection of scores will also be a problem for other omics data where the number of variables used is larger than the number of samples used for computing PCA.
+Finally, our recommendation of gathering ``populations'' of approximately the same size is relevant for other biological data as well.
+However, other pitfalls might apply when using other types of data; for example, methylation data can be confounded by factors such as age and sex, and it might be benificial to remove some probes that are associated with these confounding factors before computing PCA \cite[]{decamps2019guidelines}.
 
@@
 
-\subsection{Conclusion}
+\noindent[MORE CITATIONS IN THE DISCUSSION?]
 
@@
 
-%\newpage
+\newpage
 

### 2019-11-11 paper-bedpca.tex
save

@@
 Recently, the advent of large population-scale genetic datasets, such as the UK biobank data, has prompted research on developing scalable algorithms to compute PCA on very large data \cite[]{bycroft2017genome}.
-This is now possible to compute PCA on these very large datasets thanks to software such as FastPCA (fast mode of EIGENSOFT), FlashPCA2, PLINK 2.0 (approx mode), bigstatsr/bigsnpr and TeraPCA \cite[]{galinsky2016fast,abraham2017flashpca2,chang2015second,prive2017efficient,bose2019terapca}.
+It is now possible to efficiently approximate PCA on very large datasets thanks to software such as FastPCA (fast mode of EIGENSOFT), FlashPCA2, PLINK 2.0 (approx mode), bigstatsr/bigsnpr, TeraPCA and ProPCA \cite[]{galinsky2016fast,abraham2017flashpca2,chang2015second,prive2017efficient,bose2019terapca,agrawal2019scalable}.
 
@@
 In the following, we recall common pitfalls and explain when they are relevant.
-First, efficient methods for PCA use approximations, which can results in some lack of precision of computed PCs.
+First, some of the PCs may capture LD structure rather than population structure \cite[]{price2008long,abdellaoui2013population,prive2017efficient}.
+Including PCs that capture LD as covariates in genetic analyses can lead to reduced power, e.g.\ for detecting genetic association within these LD regions in GWAS.
+Second, another issue may arise when projecting PCs of a reference dataset to another study dataset: projected PCs are shrunk towards 0 in the new dataset \cite[]{lee2010convergence,wang2015improved,zhang2019fast}.
+This shrinkage makes it potentially dangerous to use the projected PCs for analyses such as PC regression, ancestry detection and correction for ancestry.
+Third, PC scores may capture outliers that are due to family structure, population structure or other reasons; it might be benificial to detect and remove these individuals to maximise the population structure captured by PCA (in the case of removing a few outliers) or to restrict analyses to genetically homogeneous samples (e.g.\ ``White British'' people in the UK Biobank).
+Finally, efficient methods for PCA use approximations, which can results in some lack of precision of computed PCs.
 This has been shown to be a possible issue for software such as FastPCA and PLINK 2.0, but not for FlashPCA2 and bigstatsr/bigsnpr \cite[]{abraham2017flashpca2,prive2017efficient}.
-Second, some of the PCs may capture LD structure rather than population structure \cite[]{price2008long,abdellaoui2013population,prive2017efficient}.
-Including PCs that capture LD as covariates in genetic analyses can lead to reduced power, e.g.\ for detecting genetic association within these LD regions in GWAS.
-Third, another issue may arise when projecting PCs of a reference dataset to another study dataset: projected PCs are shrunk towards 0 in the new dataset \cite[]{lee2010convergence,wang2015improved,zhang2019fast}.
-This shrinkage makes it potentially dangerous to use the projected PCs for analyses such as PC regression, ancestry detection and correction for ancestry.
-Finally, PC scores may capture outliers that are due to family structure, population structure or other reasons; it might be benificial to detect and remove these individuals to maximise the population structure captured by PCA (in the case of removing a few outliers) or to restrict analyses to genetically homogeneous samples (e.g.\ ``White British'' people in the UK Biobank).
-An overview of existing methods with their respective pros and cons is presented in table \ref{tab:method-overview}.
+An overview of existing methods with their respective advantages and limitations is presented in table \ref{tab:method-overview}.
 
-For this paper, we derive a new implementation of PCA that can be used directly on bed/bim/fam PLINK files with some missing values, which we make available in a new release of R package bigsnpr (v1.0.0).
-Then, we recall some of the pitfalls of PCA computation on genotype data and provide efficient solutions to these issues, which we implement in R packages bigsnpr and bigutilsr.
-These includes accouting for LD structure in PCA, detecting outliers in PC scores and projecting PCs to another study sample without shrinkage bias.
-
-[WHERE TO PUT THAT WE GET 40 PCs WHEN SUBSETTING?]
+For this paper, we derive implementations of truncated PCA and other useful functions for e.g.\ performing LD thinning and computing variant statistics.
+We make these available in a new release of R package bigsnpr (v1.0.0); what differs from previously available functions presented in \cite{prive2017efficient} is that these new functions can be used directly on bed/bim/fam PLINK files with some missing values.
+We use them to analyse the UK Biobank data, and show that these functions are both very fast and easy to use.
+We also point out that many PCs currently reported by the UK Biobank capture LD structure instead of population structure.
+Interestingly, subsetting the UK Biobank data enables to get more PCs capturing population structure than when using the whole sample (\textasciitilde40 instead of \textasciitilde16).
+Then, we project the other individuals that were not used in the PCA calculation, show that this projection is biased and provide an efficient solution to get unbiased projections instead.
+Finally, we talk about detecting outlier samples in PCA, either a few outlier samples that may correspond to e.g.\ batch effects, or when the goal is to restrict the data to homogeneous individuals.
 
@@
 \hline
-Analysis & Method and/or Software & Pros & Cons \\
+Analysis & Method and/or Software & Advantages & Limitations \\
 \hline
@@
 
-When there is some missing values, we compute the partial SVD similarly, except that missing values are replaced by the variant means (i.e.\ $G_{i,j} - 2 \hat{f_j} = 0$ when $G_{i,j}$ is missing) and  the $\hat{f_j}$ for each variant are estimated using only non missing values of each variant.
+When there are some missing values, we compute the partial SVD similarly, except that missing values are replaced by the variant means (i.e.\ $G_{i,j} - 2 \hat{f_j} = 0$ when $G_{i,j}$ is missing) and  the $\hat{f_j}$ for each variant are estimated using only non missing values of each variant.
 Note that this decomposition is equivalent to the decomposition presented above after imputation by the variant means.
@@
 
-In this paper, for many applications, we use the pairwise orthogonalised Gnanadesikan-Kettenrin robust Mahalanobis distance as implemented in function \texttt{covRob} of R package robust with parameter \texttt{estim = "pairwiseGK"} \cite[]{gnanadesikan1972robust,yohai1988high,maronna2002robust,todorov2009object}.
+Mahalanobis distances are computed as $d(x)^2 = (x-\mu)^T \Sigma^{-1} (x-\mu),$ where $\mu$ and $\Sigma$ are (robust) estimators of location and covariance of $x$.
+We use these distances for many applications in this paper. 
+When $x$ is multivariate Gaussian data with $K$ dimensions, the squared distances follow a $\chi^2(K)$ distribution.
+If $x$ are PC scores of centered data and if we use standard estimates, then $\mu = 0$ and $\Sigma = I_K$.
+Yet, here we use the pairwise orthogonalised Gnanadesikan-Kettenrin robust Mahalanobis distance as implemented in function \texttt{covRob} of R package robust with parameter \texttt{estim = "pairwiseGK"} \cite[]{gnanadesikan1972robust,yohai1988high,maronna2002robust,todorov2009object}.
 We reexport function \texttt{covRob} in R package bigutilsr for convenience.
-
-When computing these distances on $K$ approximately independent variables such as the first $K$ PC scores, the computed squared distances follow a $\chi^2(K)$ distribution.
 
@@
 The standard upper limit defined by Tukey's rule is $q_{75\%}(x) + 1.5 \cdot IQR(x)$, where $x$ is the vector of computed statistics and $IQR(x) = q_{75\%}(x) - q_{25\%}(x)$ is the interquartile range.
-One assumption of Tukey's rule is that the sample is normally distributed; we account for skewness in the data using the medcouple \cite[]{hubert2008adjusted}. 
+One assumption of Tukey's rule is that the sample is normally distributed; we account for skewness in the data using the medcouple as implemented in function \texttt{adjboxStats} of R package robustbase \cite[]{brys2004robust,hubert2008adjusted}.
 Standard Tukey's rule also uses a fixed coefficient (1.5) that does not account for multiple testing, which means that there are always some outliers detected when using 1.5 for large samples. 
@@
 We implement this statistic in function \texttt{prob\_dist} of R package bigutilsr.
-To make it fast, we use the fast K nearest neighbours implementation of R package nabor \cite[]{elseberg2012comparison} and paralleliseit in function \texttt{knn\_parallel} of package bigutilsr.
-Automatic outlier detection is very hard, therefore we recommend users to choose a threshold of outlierness for this statistic based on visual inspection (histogram of these statistics and PC scores colored by these statistics).
+To make it fast, we use the fast K nearest neighbours implementation of R package nabor \cite[]{elseberg2012comparison} and parallelise it in function \texttt{knn\_parallel} of package bigutilsr.
+Automatic outlier detection is difficult, therefore we recommend users to choose a threshold of outlierness for this statistic based on visual inspection (histogram of these statistics and PC scores colored by these statistics).
 
@@
 Moreover, we apply some quality control filters; we remove variants having a minor allele frequency < 0.01, variants with P-value of the Hardy-Weinberg exact test < $10^{-50}$, and non-autosomal variants.
-To remove related individuals with second-degree relationship or more, we apply KING cutoff of 0.0884 to the data using PLINK 2.0 \cite[]{manichaikul2010robust,chang2015second}.
+To remove related individuals with second-degree relationship or more, we apply KING-relatedness cutoff of 0.0884 to the data using PLINK 2.0 \cite[]{manichaikul2010robust,chang2015second}.
 This results in 2490 individuals and 1,664,852 variants of the 1000G project (phase 3) in PLINK bed/bim/fam format. 
@@
 
-\noindent[MOVE FIGURE TO MAIN TEXT? E.G. \ref{fig:UKBB-scores2}]
+\subsection{Application to the UK Biobank}
 
-\subsection{Performance}
-
-To show that we provide very fast implementations of the different methods presented in this paper, we apply them to the UK Biobank. We use 20 physical cores for most of the computations.
+To demonstrate that we provide very fast implementations of the different methods presented in this paper, we apply them to the UK Biobank. We use 20 physical cores for most of the computations.
 It takes 22 minutes to perform a first phase of clumping on 406,545 unrelated individuals genotyped over 504,139 variants, which reduces the number of variants to 261,307.
@@
 
-It takes 8 minutes to match the 1000G data to the UKBB data and compute 20 PCs of the 1000G data using the automatic LD detection technique.
-It takes 12 minutes more to do the OADP projection of all 488,371 individuals of the UKBB data to the PCA computed using the 1000G data.
+When applying our automatic procedure to remove long-range LD regions, it did not converge after 5 iterations, meaning that it kept detecting long-range LD regions at each iteration.
+Therefore, we were able to capture only 16 PCs that show stratification that is not LD structure (Figures \ref{fig:UKBB-screeplot}, \ref{fig:UKBB-scores} and \ref{fig:UKBB-loadings}).
+Interestingly, PC loadings reported by the UK Biobank clearly show that PC19 to PC40 capture LD structure, which is also the case for PC16 and PC18, although less pronounced (see peaks in figure \ref{fig:UKBB-loadings40}).
 
+\begin{figure}[!htpb]
+\centerline{\includegraphics[width=0.9\textwidth]{UKBB-loadings1-40.pdf}}
+\caption{Principal Component (PC) loadings 1 to 40 reported by the UK Biobank.
+\label{fig:UKBB-loadings40}}
+\end{figure}
+
+As for other analyses, it takes 8 minutes to match the 1000G data to the UKBB data and compute 20 PCs of the 1000G data using the automatic LD detection technique.
+It takes 12 minutes more to do the OADP projection of all 488,371 individuals of the UKBB data onto the PCA computed using the 1000G data.
 Finally, it takes only 6 minutes to compute the 30-Nearest Neighbours of 20 PC scores for 406,545 UK Biobank individuals, which is the main computational aspect of computing statistics for detecting individual outlier samples (Section \ref{outlier-sample}).
@@
 Then, we project the remaining 40\% individuals using three methods: 1/ simply multiplying the genotypes of these individuals by the previously computed loadings, 2/ correcting the simple projections using asymptotic shrinkage factors as determined by R package hdpca v1.1.3 \cite[]{dey2019asymptotic}, with all eigenvalues derived from the Genetic Relationship Matrix (GRM) computed with function \texttt{bed\_tcrossprodSelf} of R package bigsnpr, and 3/ the OADP projection (section \ref{proj}).
-When simply projecting using loadings, there is no apparent shrinkage for PC1 and PC2, a weak [REWORD?] shrinkage for PC3 and PC4, and a large shrinkage for PC5 to PC8 (Figure \ref{fig:proj1000G}).
-[USE VALUES?]
+When simply projecting using loadings, there is negligible shrinkage for PC1 and PC2, a small shrinkage for PC3 and PC4, and a large shrinkage for PC5 to PC8 (Figure \ref{fig:proj1000G}).
 In contrast, there is no apparent shrinkage when projecting new individuals with OADP (Figure \ref{fig:proj1000G}).
@@
 Finally, to assess if OADP could be used to project individuals that are related to some individuals that were used to compute PCA, we projected these 60\% individuals (as if we were projecting their monozygotic twins) using OADP.
-Projections of related individuals using OADP shows some reverse [REWORD?] shrinkage (Figure \ref{fig:proj1000G-4}).
+Projections of related individuals using OADP shows some bias in reverse direction (Figure \ref{fig:proj1000G-4}).
 
@@
 Blue points are the 40\% remaining individuals, projected using the Online Augmentation, Decomposition, and Procrustes (OADP) transformation.
+Estimated shrinkage coefficients for these 8 PCs are 1.01 (PC1), 1.02, 1.06, 1.09, 1.50 (PC5), 1.69, 1.98 and 1.39.
 \label{fig:proj1000G}}
@@
 
-\subsection{PCA of the UK Biobank}
+\subsection{Capturing subtle population structure in the UK Biobank}
 
-Principal Component loadings reported by the UK Biobank clearly show that PC19 to PC40 capture LD structure, which is also the case for PC16 and PC18, although less pronounced (see peaks in figure \ref{fig:UKBB-loadings40}).
-When applying our automatic procedure to remove long-range LD regions, it did not converge after 5 iterations, meaning that it kept detecting long-range LD regions at each iteration.
-Therefore, we were able to capture only 16 PCs that shows population stratification and not LD structure (Figures \ref{fig:UKBB-screeplot}, \ref{fig:UKBB-scores} and \ref{fig:UKBB-loadings}).
-
-Then, we computed again PCA after restricting the individuals included in the computations: we randomly subsampled UKBB data to use only 10,000 British individuals (out of 431,029) and 5000 Irish individuals (out of 12,755), while keeping all individuals with other or unkonwn self-reported ancestry.
+We recomputed PCA in the UK Biobank after restricting the individuals included in the computations: we randomly subsampled UKBB data to use only 10,000 British individuals (out of 431,029) and 5000 Irish individuals (out of 12,755), while keeping all individuals with other or unkonwn self-reported ancestry.
 We further removed all pairs of related individuals reported by the UKBB (both individuals in each pair). 
@@
 Projection of new individuals show again a clear shrinkage when using simple projection (between 1.00 for PC1 and 1.80 for PC50), but no visible bias when using OADP projection (Figure \ref{fig:projUKBB-related}).
+
+\begin{figure}[!htpb]
+\centerline{\includegraphics[width=0.9\textwidth]{UKBB-scores-restricted.png}}
+\caption{Principal Component (PC) scores 27 to 50 computed on the UK Biobank using 48,942 individuals of diverse ancestries. Different colors represent different self-reported ancestries.
+\label{fig:UKBB-scores2}}
+\end{figure}
 
@@
 
-In this work, we have compiled different pitfalls that can occur with analyses related to Principal Component Analysis (PCA) of genetic data.
-Then, we have investigated possible solutions to these pitfalls and chose the ones that we thought to be the most relevant both with respect to properties such as accuracy and robustness, but also that would be fast and easy to use.
+In this work, we have compiled different pitfalls that can arise with analyses related to Principal Component Analysis (PCA) of genetic data.
+Then, we have investigated possible solutions to these pitfalls and chose the ones that we found most relevant, both with respect to properties such as accuracy and robustness, but also that would be fast and easy to use.
 We then implemented these solutions in open-source R packages.
-These functions can be directly used on genotypes stored as PLINK bed/bim/fam files with some missing values. 
-This is in contrast with our first release of packages bigstatsr and bigsnpr that use their own ``bigSNP'' format that can store both genotype calls and (rounded) dosages, but that mostly provide functions that require no missing value.
-Yet, PCA is a useful tool on its own and does not require extensive imputed data, therefore we decided that operating directly on PLINK files with a few missing values would be more practical for the users.
+The new functions we provide can be directly applied to genotypes stored as PLINK bed/bim/fam files with some missing values.
+This contrasts with previous releases of package bigsnpr that could use only one custom data format called ``bigSNP''.
+This data format can store both genotype calls and dosages, but requires conversion from other formats and imputation of missing values using functions provided in the package \cite[]{prive2017efficient}.
+As PCA is a useful tool on its own and does not require extensive imputed data, therefore we decided that operating directly on PLINK files with a few missing values would be more practical for users.
 
-As a general recommendation for computing PCA, we show some pipeline in figure \ref{fig:pipeline}. [START WITH QC?]
-This requires removing all related individuals, for which we provide an R wrapper to PLINK's implementation of KING robust kinship coefficients \cite[]{manichaikul2010robust,chang2015second}. 
+As a general recommendation for computing PCA, we propose the pipeline shown in figure \ref{fig:pipeline}.
+Note that we have not included standard steps such as initial quality control and post-analysis checks (e.g.\ visual inspection of different plots).
+This pipeline requires removing all related individuals, for which we provide an R wrapper to PLINK's implementation of KING robust kinship coefficients \cite[]{manichaikul2010robust,chang2015second}.
 Note that one should remove both individuals in each related pairs so that projected individuals at the end are not related to the ones used for computing PCA, as we showed that relatedness is a problem when using the OADP projection.
@@
 However, other pitfalls might apply when using other types of data; for example, methylation data can be confounded by factors such as age and sex, and it might be benificial to remove some probes that are associated with these confounding factors before computing PCA \cite[]{decamps2019guidelines}.
+
+[ADD?
+``Another application of PCA is capture unmeasured confounders in RNAseq analyses, e.g.
+Pickrell et al., Natue 2010, and Kang et al., Genetics 2008.
+(It's perhaps not the state-of-the-art today, but it's still common.)'']
 
@@
 Blue points are the 40\% remaining individuals, projected using the Online Augmentation, Decomposition, and Procrustes (OADP) transformation.
+Estimated shrinkage coefficients (comparing red and blue points) for these PCs are 2.79, 3.14 (PC10), 3.64, 3.18, 2.47, 3.88, 5.31, 5.84, 3.45, 6.55, 3.68 and 6.70 (PC20).
 \label{fig:proj1000G-2}}
@@
 Black points are the 60\% individuals used for computing PCA.
-Green points are the 40\% remaining individuals, projected by multiplying their genotypes by the corresponding PC loadings, further corrected using theoritical asymptotic shrinkage factors.
+Green points are the 40\% remaining individuals, projected by multiplying their genotypes by the corresponding PC loadings, further corrected using theoritical asymptotic shrinkage factors (values for the first 12 PCs: 1.01 (PC1), 1.02, 1.07, 1.10, 1.43 (PC5), 1.54, 1.74, 1.79, 2.47, 2.77 (PC10), 2.84 and 3.15).
 Blue points are the 40\% remaining individuals, projected using the Online Augmentation, Decomposition, and Procrustes (OADP) transformation.
@@
 Black points are PC scores of 1000G individuals used for computing PCA.
+Red points are the individuals from UKBB, projected by simply multiplying their genotypes by the corresponding PC loadings.
 Blue points are the 488,371 individuals from the UK Biobank, projected using the Online Augmentation, Decomposition, and Procrustes (OADP) transformation. 
+Estimated shrinkage coefficients (comparing red and blue points) for the first 20 PCs are 1.01 (PC1), 1.02, 1.06, 1.08, 1.36 (PC5), 1.82, 2.33, 2.36, 2.78, 2.84 (PC10), 2.99, 3.51, 4.38, 4.67, 4.99, 5.31, 5.74, 6.55, 6.71 and 6.75 (PC20).
 Note that only 20,000 random projected individuals are represented in this plot.
@@
 Blue points are the individuals from 1000G, projected using the Online Augmentation, Decomposition, and Procrustes (OADP) transformation.
+Estimated shrinkage coefficients (comparing red and blue points) for the first 20 PCs are 1.00 (PC1), 1.00, 1.00, 1.01, 1.01 (PC5), 1.02, 1.03, 1.03, 1.04, 1.04 (PC10), 1.04, 1.05, 1.05, 1.06, 1.07, 1.07, 1.08, 1.08, 1.08 and 1.08 (PC20).
 \label{fig:proj1000G-6}}
@@
 \newpage
-
-\begin{figure}[!htpb]
-\centerline{\includegraphics[width=0.9\textwidth]{UKBB-loadings1-40.pdf}}
-\caption{Principal Component (PC) loadings 1 to 40 reported by the UK Biobank.
-\label{fig:UKBB-loadings40}}
-\end{figure}
 
@@
 \begin{figure}[!htpb]
-\centerline{\includegraphics[width=0.9\textwidth]{UKBB-scores-restricted.png}}
-\caption{[SHOULD BE MAIN FIGURE?] Principal Component (PC) scores 27 to 50 computed on the UK Biobank using 48,942 individuals of diverse ancestries.
-Different colors represent different self-reported ancestries.
-\label{fig:UKBB-scores2}}
-\end{figure}
-
-\begin{figure}[!htpb]
 \centerline{\includegraphics[width=0.9\textwidth]{projUKBB-related.png}}

### 2019-11-11 paper-bedpca.tex
save

@@
 
-\subsection{PCA with missing values}
+\subsection{Efficient implementation of PCA for genotype data}
 
@@
 To compute this decomposition easily and efficiently, we implement an accessor that memory-map the bed file to use it directly as if it were a standard matrix.
-Then, we apply the same algorithm for partial SVD that is used in R package bigstatsr \cite[]{prive2017efficient}.
+Then, we apply the same algorithm for partial SVD that is used in R packages bigstatsr and FlashPCA2, namely the implicitly restarted Arnoldi method \cite[]{Lehoucq1996,abraham2017flashpca2,prive2017efficient}.
+This method is implemented in R package RSpectra, and requires a function that computes the matrix-vector multiplication of the scaled genotyped matrix with a given vector, which we implement for bed files and parallelise.
 
@@
 
-To detect a few sample outliers, we compare the rule of ``6 SDs from the mean'' (6SD) to the statistic we propose in section \ref{outlier-sample}.
+To detect a few sample outliers, we compare the standard rule of ``6 SDs from the mean'' (6SD) used in e.g.\ EIGENSOFT to the statistic we propose in section \ref{outlier-sample}.
 Our statistic identifies only isolated samples or isolated pairs that seems to be outliers driving structure of PC17-20 of 1000G (Figure \ref{fig:outlier-pd}).
@@
 
-When computing the PCA on the UK Biobank using 406,545 unrelated individuals and 171,977 variants, and projecting the 1000G data onto this reference PCA space, shrinkage is much smaller ($\leq 1.08$ for all 20 first PCs, Figure \ref{fig:proj1000G-6}).
-
 \begin{figure}[!htpb]
@@
 \end{figure}
+
+When computing the PCA on the UK Biobank using 406,545 unrelated individuals and 171,977 variants, and projecting the 1000G data onto this reference PCA space, shrinkage is much smaller ($\leq 1.08$ for all 20 first PCs, Figure \ref{fig:proj1000G-6}).
+Shrinkage increases with an increased number of variants used, but decreases with an increased sample size (Table \ref{tab:shrinkage}).
+
+\begingroup
+\renewcommand*{\arraystretch}{1.8}
+\begin{table}[h]
+\centering
+\caption{Shrinkage coefficients when projecting new individuals onto reference PCA space. We list the dataset, the sample size and number of variants used to compute the final truncated PCA.} 
+\label{tab:shrinkage}
+\begin{tabular}{|c|c|c|c|}
+\hline
+Dataset & Sample size (x1000) & \makecell {Number of variants \\ (x1000, after LD removal)} & Shrinkage (PC 1 - 5 - 10 - 20) \\
+\hline
+1000G & 1.5 & 393 & 1.01 - 1.50 - 3.14 - 6.70 \\
+1000G & 2.5 & 229 & 1.01 - 1.36 - 2.84 - 6.75 \\
+UKBB & 49 & 282 & 1.00 - 1.04 - 1.12 - 1.43 \\
+UKBB & 406.5 & 172 & 1.00 - 1.01 - 1.04 - 1.08 \\
+\hline
+\end{tabular}
+\end{table}
+\endgroup
 
@@
 \centerline{\includegraphics[width=0.4\textwidth]{PCA-pipeline.pdf}}
-\caption{Pipeline for computing Principal Components (PCs) using R packages bigsnpr and bigutilsr. [ADD VISUAL INSPECTION? QC? IN THE LEGEND?]
+\caption{Pipeline for computing Principal Components (PCs) using R packages bigsnpr and bigutilsr.
 \label{fig:pipeline}}
@@
 As a general recommendation for computing PCA, we propose the pipeline shown in figure \ref{fig:pipeline}.
-Note that we have not included standard steps such as initial quality control and post-analysis checks (e.g.\ visual inspection of different plots).
+Note that we have not included standard steps such as initial quality control filters and post-analysis checks (e.g.\ visual inspection of different plots).
 This pipeline requires removing all related individuals, for which we provide an R wrapper to PLINK's implementation of KING robust kinship coefficients \cite[]{manichaikul2010robust,chang2015second}.
-Note that one should remove both individuals in each related pairs so that projected individuals at the end are not related to the ones used for computing PCA, as we showed that relatedness is a problem when using the OADP projection.
-Then, it requires to do several steps of outlier detection, one for outlier variants that capture long-range LD variation (automatic), and one for detecting outlier samples (semi-automatic and visual).
-To fasten these steps, we tried hard to come up with a solution for not recomputing PCA from scratch when removing a few samples or variants.
-Using educated guesses based on low-rank approximations seemed to be a promising approach but did not fasten computation by much, so we did not pursue this idea \cite[]{brand2003fast,wu2017primme}.
-Once PCA is done, one should look at the loadings to make sure that there is no apparent LD structure left. 
+Note that one should remove both individuals in each related pairs so that the projected individuals are not related to the ones used for computing PCA, as we showed that relatedness is a problem when using the OADP projection.
+After selecting a subset of individuals, we apply several steps of outlier detection, one for outlier variants that capture long-range LD variation (automatic), and one for detecting outlier samples (semi-automatic and visual).
+To fasten these steps, we explored solutions for not recomputing PCA from scratch when removing a few samples or variants.
+Using educated guesses in R package PRIMME based on low-rank approximations seemed to be a promising approach but did not fasten computation by much, so we did not pursue this idea \cite[]{brand2003fast,wu2017primme}.
+Once PCA is done, one should look at the loadings to make sure that there is no apparent LD structure left (peaks in PC loadings as shown in figures \ref{fig:UKBB-loadings40} and \ref{fig:UKBB-loadings}).
+Including PCs that capture LD as covariates in genetic analyses can lead to reduced power, e.g.\ for detecting genetic association within these LD regions in GWAS.  
 One should also look at PC scores to identify which PCs capture population structure.
 As in many applications, we believe that a compromise should be made between signal and noise.
-Therefore, we recommend to use only PCs that show population structure (e.g.\ first 16 PCs in figure \ref{fig:UKBB-scores}) [NOTE SURE ABOUT PC15-16..].
+Therefore, we recommend to use only PCs that show structure that is not LD (e.g.\ first 16 PCs in figure \ref{fig:UKBB-scores}).
 
@@
 
-Finally, we have showed some pitfalls when computing PCA on genotype data, yet we think most of the results in this paper are not inherent to genotype data, and can be transferred to e.g.\ other omics data as well.
-For example, outlier and homogeneous sample detection can be used on PCs of anything.
+Finally, although we have focused on PCA of genotype data in this paper, we believe most of the results presented here are not inherent to genotype data, and can be transferred to e.g.\ other omics data as well.
+For example, outlier and homogeneous sample detection can be used on PCs of other types of data.
 Moreover, projection of scores will also be a problem for other omics data where the number of variables used is larger than the number of samples used for computing PCA.
@@
 
-[ADD?
-``Another application of PCA is capture unmeasured confounders in RNAseq analyses, e.g.
-Pickrell et al., Natue 2010, and Kang et al., Genetics 2008.
-(It's perhaps not the state-of-the-art today, but it's still common.)'']
+\noindent[NEED MORE CITATIONS IN THE DISCUSSION]
 
-\noindent[DISCUSS ABOUT PC16 -> FROM DIFFERENT ASSESSMENT CENTRES -> 36\% correlation with number of NA for all PC16 > 30]
-
-\noindent[MORE CITATIONS IN THE DISCUSSION?]
+\noindent[WHAT IS PC16 IN FIG \ref{fig:UKBB-scores}?]
 

### 2019-11-12 paper-bedpca.tex
save

@@
 \title{Efficient toolkit implementing best practices for principal component analysis of population genetic data}
-\author{Florian Priv\'e,$^{\text{1,}*}$ Keurcien Luu, Michael G.B. Blum,$^{\text{4}}$ John J. McGrath,$^{\text{1,2,3}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,}*}$}
+\author{Florian Priv\'e,$^{\text{1,5,}*}$ Keurcien Luu$^{\text{5}}$, Michael G.B. Blum,$^{\text{4,5}}$ John J. McGrath,$^{\text{1,2,3}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,}*}$}
 
@@
 \noindent$^{\text{\sf 4}}$OWKIN France, Paris, 75010, France. \\
+\noindent$^{\text{\sf 5}}$Laboratoire TIMC-IMAG, UMR 5525, Univ.\ Grenoble Alpes, CNRS, La Tronche, 38700, France. \\
 \noindent$^\ast$To whom correspondence should be addressed.\\
@@
 
-For example, we show that PC19 to PC40 in the UK Biobank are capturing LD structure.
-Using our automatic algorithm for removing long-range LD region, we are able to recover only 16 PCs that capture population structure only.
+For example, we show that PC19 to PC40 in the UK Biobank capture LD structure.
+Using our automatic algorithm for removing long-range LD region, we recover 16 PCs that capture population structure only.
 Therefore, we recommend using only 16-18 PCs from the UK Biobank. 
-Another example: when projecting the PCA computed on the 1000 Genomes project, although PC1 to PC4 suffer from only moderate shrinkage (1.00-1.08), PC5 for example suffers from a shrinkage factor of 1.51 and PC10 from a shrinkage factor of 3.16.
+We provide evidence for a shrinkage bias when projecting PCs computed with data from the 1000 Genomes project.
+Although PC1 to PC4 suffer from only moderate shrinkage (1.01-1.09), PC5 for example suffers from a shrinkage factor of 1.50 and PC10 from a shrinkage factor of 3.14.
 We provide a fast way to project new individuals and that is not affected by this shrinkage bias. 
-We also show how to restrict to individuals that are genetically homogeneous based on PCA.
+We also show how to use PCA to restrict analyses to individuals of homogeneous ancestry.
 
@@
 Principal Component Analysis (PCA) has been widely used in genetics for many years and in many contexts.
-For instance, adding PCs as covariates has been used to adjust for population structure in Genome-Wide Association Studies (GWAS) \cite[]{price2006principal}.
+For instance, adding PCs as covariates has been routinely used to adjust for population structure in Genome-Wide Association Studies (GWAS) \cite[]{price2006principal}.
 PCA has also been used to detect loci under selection \cite[]{galinsky2016fast,luu2017pcadapt}.
@@
 However, some pitfalls related to PCA of genotype data have been documented and not all of the currently available software address all of these. 
-In the following, we recall common pitfalls and explain when they are relevant.
+In the following, we outline common pitfalls and explain when they are relevant.
 First, some of the PCs may capture LD structure rather than population structure \cite[]{price2008long,abdellaoui2013population,prive2017efficient}.
@@
 For this paper, we derive implementations of truncated PCA and other useful functions for e.g.\ performing LD thinning and computing variant statistics.
-We make these available in a new release of R package bigsnpr (v1.0.0); what differs from previously available functions presented in \cite{prive2017efficient} is that these new functions can be used directly on bed/bim/fam PLINK files with some missing values.
+We make these available in a new release of R package bigsnpr (v1.0.0); this package differs from previously available functions presented in \cite{prive2017efficient} is that these new functions can be used directly on bed/bim/fam PLINK files with some missing values.
 We use them to analyse the UK Biobank data, and show that these functions are both very fast and easy to use.
@@
 Interestingly, subsetting the UK Biobank data enables to get more PCs capturing population structure than when using the whole sample (\textasciitilde40 instead of \textasciitilde16).
-Then, we project the other individuals that were not used in the PCA calculation, show that this projection is biased and provide an efficient solution to get unbiased projections instead.
-Finally, we talk about detecting outlier samples in PCA, either a few outlier samples that may correspond to e.g.\ batch effects, or when the goal is to restrict the data to homogeneous individuals.
+Then, we project the other individuals that were not used in the PCA calculation, show that this projection is biased and provide an efficient solution to estimate unbiased projections instead.
+Finally, we explore options to detect outlier samples in PCA, either a few outlier samples that may correspond to e.g.\ batch effects, or when the goal is to restrict the data to individuals of homogeneous ancestry.
 
@@
 \label{tab:method-overview}
-\begin{tabular}{|c|c|c|c|}
+\begin{tabular}{|c|c|c|c|c|}
 \hline
-Analysis & Method and/or Software & Advantages & Limitations \\
+Analysis & Method and/or Software & Citation & Advantages & Limitations \\
 \hline
-\multirow{3}{*}{PCA} & bigstatsr / bigsnpr \cite[]{prive2017efficient} & \makecell{Fast and accurate + handle dosages \\ + thinning options directly included} & \makecell{Own format without missing values \\ (fast functions are available \\ for converting and imputing)} \\
- & FlashPCA2 \cite[]{abraham2017flashpca2} & Fast and accurate & Sequential implementation \\
- & \makecell{PLINK 2.0 (reimplementation of FastPCA) \\ \cite[]{galinsky2016fast,chang2015second}} & Fast & \makecell{Possible lack of accuracy \\ \cite[]{abraham2017flashpca2,prive2017efficient}} \\
+\multirow{3}{*}{PCA} & bigstatsr / bigsnpr & \cite[]{prive2017efficient} & \makecell{Fast and accurate + handle dosages \\ + thinning options directly included} & \makecell{Own format without missing values \\ (fast functions are available \\ for converting and imputing)} \\
+ & FlashPCA2 & \cite[]{abraham2017flashpca2} & Fast and accurate & Sequential implementation \\
+ & PLINK 2.0 (reimplementation of FastPCA) & \cite[]{galinsky2016fast,chang2015second} & Fast & \makecell{Possible lack of accuracy \\ \cite[]{abraham2017flashpca2,prive2017efficient}} \\
 \hline
-\makecell{Detection of \\ oultier samples} & \makecell{``6 SDs from the mean'' \\ in EIGENSOFT \cite[]{patterson2006population}} & Simple & Assume Gaussian distritbution \\
+\makecell{Detection of \\ oultier samples} & \makecell{``6 SDs from the mean'' \\ in EIGENSOFT} & \cite[]{patterson2006population} & Simple & Assume Gaussian distritbution \\
 \hline
-\makecell{Detection of \\ homogeneous \\ samples} & R package aberrant \cite[]{bellenguez2011robust} & Robust & Only two statistics at once \\
+\makecell{Detection of \\ homogeneous \\ samples} & R package aberrant & \cite[]{bellenguez2011robust} & Robust & Only two statistics at once \\
 \hline
-\multirow{5}{*}{\makecell{Projection of \\ new individuals \\ onto reference \\ PCA space}} & Simple projection (multiplication by loadings) & Simple & Shrinkage biased \\
-  & Bias-adjusted projection in R package hdpca \cite[]{dey2019asymptotic} & Independent of new samples & \makecell{Assume same shrinkage for all individuals \\ + model-based \\ + need all eigenvalues of reference} \\
-  & \makecell{Augmentation, Decomposition and Procrustes transformation \\ (ADP) in LASER 2.0 \cite[]{wang2015improved}} & Accurate & Slow (a new PCA for each new sample) \\
-  & Online ADP (OADP) in python package FRAPOSA \cite[]{zhang2019fast} & Much faster than ADP & \makecell{Does not work for related individuals} \\ 
+\multirow{5}{*}{\makecell{Projection of \\ new individuals \\ onto reference \\ PCA space}} & Simple projection (multiplication by loadings) & & Simple & Shrinkage biased \\
+  & Bias-adjusted projection in R package hdpca & \cite[]{dey2019asymptotic} & Independent of new samples & \makecell{Assume same shrinkage for all individuals \\ + model-based \\ + need all eigenvalues of reference} \\
+  & \makecell{Augmentation, Decomposition and Procrustes (ADP) \\ transformation in LASER 2.0} & \cite[]{wang2015improved} & Accurate & Slow (a new PCA for each new sample) \\
+  & Online ADP (OADP) in python package FRAPOSA & \cite[]{zhang2019fast} & Much faster than ADP & \makecell{Does not work for related individuals} \\ 
 \hline
@@
 
-When there is no missing value, we compute the partial Singular Value Decomposition (SVD) $U \Delta V^T$ of the scaled genotype matrix $\tilde{G}_{i,j} = \frac{G_{i,j} - 2 \hat{f_j}}{\sqrt{2 \hat{f_j} (1 - \hat{f_j})}}$ where $G_{i,j}$ is the allele count (genotype) of individual $i$ and variant $j$, and $\hat{f_j}$ is the estimated allele frequency of variant $j$ ($2 \hat{f_j}$ is the mean allele count of variant $j$).
+When there is no missing value, we compute the partial Singular Value Decomposition (SVD) $U \Delta V^T$ of the scaled genotype matrix of diploid individuals $\tilde{G}_{i,j} = \frac{G_{i,j} - 2 \hat{f_j}}{\sqrt{2 \hat{f_j} (1 - \hat{f_j})}}$ where $G_{i,j}$ is the allele count (genotype) of individual $i$ and variant $j$, and $\hat{f_j}$ is the estimated allele frequency of variant $j$ ($2 \hat{f_j}$ is the mean allele count of variant $j$).
 Then, $U \Delta$ are the first $K$ PC scores and $V$ are the first $K$ PC loadings, where $K$ is the number of PCs computed (e.g.\ $K=20$).
 
-When there are some missing values, we compute the partial SVD similarly, except that missing values are replaced by the variant means (i.e.\ $G_{i,j} - 2 \hat{f_j} = 0$ when $G_{i,j}$ is missing) and  the $\hat{f_j}$ for each variant are estimated using only non missing values of each variant.
+When there are some missing values, we compute the partial SVD similarly, except that missing values are replaced by the variant means (i.e.\ $G_{i,j} - 2 \hat{f_j} = 0$ when $G_{i,j}$ is missing) and  the $\hat{f_j}$ for each variant are estimated using only non-missing values of each variant.
 Note that this decomposition is equivalent to the decomposition presented above after imputation by the variant means.
 
-To compute this decomposition easily and efficiently, we implement an accessor that memory-map the bed file to use it directly as if it were a standard matrix.
+To compute this decomposition easily and efficiently, we implement an accessor that memory-map the PLINK bed file to use it directly as if it were a standard matrix.
 Then, we apply the same algorithm for partial SVD that is used in R packages bigstatsr and FlashPCA2, namely the implicitly restarted Arnoldi method \cite[]{Lehoucq1996,abraham2017flashpca2,prive2017efficient}.
-This method is implemented in R package RSpectra, and requires a function that computes the matrix-vector multiplication of the scaled genotyped matrix with a given vector, which we implement for bed files and parallelise.
+This method is implemented in R package RSpectra, and requires a function that computes the matrix-vector multiplication of the scaled genotyped matrix with a given vector, which we implement for PLINK bed files and parallelise.
 
@@
 If $x$ are PC scores of centered data and if we use standard estimates, then $\mu = 0$ and $\Sigma = I_K$.
-Yet, here we use the pairwise orthogonalised Gnanadesikan-Kettenrin robust Mahalanobis distance as implemented in function \texttt{covRob} of R package robust with parameter \texttt{estim = "pairwiseGK"} \cite[]{gnanadesikan1972robust,yohai1988high,maronna2002robust,todorov2009object}.
+Yet, here we use the pairwise orthogonalised Gnanadesikan-Kettenrin robust estimates of these parameters, as implemented in function \texttt{covRob} of R package robust with parameter \texttt{estim = "pairwiseGK"} \cite[]{gnanadesikan1972robust,yohai1988high,maronna2002robust,todorov2009object}.
 We reexport function \texttt{covRob} in R package bigutilsr for convenience.
@@
 
-Finally, to choose the threshold of outlierness based on the previously described statistics, we use a modified version of Tukey's rule, a standard rule for detecting outliers \cite[]{tukey77}. 
+Finally, to choose the threshold on the previously described statistics above which to be considered as outlier, we use a modified version of Tukey's rule, a standard rule for detecting outliers \cite[]{tukey77}.
 The standard upper limit defined by Tukey's rule is $q_{75\%}(x) + 1.5 \cdot IQR(x)$, where $x$ is the vector of computed statistics and $IQR(x) = q_{75\%}(x) - q_{25\%}(x)$ is the interquartile range.
@@
 To make it fast, we use the fast K nearest neighbours implementation of R package nabor \cite[]{elseberg2012comparison} and parallelise it in function \texttt{knn\_parallel} of package bigutilsr.
-Automatic outlier detection is difficult, therefore we recommend users to choose a threshold of outlierness for this statistic based on visual inspection (histogram of these statistics and PC scores colored by these statistics).
+Automatic outlier detection is difficult, therefore we recommend users to choose a threshold for this statistic to define outliers based on visual inspection (using the histogram of these statistics and the PC scores colored by these statistics).
 
@@
 
-To demonstrate that we provide very fast implementations of the different methods presented in this paper, we apply them to the UK Biobank. We use 20 physical cores for most of the computations.
+To demonstrate that we provide very fast implementations of the different methods presented in this paper, we apply them to the UK Biobank.
+We use 20 physical cores for most of the computations (CPU: Intel(R) Xeon(R) Silver 4114, 2.20GHz).
 It takes 22 minutes to perform a first phase of clumping on 406,545 unrelated individuals genotyped over 504,139 variants, which reduces the number of variants to 261,307.
@@
 6SD identifies a lot of outliers, of which some are part of a relatively large cluster (Figure \ref{fig:outlier-sd}).
-[WHERE TO PUT FOLLOWING?] In theory, if all PCs are normally distributed, after correcting for multiple testing of 2500 individuals and 20 PCs, 6SD has a small probability of 0.0001 of detecting one outlier or more.
+In theory, if all PCs are normally distributed, after correcting for multiple testing of 2500 individuals and 20 PCs, 6SD has a small probability of 0.0001 of detecting one outlier or more.
+
+\begin{figure}[htbp]
+\centering
+\begin{subfigure}[b]{0.5\textwidth}
+\includegraphics[width=\textwidth]{hist-outliers-1000G.pdf}
+\caption{Distribution of statistics}
+\end{subfigure}
+\\~\\
+\begin{subfigure}[b]{0.45\textwidth}
+\includegraphics[width=\textwidth]{outliers-1000G.pdf}
+\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by the statistic used to define outliers.\\~\\}
+\end{subfigure}
+~~~~~~
+\begin{subfigure}[b]{0.45\textwidth}
+\includegraphics[width=\textwidth]{be-outlier-1000G.pdf}
+\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by being detected as an outlier. Threshold of being an outlier is determined based on histogram (a).}
+\end{subfigure}
+\caption{Outlier detection in the 1000 Genomes (1000G) project, using \texttt{prob\_dist} (Section \ref{outlier-sample}).\label{fig:outlier-pd}}
+\end{figure}
 
@@
 
-We use 60\% of individuals in the 1000G data (section \ref{1000G}) to compute the PCA ($K=20$ PCs).
+We use 60\% of individuals in the 1000G data (section \ref{1000G}) to compute the PCs ($K=20$ PCs).
 Then, we project the remaining 40\% individuals using three methods: 1/ simply multiplying the genotypes of these individuals by the previously computed loadings, 2/ correcting the simple projections using asymptotic shrinkage factors as determined by R package hdpca v1.1.3 \cite[]{dey2019asymptotic}, with all eigenvalues derived from the Genetic Relationship Matrix (GRM) computed with function \texttt{bed\_tcrossprodSelf} of R package bigsnpr, and 3/ the OADP projection (section \ref{proj}).
@@
 
-When computing the PCA on the UK Biobank using 406,545 unrelated individuals and 171,977 variants, and projecting the 1000G data onto this reference PCA space, shrinkage is much smaller ($\leq 1.08$ for all 20 first PCs, Figure \ref{fig:proj1000G-6}).
+When computing the PCs on the UK Biobank using 406,545 unrelated individuals and 171,977 variants, and projecting the 1000G data onto this reference PCA space, shrinkage is much smaller ($\leq 1.08$ for all 20 first PCs, Figure \ref{fig:proj1000G-6}).
 Shrinkage increases with an increased number of variants used, but decreases with an increased sample size (Table \ref{tab:shrinkage}).
@@
 \centering
-\caption{Shrinkage coefficients when projecting new individuals onto reference PCA space. We list the dataset, the sample size and number of variants used to compute the final truncated PCA.} 
+\caption{Shrinkage coefficients when projecting new individuals onto reference PCA space.
+We list the dataset, the sample size and number of variants used to compute the final truncated PCA.
+As expected, the shrinkage bias only becomes negligible if the PCA is conducted on large samples.}
 \label{tab:shrinkage}
@@
 
-We recomputed PCA in the UK Biobank after restricting the individuals included in the computations: we randomly subsampled UKBB data to use only 10,000 British individuals (out of 431,029) and 5000 Irish individuals (out of 12,755), while keeping all individuals with other or unkonwn self-reported ancestry.
+We recomputed PCA in the UK Biobank after restricting the individuals included in the computations: we randomly subsampled UKBB data to use only 10,000 British individuals (out of 431,029) and 5000 Irish individuals (out of 12,755), while keeping all individuals with other or unknown self-reported ancestry.
 We further removed all pairs of related individuals reported by the UKBB (both individuals in each pair). 
@@
 After selecting a subset of individuals, we apply several steps of outlier detection, one for outlier variants that capture long-range LD variation (automatic), and one for detecting outlier samples (semi-automatic and visual).
-To fasten these steps, we explored solutions for not recomputing PCA from scratch when removing a few samples or variants.
-Using educated guesses in R package PRIMME based on low-rank approximations seemed to be a promising approach but did not fasten computation by much, so we did not pursue this idea \cite[]{brand2003fast,wu2017primme}.
+To make these steps more computationally efficient, we explored solutions for not recomputing PCA from scratch when removing a few samples or variants.
+Using educated guesses in R package PRIMME based on low-rank approximations seemed to be a promising approach but did not reduce computation time by much, so we did not pursue this idea \cite[]{brand2003fast,wu2017primme}.
 Once PCA is done, one should look at the loadings to make sure that there is no apparent LD structure left (peaks in PC loadings as shown in figures \ref{fig:UKBB-loadings40} and \ref{fig:UKBB-loadings}).
@@
 As in many applications, we believe that a compromise should be made between signal and noise.
-Therefore, we recommend to use only PCs that show structure that is not LD (e.g.\ first 16 PCs in figure \ref{fig:UKBB-scores}).
+Therefore, we recommend using only PCs that show structure (e.g.\ PC1-16 in figure \ref{fig:UKBB-scores}) and exclude PCs that indicate LD (e.g.\ PC19-40 in figure \ref{fig:UKBB-loadings40}).
 
@@
 Finally, although we have focused on PCA of genotype data in this paper, we believe most of the results presented here are not inherent to genotype data, and can be transferred to e.g.\ other omics data as well.
-For example, outlier and homogeneous sample detection can be used on PCs of other types of data.
+For example, PCs can be used to account for confounding in other data as well \cite[]{pickrell2010understanding}.
+Then, outlier and homogeneous sample detection can be used on PCs of other types of data.
 Moreover, projection of scores will also be a problem for other omics data where the number of variables used is larger than the number of samples used for computing PCA.
-Finally, our recommendation of gathering ``populations'' of approximately the same size is relevant for other biological data as well.
+Finally, gathering ``populations'' of approximately the same size is relevant for other biological data as well.
 However, other pitfalls might apply when using other types of data; for example, methylation data can be confounded by factors such as age and sex, and it might be benificial to remove some probes that are associated with these confounding factors before computing PCA \cite[]{decamps2019guidelines}.
@@
 \noindent[WHAT IS PC16 IN FIG \ref{fig:UKBB-scores}?]
+
+\noindent[TODO: INCLUDE CITATIONS OF EVEN SAMPLING IN PCA]
 
@@
 \begin{subfigure}[b]{0.5\textwidth}
-\includegraphics[width=\textwidth]{hist-outliers-1000G.pdf}
-\caption{Distribution of statistics}
-\end{subfigure}
-\\~\\
-\begin{subfigure}[b]{0.45\textwidth}
-\includegraphics[width=\textwidth]{outliers-1000G.pdf}
-\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by statistic of outlierness.\\~\\}
-\end{subfigure}
-~~~~~~
-\begin{subfigure}[b]{0.45\textwidth}
-\includegraphics[width=\textwidth]{be-outlier-1000G.pdf}
-\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by being detected as an outlier. Threshold of outlierness is determined based on histogram (a).}
-\end{subfigure}
-\caption{Outlier detection in the 1000 Genomes (1000G) project, using \texttt{prob\_dist} (Section \ref{outlier-sample}).\label{fig:outlier-pd}}
-\end{figure}
-
-\begin{figure}[htbp]
-\centering
-\begin{subfigure}[b]{0.5\textwidth}
 \includegraphics[width=\textwidth]{hist-outliers2-1000G.pdf}
@@
 \includegraphics[width=\textwidth]{outliers2-1000G.pdf}
-\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by statistic of outlierness (log-scale).\\~\\}
+\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by statistic used to detect outliers (log-scale).\\~\\}
 \end{subfigure}
@@
 \includegraphics[width=\textwidth]{be-outlier2-1000G.pdf}
-\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by being detected as an outlier. Threshold of outlierness is determined based on histogram (a).}
+\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by being detected as an outlier. Threshold of being considered as an outlier is determined based on histogram (a).}
 \end{subfigure}
@@
 \caption{PCs 1 to 20 of UKBB, colored by being detected as an outlier.
-Threshold of outlierness is determined based on histogram (c), where the threshold of 5 is chosen for the logarithm of the distances.}
+Threshold of being considered as an outlier is determined based on histogram (c), where the threshold of 5 is chosen for the logarithm of the distances.}
 \end{subfigure}

### 2019-11-13 paper-bedpca.tex
save

@@
 \noindent$^{\text{\sf 4}}$OWKIN France, Paris, 75010, France. \\
-\noindent$^{\text{\sf 5}}$Laboratoire TIMC-IMAG, UMR 5525, Univ.\ Grenoble Alpes, CNRS, La Tronche, 38700, France. \\
+\noindent$^{\text{\sf 5}}$Laboratoire TIMC-IMAG, UMR 5525, Univ.\ Grenoble Alpes, La Tronche, 38700, France. \\
 \noindent$^\ast$To whom correspondence should be addressed.\\
@@
 Principal Component Analysis (PCA) of genetic data is routinely used to infer ancestry and control for population structure in various genetic analyses.
-However, conducting PCA in large populations is complicated, and there are several potential pitfalls.
-These pitfalls include (1) capturing Linkage Disequilibrium (LD) structure instead of population structure, (2) projected PCs that suffer from shrinkage bias when projecting PCA from a reference dataset to another independent dataset, and (3) detecting sample outliers.
-In this work, we explore these pitfalls of using PCA, and present efficient solutions to these pitfalls.
+However, conducting PCA analyses can be complicated and has several potential pitfalls.
+These pitfalls include (1) capturing Linkage Disequilibrium (LD) structure instead of population structure, (2) projected PCs that suffer from shrinkage bias when projecting PCA from a reference dataset to another independent dataset, (3) detecting sample outliers, and (4) uneven population sizes.
+In this work, we explore these potential issues when using PCA, and present efficient solutions to these.
 Following applications to the UK Biobank and the 1000 Genomes project datasets, we make recommendations for best practices and provide efficient and user-friendly implementations of the proposed solutions in R packages bigsnpr and bigutilsr.
@@
 For example, we show that PC19 to PC40 in the UK Biobank capture LD structure.
-Using our automatic algorithm for removing long-range LD region, we recover 16 PCs that capture population structure only.
+Using our automatic algorithm for removing long-range LD regions, we recover 16 PCs that capture population structure only.
 Therefore, we recommend using only 16-18 PCs from the UK Biobank. 
 We provide evidence for a shrinkage bias when projecting PCs computed with data from the 1000 Genomes project.
-Although PC1 to PC4 suffer from only moderate shrinkage (1.01-1.09), PC5 for example suffers from a shrinkage factor of 1.50 and PC10 from a shrinkage factor of 3.14.
-We provide a fast way to project new individuals and that is not affected by this shrinkage bias. 
+Although PC1 to PC4 suffer from only moderate shrinkage (1.01-1.09), PC5 (resp.\ PC10) for example suffers from a shrinkage factor of 1.50 (resp.\ 3.14).
+We provide a fast way to project new individuals that is not affected by this shrinkage bias. 
 We also show how to use PCA to restrict analyses to individuals of homogeneous ancestry.
@@
 Principal Component Analysis (PCA) has been widely used in genetics for many years and in many contexts.
-For instance, adding PCs as covariates has been routinely used to adjust for population structure in Genome-Wide Association Studies (GWAS) \cite[]{price2006principal}.
+For instance, adding PCs as covariates has been routinely used to adjust for population structure in Genome-Wide Association Studies (GWAS) \cite[]{price2006principal,novembre2008interpreting}.
 PCA has also been used to detect loci under selection \cite[]{galinsky2016fast,luu2017pcadapt}.
@@
 For this paper, we derive implementations of truncated PCA and other useful functions for e.g.\ performing LD thinning and computing variant statistics.
-We make these available in a new release of R package bigsnpr (v1.0.0); this package differs from previously available functions presented in \cite{prive2017efficient} is that these new functions can be used directly on bed/bim/fam PLINK files with some missing values.
-We use them to analyse the UK Biobank data, and show that these functions are both very fast and easy to use.
+We make these available in a new release of R package bigsnpr (v1.0.0); what differs from previously available functions presented in \cite{prive2017efficient} is that these new functions can be used directly on PLINK bed/bim/fam files with some missing values.
+We use these new functions to analyse the UK Biobank data, and show that they are both very fast and easy to use.
 We also point out that many PCs currently reported by the UK Biobank capture LD structure instead of population structure.
@@
 \multirow{3}{*}{PCA} & bigstatsr / bigsnpr & \cite[]{prive2017efficient} & \makecell{Fast and accurate + handle dosages \\ + thinning options directly included} & \makecell{Own format without missing values \\ (fast functions are available \\ for converting and imputing)} \\
- & FlashPCA2 & \cite[]{abraham2017flashpca2} & Fast and accurate & Sequential implementation \\
+ & FlashPCA2 & \cite[]{abraham2017flashpca2} & Fast and accurate & Not parallelised \\
  & PLINK 2.0 (reimplementation of FastPCA) & \cite[]{galinsky2016fast,chang2015second} & Fast & \makecell{Possible lack of accuracy \\ \cite[]{abraham2017flashpca2,prive2017efficient}} \\
 \hline
-\makecell{Detection of \\ oultier samples} & \makecell{``6 SDs from the mean'' \\ in EIGENSOFT} & \cite[]{patterson2006population} & Simple & Assume Gaussian distritbution \\
+\makecell{Detection of \\ oultier samples} & \makecell{``6 SDs from the mean'' \\ in EIGENSOFT} & \cite[]{patterson2006population} & Simple & Assumes Gaussian distritbution \\
 \hline
@@
 \multirow{5}{*}{\makecell{Projection of \\ new individuals \\ onto reference \\ PCA space}} & Simple projection (multiplication by loadings) & & Simple & Shrinkage biased \\
-  & Bias-adjusted projection in R package hdpca & \cite[]{dey2019asymptotic} & Independent of new samples & \makecell{Assume same shrinkage for all individuals \\ + model-based \\ + need all eigenvalues of reference} \\
+  & Bias-adjusted projection in R package hdpca & \cite[]{dey2019asymptotic} & Independent of new samples & \makecell{Assumes same shrinkage for all individuals \\ + model-based \\ + need all eigenvalues of reference} \\
   & \makecell{Augmentation, Decomposition and Procrustes (ADP) \\ transformation in LASER 2.0} & \cite[]{wang2015improved} & Accurate & Slow (a new PCA for each new sample) \\
-  & Online ADP (OADP) in python package FRAPOSA & \cite[]{zhang2019fast} & Much faster than ADP & \makecell{Does not work for related individuals} \\ 
+  & Online ADP (OADP) in python package FRAPOSA & \cite[]{zhang2019fast} & Much faster than ADP & Does not work for related individuals (Section \ref{proj-related}) \\ 
 \hline
@@
 Then, we apply the same algorithm for partial SVD that is used in R packages bigstatsr and FlashPCA2, namely the implicitly restarted Arnoldi method \cite[]{Lehoucq1996,abraham2017flashpca2,prive2017efficient}.
-This method is implemented in R package RSpectra, and requires a function that computes the matrix-vector multiplication of the scaled genotyped matrix with a given vector, which we implement for PLINK bed files and parallelise.
+This algorithm is implemented in R package RSpectra, and requires a function that computes the matrix-vector multiplication of the scaled genotype matrix with a given vector, which we implement for PLINK bed files and parallelise.
 
@@
 
-Finally, to choose the threshold on the previously described statistics above which to be considered as outlier, we use a modified version of Tukey's rule, a standard rule for detecting outliers \cite[]{tukey77}.
+Finally, to choose the threshold on the previously described statistics above which variants are considered outliers, we use a modified version of Tukey's rule, a standard rule for detecting outliers \cite[]{tukey77}.
 The standard upper limit defined by Tukey's rule is $q_{75\%}(x) + 1.5 \cdot IQR(x)$, where $x$ is the vector of computed statistics and $IQR(x) = q_{75\%}(x) - q_{25\%}(x)$ is the interquartile range.
@@
 For detecting outlier samples in PCA, we use a modified version of the Probabilistic Local Outlier Factor (PLOF) statistic on PCs \cite[]{kriegel2009loop}.
-Using K nearest neighbours (KNN), this consists in comparing the distance from a point $j$ to its KNNs ($pd_j = \sqrt{\frac{1}{K} \sum_{k=1}^K d^2_{j \leftrightarrow j_k}}$, where $j_k$ is the $k$-th NN of $j$) with the distances from its KNNs to their KNNs ($\frac{1}{K} \sum_{k=1}^K pd_{j_k}$).
-The idea is that an outlier should be far from all other points, but this point is even more outlier if its KNNs are in a very dense cluster.
-In practice, we use $pd_j / \sqrt{\frac{1}{K} \sum_{k=1}^K pd_{j_k}}$. Note the square root, as it otherwise detect any point next to a very dense cluster as an outlier.
-We implement this statistic in function \texttt{prob\_dist} of R package bigutilsr.
+Using K nearest neighbours (KNN), this consists in comparing the distance from a point $j$ to its KNNs ($pd_j = \sqrt{\frac{1}{K} \sum_{k=1}^K d^2_{j \leftrightarrow j_k}}$, where $j_k$ is the $k$-th NN of $j$) with the distances from its KNNs to their respective KNNs ($\frac{1}{K} \sum_{k=1}^K pd_{j_k}$).
+The idea is that an outlier should be far from all other points, and is even more outlier if its KNNs are in a very dense cluster.
+Here, we use $pd_j / \sqrt{\frac{1}{K} \sum_{k=1}^K pd_{j_k}}$. Note the square root, as it otherwise detect any point next to a very dense cluster as an outlier.
+We implement (the two parts of) this statistic in function \texttt{prob\_dist} of R package bigutilsr.
 To make it fast, we use the fast K nearest neighbours implementation of R package nabor \cite[]{elseberg2012comparison} and parallelise it in function \texttt{knn\_parallel} of package bigutilsr.
-Automatic outlier detection is difficult, therefore we recommend users to choose a threshold for this statistic to define outliers based on visual inspection (using the histogram of these statistics and the PC scores colored by these statistics).
+Automatic outlier detection is difficult, therefore we recommend users to choose a threshold for this statistic to define outliers based on visual inspection (using the histogram of these statistics and the PC scores colored by these statistics; see e.g.\ figure \ref{fig:outlier-pd}).
 
@@
 To project individuals from the same dataset as the ones used for computing PCA, you can use function \texttt{bed\_projectSelfPCA}. 
-Note that the new individuals to be projected should not be related to the ones used for computing PCA.
+Note that the new individuals to be projected should not be related to the ones used for computing PCA (Section \ref{proj-related}).
 
@@
 
-We provide a subsetted version of the 1000 genomes (1000G) project data \cite[]{10002015global,meyer2019genotype}.
+We provide and use a subsetted version of the 1000 genomes (1000G) project data \cite[]{10002015global,meyer2019genotype}.
 Variants are restricted to the ones in common with HapMap3 or UK Biobank \cite[]{international2010integrating,bycroft2017genome}. 
@@
 This results in 2490 individuals and 1,664,852 variants of the 1000G project (phase 3) in PLINK bed/bim/fam format. 
-Resulting PLINK files and R code to generate these files are available at \url{https://doi.org/10.6084/m9.figshare.9208979.v3}. 
+Resulting PLINK files and R code to generate these files are made available at \url{https://doi.org/10.6084/m9.figshare.9208979.v3}. 
 To easily download these data, we provide function \texttt{download\_1000G} in R package bigsnpr.
 
-In this paper, we also analyse the genotyped version of the UK Biobank \cite[]{bycroft2017genome}.
+In this paper, we also analyse the genotyped version of the UK Biobank (\url{https://www.ukbiobank.ac.uk/}).
 We apply some quality control filters; we remove individuals with more than 10\% missing values, variants with more than 1\% missing values, variants having a minor allele frequency < 0.01, variants with P-value of the Hardy-Weinberg exact test < $10^{-50}$, and non-autosomal variants.
@@
 It takes 12 minutes more to do the OADP projection of all 488,371 individuals of the UKBB data onto the PCA computed using the 1000G data.
-Finally, it takes only 6 minutes to compute the 30-Nearest Neighbours of 20 PC scores for 406,545 UK Biobank individuals, which is the main computational aspect of computing statistics for detecting individual outlier samples (Section \ref{outlier-sample}).
+Finally, it takes only 6 minutes to compute the 30-Nearest Neighbours of 20 PC scores for 406,545 UK Biobank individuals, which is the main computational aspect of computing the statistics used to detect individual outlier samples (Section \ref{outlier-sample}).
 
@@
 Our statistic identifies only isolated samples or isolated pairs that seems to be outliers driving structure of PC17-20 of 1000G (Figure \ref{fig:outlier-pd}).
-6SD identifies a lot of outliers, of which some are part of a relatively large cluster (Figure \ref{fig:outlier-sd}).
-In theory, if all PCs are normally distributed, after correcting for multiple testing of 2500 individuals and 20 PCs, 6SD has a small probability of 0.0001 of detecting one outlier or more.
+In contrast, 6SD identifies a lot of outliers, of which some are part of a relatively large cluster (Figure \ref{fig:outlier-sd}).
+We recall that, in theory, if all PCs are normally distributed, after correcting for multiple testing of 2500 individuals and 20 PCs, 6SD has a small probability of 0.0001 of detecting one outlier or more.
 
@@
 \includegraphics[width=\textwidth]{outliers-1000G.pdf}
-\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by the statistic used to define outliers.\\~\\}
+\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by the statistic (S) used to define outliers.\\~\\}
 \end{subfigure}
@@
 When the goal is to restrict to homogeneous samples, we compare the use of R package aberrant used to report the ``White British'' subset of UKBB homogeneous samples, to the use of the robust Mahalanobis distance we propose.
-Here, we choose a threshold of 5 and show that this gives a similar subset of individuals than the ``White British'' subset reported by the UK Biobank (Figure \ref{fig:homogeneous}).
-Using this threshold of 5, only 3 out of 10,936 people of Asian ancestry (1 ``Chinese'' and 2 ``Indian'') are kept, and 1 ``African'' out of 7622 poeple with Black background is kept (Table \ref{tab:homogeneous}).
+Here, we visually choose a threshold of 5 on the log-distance and show that this gives a similar subset of individuals than the ``White British'' subset reported by the UK Biobank (Figure \ref{fig:homogeneous}).
+Moreover, when using this threshold, only 3 out of 10,936 people of Asian ancestry (1 ``Chinese'' and 2 ``Indian'') are kept, and 1 ``African'' out of 7622 poeple with Black background is kept (Table \ref{tab:homogeneous}).
 In contrast, 416,492 out of 431,090 ``British'' (96.6\%) and 12,620 out of 12,759 ``Irish'' (98.9\%) are kept.
@@
 Projections of related individuals using OADP shows some bias in reverse direction (Figure \ref{fig:proj1000G-4}).
+\label{proj-related}
 

### 2019-11-13 paper-bedpca.tex
save

@@
 
-We use 60\% of individuals in the 1000G data (section \ref{1000G}) to compute the PCs ($K=20$ PCs).
-Then, we project the remaining 40\% individuals using three methods: 1/ simply multiplying the genotypes of these individuals by the previously computed loadings, 2/ correcting the simple projections using asymptotic shrinkage factors as determined by R package hdpca v1.1.3 \cite[]{dey2019asymptotic}, with all eigenvalues derived from the Genetic Relationship Matrix (GRM) computed with function \texttt{bed\_tcrossprodSelf} of R package bigsnpr, and 3/ the OADP projection (section \ref{proj}).
+We use 60\% of individuals in the 1000G data (section \ref{1000G}) to compute K=20 PCs.
+Then, we project the remaining 40\% individuals using three methods: 1/ simply multiplying the genotypes of these individuals by the previously computed loadings, 2/ correcting the simple projections using asymptotic shrinkage factors as determined by R package hdpca v1.1.3 \cite[]{dey2019asymptotic}, with all eigenvalues derived from the Genetic Relationship Matrix (GRM) computed with new function \texttt{bed\_tcrossprodSelf} of R package bigsnpr, and 3/ the OADP projection (section \ref{proj}).
 When simply projecting using loadings, there is negligible shrinkage for PC1 and PC2, a small shrinkage for PC3 and PC4, and a large shrinkage for PC5 to PC8 (Figure \ref{fig:proj1000G}).
-In contrast, there is no apparent shrinkage when projecting new individuals with OADP (Figure \ref{fig:proj1000G}).
-Simple projection is affected even more by this shrinkage for PC9 to PC20, while OADP appears free of this issue (Figure \ref{fig:proj1000G-2}).
+In contrast, there is no visible shrinkage when projecting new individuals with OADP (Figure \ref{fig:proj1000G}).
+Simple projection is affected even more by this shrinkage for PC9 to PC20, while OADP still appears free of this issue (Figure \ref{fig:proj1000G-2}).
 We show the same results when projecting the full UK Biobank data onto PCA computed using 1000G data  (Figure \ref{fig:proj1000G-5}).
-When correcting projected PC scores with asymptotic shrinkage factors, bias is smaller than when simply projecting loadings, yet, there is an apparent bias for PC7-8 (Figure \ref{fig:proj1000G-3}).
+When correcting projected PC scores with asymptotic shrinkage factors, bias is smaller than when simply projecting loadings, yet, there is a visible bias for PC7-8 (Figure \ref{fig:proj1000G-3}).
 Finally, to assess if OADP could be used to project individuals that are related to some individuals that were used to compute PCA, we projected these 60\% individuals (as if we were projecting their monozygotic twins) using OADP.
@@
 When computing the PCs on the UK Biobank using 406,545 unrelated individuals and 171,977 variants, and projecting the 1000G data onto this reference PCA space, shrinkage is much smaller ($\leq 1.08$ for all 20 first PCs, Figure \ref{fig:proj1000G-6}).
-Shrinkage increases with an increased number of variants used, but decreases with an increased sample size (Table \ref{tab:shrinkage}).
+Overall, shrinkage decreases with an increased sample size (Table \ref{tab:shrinkage}).
 
@@
 \caption{Shrinkage coefficients when projecting new individuals onto reference PCA space.
-We list the dataset, the sample size and number of variants used to compute the final truncated PCA.
+We list the dataset, the sample size and number of variants used to compute the final PCA.
 As expected, the shrinkage bias only becomes negligible if the PCA is conducted on large samples.}
@@
 \hline
-Dataset & Sample size (x1000) & \makecell {Number of variants \\ (x1000, after LD removal)} & Shrinkage (PC 1 - 5 - 10 - 20) \\
+Dataset & Sample size ($\times$1000) & \makecell {Number of variants \\ ($\times$1000, after LD removal)} & Shrinkage (PC 1 - 5 - 10 - 20) \\
 \hline
-1000G & 1.5 & 393 & 1.01 - 1.50 - 3.14 - 6.70 \\
-1000G & 2.5 & 229 & 1.01 - 1.36 - 2.84 - 6.75 \\
-UKBB & 49 & 282 & 1.00 - 1.04 - 1.12 - 1.43 \\
+1000G & ~~~~1.5 & 393 & 1.01 - 1.50 - 3.14 - 6.70 \\
+1000G & ~~~~2.5 & 229 & 1.01 - 1.36 - 2.84 - 6.75 \\
+UKBB & ~~49.0 & 282 & 1.00 - 1.04 - 1.12 - 1.43 \\
 UKBB & 406.5 & 172 & 1.00 - 1.01 - 1.04 - 1.08 \\
@@
 We recomputed PCA in the UK Biobank after restricting the individuals included in the computations: we randomly subsampled UKBB data to use only 10,000 British individuals (out of 431,029) and 5000 Irish individuals (out of 12,755), while keeping all individuals with other or unknown self-reported ancestry.
-We further removed all pairs of related individuals reported by the UKBB (both individuals in each pair). 
+We further removed all pairs of related individuals reported by the UKBB (i.e.\ both individuals in each pair). 
 This resulted in 48,942 individuals that we used to compute 50 PCs, which took less than 3 hours using function \texttt{bed\_autoSVD} (that converged after 4 iterations of automatic LD removal). 
-We show that we are able to capture more PCs (at least 40) that display visual population structure (Figures \ref{fig:UKBB-scores2} and \ref{fig:UKBB-screeplot2}).
-We then projected all 439,429 remaining individuals from UKBB onto this PCA space in 21 minutes only using our implementation of the OADP projection.
+We show that we are able to capture more PCs (at least 40 instead of 16-18) that display visual population structure (Figures \ref{fig:UKBB-scores2} and \ref{fig:UKBB-screeplot2}).
+We then projected all 439,429 remaining individuals from UKBB onto this PCA space in 21 minutes only using our implementation of the OADP projection (\texttt{bed\_projectSelfPCA}).
 Note that these individuals should not be related to any of the 48,942 individuals used for training PCA because we removed both individuals from each pair of related individuals in UKBB.
@@
 \centerline{\includegraphics[width=0.9\textwidth]{UKBB-scores-restricted.png}}
-\caption{Principal Component (PC) scores 27 to 50 computed on the UK Biobank using 48,942 individuals of diverse ancestries. Different colors represent different self-reported ancestries.
+\caption{Principal Component (PC) scores 27 to 50 computed on the UK Biobank using 48,942 individuals of diverse ancestries.
+These individuals are the ones resulting from removing all related individuals and randomly subsampling the British and Irish individuals.
+Different colors represent different self-reported ancestries.
 \label{fig:UKBB-scores2}}
@@
 The new functions we provide can be directly applied to genotypes stored as PLINK bed/bim/fam files with some missing values.
-This contrasts with previous releases of package bigsnpr that could use only one custom data format called ``bigSNP''.
+This contrasts with previous releases of package bigsnpr that could only use one custom data format called ``bigSNP''.
 This data format can store both genotype calls and dosages, but requires conversion from other formats and imputation of missing values using functions provided in the package \cite[]{prive2017efficient}.
-As PCA is a useful tool on its own and does not require extensive imputed data, therefore we decided that operating directly on PLINK files with a few missing values would be more practical for users.
+As PCA is a useful tool on its own and does not require extensive imputed data, we therefore decided that operating directly on PLINK files with a few missing values would be more practical for users.
 
@@
 This pipeline requires removing all related individuals, for which we provide an R wrapper to PLINK's implementation of KING robust kinship coefficients \cite[]{manichaikul2010robust,chang2015second}.
-Note that one should remove both individuals in each related pairs so that the projected individuals are not related to the ones used for computing PCA, as we showed that relatedness is a problem when using the OADP projection.
+Note that one should remove both individuals in each related pairs so that the projected individuals are not related to the ones used for computing PCA, as we showed that relatedness is a problem when using the OADP projection (Figure \ref{fig:proj1000G-4}).
 After selecting a subset of individuals, we apply several steps of outlier detection, one for outlier variants that capture long-range LD variation (automatic), and one for detecting outlier samples (semi-automatic and visual).
@@
 Using educated guesses in R package PRIMME based on low-rank approximations seemed to be a promising approach but did not reduce computation time by much, so we did not pursue this idea \cite[]{brand2003fast,wu2017primme}.
-Once PCA is done, one should look at the loadings to make sure that there is no apparent LD structure left (peaks in PC loadings as shown in figures \ref{fig:UKBB-loadings40} and \ref{fig:UKBB-loadings}).
-Including PCs that capture LD as covariates in genetic analyses can lead to reduced power, e.g.\ for detecting genetic association within these LD regions in GWAS.  
+Once PCA is done, one should look at the loadings to make sure that there is no visible LD structure left (peaks in PC loadings as shown in figures \ref{fig:UKBB-loadings40} and \ref{fig:UKBB-loadings}), since including PCs that capture LD as covariates in genetic analyses can lead to reduced power for variants in these regions.
+These long-range LD regions are captured by PCs because variants in these regions are extensively correlated (even after a first step of LD thinning), and PCA just captures the variance in these correlated variants.
+This is a similar problem as including related individuals or even whole families in PCA, they would show up as structure in PCs. 
 One should also look at PC scores to identify which PCs capture population structure.
-As in many applications, we believe that a compromise should be made between signal and noise.
+As in many applications, we believe there should be a compromise between signal and noise.
 Therefore, we recommend using only PCs that show structure (e.g.\ PC1-16 in figure \ref{fig:UKBB-scores}) and exclude PCs that indicate LD (e.g.\ PC19-40 in figure \ref{fig:UKBB-loadings40}).
@@
 We also provide some recommendations when the dataset is composed mainly of one population (e.g.\ British people in the UK Biobank).
-We show that it is useful to actually subset these individuals so there is not too much imbalance between the number of people from different ancestries. 
-Indeed, when subsetting British and Irish people, we are able to capture a lot more PCs that show population structure.
-This makes designs such as the 1000 Genomes project, that gathered around 100 people for each of 26 populations, highly relevant for capturing population structure.
+We show that it is useful to actually subset these individuals so there is not too much imbalance between the sample sizes of the different populations.
+Previous work has also shown that uneven population sizes can distort PCs \cite[]{novembre2008interpreting,mcvean2009genealogical}.
+Indeed, when subsetting British and Irish people, we are able to capture a lot more PCs that show population structure with less than 50K individuals as compared to when using more than 400K unrelated individuals.
+This makes designs such as the 1000 Genomes project, which gathered around 100 people for each of 26 populations, highly relevant for capturing population structure.
 In contrast, when the goal is to have an homogeneous sample, we show that using the Mahalanobis distance on PC scores can efficiently achieve this goal, which we used in previous analyses \cite[]{prive2019efficient}.
-When the homogeneous sample is not predominant in the dataset, one solution is to compute the center and covariance of the robust Mahalanobis distance using only the population of interest, and then computing the distances for all individuals using these parameters.
+When the homogeneous sample is not predominant in the dataset, one solution is to compute the center and covariance of the robust Mahalanobis distance using only the population of interest, and then computing the distances for all individuals using these robust estimates.
 
@@
 Moreover, projection of scores will also be a problem for other omics data where the number of variables used is larger than the number of samples used for computing PCA.
-Finally, gathering ``populations'' of approximately the same size is relevant for other biological data as well.
+Finally, using ``populations'' with approximately the same size is relevant for other biological data as well.
 However, other pitfalls might apply when using other types of data; for example, methylation data can be confounded by factors such as age and sex, and it might be benificial to remove some probes that are associated with these confounding factors before computing PCA \cite[]{decamps2019guidelines}.
-
-\noindent[NEED MORE CITATIONS IN THE DISCUSSION]
-
-\noindent[WHAT IS PC16 IN FIG \ref{fig:UKBB-scores}?]
-
-\noindent[TODO: INCLUDE CITATIONS OF EVEN SAMPLING IN PCA]
 
@@
 
-A tutorial on the steps to perform PCA on 1000G data is presented at \url{} [TODO].
+A tutorial on the steps to perform PCA on 1000G data is available at \url{} [TODO].
 
@@
 
-F.P., J.J.M.\ and B.J.V.\ acknowledge Niels Bohr Professorship from the Danish National Research Foundation to Prof.\ John J.\ McGrath, and the Lundbeck Foundation Initiative for Integrative Psychiatric Research, iPSYCH (R248-2017-2003).
+F.P., J.M.\ and B.V.\ are supported by the Danish National Research Foundation (Niels Bohr Professorship to J.M.), and also acknowledge the Lundbeck Foundation Initiative for Integrative Psychiatric Research, iPSYCH (R248-2017-2003).
 
@@
 
-Michael Blum is an employee of OWKIN France.
+Michael Blum is now an employee of OWKIN France.
 The other authors declare no competing interests.
@@
 Then $g = \sqrt{y \cdot y^T - l \cdot l^T}$.
-\item Calculate $Q^T Q$ where $$Q = \begin{bmatrix} \Delta & l^T \\ 0 & g \end{bmatrix}$$ so that $$Q^TQ = \begin{bmatrix} \Delta^2 & \Delta \cdot l^T \\ l \cdot \Delta & g^2 + l \cdot l^T \end{bmatrix} = \begin{bmatrix} \Delta^2 & \Delta \cdot l^T \\ l \cdot \Delta & y \cdot y^T \end{bmatrix}.$$ Note that we do not actually need to compute $g$, and that we can update the last row and column of $Q^T Q$ instead of computing it from an updated version of $Q$.
+\item Calculate $Q^T Q$ where $$Q = \begin{bmatrix} \Delta & l^T \\ 0 & g \end{bmatrix}$$ so that $$Q^TQ = \begin{bmatrix} \Delta^2 & \Delta \cdot l^T \\ l \cdot \Delta & g^2 + l \cdot l^T \end{bmatrix} = \begin{bmatrix} \Delta^2 & \Delta \cdot l^T \\ l \cdot \Delta & y \cdot y^T \end{bmatrix}.$$ Note that we do not actually need to compute $g$, and that we can update only the last row and column of $Q^T Q$ instead of computing it from an updated version of $Q$.
 \item Get the eigen decomposition $Q^T Q = V' {\Delta'}^2 {V'}^T$ (truncated to $K$ components out of the $K+1$). Let us denote $V_2 = V' {\Delta'}$.
@@
 As $\widetilde{U}$ and $U \Delta$ have both their columns centered already (since $U$ does), the Procrustes transformation $\rho \widetilde{U} A$, where $\rho$ is a scaling coefficient and $A$ is an orthonomal projection matrix that minimise the Frobenius norm of $(\rho \widetilde{U} A - U)$, is given by $A = V'' {U''}^T$ and $\rho = \frac{\text{trace}(\Delta'')}{\text{trace}(\widetilde{U}^T \widetilde{U})}$ where $U'' \Delta'' {V''}^T$ is the full SVD of $(U \Delta)^T \widetilde{U}$ \cite[]{wang2015improved}.
-As $U^T U = I$, we note that $(U \Delta)^T \widetilde{U} = \Delta V_2[1{:}K,~]$ and that $\rho = \frac{\text{trace}(\Delta'')}{\text{trace}({V_2[1{:}K,~]}^T V_2[1{:}K,~])}$, therefore we do not need to explicitly compute $\widetilde{U}$.
+As $U^T U = I$, we note that $(U \Delta)^T \widetilde{U} = \Delta V_2[1{:}K,~]$ and that $\rho = \frac{\text{trace}(\Delta'')}{\text{trace}({V_2[1{:}K,~]}^T V_2[1{:}K,~])}$, therefore we do not need to explicitly compute $\widetilde{U}$ and do not need $U$.
 \item Apply the previous transformation to $\widetilde{u}$ to get the projection of $y$ in the reference PCA space (i.e.\ $\rho \widetilde{u} A$).
@@
 \includegraphics[width=\textwidth]{outliers2-1000G.pdf}
-\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by statistic used to detect outliers (log-scale).\\~\\}
+\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by statistic used to detect outliers (maximum number of SDs from the mean, log-scale).}
 \end{subfigure}
@@
 \includegraphics[width=\textwidth]{be-outlier2-1000G.pdf}
-\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by being detected as an outlier. Threshold of being considered as an outlier is determined based on histogram (a).}
+\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by being detected as an outlier.\\}
 \end{subfigure}
@@
 \includegraphics[width=\textwidth]{UKBB-Maha-dist.png}
-\caption{PCs 1 to 20 of UKBB, colored by statistic of robust Mahalanobis distances (log-scale).\\}
+\caption{PCs 1 to 20 of UKBB, colored by robust Mahalanobis distances computed on PCs (log-scale).\\}
 \end{subfigure}
@@
 \centering
-\caption{Number of UKBB individuals with (log) Mahalanobis distance lower than some threshold (top), grouped by self-reported ancestry (left).} 
+\caption{Number of UKBB individuals with (log squared) Mahalanobis distance lower than some threshold (top), and grouped by self-reported ancestry (left).
+Note that ``$<$ 12'' includes all individuals.}
 \label{tab:homogeneous}
@@
 \centerline{\includegraphics[width=0.8\textwidth]{UKBB-screeplot.pdf}}
-\caption{Scree plot: plot of singular values computed on the UK Biobank.
+\caption{Scree plot: plot of singular values computed on the UK Biobank using \texttt{bed\_autoSVD}.
 \label{fig:UKBB-screeplot}}
@@
 \centerline{\includegraphics[width=0.9\textwidth]{UKBB-PC1-20.png}}
-\caption{Principal Component (PC) scores 1 to 20 computed on the UK Biobank.
+\caption{Principal Component (PC) scores 1 to 20 computed on the UK Biobank using \texttt{bed\_autoSVD}.
 Different colors represent different self-reported ancestries.
@@
 \centerline{\includegraphics[width=0.9\textwidth]{UKBB-loadings.pdf}}
-\caption{Principal Component (PC) loadings 1 to 20 computed on the UK Biobank.
+\caption{Principal Component (PC) loadings 1 to 20 computed on the UK Biobank using \texttt{bed\_autoSVD}.
 \label{fig:UKBB-loadings}}
@@
 \caption{Scree plot: plot of singular values computed on the UK Biobank using 48,942 individuals of diverse ancestries.
+These individuals are the ones resulting from removing all related individuals and randomly subsampling the British and Irish individuals.
 \label{fig:UKBB-screeplot2}}
@@
 \caption{Principal Component (PC) scores 27 to 50 of the UK Biobank.
-Black points are the 48,942 individuals of diverse ancestries used for computing PCA.
+Black points are the 48,942 individuals of diverse ancestries () used for computing PCA.
+These individuals are the ones resulting from removing all related individuals and randomly subsampling the British and Irish individuals.
 Red points are the remaining UKBB individuals, projected by simply multiplying their genotypes by the corresponding PC loadings.

### 2019-11-13 paper-bedpca.tex
first version finished

@@
 
-\newpage
+\vspace*{5em}
 
@@
 
-All code used for the analyses of this paper is available at \url{https://github.com/privefl/paper4-bedpca/tree/master/code}.
-
+All code used for this paper is available at \url{https://github.com/privefl/paper4-bedpca/tree/master/code}.
 Package bigsnpr can be installed from GitHub (see \url{https://github.com/privefl/bigsnpr}).
-
-A tutorial on the steps to perform PCA on 1000G data is available at \url{} [TODO].
+A tutorial on the steps to perform PCA on 1000G data is available at \url{https://privefl.github.io/bigsnpr/articles/bedpca.html}.
 

### 2019-11-19 pca-toolkit.Rmd
add slides

@@
 
+## Pipeline to compute PCs
+
+```{r, out.width="55%"}
+knitr::include_graphics("figures/PCA-pipeline.svg")
+```
+
+---
+
+## Remaining problem: what is really PC16?
+
+```{r, out.width="85%"}
+knitr::include_graphics("figures/UKBB-PC1-20.png")
+```
+
+---
+
+## Facts about PC16
+
+<br>
+
+- unchanged if we impute randomly (using allele frequencies)
+
+- People with PC16 > 30
+
+    - have 36% correlation between PC16 and number of NAs per individuals (also 13% for PC1, 12% for PC3, 11% for PC13 and 14)
+
+    - from almost all assessment centers (data-field 54), batches (data-field 22000) and plates (data-field 22007)
+
+    - from different ancestries (data-field 21000)
+
+    - have between 0.2% and 3.7% of missing values
+
+- loadings are almost all positives (not symmetric)
+
+---
+
+## Possible explanation: heterozygosity
+
+```{r, out.width="90%"}
+knitr::include_graphics("figures/het-PC16.png")
+```
+
+---
+
+## Possible explanation: heterozygosity
+
+```{r, out.width="90%"}
+knitr::include_graphics("figures/het-NA.png")
+```
+
+---
+
 class: inverse, center, middle

### 2019-11-27 code/missing-values-UKBB.R
NAs and het

@@ -1,0 +1,157 @@
+library(bigsnpr)
+library(bigreadr)
+library(dplyr)
+library(ggplot2)
+library(runonce)
+
+rel <- fread2("data/ukb25589_rel_s488346.dat")
+fam <- fread2("data/ukbb_bed/ukbb_488282.fam")
+ind.row <- which(!fam$V2 %in% rel$ID2)
+obj.bed <- bed("data/ukbb_bed/ukbb_488282.bed")
+
+
+#### IMPUTATION BY THE MEAN -> see pca-UKBB.R ####
+
+obj.svd <- readRDS("tmp-results/SVD_UKBB.rds")
+ind.keep <- attr(obj.svd, "subset")
+
+counts <- bed_counts(obj.bed, ind.row, ind.keep, byrow = TRUE, ncores = nb_cores())
+het <- counts[2, ] / colSums(counts[1:3, ])
+nbNA <- counts[4, ]
+PC <- predict(obj.svd)
+summary(mylm <- lm(as.formula(paste0("het ~ (", paste0("PC[ ,", 1:6, "]", collapse = " + "), ")^2"))))
+het_adj <- mylm$coefficients[1] + mylm$residuals
+qplot(nbNA, het_adj, alpha = I(0.5), col = PC[, 16]) +
+  scale_x_log10() +
+  scale_color_viridis_c(direction = -1) +
+  theme_bigstatsr() +
+  labs(x = "Number of missing values (log-scale)",
+       y = "PC-corrected heterozygosity",  color = "PC16")
+
+ggsave("figures/het-col-PC16.png", width = 9, height = 6)
+
+
+#### USING DOSAGES ####
+
+info_snp_UKBB <- rbind_df(lapply(1:22, function(chr) {
+  file <- paste0("data/ukb_imp_mfi/ukb_mfi_chr", chr, "_v3.txt")
+  df <- fread2(file, select = c(3:5, 8), col.names = c("pos", "a0", "a1", "info"))
+  cbind.data.frame(chr = chr, df)
+}))
+
+# subset samples
+sample <- fread2("~/UKBiobank/ukb25589_imp_chr1_v3_s487327.sample")
+str(sample)
+(N <- readBin("data/ukb_imp_chr1_v3.bgen", what = 1L, size = 4, n = 4)[4])
+sample <- sample[-1, ]
+stopifnot(nrow(sample) == N)
[…]

### 2019-11-27 paper-bedpca.tex
add tables and fig need text

@@
 
-\newpage
+\clearpage
 
@@
 
-\newpage
+\clearpage
 \section*{Supplementary Materials}
@@
 
-\subsection*{Supplementary figures}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection*{Sample outlier detection}
 
@@
 
-\newpage
+\clearpage
+\subsection*{Projection onto reference PCA space}
 
@@
 
-\newpage
+\clearpage
+
+\subsection*{PCA of the UK Biobank}
 
@@
 
+\clearpage
+
+\begin{figure}[!htpb]
+\centerline{\includegraphics[width=0.8\textwidth]{het-col-PC16.png}}
+\caption{PC-corrected heterozygosity and number of missing values for individuals of the UK Biobank, colored by their value for PC16.
+\label{fig:UKBB-het}}
+\end{figure}
+
+% latex table generated in R 3.6.0 by xtable 1.8-4 package
+% Wed Nov 27 15:52:34 2019
+\begin{table}[ht]
+\caption{Correlation between first 20 PC scores of the UK Biobank, both computed using the genotyping chip and mean imputation, but wtih either no quality control on individuals based on high heterozygosity (top), or after removing some individuals based on high heterozygosity (left).
+PC16 computed without quality control on high heterozygosity completely disappears when performing this quality control.
+\label{tab:pca-het}}
+\centering
+\begin{tabular}{|r|rrrrrrrrrrrrrrrrrrrr|}
+  \hline
+ & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16 & 17 & 18 & 19 & 20 \\ 
+  \hline
+1 & 100.0 & -0.2 & 0.1 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & -0.0 & -0.0 & 0.0 & 0.0 & -0.0 & -0.0 & -0.0 & -0.3 & -0.0 & 0.0 & 0.0 & 0.0 \\ 
+  2 & 0.1 & 100.0 & 0.2 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & -0.1 & 0.0 & -0.0 & 0.0 & -0.0 & -0.0 & 0.0 & -0.0 & -0.0 & 0.0 & 0.0 \\ 
+  3 & -0.0 & -0.1 & 100.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 & -0.0 & 0.1 & -0.0 & 0.0 & 0.0 & -0.0 & -0.0 & -0.3 & 0.0 & 0.0 & -0.0 & -0.0 \\ 
+  4 & -0.0 & -0.0 & 0.0 & 100.0 & -0.0 & 0.0 & 0.0 & 0.0 & -0.0 & -0.0 & 0.0 & 0.0 & -0.0 & -0.0 & -0.0 & 0.1 & 0.0 & 0.0 & 0.0 & 0.0 \\ 
+  5 & 0.0 & -0.0 & -0.0 & 0.0 & 100.0 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & -0.0 & 0.0 & 0.0 & 0.0 & -0.0 & 0.1 & -0.0 & -0.0 & -0.0 & 0.0 \\ 
+  6 & -0.0 & -0.0 & 0.0 & -0.0 & -0.0 & 100.0 & -0.3 & 0.4 & -0.0 & 0.0 & -0.0 & 0.0 & -0.0 & -0.0 & 0.0 & -0.1 & 0.0 & -0.0 & -0.0 & 0.0 \\ 
+  7 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & 0.3 & 100.0 & -0.2 & -0.0 & 0.2 & -0.1 & -0.1 & 0.0 & 0.0 & 0.0 & 0.3 & -0.0 & -0.0 & 0.0 & -0.0 \\ 
+  8 & -0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.4 & 0.2 & 100.0 & 0.0 & 0.2 & -0.1 & -0.0 & 0.0 & 0.0 & 0.0 & 0.2 & 0.0 & 0.0 & -0.0 & 0.0 \\ 
+  9 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & 100.0 & 0.2 & -0.0 & 0.0 & -0.0 & -0.0 & 0.0 & -0.5 & -0.0 & 0.0 & -0.0 & 0.0 \\ 
+  10 & 0.0 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & -0.1 & -0.1 & -0.2 & 100.0 & 0.4 & -0.1 & 0.1 & 0.0 & 0.0 & 0.8 & -0.0 & -0.0 & 0.0 & 0.0 \\ 
+  11 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & -0.3 & 100.0 & 0.2 & -0.1 & -0.0 & -0.0 & -0.2 & 0.0 & 0.0 & -0.0 & 0.0 \\ 
+  12 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 & -0.0 & 0.0 & 0.0 & -0.0 & 0.1 & -0.1 & 100.0 & -0.3 & 0.1 & 0.0 & 1.7 & -0.0 & -0.1 & -0.0 & -0.0 \\ 
+  13 & 0.0 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & 0.3 & 100.0 & -0.4 & -0.1 & -2.8 & 0.0 & 0.0 & 0.1 & 0.0 \\ 
+  14 & 0.0 & 0.0 & 0.0 & 0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 & -0.0 & 0.0 & -0.1 & 0.4 & 100.0 & -0.0 & -1.8 & 0.0 & 0.0 & 0.0 & -0.0 \\ 
+  15 & 0.0 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & -0.0 & 100.0 & -2.1 & 0.1 & -0.0 & 0.0 & 0.0 \\ 
+  16 & 0.0 & 0.0 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & 0.0 & 0.0 & 0.0 & -0.0 & 1.3 & 100.0 & 0.6 & -0.1 & -0.3 \\ 
+  17 & -0.0 & -0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 & 0.0 & -0.0 & -0.0 & 0.0 & 0.0 & -0.0 & -0.0 & -0.1 & -5.0 & 0.7 & -99.9 & -0.8 & -0.2 \\ 
+  18 & 0.0 & 0.0 & -0.0 & 0.0 & -0.0 & -0.0 & 0.0 & 0.0 & -0.0 & 0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 & -2.9 & 0.0 & 0.9 & -99.5 & 8.7 \\ 
+  19 & 0.0 & 0.0 & -0.0 & 0.0 & 0.0 & 0.0 & -0.0 & 0.0 & 0.0 & 0.0 & 0.0 & -0.0 & 0.0 & -0.0 & 0.0 & -0.3 & -0.2 & 0.3 & -8.6 & -99.6 \\ 
+  20 & 0.0 & 0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 & 0.0 & 0.0 & 0.0 & 0.0 & -0.0 & 0.0 & 0.0 & 0.1 & 7.3 & -0.8 & -0.3 & -4.4 & -0.3 \\ 
+   \hline
+\end{tabular}
+\end{table}
+
+% latex table generated in R 3.6.0 by xtable 1.8-4 package
+% Wed Nov 27 16:04:03 2019
+\begin{table}[ht]
+\caption{Correlation between first 20 PC scores of the UK Biobank, either computed using the genotyping chip and mean imputation (top), or computed from the dosages (based on imputation from an external reference panel) of the same variants (left).
+PCs are globally the same.
+\label{tab:pca-dosage}}
+\centering
+\begin{tabular}{|r|rrrrrrrrrrrrrrrrrrrr|}
+  \hline
+ & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16 & 17 & 18 & 19 & 20 \\ 
+  \hline
+1 & -100.0 & -0.1 & -0.1 & 0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 & -0.0 & -0.0 & -0.0 & 0.0 & 0.0 & 0.0 & -0.0 \\ 
+  2 & 0.1 & -100.0 & -0.2 & -0.0 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & -0.0 & 0.0 & 0.0 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & -0.0 & -0.0 & 0.0 \\ 
+  3 & -0.1 & -0.2 & 100.0 & -0.2 & 0.0 & 0.0 & 0.1 & 0.0 & 0.0 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & -0.0 & 0.0 & 0.0 & -0.0 & 0.0 & -0.0 \\ 
+  4 & 0.0 & -0.0 & 0.2 & 100.0 & 0.1 & 0.0 & -0.1 & -0.0 & 0.0 & 0.1 & -0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & -0.0 & 0.0 & -0.0 & 0.0 \\ 
+  5 & 0.0 & 0.0 & 0.0 & 0.1 & -100.0 & 0.0 & -0.0 & -0.0 & 0.0 & 0.0 & -0.0 & -0.0 & 0.0 & 0.0 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & -0.0 \\ 
+  6 & -0.0 & -0.0 & -0.0 & -0.0 & 0.0 & 99.9 & -2.7 & 2.0 & -0.0 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & -0.0 \\ 
+  7 & -0.0 & -0.0 & -0.1 & 0.1 & -0.0 & 2.6 & 99.9 & 3.3 & -0.1 & -0.1 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & -0.0 \\ 
+  8 & 0.0 & 0.0 & -0.0 & 0.0 & -0.0 & -2.1 & -3.2 & 99.9 & 0.2 & -0.2 & 0.0 & -0.0 & -0.0 & -0.0 & -0.0 & -0.0 & -0.0 & -0.0 & 0.0 & -0.0 \\ 
+  9 & -0.0 & -0.0 & -0.0 & -0.0 & -0.0 & 0.0 & 0.1 & -0.2 & 100.0 & -0.0 & -0.0 & -0.0 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & 0.0 & 0.0 & 0.0 \\ 
+  10 & -0.0 & 0.0 & 0.1 & 0.0 & -0.0 & -0.0 & -0.1 & -0.1 & -0.0 & -99.9 & -0.3 & -0.1 & 0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 & -0.0 & -0.0 \\ 
+  11 & 0.0 & -0.0 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & 0.4 & -99.9 & 0.2 & 0.1 & 0.0 & 0.0 & -0.0 & -0.0 & -0.0 & 0.0 & -0.0 \\ 
+  12 & 0.0 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 & 0.1 & -0.2 & -99.9 & 0.7 & -0.0 & 0.1 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 \\ 
+  13 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & 0.0 & 0.0 & 0.1 & 0.7 & 99.9 & 2.1 & 0.0 & -0.0 & 0.0 & -0.1 & -0.0 & 0.0 \\ 
+  14 & -0.0 & -0.0 & 0.0 & 0.0 & 0.0 & -0.0 & -0.0 & 0.0 & 0.0 & -0.0 & 0.0 & -0.0 & -2.1 & 99.9 & -0.1 & -0.0 & 0.0 & -0.0 & 0.0 & 0.0 \\ 
+  15 & -0.0 & -0.0 & -0.0 & 0.0 & 0.0 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & 0.0 & 0.1 & -0.0 & 0.1 & 99.9 & 0.1 & 0.0 & -0.1 & 0.1 & 0.1 \\ 
+  16 & -0.0 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & -0.0 & -0.0 & -0.0 & -0.0 & -0.0 & -0.0 & -0.0 & -0.0 & 0.0 & -99.7 & 6.4 & 0.8 & 0.9 & -1.3 \\ 
+  17 & -0.0 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & 0.0 & 0.0 & 0.0 & 0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 & 6.4 & 99.6 & -1.0 & 2.4 & 0.5 \\ 
+  18 & 0.0 & -0.0 & -0.0 & 0.0 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & 0.0 & 0.0 & -0.0 & -0.1 & -0.0 & -0.1 & -0.9 & -0.9 & -99.6 & -5.0 & 4.0 \\ 
+  19 & 0.0 & -0.0 & -0.0 & 0.0 & 0.0 & -0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & -0.0 & -0.0 & 0.0 & 0.1 & -0.6 & 2.5 & 4.9 & -99.5 & -6.2 \\ 
+  20 & -0.0 & 0.0 & 0.0 & 0.0 & -0.0 & -0.0 & 0.0 & 0.0 & -0.0 & -0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 & -1.4 & -0.2 & 4.3 & -6.0 & 99.5 \\ 
+   \hline
+\end{tabular}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\clearpage
+
 \begin{figure}[!htpb]

### 2019-11-27 paper-bedpca.tex
add imputation

@@
 
-For this paper, we derive implementations of truncated PCA and other useful functions for e.g.\ performing LD thinning and computing variant statistics.
+For this paper, we derive implementations of truncated PCA and other useful functions for e.g.\ performing LD thinning and computing various statistics.
 We make these available in a new release of R package bigsnpr (v1.0.0); what differs from previously available functions presented in \cite{prive2017efficient} is that these new functions can be used directly on PLINK bed/bim/fam files with some missing values.
@@
 We also point out that many PCs currently reported by the UK Biobank capture LD structure instead of population structure.
-Interestingly, subsetting the UK Biobank data enables to get more PCs capturing population structure than when using the whole sample (\textasciitilde40 instead of \textasciitilde16).
+Interestingly, subsetting the UK Biobank data enables to get more PCs that capture population structure than when using the whole sample (\textasciitilde40 instead of \textasciitilde16).
 Then, we project the other individuals that were not used in the PCA calculation, show that this projection is biased and provide an efficient solution to estimate unbiased projections instead.
-Finally, we explore options to detect outlier samples in PCA, either a few outlier samples that may correspond to e.g.\ batch effects, or when the goal is to restrict the data to individuals of homogeneous ancestry.
+Finally, we explore options to detect outlier samples in PCA, either a few outlier samples that may correspond to e.g.\ batch effects or family structure, or when the goal is to restrict the data to individuals of homogeneous ancestry.
 
@@
 \hline
-\makecell{Detection of \\ oultier samples} & \makecell{``6 SDs from the mean'' \\ in EIGENSOFT} & \cite[]{patterson2006population} & Simple & Assumes Gaussian distritbution \\
+\makecell{Detection of \\ oultier samples} & \makecell{``6 SDs from the mean'' \\ in EIGENSOFT} & \cite[]{patterson2006population} & Simple & Assumes a Gaussian distritbution \\
 \hline
-\makecell{Detection of \\ homogeneous \\ samples} & R package aberrant & \cite[]{bellenguez2011robust} & Robust & Only two statistics at once \\
+\makecell{Detection of \\ homogeneous \\ samples} & R package aberrant & \cite[]{bellenguez2011robust} & Robust & Uses only two statistics at once \\
 \hline
@@
 The idea is that an outlier should be far from all other points, and is even more outlier if its KNNs are in a very dense cluster.
-Here, we use $pd_j / \sqrt{\frac{1}{K} \sum_{k=1}^K pd_{j_k}}$. Note the square root, as it otherwise detect any point next to a very dense cluster as an outlier.
+Here, we use $pd_j / \sqrt{\frac{1}{K} \sum_{k=1}^K pd_{j_k}}$. Note the square root, as it otherwise detects as outlier any point that is next to a very dense cluster.
 We implement (the two parts of) this statistic in function \texttt{prob\_dist} of R package bigutilsr.
@@
 
-For detecting samples that have a different ancestry from most of the samples in the data, i.e.\ for restricting to homogeneous samples, we compute the pairwise orthogonalised Gnanadesikan-Kettenrin robust Mahalanobis distances on PC scores (Section \ref{maha}).
+As for detecting samples that have a different ancestry from most of the samples in the data, i.e.\ for restricting to homogeneous samples, we compute the pairwise orthogonalised Gnanadesikan-Kettenrin robust Mahalanobis distances on PC scores (Section \ref{maha}).
 We then restrict to individuals whose log-distance (alternatively p-value) is smaller (larger) than some threshold determined based on visual inspection.
@@
 
-In this paper, we also analyse the genotyped version of the UK Biobank (\url{https://www.ukbiobank.ac.uk/}).
+In this paper, we also analyse the UK Biobank data (\url{https://www.ukbiobank.ac.uk/}).
 We apply some quality control filters; we remove individuals with more than 10\% missing values, variants with more than 1\% missing values, variants having a minor allele frequency < 0.01, variants with P-value of the Hardy-Weinberg exact test < $10^{-50}$, and non-autosomal variants.
@@
 
-When applying our automatic procedure to remove long-range LD regions, it did not converge after 5 iterations, meaning that it kept detecting long-range LD regions at each iteration.
-Therefore, we were able to capture only 16 PCs that show stratification that is not LD structure (Figures \ref{fig:UKBB-screeplot}, \ref{fig:UKBB-scores} and \ref{fig:UKBB-loadings}).
-Interestingly, PC loadings reported by the UK Biobank clearly show that PC19 to PC40 capture LD structure, which is also the case for PC16 and PC18, although less pronounced (see peaks in figure \ref{fig:UKBB-loadings40}).
+When applying our automatic procedure to remove long-range LD regions, it does not converge after 5 iterations for the UK Biobank, meaning that it keeps detecting long-range LD regions at each iteration (represented by peaks in PC loadings).
+Therefore, we are able to capture only 16 PCs that show stratification that is not LD structure (Figures \ref{fig:UKBB-screeplot}, \ref{fig:UKBB-scores} and \ref{fig:UKBB-loadings}).
+Similarly, PC loadings reported by the UK Biobank clearly show that PC19 to PC40 capture LD structure, which is also the case for PC16 and PC18, although less pronounced (see peaks in figure \ref{fig:UKBB-loadings40}).
 
@@
 Our statistic identifies only isolated samples or isolated pairs that seems to be outliers driving structure of PC17-20 of 1000G (Figure \ref{fig:outlier-pd}). 
-In contrast, 6SD identifies a lot of outliers, of which some are part of a relatively large cluster (Figure \ref{fig:outlier-sd}).
+All but one outliers are distantly-related pairs that disappear if using a more stringent threshold on relatedness (i.e.\ using a KING-relatedness cutoff of \textasciitilde0.0442 instead of \textasciitilde0.0884, see tutorial).
+In contrast, rule 6SD identifies a lot of outliers, of which some are part of a relatively large cluster (Figure \ref{fig:outlier-sd}).
 We recall that, in theory, if all PCs are normally distributed, after correcting for multiple testing of 2500 individuals and 20 PCs, 6SD has a small probability of 0.0001 of detecting one outlier or more.
@@
 
-When the goal is to restrict to homogeneous samples, we compare the use of R package aberrant used to report the ``White British'' subset of UKBB homogeneous samples, to the use of the robust Mahalanobis distance we propose.
+As for restricting to homogeneous samples, we compare the use of R package aberrant used to report the ``White British'' subset of UKBB homogeneous samples, to the use of the robust Mahalanobis distance we propose.
 Here, we visually choose a threshold of 5 on the log-distance and show that this gives a similar subset of individuals than the ``White British'' subset reported by the UK Biobank (Figure \ref{fig:homogeneous}).
-Moreover, when using this threshold, only 3 out of 10,936 people of Asian ancestry (1 ``Chinese'' and 2 ``Indian'') are kept, and 1 ``African'' out of 7622 poeple with Black background is kept (Table \ref{tab:homogeneous}).
+Moreover, when using this threshold, only 3 out of 10,936 people of self-reported Asian ancestry (1 ``Chinese'' and 2 ``Indian'') are kept, and 1 ``African'' out of 7622 people with Black background is kept (Table \ref{tab:homogeneous}).
 In contrast, 416,492 out of 431,090 ``British'' (96.6\%) and 12,620 out of 12,759 ``Irish'' (98.9\%) are kept.
@@
 We use 60\% of individuals in the 1000G data (section \ref{1000G}) to compute K=20 PCs.
-Then, we project the remaining 40\% individuals using three methods: 1/ simply multiplying the genotypes of these individuals by the previously computed loadings, 2/ correcting the simple projections using asymptotic shrinkage factors as determined by R package hdpca v1.1.3 \cite[]{dey2019asymptotic}, with all eigenvalues derived from the Genetic Relationship Matrix (GRM) computed with new function \texttt{bed\_tcrossprodSelf} of R package bigsnpr, and 3/ the OADP projection (section \ref{proj}).
+Then, we project the remaining 40\% individuals using three methods: 1/ simply multiplying the genotypes of these individuals by the previously computed loadings; 2/ correcting the simple projections using asymptotic shrinkage factors as determined by R package hdpca v1.1.3 \cite[]{dey2019asymptotic}, with all eigenvalues derived from the Genetic Relationship Matrix (GRM) computed with \texttt{bed\_tcrossprodSelf}, one of the new functions of R package bigsnpr; and 3/ the OADP projection (section \ref{proj}).
 When simply projecting using loadings, there is negligible shrinkage for PC1 and PC2, a small shrinkage for PC3 and PC4, and a large shrinkage for PC5 to PC8 (Figure \ref{fig:proj1000G}).
 In contrast, there is no visible shrinkage when projecting new individuals with OADP (Figure \ref{fig:proj1000G}).
-Simple projection is affected even more by this shrinkage for PC9 to PC20, while OADP still appears free of this issue (Figure \ref{fig:proj1000G-2}).
+Simple projection is affected even more by this shrinkage for PC9 to PC20, while OADP still appears free of this bias (Figure \ref{fig:proj1000G-2}).
 We show the same results when projecting the full UK Biobank data onto PCA computed using 1000G data  (Figure \ref{fig:proj1000G-5}).
-When correcting projected PC scores with asymptotic shrinkage factors, bias is smaller than when simply projecting loadings, yet, there is a visible bias for PC7-8 (Figure \ref{fig:proj1000G-3}).
+When correcting projected PC scores with asymptotic shrinkage factors, bias is smaller than with simple projection, yet, there is a visible bias for PC7-8 (Figure \ref{fig:proj1000G-3}).
 Finally, to assess if OADP could be used to project individuals that are related to some individuals that were used to compute PCA, we projected these 60\% individuals (as if we were projecting their monozygotic twins) using OADP.
@@
 When computing the PCs on the UK Biobank using 406,545 unrelated individuals and 171,977 variants, and projecting the 1000G data onto this reference PCA space, shrinkage is much smaller ($\leq 1.08$ for all 20 first PCs, Figure \ref{fig:proj1000G-6}).
-Overall, shrinkage decreases with an increased sample size (Table \ref{tab:shrinkage}).
+Overall, this shrinkage decreases with an increased sample size (Table \ref{tab:shrinkage}).
 
@@
 
+\subsection{Imputation of missing values}
+
+As we compute PCA on data with missing values, although we restrict to variants with less than 1\% missing values, we analyse hereinafter the effect of imputation of missing values before computing PCA.
+We compare 4 different imputation methods on two different sets of individuals, either all genotyped individuals or all imputed individuals.
+The UK Biobank imputed data is composed of approximately 1000 individuals less who have been removed because of a high number of missing values or a high heterozygosity \cite[]{bycroft2017genome}.
+When computing PCA with mean imputation and using all individuals, PC16 is capturing individuals with very high heterozygosity (Figure \ref{fig:UKBB-het}).
+When restricting to imputed individuals only, i.e.\ after removing individuals with very high heterozygosity, PC16 completely disappears and new PC16 to PC19 correspond to previous PC17 to PC20 (Table \ref{tab:pca-het}).
+When using dosage data instead of genotype data with mean imputation, PCA is globally unchanged (Table \ref{tab:pca-dosage}). 
+Overall, if we choose to use either one of the following 4 imputation methods: mean imputation, random imputation according to allele frequencies, using reported dosage data from BGEN files, or imputation of genotyped data based on machine learning using function \texttt{snp\_fastImpute} of R package bigsnpr \cite[]{prive2017efficient}, resulting PCs are always very similar (absolute correlation larger than 0.99 for the 20 computed PCs, results not shown).
+Performing PCA with mean imputation directly on PLINK bed files has the advantage to be fast as compared to imputation using \texttt{snp\_fastImpute}, which took 4 days for 406,545 individuals and 240,444 variants, or based on external reference datasets to get dosages.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 This pipeline requires removing all related individuals, for which we provide an R wrapper to PLINK's implementation of KING robust kinship coefficients \cite[]{manichaikul2010robust,chang2015second}.
-Note that one should remove both individuals in each related pairs so that the projected individuals are not related to the ones used for computing PCA, as we showed that relatedness is a problem when using the OADP projection (Figure \ref{fig:proj1000G-4}).
+Note that one should remove both individuals in each related pairs so that the projected individuals are not related to the ones used for computing PCA, since we showed that relatedness is a problem when using the OADP projection (Figure \ref{fig:proj1000G-4}).
 After selecting a subset of individuals, we apply several steps of outlier detection, one for outlier variants that capture long-range LD variation (automatic), and one for detecting outlier samples (semi-automatic and visual).
@@
 As in many applications, we believe there should be a compromise between signal and noise.
-Therefore, we recommend using only PCs that show structure (e.g.\ PC1-16 in figure \ref{fig:UKBB-scores}) and exclude PCs that indicate LD (e.g.\ PC19-40 in figure \ref{fig:UKBB-loadings40}).
+Therefore, we recommend using only PCs that show structure (e.g.\ PC1-16 in figure \ref{fig:UKBB-scores}) and excluding PCs that indicate LD (e.g.\ PC19-40 in figure \ref{fig:UKBB-loadings40}).
 
@@
 Previous work has also shown that uneven population sizes can distort PCs \cite[]{novembre2008interpreting,mcvean2009genealogical}.
-Indeed, when subsetting British and Irish people, we are able to capture a lot more PCs that show population structure with less than 50K individuals as compared to when using more than 400K unrelated individuals.
+Indeed, when subsetting British and Irish people, we are able to capture a lot more PCs that show population structure with less than 50K individuals as compared to when using more than 400K individuals who are mostly composed of British and Irish people.
 This makes designs such as the 1000 Genomes project, which gathered around 100 people for each of 26 populations, highly relevant for capturing population structure.
@@
 Finally, using ``populations'' with approximately the same size is relevant for other biological data as well.
-However, other pitfalls might apply when using other types of data; for example, methylation data can be confounded by factors such as age and sex, and it might be benificial to remove some probes that are associated with these confounding factors before computing PCA \cite[]{decamps2019guidelines}.
+However, other pitfalls might apply when using other types of data; for example, methylation data can be confounded by factors such as age and sex, and it might be benificial to remove some methylation probes that are associated with these confounding factors before computing PCA \cite[]{decamps2019guidelines}.
 
@@
 
-\vspace*{5em}
+\clearpage
 

### 2019-12-13 paper-bedpca.tex
save

@@
 However, conducting PCA analyses can be complicated and has several potential pitfalls.
-These pitfalls include (1) capturing Linkage Disequilibrium (LD) structure instead of population structure, (2) projected PCs that suffer from shrinkage bias when projecting PCA from a reference dataset to another independent dataset, (3) detecting sample outliers, and (4) uneven population sizes.
+These pitfalls include (1) capturing Linkage Disequilibrium (LD) structure instead of population structure, (2) projected PCs that suffer from shrinkage bias, (3) detecting sample outliers, and (4) uneven population sizes.
 In this work, we explore these potential issues when using PCA, and present efficient solutions to these.
@@
 
-For example, we show that PC19 to PC40 in the UK Biobank capture LD structure.
+For example, we find that PC19 to PC40 in the UK Biobank capture complex LD structure rather than population structure.
 Using our automatic algorithm for removing long-range LD regions, we recover 16 PCs that capture population structure only.
-Therefore, we recommend using only 16-18 PCs from the UK Biobank. 
-We provide evidence for a shrinkage bias when projecting PCs computed with data from the 1000 Genomes project.
-Although PC1 to PC4 suffer from only moderate shrinkage (1.01-1.09), PC5 (resp.\ PC10) for example suffers from a shrinkage factor of 1.50 (resp.\ 3.14).
-We provide a fast way to project new individuals that is not affected by this shrinkage bias. 
+Therefore, we recommend using only 16-18 PCs from the UK Biobank to account for population structure confounding.
 We also show how to use PCA to restrict analyses to individuals of homogeneous ancestry.
+Finally, when projecting individual genotypes onto the PCA computed from the 1000 Genomes project data, we find a shrinkage bias that becomes large for PC5 and beyond.
+We then demonstrate how to obtain unbiased projections efficiently using bigsnpr.
 
@@
 Principal Component Analysis (PCA) has been widely used in genetics for many years and in many contexts.
-For instance, adding PCs as covariates has been routinely used to adjust for population structure in Genome-Wide Association Studies (GWAS) \cite[]{price2006principal,novembre2008interpreting}.
-PCA has also been used to detect loci under selection \cite[]{galinsky2016fast,luu2017pcadapt}.
+For instance, adding PCs as covariates is routinely used to adjust for population structure in Genome-Wide Association Studies (GWAS) \cite[]{price2006principal,novembre2008interpreting}.
+PCA has also been used to detect loci under selection \cite[]{galinsky2016fast,luu2017pcadapt} and in heritability analyses \cite{yang2010common,loh2015contrasting}.
 Recently, the advent of large population-scale genetic datasets, such as the UK biobank data, has prompted research on developing scalable algorithms to compute PCA on very large data \cite[]{bycroft2017genome}.
@@
 
-However, some pitfalls related to PCA of genotype data have been documented and not all of the currently available software address all of these. 
-In the following, we outline common pitfalls and explain when they are relevant.
+However, in practice, conducting PCA on genotype data to capture population structure consists of more steps than simply performing singular value decomposition on the genotype matrix.
+These includes removing related individuals, pruning variants in LD, and excluding outlier samples that can suggest poor genotyping quality.
+Indeed, many pitfalls related to PCA of genotype data have been documented and none of the currently available software address all of these.
+In the following, we outline these pitfalls and explain when they are relevant.
 First, some of the PCs may capture LD structure rather than population structure \cite[]{price2008long,abdellaoui2013population,prive2017efficient}.
-Including PCs that capture LD as covariates in genetic analyses can lead to reduced power, e.g.\ for detecting genetic association within these LD regions in GWAS.
+Including PCs that capture LD as covariates in genetic analyses can lead to reduced power for detecting genetic association within these LD regions.  
 Second, another issue may arise when projecting PCs of a reference dataset to another study dataset: projected PCs are shrunk towards 0 in the new dataset \cite[]{lee2010convergence,wang2015improved,zhang2019fast}.
 This shrinkage makes it potentially dangerous to use the projected PCs for analyses such as PC regression, ancestry detection and correction for ancestry.
-Third, PC scores may capture outliers that are due to family structure, population structure or other reasons; it might be benificial to detect and remove these individuals to maximise the population structure captured by PCA (in the case of removing a few outliers) or to restrict analyses to genetically homogeneous samples (e.g.\ ``White British'' people in the UK Biobank).
+This is also an issue when projecting individuals from the same dataset that were discarded from the PCA computation (e.g.\ related individuals).
+Third, PC scores may capture outliers that are due to family structure, population structure or other reasons; it might be beneficial to detect and remove these individuals to maximise the population structure captured by PCA (in the case of removing a few outliers) or to restrict analyses to genetically homogeneous samples (e.g.\ ``White British'' people in the UK Biobank).
 Finally, efficient methods for PCA use approximations, which can results in some lack of precision of computed PCs.
@@
 \hline
-Analysis & Method and/or Software & Citation & Advantages & Limitations \\
+Analysis & Method and/or Software & Citation & Advantages & Current limitations \\
 \hline
@@
 Similarly, PC loadings reported by the UK Biobank clearly show that PC19 to PC40 capture LD structure, which is also the case for PC16 and PC18, although less pronounced (see peaks in figure \ref{fig:UKBB-loadings40}).
+These include [TODO: FIND REGION -> CLOSE TO HLA?] in PC19.
 
@@
 \caption{Principal Component (PC) loadings 1 to 40 reported by the UK Biobank.
+We expect population structure to be well distributed along the genome so that we get loadings (weights for each variant for each PC) that are normally distributed around 0 (with small effect sizes).
+In contrast, long-range LD structure is basically capturing the variation inside one long-range LD region (so localised in the genome), so that we expect the loadings to be very large in that region only (one peak).
+X-axis represent positions of variants in the data, and y-axis the value of loadings.
+Points are hex-binned to make plotting of such large number of points manageable.
 \label{fig:UKBB-loadings40}}
@@
 Our statistic identifies only isolated samples or isolated pairs that seems to be outliers driving structure of PC17-20 of 1000G (Figure \ref{fig:outlier-pd}). 
-All but one outliers are distantly-related pairs that disappear if using a more stringent threshold on relatedness (i.e.\ using a KING-relatedness cutoff of \textasciitilde0.0442 instead of \textasciitilde0.0884, see tutorial).
+All but one outliers are distantly-related pairs that disappear if using a more stringent threshold on relatedness (i.e.\ using a KING-relatedness cutoff of \textasciitilde0.0442 instead of \textasciitilde0.0884, see tutorial in section ``code availability'').
 In contrast, rule 6SD identifies a lot of outliers, of which some are part of a relatively large cluster (Figure \ref{fig:outlier-sd}).
@@
 \includegraphics[width=\textwidth]{outliers-1000G.pdf}
-\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by the statistic (S) used to define outliers.\\~\\}
+\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by the statistic (S) used to define outliers.
+A few points with higher values for this statistic S appear as outliers in PC17-20.}
 \end{subfigure}
@@
 
-\subsection{Projecting PCs from a reference dataset}
+\subsection{Projecting onto the PCA space from a reference dataset}
 
@@
 Overall, if we choose to use either one of the following 4 imputation methods: mean imputation, random imputation according to allele frequencies, using reported dosage data from BGEN files, or imputation of genotyped data based on machine learning using function \texttt{snp\_fastImpute} of R package bigsnpr \cite[]{prive2017efficient}, resulting PCs are always very similar (absolute correlation larger than 0.99 for the 20 computed PCs, results not shown).
-Performing PCA with mean imputation directly on PLINK bed files has the advantage to be fast as compared to imputation using \texttt{snp\_fastImpute}, which took 4 days for 406,545 individuals and 240,444 variants, or based on external reference datasets to get dosages.
+This justifies performing PCA with mean imputation directly on PLINK bed files with a few missing values; this has the advantage to be much faster than having to impute data using \texttt{snp\_fastImpute}, which took 4 days for 406,545 individuals and 240,444 variants, or based on external reference datasets.
 
@@
 \centerline{\includegraphics[width=0.4\textwidth]{PCA-pipeline.pdf}}
-\caption{Pipeline for computing Principal Components (PCs) using R packages bigsnpr and bigutilsr.
+\caption{Proposed pipeline for computing Principal Components (PCs) using R packages bigsnpr and bigutilsr.
 \label{fig:pipeline}}
@@
 The new functions we provide can be directly applied to genotypes stored as PLINK bed/bim/fam files with some missing values.
-This contrasts with previous releases of package bigsnpr that could only use one custom data format called ``bigSNP''.
+This contrasts with previous releases of R package bigsnpr that could only use format ``bigSNP''.
 This data format can store both genotype calls and dosages, but requires conversion from other formats and imputation of missing values using functions provided in the package \cite[]{prive2017efficient}.
@@
 Indeed, when subsetting British and Irish people, we are able to capture a lot more PCs that show population structure with less than 50K individuals as compared to when using more than 400K individuals who are mostly composed of British and Irish people.
-This makes designs such as the 1000 Genomes project, which gathered around 100 people for each of 26 populations, highly relevant for capturing population structure.
+This makes designs such as the 1000 Genomes project, which gathered around 100 people for each of 26 populations, highly relevant for capturing population structure \cite[]{10002015global}.
 In contrast, when the goal is to have an homogeneous sample, we show that using the Mahalanobis distance on PC scores can efficiently achieve this goal, which we used in previous analyses \cite[]{prive2019efficient}.
 When the homogeneous sample is not predominant in the dataset, one solution is to compute the center and covariance of the robust Mahalanobis distance using only the population of interest, and then computing the distances for all individuals using these robust estimates.
+
+[TODO: ADD PART ON MORE WORK NEEDED -> optimal subsetting for populations and effect of including long-range LD regions in e.g.\ GWAS.]
 
@@
 \caption{Principal Component (PC) loadings 1 to 20 computed on the UK Biobank using \texttt{bed\_autoSVD}.
+X-axis represent positions of variants in the data, and y-axis the value of loadings.
+Points are hex-binned to make plotting of such large number of points manageable.
 \label{fig:UKBB-loadings}}
@@
 \caption{Principal Component (PC) scores 27 to 50 of the UK Biobank.
-Black points are the 48,942 individuals of diverse ancestries () used for computing PCA.
+Black points are the 48,942 individuals of diverse ancestries used for computing PCA.
 These individuals are the ones resulting from removing all related individuals and randomly subsampling the British and Irish individuals.

### 2019-12-13 pdf2png.R

@@ -6,3 +6,10 @@ library(magick)
   image_write(img, svg, format = "SVG")
   print(svg)
 }
+
+"figures/outliers-1000G.pdf" %>% {
+  svg <- sub("\\.pdf$", ".png", .)
+  img <- image_read_pdf(.)
+  image_write(img, svg, format = "PNG")
+  print(svg)
+}

### 2019-12-14 code/pca-UKBB.R
find LDLDR of PC19

@@ -8,8 +8,9 @@ fam <- fread2("data/ukbb_bed/ukbb_488282.fam")
 ind.row <- which(!fam$V2 %in% rel$ID2)
 obj.bed <- bed("data/ukbb_bed/ukbb_488282.bed")
 
-system.time(
-  obj.svd <- bed_autoSVD(obj.bed, ind.row = ind.row, k = 20, ncores = nb_cores())
+obj.svd <- runonce::save_run(
+  bed_autoSVD(obj.bed, ind.row = ind.row, k = 20, ncores = nb_cores()),
+  file = "tmp-results/SVD_UKBB.rds"
 ) # 5h
 # Phase of clumping (on MAC) at r^2 > 0.2.. keep 261306 variants.
 # Discarding 0 variant with MAC < 10.
@@ -43,7 +44,6 @@ system.time(
 # Computing SVD..
 # Maximum number of iterations reached.
 
-# saveRDS(obj.svd, "tmp-results/SVD_UKBB.rds")
 
 system.time(
   ind.keep2 <- bed_clumping(obj.bed, ind.row = ind.row, ncores = nb_cores())
@@ -106,6 +106,12 @@ plot_grid(plotlist = lapply(1:40, function(i) {
 }), align = "hv", ncol = 5, scale = 0.95)
 
 # ggsave("figures/UKBB-loadings1-40.pdf", width = 15, height = 15)
+
+# Long-range LD region of PC19
+load_PC19 <- loadings$PC9_loading.1
+ind_large <- which(abs(load_PC19) > 0.01)
+plot(ind_large, load_PC19[ind_large], pch = 20)
+snp_qc[ind_large, c("chromosome", "position")] # chr6 - [70-91]
 
 nPC <- 20
 PC <- fread2(

### 2019-12-19 code/outlier-samples.R
common title and legend

@@ -18,21 +18,22 @@ ggplot() +
 
 # ggsave("figures/hist-outliers-1000G.pdf", width = 9, height = 6)
 
-plot_grid(plotlist = lapply(7:10, function(k) {
+source("code/plot_grid2.R")
+plot_grid2(plotlist = lapply(7:10, function(k) {
   plot(obj.svd, type = "scores", scores = 2 * k - 1:0, coeff = 0.6) +
     aes(color = S) +
     scale_colour_viridis_c()
 }), scale = 0.95)
 
-# ggsave("figures/outliers-1000G.pdf", width = 7, height = 5)
+# ggsave("figures/outliers-1000G.pdf", width = 6, height = 5)
 
-plot_grid(plotlist = lapply(7:10, function(k) {
+plot_grid2(plotlist = lapply(7:10, function(k) {
   plot(obj.svd, type = "scores", scores = 2 * k - 1:0, coeff = 0.6) +
     aes(color = S > 0.4) +
     scale_colour_viridis_d()
 }), scale = 0.95)
 
-# ggsave("figures/be-outlier-1000G.pdf", width = 7, height = 5)
+# ggsave("figures/be-outlier-1000G.pdf", width = 6, height = 5)
 
 library(magrittr)
 S2 <- apply(obj.svd$u, 2, function(x) abs(x - mean(x)) / sd(x)) %>%
@@ -47,18 +48,18 @@ ggplot() +
 
 # ggsave("figures/hist-outliers2-1000G.pdf", width = 9, height = 6)
 
-plot_grid(plotlist = lapply(7:10, function(k) {
+plot_grid2(plotlist = lapply(7:10, function(k) {
   plot(obj.svd, type = "scores", scores = 2 * k - 1:0, coeff = 0.6) +
     aes(color = S2) +
     scale_colour_viridis_c(trans = "log", breaks = c(1, 3, 6, 20))
 }), scale = 0.95)
 
-# ggsave("figures/outliers2-1000G.pdf", width = 7, height = 5)
+# ggsave("figures/outliers2-1000G.pdf", width = 6, height = 5)
 
-plot_grid(plotlist = lapply(7:10, function(k) {
+plot_grid2(plotlist = lapply(7:10, function(k) {
   plot(obj.svd, type = "scores", scores = 2 * k - 1:0, coeff = 0.6) +
     aes(color = S2 > 6) +
     scale_colour_viridis_d()
 }), scale = 0.95)
 
[…]

### 2019-12-19 code/plot_grid2.R

@@ -1,0 +1,13 @@
+plot_grid2 <- function(plotlist, ..., title_ratio = 1/8, legend_ratio = 1/6) {
+
+  main_grid <- plot_grid(plotlist = lapply(plotlist, function(p) {
+    p + theme(legend.position = "none") + ggtitle(NULL)
+  }), ...)
+
+  plot_grid(
+    cowplot::get_title(plotlist[[1]]),  ggplot() + theme_minimal(),
+    main_grid,                          cowplot::get_legend(plotlist[[1]]),
+    rel_heights = c(title_ratio, 1), ncol = 2,
+    rel_widths = c(1, legend_ratio), nrow = 2
+  )
+}

### 2019-12-19 paper-bedpca.tex
merge some suggestions

@@
 Similarly, PC loadings reported by the UK Biobank clearly show that PC19 to PC40 capture LD structure, which is also the case for PC16 and PC18, although less pronounced (see peaks in figure \ref{fig:UKBB-loadings40}).
-These include [TODO: FIND REGION -> CLOSE TO HLA?] in PC19.
+These include e.g.\ one region on chromosome 6 (70-91 Mbp) that is captured in PC19 (Figure \ref{fig:UKBB-loadings}) and that was not previously reported in \cite{price2008long}.
 
@@
 \centering
-\begin{subfigure}[b]{0.5\textwidth}
+\begin{subfigure}[b]{0.55\textwidth}
 \includegraphics[width=\textwidth]{hist-outliers-1000G.pdf}
-\caption{Distribution of statistics}
+\caption{Distribution of statistics (S)}
 \end{subfigure}
@@
 
-In this work, we have compiled different pitfalls that can arise with analyses related to Principal Component Analysis (PCA) of genetic data.
-Then, we have investigated possible solutions to these pitfalls and chose the ones that we found most relevant, both with respect to properties such as accuracy and robustness, but also that would be fast and easy to use.
-We then implemented these solutions in open-source R packages.
-The new functions we provide can be directly applied to genotypes stored as PLINK bed/bim/fam files with some missing values.
-This contrasts with previous releases of R package bigsnpr that could only use format ``bigSNP''.
+In this work, we have compiled different pitfalls that can arise with Principal Component Analysis (PCA) of genetic data.
+Then, we have investigated possible solutions to these pitfalls and selected the ones that we found most advantageous, both with respect to properties such as accuracy and robustness, but also computational efficiency and ease of use.
+We then implemented these solutions in R packages bigsnpr and bigutilsr. 
+The new functions we provide in R package bigsnpr can be directly applied to genotypes stored as PLINK bed/bim/fam files with some missing values.
+This contrasts with previous releases of bigsnpr that could only use format ``bigSNP''.
 This data format can store both genotype calls and dosages, but requires conversion from other formats and imputation of missing values using functions provided in the package \cite[]{prive2017efficient}.
@@
 
-As a general recommendation for computing PCA, we propose the pipeline shown in figure \ref{fig:pipeline}.
+We summarise our work into several recommendations for computing PCA, and propose the pipeline shown in figure \ref{fig:pipeline}.
 Note that we have not included standard steps such as initial quality control filters and post-analysis checks (e.g.\ visual inspection of different plots).
 This pipeline requires removing all related individuals, for which we provide an R wrapper to PLINK's implementation of KING robust kinship coefficients \cite[]{manichaikul2010robust,chang2015second}.
-Note that one should remove both individuals in each related pairs so that the projected individuals are not related to the ones used for computing PCA, since we showed that relatedness is a problem when using the OADP projection (Figure \ref{fig:proj1000G-4}).
+Note that one should remove both individuals in each pair of related individuals.
+This ensures that the projected individuals are not related to the ones used for computing PCA, since we showed that relatedness is a problem when using the OADP projection (Figure \ref{fig:proj1000G-4}).
 After selecting a subset of individuals, we apply several steps of outlier detection, one for outlier variants that capture long-range LD variation (automatic), and one for detecting outlier samples (semi-automatic and visual).
@@
 
-We also provide some recommendations when the dataset is composed mainly of one population (e.g.\ British people in the UK Biobank).
-We show that it is useful to actually subset these individuals so there is not too much imbalance between the sample sizes of the different populations.
+When analysing a dataset that is composed mainly of one population (e.g.\ British people in the UK Biobank), we found that it is useful to subset these individuals to reduce the imbalance between the different population sizes.
 Previous work has also shown that uneven population sizes can distort PCs \cite[]{novembre2008interpreting,mcvean2009genealogical}. 
 Indeed, when subsetting British and Irish people, we are able to capture a lot more PCs that show population structure with less than 50K individuals as compared to when using more than 400K individuals who are mostly composed of British and Irish people.
-This makes designs such as the 1000 Genomes project, which gathered around 100 people for each of 26 populations, highly relevant for capturing population structure \cite[]{10002015global}.
-In contrast, when the goal is to have an homogeneous sample, we show that using the Mahalanobis distance on PC scores can efficiently achieve this goal, which we used in previous analyses \cite[]{prive2019efficient}.
+This suggests that designs such as the 1000 Genomes project, which gathered around 100 people for each of 26 different populations, are highly relevant for capturing population structure \cite[]{10002015global}.
+
+In contrast, a common strategy in genetic analyses is to restrict the analysis to a homogeneous sample to reduce risk of confounding due to population stratification.
+For that purpose, we show that using the Mahalanobis distance on PC scores can efficiently achieve this goal, which we used in previous analyses \cite[]{prive2019efficient}.
 When the homogeneous sample is not predominant in the dataset, one solution is to compute the center and covariance of the robust Mahalanobis distance using only the population of interest, and then computing the distances for all individuals using these robust estimates.
+
+We tackled many problems related to computing PCA on genotype data in this paper. 
+Yet, more work is still needed. 
+For example, the effect of including long-range LD regions in GWAS should be investigated quantitatively.
+These are directions of future work.
 
@@
 Finally, using ``populations'' with approximately the same size is relevant for other biological data as well.
-However, other pitfalls might apply when using other types of data; for example, methylation data can be confounded by factors such as age and sex, and it might be benificial to remove some methylation probes that are associated with these confounding factors before computing PCA \cite[]{decamps2019guidelines}.
+However, other pitfalls might apply when using other types of data; for example, methylation data can be confounded by factors such as age and sex, and it might be beneficial to remove some methylation probes that are associated with these confounding factors before computing PCA \cite[]{decamps2019guidelines}.
 
@@
 
-\section*{Declaraction of Interests}
+\section*{Declaration of Interests}
 
@@
 \centering
-\begin{subfigure}[b]{0.5\textwidth}
+\begin{subfigure}[b]{0.55\textwidth}
 \includegraphics[width=\textwidth]{hist-outliers2-1000G.pdf}
-\caption{Distribution of statistics and default threshold (6, in red)}
+\caption{Distribution of statistics (S2) and default threshold (6, in red)}
 \end{subfigure}

### 2019-12-20 paper-bedpca.tex
add future directions

@@
 \caption{Principal Component (PC) loadings 1 to 40 reported by the UK Biobank.
-We expect population structure to be well distributed along the genome so that we get loadings (weights for each variant for each PC) that are normally distributed around 0 (with small effect sizes).
-In contrast, long-range LD structure is basically capturing the variation inside one long-range LD region (so localised in the genome), so that we expect the loadings to be very large in that region only (one peak).
 X-axis represent positions of variants in the data, and y-axis the value of loadings.
@@
 Using educated guesses in R package PRIMME based on low-rank approximations seemed to be a promising approach but did not reduce computation time by much, so we did not pursue this idea \cite[]{brand2003fast,wu2017primme}.
-Once PCA is done, one should look at the loadings to make sure that there is no visible LD structure left (peaks in PC loadings as shown in figures \ref{fig:UKBB-loadings40} and \ref{fig:UKBB-loadings}), since including PCs that capture LD as covariates in genetic analyses can lead to reduced power for variants in these regions.
-These long-range LD regions are captured by PCs because variants in these regions are extensively correlated (even after a first step of LD thinning), and PCA just captures the variance in these correlated variants.
-This is a similar problem as including related individuals or even whole families in PCA, they would show up as structure in PCs. 
-One should also look at PC scores to identify which PCs capture population structure.
-As in many applications, we believe there should be a compromise between signal and noise.
-Therefore, we recommend using only PCs that show structure (e.g.\ PC1-16 in figure \ref{fig:UKBB-scores}) and excluding PCs that indicate LD (e.g.\ PC19-40 in figure \ref{fig:UKBB-loadings40}).
+
+Once PCA is done, one should check the PC scores (scores of each individual for each PCA dimension) and PC loadings (weights for each variant for each PCA dimension). 
+We differentiate PCs in 3 broad types: the ones capturing LD structure, the ones capturing population structure, and noise.
+We expect population structure to be well distributed along the genome so that we get loadings that are normally distributed around 0 (with small effect sizes).
+In contrast, long-range LD structure is basically capturing the variation inside one long-range LD region (so localised in the genome), so that we expect the loadings to be very large in that region only (one peak).
+Therefore, PCs capturing LD structure can be identified by looking for peaks in PC loadings (e.g.\ PC17-20 in figure \ref{fig:UKBB-loadings}). 
+To identify which PCs capture population structure, and which ones are probably just noise, one should also look at PC scores (colored by ancestry if possible).
+As in many applications, we believe a compromise between signal and noise should be preferred.
+Therefore, we recommend using only PCs that show structure (e.g.\ PC1-14, and probably also PC15-16, in figure \ref{fig:UKBB-scores}) and excluding PCs that does not seem to capture any population structure (e.g.\ PC17-20 in figure \ref{fig:UKBB-scores}).
 
 When analysing a dataset that is composed mainly of one population (e.g.\ British people in the UK Biobank), we found that it is useful to subset these individuals to reduce the imbalance between the different population sizes.
-Previous work has also shown that uneven population sizes can distort PCs \cite[]{novembre2008interpreting,mcvean2009genealogical}. 
+Likewise, previous works have shown that uneven population sizes can distort PCs \cite[]{novembre2008interpreting,mcvean2009genealogical}. 
 Indeed, when subsetting British and Irish people, we are able to capture a lot more PCs that show population structure with less than 50K individuals as compared to when using more than 400K individuals who are mostly composed of British and Irish people.
+Determining how much overdominant populations should be subsampled to maximise population structure captured by PCA is a direction of future work.
+The remaining individuals can then be projected onto the resulting PCA space using the OADP projection we recommend in this paper.
 This suggests that designs such as the 1000 Genomes project, which gathered around 100 people for each of 26 different populations, are highly relevant for capturing population structure \cite[]{10002015global}.
@@
 
-We tackled many problems related to computing PCA on genotype data in this paper. 
-Yet, more work is still needed. 
-For example, the effect of including long-range LD regions in GWAS should be investigated quantitatively.
+The ubiquitous use of PCA in a wide variety of genomic analyses makes it difficult to establish universal guidelines for such analysis. 
+Although we have tackled many problems related to computing PCA on genotype data in this paper, we do not answer other important problems, such as how to best control for population structure in genomic analyses.
+For example, when conducting a GWAS, should one restrict to a homogenous sample, or is it enough to just include PCs that capture population structure as covariates, or should one also use PCs as covariates in mixed linear models \cite[]{price2010new,loh2015efficient}?
+Similarly, in some analyses, it may be beneficial to include PCs that capture long-range or even inter-chromosomal LD.
+More work is needed to understand these fundamental problems, and to provide precise guidelines for conducting successful GWAS, heritability, and other genomic analyses where PCA is used.
 These are directions of future work.
-
-[TODO: ADD PART ON MORE WORK NEEDED -> optimal subsetting for populations and effect of including long-range LD regions in e.g.\ GWAS.]
 
@@
 
-\clearpage
+%\clearpage
+\vspace*{5em}
 

### 2020-02-18 paper-bedpca.tex
loadings ordered
minor corrections on text

@@
 For instance, adding PCs as covariates is routinely used to adjust for population structure in Genome-Wide Association Studies (GWAS) \cite[]{price2006principal,novembre2008interpreting}.
-PCA has also been used to detect loci under selection \cite[]{galinsky2016fast,luu2017pcadapt} and in heritability analyses \cite{yang2010common,loh2015contrasting}.
+PCA has also been used to detect loci under selection \cite[]{galinsky2016fast,luu2017pcadapt} and in heritability analyses \cite[]{yang2010common,loh2015contrasting}.
 Recently, the advent of large population-scale genetic datasets, such as the UK biobank data, has prompted research on developing scalable algorithms to compute PCA on very large data \cite[]{bycroft2017genome}.
@@
 However, in practice, conducting PCA on genotype data to capture population structure consists of more steps than simply performing singular value decomposition on the genotype matrix.
-These includes removing related individuals, pruning variants in LD, and excluding outlier samples that can suggest poor genotyping quality.
+These steps includes removing related individuals, pruning variants in LD, and excluding outlier samples that can suggest poor genotyping quality or distant relatedness.
 Indeed, many pitfalls related to PCA of genotype data have been documented and none of the currently available software address all of these.
 In the following, we outline these pitfalls and explain when they are relevant.
-First, some of the PCs may capture LD structure rather than population structure \cite[]{price2008long,abdellaoui2013population,prive2017efficient}.
-Including PCs that capture LD as covariates in genetic analyses can lead to reduced power for detecting genetic association within these LD regions.  
-Second, another issue may arise when projecting PCs of a reference dataset to another study dataset: projected PCs are shrunk towards 0 in the new dataset \cite[]{lee2010convergence,wang2015improved,zhang2019fast}.
+First, some of the PCs may capture LD structure rather than population structure \cite[]{price2008long,zou2010quantification,abdellaoui2013population,prive2017efficient}.
+Including PCs that capture LD as covariates in genetic analyses can lead to reduced power for detecting genetic associations within these LD regions \cite[]{zou2010quantification}.  
+Second, another issue may arise when projecting a new study dataset to the PCA space computed from a reference dataset: projected PCs are shrunk towards 0 in the new dataset \cite[]{lee2010convergence,wang2015improved,zhang2019fast}.
 This shrinkage makes it potentially dangerous to use the projected PCs for analyses such as PC regression, ancestry detection and correction for ancestry.
-This is also an issue when projecting individuals from the same dataset that were discarded from the PCA computation (e.g.\ related individuals).
+This same issue also arises when projecting individuals from the same dataset that were discarded from the PCA computation (e.g.\ related individuals).
 Third, PC scores may capture outliers that are due to family structure, population structure or other reasons; it might be beneficial to detect and remove these individuals to maximise the population structure captured by PCA (in the case of removing a few outliers) or to restrict analyses to genetically homogeneous samples (e.g.\ ``White British'' people in the UK Biobank).
 Finally, efficient methods for PCA use approximations, which can results in some lack of precision of computed PCs.
-This has been shown to be a possible issue for software such as FastPCA and PLINK 2.0, but not for FlashPCA2 and bigstatsr/bigsnpr \cite[]{abraham2017flashpca2,prive2017efficient}.
+This potential issue has been demonstrated for software such as FastPCA and PLINK 2.0, but not for FlashPCA2 and bigstatsr/bigsnpr \cite[]{abraham2017flashpca2,prive2017efficient}.
 An overview of existing methods with their respective advantages and limitations is presented in table \ref{tab:method-overview}.
@@
 We make these available in a new release of R package bigsnpr (v1.0.0); what differs from previously available functions presented in \cite{prive2017efficient} is that these new functions can be used directly on PLINK bed/bim/fam files with some missing values.
-We use these new functions to analyse the UK Biobank data, and show that they are both very fast and easy to use.
+We use these new functions to analyse the UK Biobank data, and show that these functions are both very fast and easy to use.
 We also point out that many PCs currently reported by the UK Biobank capture LD structure instead of population structure.
 Interestingly, subsetting the UK Biobank data enables to get more PCs that capture population structure than when using the whole sample (\textasciitilde40 instead of \textasciitilde16).
-Then, we project the other individuals that were not used in the PCA calculation, show that this projection is biased and provide an efficient solution to estimate unbiased projections instead.
-Finally, we explore options to detect outlier samples in PCA, either a few outlier samples that may correspond to e.g.\ batch effects or family structure, or when the goal is to restrict the data to individuals of homogeneous ancestry.
+Then, we project the other individuals that were not used in the PCA calculation, show that this projection is biased and provide an efficient solution to get unbiased projections instead.
+Finally, we explore options to detect outlier samples in PCA, either a few outlier samples that may correspond to e.g.\ batch effects or distant family structure, or when the goal is to restrict the data to individuals of homogeneous ancestry.
 
@@
 
-\subsection{Efficient implementation of PCA for genotype data}
+\subsection{Efficient implementation of PCA for genotype data \label{randomSVD}}
 
-When there is no missing value, we compute the partial Singular Value Decomposition (SVD) $U \Delta V^T$ of the scaled genotype matrix of diploid individuals $\tilde{G}_{i,j} = \frac{G_{i,j} - 2 \hat{f_j}}{\sqrt{2 \hat{f_j} (1 - \hat{f_j})}}$ where $G_{i,j}$ is the allele count (genotype) of individual $i$ and variant $j$, and $\hat{f_j}$ is the estimated allele frequency of variant $j$ ($2 \hat{f_j}$ is the mean allele count of variant $j$).
+When there is no missing value, we compute the truncated Singular Value Decomposition (SVD) $U \Delta V^T$ of the scaled genotype matrix of diploid individuals $\tilde{G}_{i,j} = \frac{G_{i,j} - 2 \hat{f_j}}{\sqrt{2 \hat{f_j} (1 - \hat{f_j})}}$ where $G_{i,j}$ is the allele count (genotype) of individual $i$ and variant $j$, and $\hat{f_j}$ is the estimated allele frequency of variant $j$ ($2 \hat{f_j}$ is the mean allele count of variant $j$).
 Then, $U \Delta$ are the first $K$ PC scores and $V$ are the first $K$ PC loadings, where $K$ is the number of PCs computed (e.g.\ $K=20$).
 
-When there are some missing values, we compute the partial SVD similarly, except that missing values are replaced by the variant means (i.e.\ $G_{i,j} - 2 \hat{f_j} = 0$ when $G_{i,j}$ is missing) and  the $\hat{f_j}$ for each variant are estimated using only non-missing values of each variant.
+When there are some missing values, we compute the partial SVD similarly, except that missing values are replaced by the variant means (i.e.\ $G_{i,j} - 2 \hat{f_j} = 0$ when $G_{i,j}$ is missing) and  the $\hat{f_j}$ for each variant are estimated using only non-missing genotypes.
 Note that this decomposition is equivalent to the decomposition presented above after imputation by the variant means.
@@
 Then, we apply the same algorithm for partial SVD that is used in R packages bigstatsr and FlashPCA2, namely the implicitly restarted Arnoldi method \cite[]{Lehoucq1996,abraham2017flashpca2,prive2017efficient}.
-This algorithm is implemented in R package RSpectra, and requires a function that computes the matrix-vector multiplication of the scaled genotype matrix with a given vector, which we implement for PLINK bed files and parallelise.
+This algorithm, implemented in R package RSpectra, requires a function that computes the matrix-vector multiplication of the scaled genotype matrix with a given vector.
+We implement such multiplication in parallel from a PLINK bed file.
 
@@
 
-Mahalanobis distances are computed as $d(x)^2 = (x-\mu)^T \Sigma^{-1} (x-\mu),$ where $\mu$ and $\Sigma$ are (robust) estimators of location and covariance of $x$.
+Mahalanobis distances are computed as $d(x)^2 = (x-\mu)^T \Sigma^{-1} (x-\mu),$ where $\mu$ and $\Sigma$ are (robust) estimators of location and covariance.
 We use these distances for many applications in this paper. 
 When $x$ is multivariate Gaussian data with $K$ dimensions, the squared distances follow a $\chi^2(K)$ distribution.
-If $x$ are PC scores of centered data and if we use standard estimates, then $\mu = 0$ and $\Sigma = I_K$.
-Yet, here we use the pairwise orthogonalised Gnanadesikan-Kettenrin robust estimates of these parameters, as implemented in function \texttt{covRob} of R package robust with parameter \texttt{estim = "pairwiseGK"} \cite[]{gnanadesikan1972robust,yohai1988high,maronna2002robust,todorov2009object}.
-We reexport function \texttt{covRob} in R package bigutilsr for convenience.
+If $x$ represents PC scores of centered data and if we use standard estimates, then $\mu = 0$ and $\Sigma = I_K$.
+Yet, here we use the pairwise orthogonalised Gnanadesikan-Kettenrin robust estimates of these parameters \cite[]{gnanadesikan1972robust,yohai1988high,maronna2002robust}.
+We implement the estimation of these robust parameters in function \texttt{covrob\_ogk} of R package bigutilsr.
 
@@
 We first apply a first round of clumping at e.g.\ $r^2$ > 0.2, prioritising variants by higher minor allele count.
-Then, we compute $K$ PC scores and loadings.
-To summarise the contribution of each variant in all $K$ PC loadings, we compute the robust Mahalanobis distances (Section \ref{maha}) of these PC loadings.  
-To capture consecutive outliers that corresponds to long-range LD regions, we apply a Gaussian smoothing to these statistics (moving average with a Gaussian filter over a window of 50 variants as radius by default).
+Then, we compute $K$ PC scores and loadings (Section \ref{randomSVD}).
+To summarise the contribution of each variant in all $K$ PC loadings, we compute the robust Mahalanobis distances of these PC loadings (Section \ref{maha}).  
+To capture consecutive outliers that corresponds to long-range LD regions, we apply a Gaussian smoothing to these statistics (moving average with a Gaussian filter over a window with a radius of 50 variants by default).
 
@@
 Standard Tukey's rule also uses a fixed coefficient (1.5) that does not account for multiple testing, which means that there are always some outliers detected when using 1.5 for large samples. 
-To solve these two potential issues, we implement \texttt{tukey\_mc\_up} in R package bigutilsr and use it here, which accounts both for skewness and multiple testing by default.
+To solve these two potential issues, we implement \texttt{tukey\_mc\_up} in R package bigutilsr and use it here, which accounts for both skewness and multiple testing by default.
 
@@
 Using K nearest neighbours (KNN), this consists in comparing the distance from a point $j$ to its KNNs ($pd_j = \sqrt{\frac{1}{K} \sum_{k=1}^K d^2_{j \leftrightarrow j_k}}$, where $j_k$ is the $k$-th NN of $j$) with the distances from its KNNs to their respective KNNs ($\frac{1}{K} \sum_{k=1}^K pd_{j_k}$).
-The idea is that an outlier should be far from all other points, and is even more outlier if its KNNs are in a very dense cluster.
-Here, we use $pd_j / \sqrt{\frac{1}{K} \sum_{k=1}^K pd_{j_k}}$. Note the square root, as it otherwise detects as outlier any point that is next to a very dense cluster.
+Intuitively, an outlier should be far from all other points, and is even more outlier if its KNNs are in a very dense cluster.
+Here, we use $pd_j / \sqrt{\frac{1}{K} \sum_{k=1}^K pd_{j_k}}$ as statistic to detect individual outliers in PCA.
+Note the square root, as it otherwise detects as outlier any point that is next to a very dense cluster.
 We implement (the two parts of) this statistic in function \texttt{prob\_dist} of R package bigutilsr.
@@
 
-To project PCs of a reference dataset (e.g.\ 1000G) to a target genotype dataset, we implement the following 3 steps in function \texttt{bed\_projectPCA} of package bigsnpr: 1) matching the variants of each dataset, including removing ambiguous alleles [A/T] and [C/G], and matching strand and direction of the alleles; 2) computing PCA of the reference dataset using the matched variants only; 3) projecting computed PCs to the target data using an optimised implementation (see Supplementary Materials) of the Online Augmentation, Decomposition, and Procrustes (OADP) transformation \cite[]{zhang2019fast}.
-To project individuals from the same dataset as the ones used for computing PCA, you can use function \texttt{bed\_projectSelfPCA}. 
+To project a target genotype dataset to the PCA space from a reference dataset (e.g.\ the 1000 genomes data), we implement the following 3 steps in function \texttt{bed\_projectPCA} of package bigsnpr: 1) matching the variants of each dataset, including removing ambiguous alleles [A/T] and [C/G], and matching strand and direction of the alleles; 2) computing PCA of the reference dataset using the matched variants only; 3) projecting computed PCs to the target data using an optimised implementation (see Supplementary Materials) of the Online Augmentation, Decomposition, and Procrustes (OADP) transformation \cite[]{zhang2019fast}.
+To project individuals from the same dataset as the individuals used for computing PCA, we provide function \texttt{bed\_projectSelfPCA}. 
 Note that the new individuals to be projected should not be related to the ones used for computing PCA (Section \ref{proj-related}).
@@
 Resulting PLINK files and R code to generate these files are made available at \url{https://doi.org/10.6084/m9.figshare.9208979.v3}. 
-To easily download these data, we provide function \texttt{download\_1000G} in R package bigsnpr.
+To easily download this data, we provide function \texttt{download\_1000G} in R package bigsnpr.
 
@@
 \caption{Principal Component (PC) loadings 1 to 40 reported by the UK Biobank.
-X-axis represent positions of variants in the data, and y-axis the value of loadings.
-Points are hex-binned to make plotting of such large number of points manageable.
+Column indices of variants in the data, ordered by chromosome and physical position, are represented on the x-axis, and the value of loadings are represented on the y-axis.
+Points are hex-binned.
 \label{fig:UKBB-loadings40}}
@@
 It takes 12 minutes more to do the OADP projection of all 488,371 individuals of the UKBB data onto the PCA computed using the 1000G data.
-Finally, it takes only 6 minutes to compute the 30-Nearest Neighbours of 20 PC scores for 406,545 UK Biobank individuals, which is the main computational aspect of computing the statistics used to detect individual outlier samples (Section \ref{outlier-sample}).
+Finally, it takes only 6 minutes to compute the 30-Nearest Neighbours of 20 PC scores for 406,545 UK Biobank individuals, which is the most computationally demanding step when computing the statistics used to detect individual outlier samples (Section \ref{outlier-sample}).
 
@@
 
-To detect a few sample outliers, we compare the standard rule of ``6 SDs from the mean'' (6SD) used in e.g.\ EIGENSOFT to the statistic we propose in section \ref{outlier-sample}.
+To detect a few outlier samples, we compare the standard rule of ``6 SDs from the mean'' (6SD) used in e.g.\ EIGENSOFT to the statistic we propose in section \ref{outlier-sample}.
 Our statistic identifies only isolated samples or isolated pairs that seems to be outliers driving structure of PC17-20 of 1000G (Figure \ref{fig:outlier-pd}). 
-All but one outliers are distantly-related pairs that disappear if using a more stringent threshold on relatedness (i.e.\ using a KING-relatedness cutoff of \textasciitilde0.0442 instead of \textasciitilde0.0884, see tutorial in section ``code availability'').
+All but one outlier are distantly-related pairs that disappear if using a more stringent threshold on relatedness (i.e.\ using a KING-relatedness cutoff of \textasciitilde0.0442 instead of \textasciitilde0.0884, see tutorial in section ``code availability'').
 In contrast, rule 6SD identifies a lot of outliers, of which some are part of a relatively large cluster (Figure \ref{fig:outlier-sd}).
-We recall that, in theory, if all PCs are normally distributed, after correcting for multiple testing of 2500 individuals and 20 PCs, 6SD has a small probability of 0.0001 of detecting one outlier or more.
+We recall that, if all PCs are normally distributed, after correcting for multiple testing of 2500 individuals and 20 PCs, the probability of detecting one outlier or more using 6SD is only of 0.0001.
 
@@
 \includegraphics[width=\textwidth]{hist-outliers-1000G.pdf}
-\caption{Distribution of statistics (S)}
+\caption{Distribution of statistics (S).}
 \end{subfigure}
@@
 
-As for restricting to homogeneous samples, we compare the use of R package aberrant used to report the ``White British'' subset of UKBB homogeneous samples, to the use of the robust Mahalanobis distance we propose.
-Here, we visually choose a threshold of 5 on the log-distance and show that this gives a similar subset of individuals than the ``White British'' subset reported by the UK Biobank (Figure \ref{fig:homogeneous}).
+As for restricting to homogeneous samples, we compare the use of R package aberrant, which was used to report the homogeneous ``White British'' subset in UKBB, to the use of the robust Mahalanobis distance we propose here.
+We visually choose a threshold of 5 on the log-distance and show that this gives a similar subset of individuals than the ``White British'' subset reported by the UK Biobank (Figure \ref{fig:homogeneous}).
 Moreover, when using this threshold, only 3 out of 10,936 people of self-reported Asian ancestry (1 ``Chinese'' and 2 ``Indian'') are kept, and 1 ``African'' out of 7622 people with Black background is kept (Table \ref{tab:homogeneous}).
@@
 Finally, to assess if OADP could be used to project individuals that are related to some individuals that were used to compute PCA, we projected these 60\% individuals (as if we were projecting their monozygotic twins) using OADP.
-Projections of related individuals using OADP shows some bias in reverse direction (Figure \ref{fig:proj1000G-4}).
+Projections of related individuals using OADP suffers from some bias in reverse direction (Figure \ref{fig:proj1000G-4}).
 \label{proj-related}
@@
 When computing the PCs on the UK Biobank using 406,545 unrelated individuals and 171,977 variants, and projecting the 1000G data onto this reference PCA space, shrinkage is much smaller ($\leq 1.08$ for all 20 first PCs, Figure \ref{fig:proj1000G-6}).
-Overall, this shrinkage decreases with an increased sample size (Table \ref{tab:shrinkage}).
+Overall, this shrinkage for simple projection decreases with an increased sample size (Table \ref{tab:shrinkage}).
 
@@
 We show that we are able to capture more PCs (at least 40 instead of 16-18) that display visual population structure (Figures \ref{fig:UKBB-scores2} and \ref{fig:UKBB-screeplot2}).
-We then projected all 439,429 remaining individuals from UKBB onto this PCA space in 21 minutes only using our implementation of the OADP projection (\texttt{bed\_projectSelfPCA}).
-Note that these individuals should not be related to any of the 48,942 individuals used for training PCA because we removed both individuals from each pair of related individuals in UKBB.
+We then projected all 439,429 remaining individuals from UKBB onto this PCA space in 21 minutes only using our implementation of the OADP projection (function \texttt{bed\_projectSelfPCA}).
+Note that these individuals should not be related to any of the 48,942 individuals used for training PCA because we removed both individuals from each pair of related individuals in the UKBB.
 Projection of new individuals show again a clear shrinkage when using simple projection (between 1.00 for PC1 and 1.80 for PC50), but no visible bias when using OADP projection (Figure \ref{fig:projUKBB-related}).
@@
 
-\subsection{Imputation of missing values}
+\subsection{PCA \& missing value imputation}
 
 As we compute PCA on data with missing values, although we restrict to variants with less than 1\% missing values, we analyse hereinafter the effect of imputation of missing values before computing PCA.
-We compare 4 different imputation methods on two different sets of individuals, either all genotyped individuals or all imputed individuals.
-The UK Biobank imputed data is composed of approximately 1000 individuals less who have been removed because of a high number of missing values or a high heterozygosity \cite[]{bycroft2017genome}.
-When computing PCA with mean imputation and using all individuals, PC16 is capturing individuals with very high heterozygosity (Figure \ref{fig:UKBB-het}).
+We compare 4 different imputation methods and two different sets of individuals.
+In the UK Biobank imputed data, approximately 1000 individuals have been removed because of a high number of missing values or a high heterozygosity, as compared to the genotyped data \cite[]{bycroft2017genome}.
+When computing PCA with mean imputation and using all genotyped individuals, PC16 captures individuals with very high heterozygosity (Figure \ref{fig:UKBB-het}).
 When restricting to imputed individuals only, i.e.\ after removing individuals with very high heterozygosity, PC16 completely disappears and new PC16 to PC19 correspond to previous PC17 to PC20 (Table \ref{tab:pca-het}).
 When using dosage data instead of genotype data with mean imputation, PCA is globally unchanged (Table \ref{tab:pca-dosage}). 
-Overall, if we choose to use either one of the following 4 imputation methods: mean imputation, random imputation according to allele frequencies, using reported dosage data from BGEN files, or imputation of genotyped data based on machine learning using function \texttt{snp\_fastImpute} of R package bigsnpr \cite[]{prive2017efficient}, resulting PCs are always very similar (absolute correlation larger than 0.99 for the 20 computed PCs, results not shown).
-This justifies performing PCA with mean imputation directly on PLINK bed files with a few missing values; this has the advantage to be much faster than having to impute data using \texttt{snp\_fastImpute}, which took 4 days for 406,545 individuals and 240,444 variants, or based on external reference datasets.
+Overall, if we choose to use either one of the following 4 imputation methods: mean imputation, random imputation according to allele frequencies, using reported dosage data from BGEN files, or imputation of genotyped data based on machine learning using function \texttt{snp\_fastImpute} of R package bigsnpr \cite[]{prive2017efficient}, resulting PCs are always very similar (absolute correlation larger than 0.99 for the 20 computed PCs; results partially shown in tables \ref{tab:pca-het} and \ref{tab:pca-dosage}).
+This justifies performing PCA with mean imputation directly on PLINK bed files with a few missing values; this has the advantage to be much faster than having to impute genotyped data using \texttt{snp\_fastImpute}, which took 4 days for 406,545 individuals and 240,444 variants, or based on external reference datasets.
 
@@
 The new functions we provide in R package bigsnpr can be directly applied to genotypes stored as PLINK bed/bim/fam files with some missing values.
-This contrasts with previous releases of bigsnpr that could only use format ``bigSNP''.
+This contrasts with previous releases of package bigsnpr that could only use format ``bigSNP''.
 This data format can store both genotype calls and dosages, but requires conversion from other formats and imputation of missing values using functions provided in the package \cite[]{prive2017efficient}.
@@
 After selecting a subset of individuals, we apply several steps of outlier detection, one for outlier variants that capture long-range LD variation (automatic), and one for detecting outlier samples (semi-automatic and visual).
-To make these steps more computationally efficient, we explored solutions for not recomputing PCA from scratch when removing a few samples or variants.
-Using educated guesses in R package PRIMME based on low-rank approximations seemed to be a promising approach but did not reduce computation time by much, so we did not pursue this idea \cite[]{brand2003fast,wu2017primme}.
+To make these steps more computationally efficient, we explored solutions for not recomputing PCA from scratch when removing a few samples or a few variants.
+Using educated guesses in R package PRIMME based on low-rank approximations of the updated PCA seemed to be a promising approach but did not reduce computation time by much, so we did not pursue this idea \cite[]{brand2003fast,wu2017primme}.
 
@@
 We differentiate PCs in 3 broad types: the ones capturing LD structure, the ones capturing population structure, and noise.
-We expect population structure to be well distributed along the genome so that we get loadings that are normally distributed around 0 (with small effect sizes).
-In contrast, long-range LD structure is basically capturing the variation inside one long-range LD region (so localised in the genome), so that we expect the loadings to be very large in that region only (one peak).
+We expect population structure to be evenly distributed along the genome so that loadings are normally distributed around 0 (with small effect sizes).
+In contrast, long-range LD structure is essentially capturing the variation inside one long-range LD region (so localised in the genome), so that we expect the loadings to be very large in that region only (one peak).
 Therefore, PCs capturing LD structure can be identified by looking for peaks in PC loadings (e.g.\ PC17-20 in figure \ref{fig:UKBB-loadings}). 
@@
 As in many applications, we believe a compromise between signal and noise should be preferred.
-Therefore, we recommend using only PCs that show structure (e.g.\ PC1-14, and probably also PC15-16, in figure \ref{fig:UKBB-scores}) and excluding PCs that does not seem to capture any population structure (e.g.\ PC17-20 in figure \ref{fig:UKBB-scores}).
+Therefore, we recommend using only PCs that show structure (e.g.\ PC1-16 in figure \ref{fig:UKBB-scores}) and excluding PCs that do not seem to capture any population structure (e.g.\ PC17-20 in figure \ref{fig:UKBB-scores}).
 
@@
 Likewise, previous works have shown that uneven population sizes can distort PCs \cite[]{novembre2008interpreting,mcvean2009genealogical}. 
-Indeed, when subsetting British and Irish people, we are able to capture a lot more PCs that show population structure with less than 50K individuals as compared to when using more than 400K individuals who are mostly composed of British and Irish people.
+Indeed, when subsetting British and Irish people in the UK Biobank data, we are able to capture a lot more PCs that show population structure with less than 50K individuals compared to when using more than 400K individuals who are mostly composed of British and Irish people.
 Determining how much overdominant populations should be subsampled to maximise population structure captured by PCA is a direction of future work.
@@
 Finally, using ``populations'' with approximately the same size is relevant for other biological data as well.
-However, other pitfalls might apply when using other types of data; for example, methylation data can be confounded by factors such as age and sex, and it might be beneficial to remove some methylation probes that are associated with these confounding factors before computing PCA \cite[]{decamps2019guidelines}.
+However, other pitfalls might apply when using other types of data; for example, methylation data can be confounded by factors such as age and sex, and it might be beneficial to remove the methylation probes that are associated with these confounding factors before computing PCA \cite[]{decamps2019guidelines}.
 
@@
 We implement an optimised version of the Online Augmentation, Decomposition, and Procrustes (OADP) transformation when using $K''=K'=K$ \cite[]{zhang2019fast}.
-We assume that the $K$-partial Singular Value Decomposition (SVD) of the reference matrix $X$ (of size $n \times p$) has been computed as $U \Delta V^T$.
+We assume that the $K$-partial Singular Value Decomposition (SVD) of the scaled reference matrix $X$ (of size $n \times p$) has been computed as $U \Delta V^T$.
 There are several steps to perform OADP transformation for each sample $y$ (of size $1 \times p$) of target matrix $Y$ (of size $m \times p$):
@@
 \caption{Principal Component (PC) loadings 1 to 20 computed on the UK Biobank using \texttt{bed\_autoSVD}.
-X-axis represent positions of variants in the data, and y-axis the value of loadings.
-Points are hex-binned to make plotting of such large number of points manageable.
+Column indices of variants in the data, ordered by chromosome and physical position, are represented on the x-axis, and the value of loadings are represented on the y-axis.
+Points are hex-binned.
 \label{fig:UKBB-loadings}}

### 2020-02-18 paper-bedpca.tex
minor corrections

@@
 
-As for restricting to homogeneous samples, we compare the use of R package aberrant, which was used to report the homogeneous ``White British'' subset in UKBB, to the use of the robust Mahalanobis distance we propose here.
+As for restricting to homogeneous samples, we compare the use of the robust Mahalanobis distance we propose here to the use of R package aberrant, which was used to report the homogeneous ``White British'' subset in the UKBB \cite[]{bellenguez2011robust,bycroft2017genome}.
 We visually choose a threshold of 5 on the log-distance and show that this gives a similar subset of individuals than the ``White British'' subset reported by the UK Biobank (Figure \ref{fig:homogeneous}).
@@
 In contrast, 416,492 out of 431,090 ``British'' (96.6\%) and 12,620 out of 12,759 ``Irish'' (98.9\%) are kept.
+Results are very similar to the set of ``White British'' made using R package aberrant (Figure \ref{fig:homogeneous}).
 
@@
 All code used for this paper is available at \url{https://github.com/privefl/paper4-bedpca/tree/master/code}.
-Package bigsnpr can be installed from GitHub (see \url{https://github.com/privefl/bigsnpr}).
+R packages bigsnpr and bigutilsr can be installed from either CRAN or GitHub (see \url{https://github.com/privefl/bigsnpr}).
 A tutorial on the steps to perform PCA on 1000G data is available at \url{https://privefl.github.io/bigsnpr/articles/bedpca.html}.

### 2020-02-18 code/homogeneous-samples.R

@@ -21,7 +21,7 @@ ind_na <- attr(PC, "na.action")
 pop_UKBB <- df0$pop[-ind_na]
 
 
-dist <- bigutilsr::covRob(PC, estim = "pairwiseGK")$dist
+dist <- bigutilsr::dist_ogk(PC)
 
 ggplot() +
   geom_histogram(aes(log(dist)), color = "#000000", fill = "#000000", alpha = 0.5) +

### 2020-02-20 paper-bedpca-supp.tex
split main and supp

--- 2020-02-18 paper-bedpca.tex
@@
 
-\title{Efficient toolkit implementing best practices for principal component analysis of population genetic data}
-\author{Florian Priv\'e,$^{\text{1,5,}*}$ Keurcien Luu$^{\text{5}}$, Michael G.B. Blum,$^{\text{4,5}}$ John J. McGrath,$^{\text{1,2,3}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,}*}$}
-
-\date{~ }
-\maketitle
-
-\noindent$^{\text{\sf 1}}$National Centre for Register-Based Research, Aarhus University, Aarhus, 8210, Denmark. \\
-\noindent$^{\text{\sf 2}}$Queensland Brain Institute, University of Queensland, St. Lucia, 4072, Queensland, Australia. \\
-\noindent$^{\text{\sf 3}}$Queensland Centre for Mental Health Research, The Park Centre for Mental Health, Wacol, 4076, Queensland, Australia. \\
-\noindent$^{\text{\sf 4}}$OWKIN France, Paris, 75010, France. \\
-\noindent$^{\text{\sf 5}}$Laboratoire TIMC-IMAG, UMR 5525, Univ.\ Grenoble Alpes, La Tronche, 38700, France. \\
-\noindent$^\ast$To whom correspondence should be addressed.\\
-
-\noindent Contacts:
-\begin{itemize}
-\item \url{florian.prive.21@gmail.com}
-\item \url{bjv@econ.au.dk}
-\end{itemize}
-
-\clearpage
-
-\abstract{	
-Principal Component Analysis (PCA) of genetic data is routinely used to infer ancestry and control for population structure in various genetic analyses.
-However, conducting PCA analyses can be complicated and has several potential pitfalls.
-These pitfalls include (1) capturing Linkage Disequilibrium (LD) structure instead of population structure, (2) projected PCs that suffer from shrinkage bias, (3) detecting sample outliers, and (4) uneven population sizes.
-In this work, we explore these potential issues when using PCA, and present efficient solutions to these.
-Following applications to the UK Biobank and the 1000 Genomes project datasets, we make recommendations for best practices and provide efficient and user-friendly implementations of the proposed solutions in R packages bigsnpr and bigutilsr.
-
-For example, we find that PC19 to PC40 in the UK Biobank capture complex LD structure rather than population structure.
-Using our automatic algorithm for removing long-range LD regions, we recover 16 PCs that capture population structure only.
-Therefore, we recommend using only 16-18 PCs from the UK Biobank to account for population structure confounding.
-We also show how to use PCA to restrict analyses to individuals of homogeneous ancestry.
-Finally, when projecting individual genotypes onto the PCA computed from the 1000 Genomes project data, we find a shrinkage bias that becomes large for PC5 and beyond.
-We then demonstrate how to obtain unbiased projections efficiently using bigsnpr.
-
-Overall, we believe this work would be of interest for anyone using PCA in their analyses of genetic data, as well as for other omics data.
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\clearpage
-
-\section{Introduction}
-
-Principal Component Analysis (PCA) has been widely used in genetics for many years and in many contexts.
-For instance, adding PCs as covariates is routinely used to adjust for population structure in Genome-Wide Association Studies (GWAS) \cite[]{price2006principal,novembre2008interpreting}.
-PCA has also been used to detect loci under selection \cite[]{galinsky2016fast,luu2017pcadapt} and in heritability analyses \cite[]{yang2010common,loh2015contrasting}.
-Recently, the advent of large population-scale genetic datasets, such as the UK biobank data, has prompted research on developing scalable algorithms to compute PCA on very large data \cite[]{bycroft2017genome}.
-It is now possible to efficiently approximate PCA on very large datasets thanks to software such as FastPCA (fast mode of EIGENSOFT), FlashPCA2, PLINK 2.0 (approx mode), bigstatsr/bigsnpr, TeraPCA and ProPCA \cite[]{galinsky2016fast,abraham2017flashpca2,chang2015second,prive2017efficient,bose2019terapca,agrawal2019scalable}.
-
-However, in practice, conducting PCA on genotype data to capture population structure consists of more steps than simply performing singular value decomposition on the genotype matrix.
-These steps includes removing related individuals, pruning variants in LD, and excluding outlier samples that can suggest poor genotyping quality or distant relatedness.
-Indeed, many pitfalls related to PCA of genotype data have been documented and none of the currently available software address all of these.
-In the following, we outline these pitfalls and explain when they are relevant.
-First, some of the PCs may capture LD structure rather than population structure \cite[]{price2008long,zou2010quantification,abdellaoui2013population,prive2017efficient}.
-Including PCs that capture LD as covariates in genetic analyses can lead to reduced power for detecting genetic associations within these LD regions \cite[]{zou2010quantification}.  
-Second, another issue may arise when projecting a new study dataset to the PCA space computed from a reference dataset: projected PCs are shrunk towards 0 in the new dataset \cite[]{lee2010convergence,wang2015improved,zhang2019fast}.
-This shrinkage makes it potentially dangerous to use the projected PCs for analyses such as PC regression, ancestry detection and correction for ancestry.
-This same issue also arises when projecting individuals from the same dataset that were discarded from the PCA computation (e.g.\ related individuals).
-Third, PC scores may capture outliers that are due to family structure, population structure or other reasons; it might be beneficial to detect and remove these individuals to maximise the population structure captured by PCA (in the case of removing a few outliers) or to restrict analyses to genetically homogeneous samples (e.g.\ ``White British'' people in the UK Biobank).
-Finally, efficient methods for PCA use approximations, which can results in some lack of precision of computed PCs.
-This potential issue has been demonstrated for software such as FastPCA and PLINK 2.0, but not for FlashPCA2 and bigstatsr/bigsnpr \cite[]{abraham2017flashpca2,prive2017efficient}.
-An overview of existing methods with their respective advantages and limitations is presented in table \ref{tab:method-overview}.
-
-For this paper, we derive implementations of truncated PCA and other useful functions for e.g.\ performing LD thinning and computing various statistics.
-We make these available in a new release of R package bigsnpr (v1.0.0); what differs from previously available functions presented in \cite{prive2017efficient} is that these new functions can be used directly on PLINK bed/bim/fam files with some missing values.
-We use these new functions to analyse the UK Biobank data, and show that these functions are both very fast and easy to use.
-We also point out that many PCs currently reported by the UK Biobank capture LD structure instead of population structure.
-Interestingly, subsetting the UK Biobank data enables to get more PCs that capture population structure than when using the whole sample (\textasciitilde40 instead of \textasciitilde16).
-Then, we project the other individuals that were not used in the PCA calculation, show that this projection is biased and provide an efficient solution to get unbiased projections instead.
-Finally, we explore options to detect outlier samples in PCA, either a few outlier samples that may correspond to e.g.\ batch effects or distant family structure, or when the goal is to restrict the data to individuals of homogeneous ancestry.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Material and Methods}
-
-\begingroup
-\renewcommand*{\arraystretch}{2}
-\begin{table}[h]
-\centering
-\caption{Overview of existing methods.} 
-\label{tab:method-overview}
-\begin{tabular}{|c|c|c|c|c|}
-\hline
-Analysis & Method and/or Software & Citation & Advantages & Current limitations \\
-\hline
-\multirow{3}{*}{PCA} & bigstatsr / bigsnpr & \cite[]{prive2017efficient} & \makecell{Fast and accurate + handle dosages \\ + thinning options directly included} & \makecell{Own format without missing values \\ (fast functions are available \\ for converting and imputing)} \\
- & FlashPCA2 & \cite[]{abraham2017flashpca2} & Fast and accurate & Not parallelised \\
- & PLINK 2.0 (reimplementation of FastPCA) & \cite[]{galinsky2016fast,chang2015second} & Fast & \makecell{Possible lack of accuracy \\ \cite[]{abraham2017flashpca2,prive2017efficient}} \\
-\hline
-\makecell{Detection of \\ oultier samples} & \makecell{``6 SDs from the mean'' \\ in EIGENSOFT} & \cite[]{patterson2006population} & Simple & Assumes a Gaussian distritbution \\
-\hline
-\makecell{Detection of \\ homogeneous \\ samples} & R package aberrant & \cite[]{bellenguez2011robust} & Robust & Uses only two statistics at once \\
-\hline
-\multirow{5}{*}{\makecell{Projection of \\ new individuals \\ onto reference \\ PCA space}} & Simple projection (multiplication by loadings) & & Simple & Shrinkage biased \\
-  & Bias-adjusted projection in R package hdpca & \cite[]{dey2019asymptotic} & Independent of new samples & \makecell{Assumes same shrinkage for all individuals \\ + model-based \\ + need all eigenvalues of reference} \\
-  & \makecell{Augmentation, Decomposition and Procrustes (ADP) \\ transformation in LASER 2.0} & \cite[]{wang2015improved} & Accurate & Slow (a new PCA for each new sample) \\
-  & Online ADP (OADP) in python package FRAPOSA & \cite[]{zhang2019fast} & Much faster than ADP & Does not work for related individuals (Section \ref{proj-related}) \\ 
-\hline
-\end{tabular}
-\end{table}
-\endgroup
-
-\subsection{Efficient implementation of PCA for genotype data \label{randomSVD}}
-
-When there is no missing value, we compute the truncated Singular Value Decomposition (SVD) $U \Delta V^T$ of the scaled genotype matrix of diploid individuals $\tilde{G}_{i,j} = \frac{G_{i,j} - 2 \hat{f_j}}{\sqrt{2 \hat{f_j} (1 - \hat{f_j})}}$ where $G_{i,j}$ is the allele count (genotype) of individual $i$ and variant $j$, and $\hat{f_j}$ is the estimated allele frequency of variant $j$ ($2 \hat{f_j}$ is the mean allele count of variant $j$).
-Then, $U \Delta$ are the first $K$ PC scores and $V$ are the first $K$ PC loadings, where $K$ is the number of PCs computed (e.g.\ $K=20$).
-
-When there are some missing values, we compute the partial SVD similarly, except that missing values are replaced by the variant means (i.e.\ $G_{i,j} - 2 \hat{f_j} = 0$ when $G_{i,j}$ is missing) and  the $\hat{f_j}$ for each variant are estimated using only non-missing genotypes.
-Note that this decomposition is equivalent to the decomposition presented above after imputation by the variant means.
-
-To compute this decomposition easily and efficiently, we implement an accessor that memory-map the PLINK bed file to use it directly as if it were a standard matrix.
-Then, we apply the same algorithm for partial SVD that is used in R packages bigstatsr and FlashPCA2, namely the implicitly restarted Arnoldi method \cite[]{Lehoucq1996,abraham2017flashpca2,prive2017efficient}.
-This algorithm, implemented in R package RSpectra, requires a function that computes the matrix-vector multiplication of the scaled genotype matrix with a given vector.
-We implement such multiplication in parallel from a PLINK bed file.
-
-\subsection{Robust Mahalanobis distance \label{maha}}
-
-Mahalanobis distances are computed as $d(x)^2 = (x-\mu)^T \Sigma^{-1} (x-\mu),$ where $\mu$ and $\Sigma$ are (robust) estimators of location and covariance.
-We use these distances for many applications in this paper. 
-When $x$ is multivariate Gaussian data with $K$ dimensions, the squared distances follow a $\chi^2(K)$ distribution.
-If $x$ represents PC scores of centered data and if we use standard estimates, then $\mu = 0$ and $\Sigma = I_K$.
-Yet, here we use the pairwise orthogonalised Gnanadesikan-Kettenrin robust estimates of these parameters \cite[]{gnanadesikan1972robust,yohai1988high,maronna2002robust}.
-We implement the estimation of these robust parameters in function \texttt{covrob\_ogk} of R package bigutilsr.
-
-\subsection{Detecting LD structure in PCA}
-
-For detecting outlier variants in PCA that are due to long-range Linkage Disequilibrium (LD) regions, we use a similar procedure as described by \cite{prive2017efficient}.
-We first apply a first round of clumping at e.g.\ $r^2$ > 0.2, prioritising variants by higher minor allele count.
-Then, we compute $K$ PC scores and loadings (Section \ref{randomSVD}).
-To summarise the contribution of each variant in all $K$ PC loadings, we compute the robust Mahalanobis distances of these PC loadings (Section \ref{maha}).  
-To capture consecutive outliers that corresponds to long-range LD regions, we apply a Gaussian smoothing to these statistics (moving average with a Gaussian filter over a window with a radius of 50 variants by default).
-
-Finally, to choose the threshold on the previously described statistics above which variants are considered outliers, we use a modified version of Tukey's rule, a standard rule for detecting outliers \cite[]{tukey77}.
-The standard upper limit defined by Tukey's rule is $q_{75\%}(x) + 1.5 \cdot IQR(x)$, where $x$ is the vector of computed statistics and $IQR(x) = q_{75\%}(x) - q_{25\%}(x)$ is the interquartile range.
-One assumption of Tukey's rule is that the sample is normally distributed; we account for skewness in the data using the medcouple as implemented in function \texttt{adjboxStats} of R package robustbase \cite[]{brys2004robust,hubert2008adjusted}.
-Standard Tukey's rule also uses a fixed coefficient (1.5) that does not account for multiple testing, which means that there are always some outliers detected when using 1.5 for large samples. 
-To solve these two potential issues, we implement \texttt{tukey\_mc\_up} in R package bigutilsr and use it here, which accounts for both skewness and multiple testing by default.
-
-We remove the detected outlier variants, compute the PC scores and loadings again, and iterate until there is no detected outlier variant anymore.
-This procedure is implemented in function \texttt{bed\_autoSVD} of R package bigsnpr.
-
-\subsection{Detecting outlier samples in PCA \label{outlier-sample}}
-
-For detecting outlier samples in PCA, we use a modified version of the Probabilistic Local Outlier Factor (PLOF) statistic on PCs \cite[]{kriegel2009loop}.
-Using K nearest neighbours (KNN), this consists in comparing the distance from a point $j$ to its KNNs ($pd_j = \sqrt{\frac{1}{K} \sum_{k=1}^K d^2_{j \leftrightarrow j_k}}$, where $j_k$ is the $k$-th NN of $j$) with the distances from its KNNs to their respective KNNs ($\frac{1}{K} \sum_{k=1}^K pd_{j_k}$).
-Intuitively, an outlier should be far from all other points, and is even more outlier if its KNNs are in a very dense cluster.
-Here, we use $pd_j / \sqrt{\frac{1}{K} \sum_{k=1}^K pd_{j_k}}$ as statistic to detect individual outliers in PCA.
-Note the square root, as it otherwise detects as outlier any point that is next to a very dense cluster.
-We implement (the two parts of) this statistic in function \texttt{prob\_dist} of R package bigutilsr.
-To make it fast, we use the fast K nearest neighbours implementation of R package nabor \cite[]{elseberg2012comparison} and parallelise it in function \texttt{knn\_parallel} of package bigutilsr.
-Automatic outlier detection is difficult, therefore we recommend users to choose a threshold for this statistic to define outliers based on visual inspection (using the histogram of these statistics and the PC scores colored by these statistics; see e.g.\ figure \ref{fig:outlier-pd}).
-
-As for detecting samples that have a different ancestry from most of the samples in the data, i.e.\ for restricting to homogeneous samples, we compute the pairwise orthogonalised Gnanadesikan-Kettenrin robust Mahalanobis distances on PC scores (Section \ref{maha}).
-We then restrict to individuals whose log-distance (alternatively p-value) is smaller (larger) than some threshold determined based on visual inspection.
-
-\subsection{Projecting PCs from a reference dataset \label{proj}}
-
-To project a target genotype dataset to the PCA space from a reference dataset (e.g.\ the 1000 genomes data), we implement the following 3 steps in function \texttt{bed\_projectPCA} of package bigsnpr: 1) matching the variants of each dataset, including removing ambiguous alleles [A/T] and [C/G], and matching strand and direction of the alleles; 2) computing PCA of the reference dataset using the matched variants only; 3) projecting computed PCs to the target data using an optimised implementation (see Supplementary Materials) of the Online Augmentation, Decomposition, and Procrustes (OADP) transformation \cite[]{zhang2019fast}.
-To project individuals from the same dataset as the individuals used for computing PCA, we provide function \texttt{bed\_projectSelfPCA}. 
-Note that the new individuals to be projected should not be related to the ones used for computing PCA (Section \ref{proj-related}).
-
-\subsection{Data\label{1000G}}
-
-We provide and use a subsetted version of the 1000 genomes (1000G) project data \cite[]{10002015global,meyer2019genotype}.
-Variants are restricted to the ones in common with HapMap3 or UK Biobank \cite[]{international2010integrating,bycroft2017genome}. 
-Moreover, we apply some quality control filters; we remove variants having a minor allele frequency < 0.01, variants with P-value of the Hardy-Weinberg exact test < $10^{-50}$, and non-autosomal variants.
-To remove related individuals with second-degree relationship or more, we apply KING-relatedness cutoff of 0.0884 to the data using PLINK 2.0 \cite[]{manichaikul2010robust,chang2015second}.
-This results in 2490 individuals and 1,664,852 variants of the 1000G project (phase 3) in PLINK bed/bim/fam format. 
-Resulting PLINK files and R code to generate these files are made available at \url{https://doi.org/10.6084/m9.figshare.9208979.v3}. 
-To easily download this data, we provide function \texttt{download\_1000G} in R package bigsnpr.
-
-In this paper, we also analyse the UK Biobank data (\url{https://www.ukbiobank.ac.uk/}).
-We apply some quality control filters; we remove individuals with more than 10\% missing values, variants with more than 1\% missing values, variants having a minor allele frequency < 0.01, variants with P-value of the Hardy-Weinberg exact test < $10^{-50}$, and non-autosomal variants.
-This results in 488,371 individuals and 504,139 variants.
-When removing related individuals, we use the list of individual pairs reported by the UK Biobank.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Results}
-
-\subsection{Application to the UK Biobank}
-
-To demonstrate that we provide very fast implementations of the different methods presented in this paper, we apply them to the UK Biobank.
-We use 20 physical cores for most of the computations (CPU: Intel(R) Xeon(R) Silver 4114, 2.20GHz).
-It takes 22 minutes to perform a first phase of clumping on 406,545 unrelated individuals genotyped over 504,139 variants, which reduces the number of variants to 261,307.
-It then takes 34 minutes to compute the first 20 PCs using these 261,307 variants.
-When performing the automatic procedure for LD detection, it takes 5 hours to perform the initial clumping step, 6 rounds of computation of PCs and 5 rounds of outlier variant detection (i.e.\ 5 iterations of outlier detection and one final computation of PCs).
-
-When applying our automatic procedure to remove long-range LD regions, it does not converge after 5 iterations for the UK Biobank, meaning that it keeps detecting long-range LD regions at each iteration (represented by peaks in PC loadings).
-Therefore, we are able to capture only 16 PCs that show stratification that is not LD structure (Figures \ref{fig:UKBB-screeplot}, \ref{fig:UKBB-scores} and \ref{fig:UKBB-loadings}).
-Similarly, PC loadings reported by the UK Biobank clearly show that PC19 to PC40 capture LD structure, which is also the case for PC16 and PC18, although less pronounced (see peaks in figure \ref{fig:UKBB-loadings40}).
-These include e.g.\ one region on chromosome 6 (70-91 Mbp) that is captured in PC19 (Figure \ref{fig:UKBB-loadings}) and that was not previously reported in \cite{price2008long}.
-
-\begin{figure}[!htpb]
-\centerline{\includegraphics[width=0.9\textwidth]{UKBB-loadings1-40.pdf}}
-\caption{Principal Component (PC) loadings 1 to 40 reported by the UK Biobank.
-Column indices of variants in the data, ordered by chromosome and physical position, are represented on the x-axis, and the value of loadings are represented on the y-axis.
-Points are hex-binned.
-\label{fig:UKBB-loadings40}}
-\end{figure}
-
-As for other analyses, it takes 8 minutes to match the 1000G data to the UKBB data and compute 20 PCs of the 1000G data using the automatic LD detection technique.
-It takes 12 minutes more to do the OADP projection of all 488,371 individuals of the UKBB data onto the PCA computed using the 1000G data.
-Finally, it takes only 6 minutes to compute the 30-Nearest Neighbours of 20 PC scores for 406,545 UK Biobank individuals, which is the most computationally demanding step when computing the statistics used to detect individual outlier samples (Section \ref{outlier-sample}).
-
-\subsection{Outlier sample detection}
-
-To detect a few outlier samples, we compare the standard rule of ``6 SDs from the mean'' (6SD) used in e.g.\ EIGENSOFT to the statistic we propose in section \ref{outlier-sample}.
-Our statistic identifies only isolated samples or isolated pairs that seems to be outliers driving structure of PC17-20 of 1000G (Figure \ref{fig:outlier-pd}). 
-All but one outlier are distantly-related pairs that disappear if using a more stringent threshold on relatedness (i.e.\ using a KING-relatedness cutoff of \textasciitilde0.0442 instead of \textasciitilde0.0884, see tutorial in section ``code availability'').
-In contrast, rule 6SD identifies a lot of outliers, of which some are part of a relatively large cluster (Figure \ref{fig:outlier-sd}).
-We recall that, if all PCs are normally distributed, after correcting for multiple testing of 2500 individuals and 20 PCs, the probability of detecting one outlier or more using 6SD is only of 0.0001.
-
-\begin{figure}[htbp]
-\centering
-\begin{subfigure}[b]{0.55\textwidth}
-\includegraphics[width=\textwidth]{hist-outliers-1000G.pdf}
-\caption{Distribution of statistics (S).}
-\end{subfigure}
-\\~\\
-\begin{subfigure}[b]{0.45\textwidth}
-\includegraphics[width=\textwidth]{outliers-1000G.pdf}
-\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by the statistic (S) used to define outliers.
-A few points with higher values for this statistic S appear as outliers in PC17-20.}
-\end{subfigure}
-~~~~~~
-\begin{subfigure}[b]{0.45\textwidth}
-\includegraphics[width=\textwidth]{be-outlier-1000G.pdf}
-\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by being detected as an outlier. Threshold of being an outlier is determined based on histogram (a).}
-\end{subfigure}
-\caption{Outlier detection in the 1000 Genomes (1000G) project, using \texttt{prob\_dist} (Section \ref{outlier-sample}).\label{fig:outlier-pd}}
-\end{figure}
-
-As for restricting to homogeneous samples, we compare the use of R package aberrant, which was used to report the homogeneous ``White British'' subset in UKBB, to the use of the robust Mahalanobis distance we propose here.
-We visually choose a threshold of 5 on the log-distance and show that this gives a similar subset of individuals than the ``White British'' subset reported by the UK Biobank (Figure \ref{fig:homogeneous}).
-Moreover, when using this threshold, only 3 out of 10,936 people of self-reported Asian ancestry (1 ``Chinese'' and 2 ``Indian'') are kept, and 1 ``African'' out of 7622 people with Black background is kept (Table \ref{tab:homogeneous}).
-In contrast, 416,492 out of 431,090 ``British'' (96.6\%) and 12,620 out of 12,759 ``Irish'' (98.9\%) are kept.
-
-\subsection{Projecting onto the PCA space from a reference dataset}
-
-We use 60\% of individuals in the 1000G data (section \ref{1000G}) to compute K=20 PCs.
-Then, we project the remaining 40\% individuals using three methods: 1/ simply multiplying the genotypes of these individuals by the previously computed loadings; 2/ correcting the simple projections using asymptotic shrinkage factors as determined by R package hdpca v1.1.3 \cite[]{dey2019asymptotic}, with all eigenvalues derived from the Genetic Relationship Matrix (GRM) computed with \texttt{bed\_tcrossprodSelf}, one of the new functions of R package bigsnpr; and 3/ the OADP projection (section \ref{proj}).
-When simply projecting using loadings, there is negligible shrinkage for PC1 and PC2, a small shrinkage for PC3 and PC4, and a large shrinkage for PC5 to PC8 (Figure \ref{fig:proj1000G}).
-In contrast, there is no visible shrinkage when projecting new individuals with OADP (Figure \ref{fig:proj1000G}).
-Simple projection is affected even more by this shrinkage for PC9 to PC20, while OADP still appears free of this bias (Figure \ref{fig:proj1000G-2}).
-We show the same results when projecting the full UK Biobank data onto PCA computed using 1000G data  (Figure \ref{fig:proj1000G-5}).
-When correcting projected PC scores with asymptotic shrinkage factors, bias is smaller than with simple projection, yet, there is a visible bias for PC7-8 (Figure \ref{fig:proj1000G-3}).
-Finally, to assess if OADP could be used to project individuals that are related to some individuals that were used to compute PCA, we projected these 60\% individuals (as if we were projecting their monozygotic twins) using OADP.
-Projections of related individuals using OADP suffers from some bias in reverse direction (Figure \ref{fig:proj1000G-4}).
-\label{proj-related}
-
-\begin{figure}[!htpb]
-\centerline{\includegraphics[width=0.8\textwidth]{proj1000G-PC1-8.pdf}}
-\caption{Principal Component (PC) scores 1 to 8 of the 1000 Genomes project.
-Black points are the 60\% individuals used for computing PCA.
-Red points are the 40\% remaining individuals, projected by simply multiplying their genotypes by the corresponding PC loadings.
-Blue points are the 40\% remaining individuals, projected using the Online Augmentation, Decomposition, and Procrustes (OADP) transformation.
-Estimated shrinkage coefficients for these 8 PCs are 1.01 (PC1), 1.02, 1.06, 1.09, 1.50 (PC5), 1.69, 1.98 and 1.39.
-\label{fig:proj1000G}}
-\end{figure}
-
-When computing the PCs on the UK Biobank using 406,545 unrelated individuals and 171,977 variants, and projecting the 1000G data onto this reference PCA space, shrinkage is much smaller ($\leq 1.08$ for all 20 first PCs, Figure \ref{fig:proj1000G-6}).
-Overall, this shrinkage for simple projection decreases with an increased sample size (Table \ref{tab:shrinkage}).
-
-\begingroup
-\renewcommand*{\arraystretch}{1.8}
-\begin{table}[h]
-\centering
-\caption{Shrinkage coefficients when projecting new individuals onto reference PCA space.
-We list the dataset, the sample size and number of variants used to compute the final PCA.
-As expected, the shrinkage bias only becomes negligible if the PCA is conducted on large samples.}
-\label{tab:shrinkage}
-\begin{tabular}{|c|c|c|c|}
-\hline
-Dataset & Sample size ($\times$1000) & \makecell {Number of variants \\ ($\times$1000, after LD removal)} & Shrinkage (PC 1 - 5 - 10 - 20) \\
-\hline
-1000G & ~~~~1.5 & 393 & 1.01 - 1.50 - 3.14 - 6.70 \\
-1000G & ~~~~2.5 & 229 & 1.01 - 1.36 - 2.84 - 6.75 \\
-UKBB & ~~49.0 & 282 & 1.00 - 1.04 - 1.12 - 1.43 \\
-UKBB & 406.5 & 172 & 1.00 - 1.01 - 1.04 - 1.08 \\
-\hline
-\end{tabular}
-\end{table}
-\endgroup
-
-\subsection{Capturing subtle population structure in the UK Biobank}
-
-We recomputed PCA in the UK Biobank after restricting the individuals included in the computations: we randomly subsampled UKBB data to use only 10,000 British individuals (out of 431,029) and 5000 Irish individuals (out of 12,755), while keeping all individuals with other or unknown self-reported ancestry.
-We further removed all pairs of related individuals reported by the UKBB (i.e.\ both individuals in each pair). 
-This resulted in 48,942 individuals that we used to compute 50 PCs, which took less than 3 hours using function \texttt{bed\_autoSVD} (that converged after 4 iterations of automatic LD removal). 
-We show that we are able to capture more PCs (at least 40 instead of 16-18) that display visual population structure (Figures \ref{fig:UKBB-scores2} and \ref{fig:UKBB-screeplot2}).
-We then projected all 439,429 remaining individuals from UKBB onto this PCA space in 21 minutes only using our implementation of the OADP projection (function \texttt{bed\_projectSelfPCA}).
-Note that these individuals should not be related to any of the 48,942 individuals used for training PCA because we removed both individuals from each pair of related individuals in the UKBB.
-Projection of new individuals show again a clear shrinkage when using simple projection (between 1.00 for PC1 and 1.80 for PC50), but no visible bias when using OADP projection (Figure \ref{fig:projUKBB-related}).
-
-\begin{figure}[!htpb]
-\centerline{\includegraphics[width=0.9\textwidth]{UKBB-scores-restricted.png}}
-\caption{Principal Component (PC) scores 27 to 50 computed on the UK Biobank using 48,942 individuals of diverse ancestries.
-These individuals are the ones resulting from removing all related individuals and randomly subsampling the British and Irish individuals.
-Different colors represent different self-reported ancestries.
-\label{fig:UKBB-scores2}}
-\end{figure}
-
-\subsection{PCA \& missing value imputation}
-
-As we compute PCA on data with missing values, although we restrict to variants with less than 1\% missing values, we analyse hereinafter the effect of imputation of missing values before computing PCA.
-We compare 4 different imputation methods and two different sets of individuals.
-In the UK Biobank imputed data, approximately 1000 individuals have been removed because of a high number of missing values or a high heterozygosity, as compared to the genotyped data \cite[]{bycroft2017genome}.
-When computing PCA with mean imputation and using all genotyped individuals, PC16 captures individuals with very high heterozygosity (Figure \ref{fig:UKBB-het}).
-When restricting to imputed individuals only, i.e.\ after removing individuals with very high heterozygosity, PC16 completely disappears and new PC16 to PC19 correspond to previous PC17 to PC20 (Table \ref{tab:pca-het}).
-When using dosage data instead of genotype data with mean imputation, PCA is globally unchanged (Table \ref{tab:pca-dosage}). 
-Overall, if we choose to use either one of the following 4 imputation methods: mean imputation, random imputation according to allele frequencies, using reported dosage data from BGEN files, or imputation of genotyped data based on machine learning using function \texttt{snp\_fastImpute} of R package bigsnpr \cite[]{prive2017efficient}, resulting PCs are always very similar (absolute correlation larger than 0.99 for the 20 computed PCs; results partially shown in tables \ref{tab:pca-het} and \ref{tab:pca-dosage}).
-This justifies performing PCA with mean imputation directly on PLINK bed files with a few missing values; this has the advantage to be much faster than having to impute genotyped data using \texttt{snp\_fastImpute}, which took 4 days for 406,545 individuals and 240,444 variants, or based on external reference datasets.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Discussion}
-
-\begin{figure}[!htpb]
-\centerline{\includegraphics[width=0.4\textwidth]{PCA-pipeline.pdf}}
-\caption{Proposed pipeline for computing Principal Components (PCs) using R packages bigsnpr and bigutilsr.
-\label{fig:pipeline}}
-\end{figure}
-
-In this work, we have compiled different pitfalls that can arise with Principal Component Analysis (PCA) of genetic data.
-Then, we have investigated possible solutions to these pitfalls and selected the ones that we found most advantageous, both with respect to properties such as accuracy and robustness, but also computational efficiency and ease of use.
-We then implemented these solutions in R packages bigsnpr and bigutilsr. 
-The new functions we provide in R package bigsnpr can be directly applied to genotypes stored as PLINK bed/bim/fam files with some missing values.
-This contrasts with previous releases of package bigsnpr that could only use format ``bigSNP''.
-This data format can store both genotype calls and dosages, but requires conversion from other formats and imputation of missing values using functions provided in the package \cite[]{prive2017efficient}.
-As PCA is a useful tool on its own and does not require extensive imputed data, we therefore decided that operating directly on PLINK files with a few missing values would be more practical for users.
-
-We summarise our work into several recommendations for computing PCA, and propose the pipeline shown in figure \ref{fig:pipeline}.
-Note that we have not included standard steps such as initial quality control filters and post-analysis checks (e.g.\ visual inspection of different plots).
-This pipeline requires removing all related individuals, for which we provide an R wrapper to PLINK's implementation of KING robust kinship coefficients \cite[]{manichaikul2010robust,chang2015second}.
-Note that one should remove both individuals in each pair of related individuals.
-This ensures that the projected individuals are not related to the ones used for computing PCA, since we showed that relatedness is a problem when using the OADP projection (Figure \ref{fig:proj1000G-4}).
-After selecting a subset of individuals, we apply several steps of outlier detection, one for outlier variants that capture long-range LD variation (automatic), and one for detecting outlier samples (semi-automatic and visual).
-To make these steps more computationally efficient, we explored solutions for not recomputing PCA from scratch when removing a few samples or a few variants.
-Using educated guesses in R package PRIMME based on low-rank approximations of the updated PCA seemed to be a promising approach but did not reduce computation time by much, so we did not pursue this idea \cite[]{brand2003fast,wu2017primme}.
-
-Once PCA is done, one should check the PC scores (scores of each individual for each PCA dimension) and PC loadings (weights for each variant for each PCA dimension). 
-We differentiate PCs in 3 broad types: the ones capturing LD structure, the ones capturing population structure, and noise.
-We expect population structure to be evenly distributed along the genome so that loadings are normally distributed around 0 (with small effect sizes).
-In contrast, long-range LD structure is essentially capturing the variation inside one long-range LD region (so localised in the genome), so that we expect the loadings to be very large in that region only (one peak).
-Therefore, PCs capturing LD structure can be identified by looking for peaks in PC loadings (e.g.\ PC17-20 in figure \ref{fig:UKBB-loadings}). 
-To identify which PCs capture population structure, and which ones are probably just noise, one should also look at PC scores (colored by ancestry if possible).
-As in many applications, we believe a compromise between signal and noise should be preferred.
-Therefore, we recommend using only PCs that show structure (e.g.\ PC1-16 in figure \ref{fig:UKBB-scores}) and excluding PCs that do not seem to capture any population structure (e.g.\ PC17-20 in figure \ref{fig:UKBB-scores}).
-
-When analysing a dataset that is composed mainly of one population (e.g.\ British people in the UK Biobank), we found that it is useful to subset these individuals to reduce the imbalance between the different population sizes.
-Likewise, previous works have shown that uneven population sizes can distort PCs \cite[]{novembre2008interpreting,mcvean2009genealogical}. 
-Indeed, when subsetting British and Irish people in the UK Biobank data, we are able to capture a lot more PCs that show population structure with less than 50K individuals compared to when using more than 400K individuals who are mostly composed of British and Irish people.
-Determining how much overdominant populations should be subsampled to maximise population structure captured by PCA is a direction of future work.
-The remaining individuals can then be projected onto the resulting PCA space using the OADP projection we recommend in this paper.
-This suggests that designs such as the 1000 Genomes project, which gathered around 100 people for each of 26 different populations, are highly relevant for capturing population structure \cite[]{10002015global}.
-
-In contrast, a common strategy in genetic analyses is to restrict the analysis to a homogeneous sample to reduce risk of confounding due to population stratification.
-For that purpose, we show that using the Mahalanobis distance on PC scores can efficiently achieve this goal, which we used in previous analyses \cite[]{prive2019efficient}.
-When the homogeneous sample is not predominant in the dataset, one solution is to compute the center and covariance of the robust Mahalanobis distance using only the population of interest, and then computing the distances for all individuals using these robust estimates.
-
-The ubiquitous use of PCA in a wide variety of genomic analyses makes it difficult to establish universal guidelines for such analysis. 
-Although we have tackled many problems related to computing PCA on genotype data in this paper, we do not answer other important problems, such as how to best control for population structure in genomic analyses.
-For example, when conducting a GWAS, should one restrict to a homogenous sample, or is it enough to just include PCs that capture population structure as covariates, or should one also use PCs as covariates in mixed linear models \cite[]{price2010new,loh2015efficient}?
-Similarly, in some analyses, it may be beneficial to include PCs that capture long-range or even inter-chromosomal LD.
-More work is needed to understand these fundamental problems, and to provide precise guidelines for conducting successful GWAS, heritability, and other genomic analyses where PCA is used.
-These are directions of future work.
-
-Finally, although we have focused on PCA of genotype data in this paper, we believe most of the results presented here are not inherent to genotype data, and can be transferred to e.g.\ other omics data as well.
-For example, PCs can be used to account for confounding in other data as well \cite[]{pickrell2010understanding}.
-Then, outlier and homogeneous sample detection can be used on PCs of other types of data.
-Moreover, projection of scores will also be a problem for other omics data where the number of variables used is larger than the number of samples used for computing PCA.
-Finally, using ``populations'' with approximately the same size is relevant for other biological data as well.
-However, other pitfalls might apply when using other types of data; for example, methylation data can be confounded by factors such as age and sex, and it might be beneficial to remove the methylation probes that are associated with these confounding factors before computing PCA \cite[]{decamps2019guidelines}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\clearpage
-\vspace*{5em}
-
-\section*{Software and code availability}
-
-All code used for this paper is available at \url{https://github.com/privefl/paper4-bedpca/tree/master/code}.
-Package bigsnpr can be installed from GitHub (see \url{https://github.com/privefl/bigsnpr}).
-A tutorial on the steps to perform PCA on 1000G data is available at \url{https://privefl.github.io/bigsnpr/articles/bedpca.html}.
-
-\section*{Acknowledgements}
-
-Authors thank Rounak Dey and Seunggeun Lee for helpful discussions about PCA projection.
-This research has been conducted using the UK Biobank Resource under Application Number 25589.
-
-F.P., J.M.\ and B.V.\ are supported by the Danish National Research Foundation (Niels Bohr Professorship to J.M.), and also acknowledge the Lundbeck Foundation Initiative for Integrative Psychiatric Research, iPSYCH (R248-2017-2003).
-
-\section*{Declaration of Interests}
-
-Michael Blum is now an employee of OWKIN France.
-The other authors declare no competing interests.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\clearpage
-
-\bibliographystyle{natbib}
-\bibliography{refs}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\clearpage
 \section*{Supplementary Materials}

### 2020-02-20 paper-bedpca.tex
split main and supp

@@
 
-%\usepackage{xr}
-%\externaldocument{SCT-supp}
+\usepackage{xr}
+\externaldocument{paper-bedpca-supp}
 
@@
 
-\clearpage
-\section*{Supplementary Materials}
-
-\renewcommand{\thefigure}{S\arabic{figure}}
-\setcounter{figure}{0}
-\renewcommand{\thetable}{S\arabic{table}}
-\setcounter{table}{0}
-
-\subsection*{Optimised OADP transformation}
-
-We implement an optimised version of the Online Augmentation, Decomposition, and Procrustes (OADP) transformation when using $K''=K'=K$ \cite[]{zhang2019fast}.
-We assume that the $K$-partial Singular Value Decomposition (SVD) of the scaled reference matrix $X$ (of size $n \times p$) has been computed as $U \Delta V^T$.
-There are several steps to perform OADP transformation for each sample $y$ (of size $1 \times p$) of target matrix $Y$ (of size $m \times p$):
-\begin{enumerate}
-\item Calculate $l = y \cdot V$ (of size $1 \times K$), where $V$ are the K PC loadings.
-And $g = y \cdot h^T$ (of size $1 \times 1$), where $h = (y - l \cdot V^T) / ||y - l \cdot V^T||_2$.
-Actually, $||y - l \cdot V^T||_2^2 = y \cdot y^T - 2 \cdot y \cdot V \cdot l^T  + l \cdot V^T \cdot V \cdot l^T = y \cdot y^T - l \cdot l^T$ and $y \cdot (y - l \cdot V^T)^T = y \cdot y^T - y \cdot V \cdot l^T = y \cdot y^T - l \cdot l^T$.
-Then $g = \sqrt{y \cdot y^T - l \cdot l^T}$.
-\item Calculate $Q^T Q$ where $$Q = \begin{bmatrix} \Delta & l^T \\ 0 & g \end{bmatrix}$$ so that $$Q^TQ = \begin{bmatrix} \Delta^2 & \Delta \cdot l^T \\ l \cdot \Delta & g^2 + l \cdot l^T \end{bmatrix} = \begin{bmatrix} \Delta^2 & \Delta \cdot l^T \\ l \cdot \Delta & y \cdot y^T \end{bmatrix}.$$ Note that we do not actually need to compute $g$, and that we can update only the last row and column of $Q^T Q$ instead of computing it from an updated version of $Q$.
-\item Get the eigen decomposition $Q^T Q = V' {\Delta'}^2 {V'}^T$ (truncated to $K$ components out of the $K+1$). Let us denote $V_2 = V' {\Delta'}$.
-\item Calculate $$\begin{bmatrix} \widetilde{U} \\ \widetilde{u}  \end{bmatrix} = \begin{bmatrix} U & 0 \\ 0 & 1 \end{bmatrix} V_2 = \begin{bmatrix} U V_2[1{:}K,~] \\ V_2[K{+}1,~] \end{bmatrix} $$
-\item Find the Procrustes transformation from $\widetilde{U}$ to $U \Delta$.
-As $\widetilde{U}$ and $U \Delta$ have both their columns centered already (since $U$ does), the Procrustes transformation $\rho \widetilde{U} A$, where $\rho$ is a scaling coefficient and $A$ is an orthonomal projection matrix that minimise the Frobenius norm of $(\rho \widetilde{U} A - U)$, is given by $A = V'' {U''}^T$ and $\rho = \frac{\text{trace}(\Delta'')}{\text{trace}(\widetilde{U}^T \widetilde{U})}$ where $U'' \Delta'' {V''}^T$ is the full SVD of $(U \Delta)^T \widetilde{U}$ \cite[]{wang2015improved}.
-As $U^T U = I$, we note that $(U \Delta)^T \widetilde{U} = \Delta V_2[1{:}K,~]$ and that $\rho = \frac{\text{trace}(\Delta'')}{\text{trace}({V_2[1{:}K,~]}^T V_2[1{:}K,~])}$, therefore we do not need to explicitly compute $\widetilde{U}$ and do not need $U$.
-\item Apply the previous transformation to $\widetilde{u}$ to get the projection of $y$ in the reference PCA space (i.e.\ $\rho \widetilde{u} A$).
-\end{enumerate}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\clearpage
-
-\subsection*{Sample outlier detection}
-
-\begin{figure}[htbp]
-\centering
-\begin{subfigure}[b]{0.55\textwidth}
-\includegraphics[width=\textwidth]{hist-outliers2-1000G.pdf}
-\caption{Distribution of statistics (S2) and default threshold (6, in red)}
-\end{subfigure}
-\\~\\
-\begin{subfigure}[b]{0.45\textwidth}
-\includegraphics[width=\textwidth]{outliers2-1000G.pdf}
-\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by statistic used to detect outliers (maximum number of SDs from the mean, log-scale).}
-\end{subfigure}
-~~~~~~
-\begin{subfigure}[b]{0.45\textwidth}
-\includegraphics[width=\textwidth]{be-outlier2-1000G.pdf}
-\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by being detected as an outlier.\\}
-\end{subfigure}
-\caption{Outlier detection in the 1000 Genomes (1000G) project, using the rule ``6 SDs from the mean'', i.e.\ where S2 is the maximum (for all PCs) number of SDs from the mean (Section \ref{outlier-sample}).\label{fig:outlier-sd}}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\clearpage
-
-\begin{figure}[htbp]
-\centering
-\begin{subfigure}[b]{0.45\textwidth}
-\includegraphics[width=\textwidth]{UKBB-White-British.png}
-\caption{PCs 1 to 20 of UKBB, colored by whether it is in the group of ``White British'' reported by the UKBB (blue).}
-\end{subfigure}
-~~~~~~
-\begin{subfigure}[b]{0.5\textwidth}
-\includegraphics[width=\textwidth]{UKBB-Maha-dist.png}
-\caption{PCs 1 to 20 of UKBB, colored by robust Mahalanobis distances computed on PCs (log-scale).\\}
-\end{subfigure}
-\\~\\
-\begin{subfigure}[b]{0.45\textwidth}
-\includegraphics[width=\textwidth]{hist-Maha-dist.pdf}
-\caption{Distribution of (log squared) distances.\\~\\~\\~\\~\\}
-\end{subfigure}
-~~~~~~
-\begin{subfigure}[b]{0.45\textwidth}
-\includegraphics[width=\textwidth]{UKBB-Maha-outlier.png}
-\caption{PCs 1 to 20 of UKBB, colored by being detected as an outlier.
-Threshold of being considered as an outlier is determined based on histogram (c), where the threshold of 5 is chosen for the logarithm of the distances.}
-\end{subfigure}
-\caption{Homogeneous sample detection in the UK Biobank (UKBB), using robust Mahalanobis distances computed on the first 20 Principal Component scores (PCs) of UKBB.).\label{fig:homogeneous}}
-\end{figure}
-
-% latex table generated in R 3.6.0 by xtable 1.8-4 package
-% Sat Nov  9 10:17:25 2019
-\begin{table}[ht]
-\centering
-\caption{Number of UKBB individuals with (log squared) Mahalanobis distance lower than some threshold (top), and grouped by self-reported ancestry (left).
-Note that ``$<$ 12'' includes all individuals.}
-\label{tab:homogeneous}
-\begin{tabular}{|l|r|r|r|r|r|r|r|r|r|r|}
-  \hline
- & $<$ 3 & $<$ 4 & $<$ 5 & $<$ 6 & $<$ 7 & $<$ 8 & $<$ 9 & $<$ 10 & $<$ 11 & $<$ 12 \\ 
-  \hline
-Prefer not to answer & 484 & 1013 & 1062 & 1099 & 1139 & 1177 & 1279 & 1405 & 1471 & 1583 \\ 
-  Do not know & 36 & 68 & 76 & 84 & 92 & 118 & 155 & 188 & 196 & 204 \\ 
-  White & 186 & 422 & 457 & 483 & 513 & 533 & 543 & 543 & 545 & 546 \\ 
-  Mixed & 2 & 6 & 6 & 7 & 8 & 15 & 26 & 42 & 46 & 46 \\ 
-  Asian or Asian British &  &  &  &  &  & 3 & 20 & 40 & 42 & 42 \\ 
-  Black or Black British & 1 & 2 & 2 & 2 & 2 & 2 & 2 & 4 & 6 & 26 \\ 
-  Chinese &  & 1 & 1 & 1 & 1 & 2 & 5 & 21 & 1423 & 1504 \\ 
-  Other ethnic group & 57 & 230 & 261 & 314 & 469 & 885 & 1939 & 2761 & 3681 & 4356 \\ 
-  British & 191713 & 400516 & 416492 & 424490 & 427769 & 429172 & 431026 & 431082 & 431089 & 431090 \\ 
-  Irish & 1416 & 12039 & 12620 & 12700 & 12734 & 12743 & 12759 & 12759 & 12759 & 12759 \\ 
-  Any other white background & 1468 & 4747 & 6953 & 9341 & 12979 & 14613 & 15741 & 15810 & 15820 & 15820 \\ 
-  White and Black Caribbean & 1 & 4 & 4 & 4 & 9 & 35 & 142 & 537 & 589 & 597 \\ 
-  White and Black African & 1 & 3 & 3 & 4 & 6 & 29 & 99 & 333 & 400 & 402 \\ 
-  White and Asian & 4 & 7 & 13 & 23 & 79 & 350 & 651 & 790 & 802 & 802 \\ 
-  Any other mixed background & 24 & 66 & 87 & 155 & 274 & 391 & 595 & 884 & 990 & 996 \\ 
-  Indian &  & 2 & 2 & 5 & 6 & 29 & 1682 & 5700 & 5716 & 5716 \\ 
-  Pakistani &  &  &  &  & 1 & 13 & 532 & 1747 & 1748 & 1748 \\ 
-  Bangladeshi &  &  &  &  &  &  & 2 & 220 & 221 & 221 \\ 
-  Any other Asian background &  &  &  & 1 & 6 & 66 & 427 & 1364 & 1730 & 1747 \\ 
-  Caribbean &  &  &  &  &  &  & 3 & 113 & 1323 & 4299 \\ 
-  African &  & 1 & 1 & 1 & 1 & 1 & 3 & 58 & 350 & 3205 \\ 
-  Any other Black background &  &  &  &  &  & 1 & 3 & 22 & 49 & 118 \\ 
-   \hline
-  All & 195393 & 419127 & 438040 & 448714 & 456088 & 460178 & 467634 & 476423 & 480996 & 487827 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\clearpage
-\subsection*{Projection onto reference PCA space}
-
-\begin{figure}[!htpb]
-\centerline{\includegraphics[width=0.9\textwidth]{proj1000G-PC9-20.pdf}}
-\caption{Principal Component (PC) scores 9 to 20 of the 1000 Genomes project.
-Black points are the 60\% individuals used for computing PCA.
-Red points are the 40\% remaining individuals, projected by simply multiplying their genotypes by the corresponding PC loadings.
-Blue points are the 40\% remaining individuals, projected using the Online Augmentation, Decomposition, and Procrustes (OADP) transformation.
-Estimated shrinkage coefficients (comparing red and blue points) for these PCs are 2.79, 3.14 (PC10), 3.64, 3.18, 2.47, 3.88, 5.31, 5.84, 3.45, 6.55, 3.68 and 6.70 (PC20).
-\label{fig:proj1000G-2}}
-\end{figure}
-
-\vspace*{1em}
-
-\begin{figure}[!htpb]
-\centerline{\includegraphics[width=0.9\textwidth]{proj1000G-PC5-12.pdf}}
-\caption{Principal Component (PC) scores 5 to 12 of the 1000 Genomes project.
-Black points are the 60\% individuals used for computing PCA.
-Green points are the 40\% remaining individuals, projected by multiplying their genotypes by the corresponding PC loadings, further corrected using theoritical asymptotic shrinkage factors (values for the first 12 PCs: 1.01 (PC1), 1.02, 1.07, 1.10, 1.43 (PC5), 1.54, 1.74, 1.79, 2.47, 2.77 (PC10), 2.84 and 3.15).
-Blue points are the 40\% remaining individuals, projected using the Online Augmentation, Decomposition, and Procrustes (OADP) transformation.
-\label{fig:proj1000G-3}}
-\end{figure}
-
-\begin{figure}[!htpb]
-\centerline{\includegraphics[width=0.9\textwidth]{proj1000G-related.pdf}}
-\caption{Principal Component (PC) scores 5 to 12 of the 1000 Genomes project.
-Black points are the 60\% individuals used for computing PCA.
-Blue points are the same 60\% individuals, projected using the Online Augmentation, Decomposition, and Procrustes (OADP) transformation.
-\label{fig:proj1000G-4}}
-\end{figure}
-
-\begin{figure}[!htpb]
-\centerline{\includegraphics[width=0.9\textwidth]{proj1000G-UKBB.png}}
-\caption{Principal Component (PC) scores 1 to 16 of the 1000 Genomes project and projected individuals from the UK Biobank.
-Black points are PC scores of 1000G individuals used for computing PCA.
-Red points are the individuals from UKBB, projected by simply multiplying their genotypes by the corresponding PC loadings.
-Blue points are the 488,371 individuals from the UK Biobank, projected using the Online Augmentation, Decomposition, and Procrustes (OADP) transformation. 
-Estimated shrinkage coefficients (comparing red and blue points) for the first 20 PCs are 1.01 (PC1), 1.02, 1.06, 1.08, 1.36 (PC5), 1.82, 2.33, 2.36, 2.78, 2.84 (PC10), 2.99, 3.51, 4.38, 4.67, 4.99, 5.31, 5.74, 6.55, 6.71 and 6.75 (PC20).
-Note that only 20,000 random projected individuals are represented in this plot.
-\label{fig:proj1000G-5}}
-\end{figure}
-
-\begin{figure}[!htpb]
-\centerline{\includegraphics[width=0.8\textwidth]{proj1000G-UKBB2.png}}
-\caption{Principal Component (PC) scores 1 to 16 from the UK Biobank and projected individuals of the 1000 Genomes (1000G) project.
-Black points are the UK Biobank individuals used for computing PCA.
-Red points are the individuals from 1000G, projected by simply multiplying their genotypes by the corresponding PC loadings.
-Blue points are the individuals from 1000G, projected using the Online Augmentation, Decomposition, and Procrustes (OADP) transformation.
-Estimated shrinkage coefficients (comparing red and blue points) for the first 20 PCs are 1.00 (PC1), 1.00, 1.00, 1.01, 1.01 (PC5), 1.02, 1.03, 1.03, 1.04, 1.04 (PC10), 1.04, 1.05, 1.05, 1.06, 1.07, 1.07, 1.08, 1.08, 1.08 and 1.08 (PC20).
-\label{fig:proj1000G-6}}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\clearpage
-
-\subsection*{PCA of the UK Biobank}
-
-\begin{figure}[!htpb]
-\centerline{\includegraphics[width=0.8\textwidth]{UKBB-screeplot.pdf}}
-\caption{Scree plot: plot of singular values computed on the UK Biobank using \texttt{bed\_autoSVD}.
-\label{fig:UKBB-screeplot}}
-\end{figure}
-
-\begin{figure}[!htpb]
-\centerline{\includegraphics[width=0.9\textwidth]{UKBB-PC1-20.png}}
-\caption{Principal Component (PC) scores 1 to 20 computed on the UK Biobank using \texttt{bed\_autoSVD}.
-Different colors represent different self-reported ancestries.
-\label{fig:UKBB-scores}}
-\end{figure}
-
-\begin{figure}[!htpb]
-\centerline{\includegraphics[width=0.9\textwidth]{UKBB-loadings.pdf}}
-\caption{Principal Component (PC) loadings 1 to 20 computed on the UK Biobank using \texttt{bed\_autoSVD}.
-Column indices of variants in the data, ordered by chromosome and physical position, are represented on the x-axis, and the value of loadings are represented on the y-axis.
-Points are hex-binned.
-\label{fig:UKBB-loadings}}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\clearpage
-
-\begin{figure}[!htpb]
-\centerline{\includegraphics[width=0.8\textwidth]{het-col-PC16.png}}
-\caption{PC-corrected heterozygosity and number of missing values for individuals of the UK Biobank, colored by their value for PC16.
-\label{fig:UKBB-het}}
-\end{figure}
-
-% latex table generated in R 3.6.0 by xtable 1.8-4 package
-% Wed Nov 27 15:52:34 2019
-\begin{table}[ht]
-\caption{Correlation between first 20 PC scores of the UK Biobank, both computed using the genotyping chip and mean imputation, but wtih either no quality control on individuals based on high heterozygosity (top), or after removing some individuals based on high heterozygosity (left).
-PC16 computed without quality control on high heterozygosity completely disappears when performing this quality control.
-\label{tab:pca-het}}
-\centering
-\begin{tabular}{|r|rrrrrrrrrrrrrrrrrrrr|}
-  \hline
- & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16 & 17 & 18 & 19 & 20 \\ 
-  \hline
-1 & 100.0 & -0.2 & 0.1 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & -0.0 & -0.0 & 0.0 & 0.0 & -0.0 & -0.0 & -0.0 & -0.3 & -0.0 & 0.0 & 0.0 & 0.0 \\ 
-  2 & 0.1 & 100.0 & 0.2 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & -0.1 & 0.0 & -0.0 & 0.0 & -0.0 & -0.0 & 0.0 & -0.0 & -0.0 & 0.0 & 0.0 \\ 
-  3 & -0.0 & -0.1 & 100.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 & -0.0 & 0.1 & -0.0 & 0.0 & 0.0 & -0.0 & -0.0 & -0.3 & 0.0 & 0.0 & -0.0 & -0.0 \\ 
-  4 & -0.0 & -0.0 & 0.0 & 100.0 & -0.0 & 0.0 & 0.0 & 0.0 & -0.0 & -0.0 & 0.0 & 0.0 & -0.0 & -0.0 & -0.0 & 0.1 & 0.0 & 0.0 & 0.0 & 0.0 \\ 
-  5 & 0.0 & -0.0 & -0.0 & 0.0 & 100.0 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & -0.0 & 0.0 & 0.0 & 0.0 & -0.0 & 0.1 & -0.0 & -0.0 & -0.0 & 0.0 \\ 
-  6 & -0.0 & -0.0 & 0.0 & -0.0 & -0.0 & 100.0 & -0.3 & 0.4 & -0.0 & 0.0 & -0.0 & 0.0 & -0.0 & -0.0 & 0.0 & -0.1 & 0.0 & -0.0 & -0.0 & 0.0 \\ 
-  7 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & 0.3 & 100.0 & -0.2 & -0.0 & 0.2 & -0.1 & -0.1 & 0.0 & 0.0 & 0.0 & 0.3 & -0.0 & -0.0 & 0.0 & -0.0 \\ 
-  8 & -0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.4 & 0.2 & 100.0 & 0.0 & 0.2 & -0.1 & -0.0 & 0.0 & 0.0 & 0.0 & 0.2 & 0.0 & 0.0 & -0.0 & 0.0 \\ 
-  9 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & 100.0 & 0.2 & -0.0 & 0.0 & -0.0 & -0.0 & 0.0 & -0.5 & -0.0 & 0.0 & -0.0 & 0.0 \\ 
-  10 & 0.0 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & -0.1 & -0.1 & -0.2 & 100.0 & 0.4 & -0.1 & 0.1 & 0.0 & 0.0 & 0.8 & -0.0 & -0.0 & 0.0 & 0.0 \\ 
-  11 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & -0.3 & 100.0 & 0.2 & -0.1 & -0.0 & -0.0 & -0.2 & 0.0 & 0.0 & -0.0 & 0.0 \\ 
-  12 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 & -0.0 & 0.0 & 0.0 & -0.0 & 0.1 & -0.1 & 100.0 & -0.3 & 0.1 & 0.0 & 1.7 & -0.0 & -0.1 & -0.0 & -0.0 \\ 
-  13 & 0.0 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & 0.3 & 100.0 & -0.4 & -0.1 & -2.8 & 0.0 & 0.0 & 0.1 & 0.0 \\ 
-  14 & 0.0 & 0.0 & 0.0 & 0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 & -0.0 & 0.0 & -0.1 & 0.4 & 100.0 & -0.0 & -1.8 & 0.0 & 0.0 & 0.0 & -0.0 \\ 
-  15 & 0.0 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & -0.0 & 100.0 & -2.1 & 0.1 & -0.0 & 0.0 & 0.0 \\ 
-  16 & 0.0 & 0.0 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & 0.0 & 0.0 & 0.0 & -0.0 & 1.3 & 100.0 & 0.6 & -0.1 & -0.3 \\ 
-  17 & -0.0 & -0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 & 0.0 & -0.0 & -0.0 & 0.0 & 0.0 & -0.0 & -0.0 & -0.1 & -5.0 & 0.7 & -99.9 & -0.8 & -0.2 \\ 
-  18 & 0.0 & 0.0 & -0.0 & 0.0 & -0.0 & -0.0 & 0.0 & 0.0 & -0.0 & 0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 & -2.9 & 0.0 & 0.9 & -99.5 & 8.7 \\ 
-  19 & 0.0 & 0.0 & -0.0 & 0.0 & 0.0 & 0.0 & -0.0 & 0.0 & 0.0 & 0.0 & 0.0 & -0.0 & 0.0 & -0.0 & 0.0 & -0.3 & -0.2 & 0.3 & -8.6 & -99.6 \\ 
-  20 & 0.0 & 0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 & 0.0 & 0.0 & 0.0 & 0.0 & -0.0 & 0.0 & 0.0 & 0.1 & 7.3 & -0.8 & -0.3 & -4.4 & -0.3 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-% latex table generated in R 3.6.0 by xtable 1.8-4 package
-% Wed Nov 27 16:04:03 2019
-\begin{table}[ht]
-\caption{Correlation between first 20 PC scores of the UK Biobank, either computed using the genotyping chip and mean imputation (top), or computed from the dosages (based on imputation from an external reference panel) of the same variants (left).
-PCs are globally the same.
-\label{tab:pca-dosage}}
-\centering
-\begin{tabular}{|r|rrrrrrrrrrrrrrrrrrrr|}
-  \hline
- & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16 & 17 & 18 & 19 & 20 \\ 
-  \hline
-1 & -100.0 & -0.1 & -0.1 & 0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 & -0.0 & -0.0 & -0.0 & 0.0 & 0.0 & 0.0 & -0.0 \\ 
-  2 & 0.1 & -100.0 & -0.2 & -0.0 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & -0.0 & 0.0 & 0.0 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & -0.0 & -0.0 & 0.0 \\ 
-  3 & -0.1 & -0.2 & 100.0 & -0.2 & 0.0 & 0.0 & 0.1 & 0.0 & 0.0 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & -0.0 & 0.0 & 0.0 & -0.0 & 0.0 & -0.0 \\ 
-  4 & 0.0 & -0.0 & 0.2 & 100.0 & 0.1 & 0.0 & -0.1 & -0.0 & 0.0 & 0.1 & -0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & -0.0 & 0.0 & -0.0 & 0.0 \\ 
-  5 & 0.0 & 0.0 & 0.0 & 0.1 & -100.0 & 0.0 & -0.0 & -0.0 & 0.0 & 0.0 & -0.0 & -0.0 & 0.0 & 0.0 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & -0.0 \\ 
-  6 & -0.0 & -0.0 & -0.0 & -0.0 & 0.0 & 99.9 & -2.7 & 2.0 & -0.0 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & -0.0 \\ 
-  7 & -0.0 & -0.0 & -0.1 & 0.1 & -0.0 & 2.6 & 99.9 & 3.3 & -0.1 & -0.1 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & -0.0 \\ 
-  8 & 0.0 & 0.0 & -0.0 & 0.0 & -0.0 & -2.1 & -3.2 & 99.9 & 0.2 & -0.2 & 0.0 & -0.0 & -0.0 & -0.0 & -0.0 & -0.0 & -0.0 & -0.0 & 0.0 & -0.0 \\ 
-  9 & -0.0 & -0.0 & -0.0 & -0.0 & -0.0 & 0.0 & 0.1 & -0.2 & 100.0 & -0.0 & -0.0 & -0.0 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & 0.0 & 0.0 & 0.0 \\ 
-  10 & -0.0 & 0.0 & 0.1 & 0.0 & -0.0 & -0.0 & -0.1 & -0.1 & -0.0 & -99.9 & -0.3 & -0.1 & 0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 & -0.0 & -0.0 \\ 
-  11 & 0.0 & -0.0 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & 0.4 & -99.9 & 0.2 & 0.1 & 0.0 & 0.0 & -0.0 & -0.0 & -0.0 & 0.0 & -0.0 \\ 
-  12 & 0.0 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 & 0.1 & -0.2 & -99.9 & 0.7 & -0.0 & 0.1 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 \\ 
-  13 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & 0.0 & 0.0 & 0.1 & 0.7 & 99.9 & 2.1 & 0.0 & -0.0 & 0.0 & -0.1 & -0.0 & 0.0 \\ 
-  14 & -0.0 & -0.0 & 0.0 & 0.0 & 0.0 & -0.0 & -0.0 & 0.0 & 0.0 & -0.0 & 0.0 & -0.0 & -2.1 & 99.9 & -0.1 & -0.0 & 0.0 & -0.0 & 0.0 & 0.0 \\ 
-  15 & -0.0 & -0.0 & -0.0 & 0.0 & 0.0 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & 0.0 & 0.1 & -0.0 & 0.1 & 99.9 & 0.1 & 0.0 & -0.1 & 0.1 & 0.1 \\ 
-  16 & -0.0 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & -0.0 & -0.0 & -0.0 & -0.0 & -0.0 & -0.0 & -0.0 & -0.0 & 0.0 & -99.7 & 6.4 & 0.8 & 0.9 & -1.3 \\ 
-  17 & -0.0 & 0.0 & -0.0 & 0.0 & 0.0 & -0.0 & 0.0 & 0.0 & 0.0 & 0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 & 6.4 & 99.6 & -1.0 & 2.4 & 0.5 \\ 
-  18 & 0.0 & -0.0 & -0.0 & 0.0 & -0.0 & -0.0 & 0.0 & -0.0 & 0.0 & 0.0 & 0.0 & -0.0 & -0.1 & -0.0 & -0.1 & -0.9 & -0.9 & -99.6 & -5.0 & 4.0 \\ 
-  19 & 0.0 & -0.0 & -0.0 & 0.0 & 0.0 & -0.0 & 0.0 & 0.0 & 0.0 & 0.0 & 0.0 & -0.0 & -0.0 & 0.0 & 0.1 & -0.6 & 2.5 & 4.9 & -99.5 & -6.2 \\ 
-  20 & -0.0 & 0.0 & 0.0 & 0.0 & -0.0 & -0.0 & 0.0 & 0.0 & -0.0 & -0.0 & -0.0 & 0.0 & -0.0 & -0.0 & -0.0 & -1.4 & -0.2 & 4.3 & -6.0 & 99.5 \\ 
-   \hline
-\end{tabular}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\clearpage
-
-\begin{figure}[!htpb]
-\centerline{\includegraphics[width=0.8\textwidth]{UKBB-screeplot-restricted.pdf}}
-\caption{Scree plot: plot of singular values computed on the UK Biobank using 48,942 individuals of diverse ancestries.
-These individuals are the ones resulting from removing all related individuals and randomly subsampling the British and Irish individuals.
-\label{fig:UKBB-screeplot2}}
-\end{figure}
-
-\begin{figure}[!htpb]
-\centerline{\includegraphics[width=0.9\textwidth]{projUKBB-related.png}}
-\caption{Principal Component (PC) scores 27 to 50 of the UK Biobank.
-Black points are the 48,942 individuals of diverse ancestries used for computing PCA.
-These individuals are the ones resulting from removing all related individuals and randomly subsampling the British and Irish individuals.
-Red points are the remaining UKBB individuals, projected by simply multiplying their genotypes by the corresponding PC loadings.
-Blue points are the remaining UKBB individuals, projected using the Online Augmentation, Decomposition, and Procrustes (OADP) transformation.
-Note that only 20,000 random projected individuals are represented in this plot.
-\label{fig:projUKBB-related}}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \end{document}

### 2020-05-04 paper-bedpca.tex
need to change format now

@@
 For instance, adding PCs as covariates is routinely used to adjust for population structure in Genome-Wide Association Studies (GWAS) \cite[]{price2006principal,novembre2008interpreting}.
-PCA has also been used to detect loci under selection \cite[]{galinsky2016fast,luu2017pcadapt} and in heritability analyses \cite[]{yang2010common,loh2015contrasting}.
+PCA has also been used to detect loci under selection \cite[]{galinsky2016fast,luu2017pcadapt,prive2020performing} and in heritability analyses \cite[]{yang2010common,loh2015contrasting}.
 Recently, the advent of large population-scale genetic datasets, such as the UK biobank data, has prompted research on developing scalable algorithms to compute PCA on very large data \cite[]{bycroft2017genome}.
@@
 However, in practice, conducting PCA on genotype data to capture population structure consists of more steps than simply performing singular value decomposition on the genotype matrix.
-These steps includes removing related individuals, pruning variants in LD, and excluding outlier samples that can suggest poor genotyping quality or distant relatedness.
-Indeed, many pitfalls related to PCA of genotype data have been documented and none of the currently available software address all of these.
+These steps include removing related individuals, pruning variants in LD, and excluding outlier samples that can suggest poor genotyping quality or distant relatedness.
+Some genetic analyses may also require to restrict to individuals of homogeneous ancestry.
+Many pitfalls related to PCA of genotype data have been documented and none of the currently available software address all of these. 
 In the following, we outline these pitfalls and explain when they are relevant.
@@
 Yet, here we use the pairwise orthogonalised Gnanadesikan-Kettenrin robust estimates of these parameters \cite[]{gnanadesikan1972robust,yohai1988high,maronna2002robust}.
-We implement the estimation of these robust parameters in function \texttt{covrob\_ogk} of R package bigutilsr.
+We implement the estimation of these robust parameters in function \texttt{covrob\_ogk} of R package bigutilsr, and the direct computation of these robust distances in function \texttt{dist\_ogk}.
 
@@
 For detecting outlier variants in PCA that are due to long-range Linkage Disequilibrium (LD) regions, we use a similar procedure as described by \cite{prive2017efficient}.
+Note that this procedure does not require removing any known long-range LD region a priori.
 We first apply a first round of clumping at e.g.\ $r^2$ > 0.2, prioritising variants by higher minor allele count.
@@
 To project individuals from the same dataset as the individuals used for computing PCA, we provide function \texttt{bed\_projectSelfPCA}. 
-Note that the new individuals to be projected should not be related to the ones used for computing PCA (Section \ref{proj-related}).
+Note that the new individuals to be projected should not be related to the ones used for computing PCA (cf.\ section \ref{proj-related}).
 
@@
 
-When applying our automatic procedure to remove long-range LD regions, it does not converge after 5 iterations for the UK Biobank, meaning that it keeps detecting long-range LD regions at each iteration (represented by peaks in PC loadings).
+When applying our automatic procedure to remove long-range LD regions, it does not converge after 5 iterations for the UK Biobank, i.e.\ it keeps detecting long-range LD regions at each iteration (represented by peaks in PC loadings).
 Therefore, we are able to capture only 16 PCs that show stratification that is not LD structure (Figures \ref{fig:UKBB-screeplot}, \ref{fig:UKBB-scores} and \ref{fig:UKBB-loadings}).
@@
 As for other analyses, it takes 8 minutes to match the 1000G data to the UKBB data and compute 20 PCs of the 1000G data using the automatic LD detection technique.
-It takes 12 minutes more to do the OADP projection of all 488,371 individuals of the UKBB data onto the PCA computed using the 1000G data.
+It takes 12 minutes more to perform the OADP projection of all 488,371 UKBB individuals onto the PCA space computed using the 1000G data.
 Finally, it takes only 6 minutes to compute the 30-Nearest Neighbours of 20 PC scores for 406,545 UK Biobank individuals, which is the most computationally demanding step when computing the statistics used to detect individual outlier samples (Section \ref{outlier-sample}).
@@
 To identify which PCs capture population structure, and which ones are probably just noise, one should also look at PC scores (colored by ancestry if possible).
+PCs with no visible population stratification, i.e.\ where all individuals are normally distributed around 0, can be considered as noise.   
 As in many applications, we believe a compromise between signal and noise should be preferred.

### 2020-05-04 paper-bedpca-supp.tex
reformat & submit

@@
 \usepackage{graphicx}
-\usepackage{subcaption}
 \usepackage{amsmath, amsthm, amssymb}
@@
 
+\usepackage{float}
+\usepackage[caption = false]{subfig}
+
 \let\tabbeg\tabular
@@
 
-\section*{Supplementary Materials}
+%\section*{Supplementary Materials}
 
@@
 
+\bibliographystyle{natbib}
+\bibliography{refs.bib}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 	\centering
-\begin{subfigure}[b]{0.55\textwidth}
-\includegraphics[width=\textwidth]{hist-outliers2-1000G.pdf}
-\caption{Distribution of statistics (S2) and default threshold (6, in red)}
-\end{subfigure}
-\\~\\
-\begin{subfigure}[b]{0.45\textwidth}
-\includegraphics[width=\textwidth]{outliers2-1000G.pdf}
-\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by statistic used to detect outliers (maximum number of SDs from the mean, log-scale).}
-\end{subfigure}
-~~~~~~
-\begin{subfigure}[b]{0.45\textwidth}
-\includegraphics[width=\textwidth]{be-outlier2-1000G.pdf}
-\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by being detected as an outlier.\\}
-\end{subfigure}
-\caption{Outlier detection in the 1000 Genomes (1000G) project, using the rule ``6 SDs from the mean'', i.e.\ where S2 is the maximum (for all PCs) number of SDs from the mean (Section \ref{outlier-sample}).\label{fig:outlier-sd}}
+	\subfloat[Distribution of statistics (S2) and default threshold (6, in red)]{\includegraphics[width=0.55\textwidth]{hist-outliers2-1000G.pdf}} \\~\\
+	\subfloat[Principal Component (PC) scores 13 to 20 of 1000G, colored by statistic used to detect outliers (maximum number of SDs from the mean, log-scale).]{\includegraphics[width=0.45\textwidth]{outliers2-1000G.pdf}} $~~~~~~$
+	\subfloat[Principal Component (PC) scores 13 to 20 of 1000G, colored by being detected as an outlier.\\]{\includegraphics[width=0.45\textwidth]{be-outlier2-1000G.pdf}}
+	\caption{Outlier detection in the 1000 Genomes (1000G) project, using the rule ``6 SDs from the mean'', i.e.\ where S2 is the maximum (for all PCs) number of SDs from the mean (Section 2.4).\label{fig:outlier-sd}}
 \end{figure}
@@
 	\centering
-\begin{subfigure}[b]{0.45\textwidth}
-\includegraphics[width=\textwidth]{UKBB-White-British.png}
-\caption{PCs 1 to 20 of UKBB, colored by whether it is in the group of ``White British'' reported by the UKBB (blue).}
-\end{subfigure}
-~~~~~~
-\begin{subfigure}[b]{0.5\textwidth}
-\includegraphics[width=\textwidth]{UKBB-Maha-dist.png}
-\caption{PCs 1 to 20 of UKBB, colored by robust Mahalanobis distances computed on PCs (log-scale).\\}
-\end{subfigure}
-\\~\\
-\begin{subfigure}[b]{0.45\textwidth}
-\includegraphics[width=\textwidth]{hist-Maha-dist.pdf}
-\caption{Distribution of (log squared) distances.\\~\\~\\~\\~\\}
-\end{subfigure}
-~~~~~~
-\begin{subfigure}[b]{0.45\textwidth}
-\includegraphics[width=\textwidth]{UKBB-Maha-outlier.png}
-\caption{PCs 1 to 20 of UKBB, colored by being detected as an outlier.
-Threshold of being considered as an outlier is determined based on histogram (c), where the threshold of 5 is chosen for the logarithm of the distances.}
-\end{subfigure}
+	\subfloat[PCs 1 to 20 of UKBB, colored by whether it is in the group of ``White British'' reported by the UKBB (blue).]{\includegraphics[width=0.45\textwidth]{UKBB-White-British.png}} $~~~~~~$
+	\subfloat[PCs 1 to 20 of UKBB, colored by robust Mahalanobis distances computed on PCs (log-scale).\\]{\includegraphics[width=0.45\textwidth]{UKBB-Maha-dist.png}} \\~\\
+	\subfloat[Distribution of (log squared) distances.\\~\\~\\~\\~\\).]{\includegraphics[width=0.45\textwidth]{hist-Maha-dist.pdf}} $~~~~~~$
+	\subfloat[PCs 1 to 20 of UKBB, colored by being detected as an outlier.
+Threshold of being considered as an outlier is determined based on histogram (c), where the threshold of 5 is chosen for the logarithm of the distances.]{\includegraphics[width=0.45\textwidth]{UKBB-Maha-outlier.png}}
 	\caption{Homogeneous sample detection in the UK Biobank (UKBB), using robust Mahalanobis distances computed on the first 20 Principal Component scores (PCs) of UKBB.).\label{fig:homogeneous}}

### 2020-05-04 paper-bedpca.tex
reformat & submit

@@
-%% LyX 1.3 created this file.  For more info, see http://www.lyx.org/.
-%% Do not edit unless you really know what you are doing.
-\documentclass[english, 12pt]{article}
-\usepackage{times}
+\documentclass{bioinfo}
+\copyrightyear{2015} \pubyear{2015}
+
+\access{Advance Access Publication Date: Day Month Year}
+\appnotes{Manuscript Category}
+
+%\usepackage{times}
 %\usepackage{algorithm2e}
-\usepackage{url}
-\usepackage{bbm}
-\usepackage[T1]{fontenc}
-\usepackage[latin1]{inputenc}
-\usepackage{geometry}
-\geometry{verbose,letterpaper,tmargin=2cm,bmargin=2cm,lmargin=1.5cm,rmargin=1.5cm}
-\usepackage{rotating}
-\usepackage{color}
-\usepackage{graphicx}
-\usepackage{subcaption}
-\usepackage{amsmath, amsthm, amssymb}
-\usepackage{setspace}
-\usepackage{lineno}
+%\usepackage{url}
+%\usepackage{bbm}
+%\usepackage[T1]{fontenc}
+%\usepackage[latin1]{inputenc}
+%\usepackage{geometry}
+%\geometry{verbose,letterpaper,tmargin=2cm,bmargin=2cm,lmargin=1.5cm,rmargin=1.5cm}
+%\usepackage{rotating}
+%\usepackage{color}
+%\usepackage{graphicx}
+%\usepackage{amsmath, amsthm, amssymb}
+%\usepackage{setspace}
+%\usepackage{lineno}
 \usepackage{hyperref}
-\usepackage{bbm}
-\usepackage{makecell}
 
@@
 
-\usepackage{xr}
-\externaldocument{paper-bedpca-supp}
+%\usepackage{xr}
+%\externaldocument{paper-bedpca-supp}
 
@@
 %\doublespacing
-\onehalfspacing
+%\onehalfspacing
 %\usepackage[authoryear]{natbib}
-\usepackage{natbib} \bibpunct{(}{)}{;}{author-year}{}{,}
+%\usepackage{natbib} \bibpunct{(}{)}{;}{author-year}{}{,}
 
 %Pour les rajouts
-\usepackage{color}
-\definecolor{trustcolor}{rgb}{0,0,1}
+%\usepackage{color}
+%\definecolor{trustcolor}{rgb}{0,0,1}
 
-\usepackage{dsfont}
-\usepackage[warn]{textcomp}
+%\usepackage{dsfont}
+%\usepackage[warn]{textcomp}
 \usepackage{adjustbox}
+\usepackage{makecell}
 \usepackage{multirow}
-\usepackage{graphicx}
-\graphicspath{{figures/}}
-\DeclareMathOperator*{\argmin}{\arg\!\min}
+%\usepackage{graphicx}
+%\graphicspath{{figures/}}
+%\DeclareMathOperator*{\argmin}{\arg\!\min}
 
@@
 \let\tabend\endtabular
-\renewenvironment{tabular}{\begin{adjustbox}{max width=0.9\textwidth}\tabbeg}{\tabend\end{adjustbox}}
+\renewenvironment{tabular}{\begin{adjustbox}{max width=0.95\linewidth}\tabbeg}{\tabend\end{adjustbox}}
+
+\usepackage{float}
+\usepackage[caption = false]{subfig}
+
+\usepackage{natbib}
 
@@
 %% Because html converters don't know tabularnewline
-\providecommand{\tabularnewline}{\\}
+%\providecommand{\tabularnewline}{\\}
+%\providecommand{\@LN}[2]{}
 
-\usepackage{babel}
+%\usepackage{babel}
 \makeatother
@@
 
-\title{Efficient toolkit implementing best practices for principal component analysis of population genetic data}
-\author{Florian Priv\'e,$^{\text{1,5,}*}$ Keurcien Luu$^{\text{5}}$, Michael G.B. Blum,$^{\text{4,5}}$ John J. McGrath,$^{\text{1,2,3}}$ and Bjarni J. Vilhj\'almsson$^{\text{1,}*}$}
+\firstpage{1}
 
-\date{~ }
-\maketitle
+\subtitle{Subject Section}
 
-\noindent$^{\text{\sf 1}}$National Centre for Register-Based Research, Aarhus University, Aarhus, 8210, Denmark. \\
-\noindent$^{\text{\sf 2}}$Queensland Brain Institute, University of Queensland, St. Lucia, 4072, Queensland, Australia. \\
-\noindent$^{\text{\sf 3}}$Queensland Centre for Mental Health Research, The Park Centre for Mental Health, Wacol, 4076, Queensland, Australia. \\
-\noindent$^{\text{\sf 4}}$OWKIN France, Paris, 75010, France. \\
-\noindent$^{\text{\sf 5}}$Laboratoire TIMC-IMAG, UMR 5525, Univ.\ Grenoble Alpes, La Tronche, 38700, France. \\
-\noindent$^\ast$To whom correspondence should be addressed.\\
+\title[Efficient PCA Toolkit for Genetic Data]{Efficient toolkit implementing best practices for principal component analysis of population genetic data}
 
-\noindent Contacts:
-\begin{itemize}
-\item \url{florian.prive.21@gmail.com}
-\item \url{bjv@econ.au.dk}
-\end{itemize}
+\author[Priv\'e \textit{et~al}.]{Florian Priv\'e,$^{\text{1,5,}*}$ Keurcien Luu,$^{\text{5}}$ Michael G.B.
+Blum,$^{\text{4,5}}$ John J.
+McGrath,$^{\text{1,2,3}}$ and Bjarni J.
+Vilhj\'almsson$^{\text{1,}*}$}
 
-\clearpage
+\address{$^{\text{\sf 1}}$National Centre for Register-Based Research, Aarhus University, Aarhus, 8210, Denmark. \\
+$^{\text{\sf 2}}$Queensland Brain Institute, University of Queensland, St. Lucia, 4072, Queensland, Australia. \\
+$^{\text{\sf 3}}$Queensland Centre for Mental Health Research, The Park Centre for Mental Health, Wacol, 4076, Queensland, Australia. \\
+$^{\text{\sf 4}}$OWKIN France, Paris, 75010, France. \\
+$^{\text{\sf 5}}$Laboratoire TIMC-IMAG, UMR 5525, Univ.\ Grenoble Alpes, La Tronche, 38700, France.}
 
-\abstract{	
-Principal Component Analysis (PCA) of genetic data is routinely used to infer ancestry and control for population structure in various genetic analyses.
+\corresp{$^\ast$To whom correspondence should be addressed.}
+
+\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
+
+\editor{Associate Editor: XXXXXXX}
+
+\abstract{\textbf{Motivation:} Principal Component Analysis (PCA) of genetic data is routinely used to infer ancestry and control for population structure in various genetic analyses.
 However, conducting PCA analyses can be complicated and has several potential pitfalls.
@@
 Following applications to the UK Biobank and the 1000 Genomes project datasets, we make recommendations for best practices and provide efficient and user-friendly implementations of the proposed solutions in R packages bigsnpr and bigutilsr.
-
-For example, we find that PC19 to PC40 in the UK Biobank capture complex LD structure rather than population structure.
+\\
+\textbf{Results:} For example, we find that PC19 to PC40 in the UK Biobank capture complex LD structure rather than population structure.
 Using our automatic algorithm for removing long-range LD regions, we recover 16 PCs that capture population structure only.
@@
 We then demonstrate how to obtain unbiased projections efficiently using bigsnpr.
+\\
+\textbf{Conclusion:} Overall, we believe this work would be of interest for anyone using PCA in their analyses of genetic data, as well as for other omics data. \\
+\textbf{Availability:} R packages bigsnpr and bigutilsr can be installed from either CRAN or GitHub (see \url{https://github.com/privefl/bigsnpr}).
+A tutorial on the steps to perform PCA on 1000G data is available at \url{https://privefl.github.io/bigsnpr/articles/bedpca.html}.
+All code used for this paper is available at \url{https://github.com/privefl/paper4-bedpca/tree/master/code}.\\
+\textbf{Contact:} \href{florian.prive.21@gmail.com}{florian.prive.21@gmail.com} \& \href{bjv@econ.au.dk}{bjv@econ.au.dk}\\
+\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
+online.}
 
-Overall, we believe this work would be of interest for anyone using PCA in their analyses of genetic data, as well as for other omics data.
-}
+\maketitle
 
@@
 
-\clearpage
+%\clearpage
 
@@
 
+\section{Approach}
+
 For this paper, we derive implementations of truncated PCA and other useful functions for e.g.\ performing LD thinning and computing various statistics.
@@
 
-\section{Material and Methods}
+\begin{methods}
+\section{Methods}
 
@@
 \renewcommand*{\arraystretch}{2}
-\begin{table}[h]
+\begin{table*}[htb]
 \centering
@@
 \end{tabular}
-\end{table}
+\end{table*}
 \endgroup
@@
 
+\end{methods}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 When applying our automatic procedure to remove long-range LD regions, it does not converge after 5 iterations for the UK Biobank, i.e.\ it keeps detecting long-range LD regions at each iteration (represented by peaks in PC loadings).
-Therefore, we are able to capture only 16 PCs that show stratification that is not LD structure (Figures \ref{fig:UKBB-screeplot}, \ref{fig:UKBB-scores} and \ref{fig:UKBB-loadings}).
+Therefore, we are able to capture only 16 PCs that show stratification that is not LD structure (Figures S8, S9 and S10).
 Similarly, PC loadings reported by the UK Biobank clearly show that PC19 to PC40 capture LD structure, which is also the case for PC16 and PC18, although less pronounced (see peaks in figure \ref{fig:UKBB-loadings40}).
-These include e.g.\ one region on chromosome 6 (70-91 Mbp) that is captured in PC19 (Figure \ref{fig:UKBB-loadings}) and that was not previously reported in \cite{price2008long}.
+These include e.g.\ one region on chromosome 6 (70-91 Mbp) that is captured in PC19 (Figure S10) and that was not previously reported in \cite{price2008long}.
 
-\begin{figure}[!htpb]
-\centerline{\includegraphics[width=0.9\textwidth]{UKBB-loadings1-40.pdf}}
+\begin{figure*}[htb]
+\centerline{\includegraphics[width=0.8\textwidth]{UKBB-loadings1-40.pdf}}
 \caption{Principal Component (PC) loadings 1 to 40 reported by the UK Biobank.
@@
 \label{fig:UKBB-loadings40}}
-\end{figure}
+\end{figure*}
 
@@
 All but one outlier are distantly-related pairs that disappear if using a more stringent threshold on relatedness (i.e.\ using a KING-relatedness cutoff of \textasciitilde0.0442 instead of \textasciitilde0.0884, see tutorial in section ``code availability'').
-In contrast, rule 6SD identifies a lot of outliers, of which some are part of a relatively large cluster (Figure \ref{fig:outlier-sd}).
+In contrast, rule 6SD identifies a lot of outliers, of which some are part of a relatively large cluster (Figure S1).
 We recall that, if all PCs are normally distributed, after correcting for multiple testing of 2500 individuals and 20 PCs, the probability of detecting one outlier or more using 6SD is only of 0.0001.
 
-\begin{figure}[htbp]
+\begin{figure*}[htb]
 	\centering
-\begin{subfigure}[b]{0.55\textwidth}
-\includegraphics[width=\textwidth]{hist-outliers-1000G.pdf}
-\caption{Distribution of statistics (S).}
-\end{subfigure}
-\\~\\
-\begin{subfigure}[b]{0.45\textwidth}
-\includegraphics[width=\textwidth]{outliers-1000G.pdf}
-\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by the statistic (S) used to define outliers.
-A few points with higher values for this statistic S appear as outliers in PC17-20.}
-\end{subfigure}
-~~~~~~
-\begin{subfigure}[b]{0.45\textwidth}
-\includegraphics[width=\textwidth]{be-outlier-1000G.pdf}
-\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by being detected as an outlier. Threshold of being an outlier is determined based on histogram (a).}
-\end{subfigure}
+	\subfloat[Distribution of statistics (S).]{\includegraphics[width=0.35\linewidth]{hist-outliers-1000G.pdf}} $~~~$
+	\subfloat[Principal Component (PC) scores 13 to 20 of 1000G, colored by the statistic (S) used to define outliers.
+A few points with higher values for this statistic S appear as outliers in PC17-20.]{\includegraphics[width=0.3\linewidth]{outliers-1000G.pdf}} $~~~~~$
+	\subfloat[Principal Component (PC) scores 13 to 20 of 1000G, colored by being detected as an outlier.
+Threshold of being an outlier is determined based on histogram (a).]{\includegraphics[width=0.3\linewidth]{be-outlier-1000G.pdf}}
 	\caption{Outlier detection in the 1000 Genomes (1000G) project, using \texttt{prob\_dist} (Section \ref{outlier-sample}).\label{fig:outlier-pd}}
-\end{figure}
+\end{figure*}
 
 As for restricting to homogeneous samples, we compare the use of the robust Mahalanobis distance we propose here to the use of R package aberrant, which was used to report the homogeneous ``White British'' subset in the UKBB \cite[]{bellenguez2011robust,bycroft2017genome}.
-We visually choose a threshold of 5 on the log-distance and show that this gives a similar subset of individuals than the ``White British'' subset reported by the UK Biobank (Figure \ref{fig:homogeneous}).
-Moreover, when using this threshold, only 3 out of 10,936 people of self-reported Asian ancestry (1 ``Chinese'' and 2 ``Indian'') are kept, and 1 ``African'' out of 7622 people with Black background is kept (Table \ref{tab:homogeneous}).
+We visually choose a threshold of 5 on the log-distance and show that this gives a similar subset of individuals than the ``White British'' subset reported by the UK Biobank (Figure S2).
+Moreover, when using this threshold, only 3 out of 10,936 people of self-reported Asian ancestry (1 ``Chinese'' and 2 ``Indian'') are kept, and 1 ``African'' out of 7622 people with Black background is kept (Table S1).
 In contrast, 416,492 out of 431,090 ``British'' (96.6\%) and 12,620 out of 12,759 ``Irish'' (98.9\%) are kept.
-Results are very similar to the set of ``White British'' made using R package aberrant (Figure \ref{fig:homogeneous}).
+Results are very similar to the set of ``White British'' made using R package aberrant (Figure S2).
 
@@
 In contrast, there is no visible shrinkage when projecting new individuals with OADP (Figure \ref{fig:proj1000G}).
-Simple projection is affected even more by this shrinkage for PC9 to PC20, while OADP still appears free of this bias (Figure \ref{fig:proj1000G-2}).
-We show the same results when projecting the full UK Biobank data onto PCA computed using 1000G data  (Figure \ref{fig:proj1000G-5}).
-When correcting projected PC scores with asymptotic shrinkage factors, bias is smaller than with simple projection, yet, there is a visible bias for PC7-8 (Figure \ref{fig:proj1000G-3}).
+Simple projection is affected even more by this shrinkage for PC9 to PC20, while OADP still appears free of this bias (Figure S3).
+We show the same results when projecting the full UK Biobank data onto PCA computed using 1000G data  (Figure S6).
+When correcting projected PC scores with asymptotic shrinkage factors, bias is smaller than with simple projection, yet, there is a visible bias for PC7-8 (Figure S4).
 Finally, to assess if OADP could be used to project individuals that are related to some individuals that were used to compute PCA, we projected these 60\% individuals (as if we were projecting their monozygotic twins) using OADP.
-Projections of related individuals using OADP suffers from some bias in reverse direction (Figure \ref{fig:proj1000G-4}).
+Projections of related individuals using OADP suffers from some bias in reverse direction (Figure S5).
 \label{proj-related}
 
-\begin{figure}[!htpb]
+\begin{figure*}[htb]
 \centerline{\includegraphics[width=0.8\textwidth]{proj1000G-PC1-8.pdf}}
@@
 \label{fig:proj1000G}}
-\end{figure}
+\end{figure*}
 
-When computing the PCs on the UK Biobank using 406,545 unrelated individuals and 171,977 variants, and projecting the 1000G data onto this reference PCA space, shrinkage is much smaller ($\leq 1.08$ for all 20 first PCs, Figure \ref{fig:proj1000G-6}).
+When computing the PCs on the UK Biobank using 406,545 unrelated individuals and 171,977 variants, and projecting the 1000G data onto this reference PCA space, shrinkage is much smaller ($\leq 1.08$ for all 20 first PCs, Figure S7).
 Overall, this shrinkage for simple projection decreases with an increased sample size (Table \ref{tab:shrinkage}).
@@
 \renewcommand*{\arraystretch}{1.8}
-\begin{table}[h]
+\begin{table}[htb]
 \centering
@@
 This resulted in 48,942 individuals that we used to compute 50 PCs, which took less than 3 hours using function \texttt{bed\_autoSVD} (that converged after 4 iterations of automatic LD removal). 
-We show that we are able to capture more PCs (at least 40 instead of 16-18) that display visual population structure (Figures \ref{fig:UKBB-scores2} and \ref{fig:UKBB-screeplot2}).
+We show that we are able to capture more PCs (at least 40 instead of 16-18) that display visual population structure (Figures \ref{fig:UKBB-scores2} and S12).
 We then projected all 439,429 remaining individuals from UKBB onto this PCA space in 21 minutes only using our implementation of the OADP projection (function \texttt{bed\_projectSelfPCA}).
 Note that these individuals should not be related to any of the 48,942 individuals used for training PCA because we removed both individuals from each pair of related individuals in the UKBB.
-Projection of new individuals show again a clear shrinkage when using simple projection (between 1.00 for PC1 and 1.80 for PC50), but no visible bias when using OADP projection (Figure \ref{fig:projUKBB-related}).
+Projection of new individuals show again a clear shrinkage when using simple projection (between 1.00 for PC1 and 1.80 for PC50), but no visible bias when using OADP projection (Figure S13).
 
-\begin{figure}[!htpb]
-\centerline{\includegraphics[width=0.9\textwidth]{UKBB-scores-restricted.png}}
+\begin{figure*}[htb]
+\centerline{\includegraphics[width=0.8\textwidth]{UKBB-scores-restricted.png}}
 \caption{Principal Component (PC) scores 27 to 50 computed on the UK Biobank using 48,942 individuals of diverse ancestries.
@@
 \label{fig:UKBB-scores2}}
-\end{figure}
+\end{figure*}
 
@@
 In the UK Biobank imputed data, approximately 1000 individuals have been removed because of a high number of missing values or a high heterozygosity, as compared to the genotyped data \cite[]{bycroft2017genome}.
-When computing PCA with mean imputation and using all genotyped individuals, PC16 captures individuals with very high heterozygosity (Figure \ref{fig:UKBB-het}).
-When restricting to imputed individuals only, i.e.\ after removing individuals with very high heterozygosity, PC16 completely disappears and new PC16 to PC19 correspond to previous PC17 to PC20 (Table \ref{tab:pca-het}).
-When using dosage data instead of genotype data with mean imputation, PCA is globally unchanged (Table \ref{tab:pca-dosage}). 
-Overall, if we choose to use either one of the following 4 imputation methods: mean imputation, random imputation according to allele frequencies, using reported dosage data from BGEN files, or imputation of genotyped data based on machine learning using function \texttt{snp\_fastImpute} of R package bigsnpr \cite[]{prive2017efficient}, resulting PCs are always very similar (absolute correlation larger than 0.99 for the 20 computed PCs; results partially shown in tables \ref{tab:pca-het} and \ref{tab:pca-dosage}).
+When computing PCA with mean imputation and using all genotyped individuals, PC16 captures individuals with very high heterozygosity (Figure S11).
+When restricting to imputed individuals only, i.e.\ after removing individuals with very high heterozygosity, PC16 completely disappears and new PC16 to PC19 correspond to previous PC17 to PC20 (Table S2).
+When using dosage data instead of genotype data with mean imputation, PCA is globally unchanged (Table S3). 
+Overall, if we choose to use either one of the following 4 imputation methods: mean imputation, random imputation according to allele frequencies, using reported dosage data from BGEN files, or imputation of genotyped data based on machine learning using function \texttt{snp\_fastImpute} of R package bigsnpr \cite[]{prive2017efficient}, resulting PCs are always very similar (absolute correlation larger than 0.99 for the 20 computed PCs; results partially shown in tables S2 and S3).
 This justifies performing PCA with mean imputation directly on PLINK bed files with a few missing values; this has the advantage to be much faster than having to impute genotyped data using \texttt{snp\_fastImpute}, which took 4 days for 406,545 individuals and 240,444 variants, or based on external reference datasets.
@@
 
-\begin{figure}[!htpb]
-\centerline{\includegraphics[width=0.4\textwidth]{PCA-pipeline.pdf}}
+\begin{figure}[htb]
+\centerline{\includegraphics[width=0.8\linewidth]{PCA-pipeline.pdf}}
 \caption{Proposed pipeline for computing Principal Components (PCs) using R packages bigsnpr and bigutilsr.
@@
 Note that one should remove both individuals in each pair of related individuals.
-This ensures that the projected individuals are not related to the ones used for computing PCA, since we showed that relatedness is a problem when using the OADP projection (Figure \ref{fig:proj1000G-4}).
+This ensures that the projected individuals are not related to the ones used for computing PCA, since we showed that relatedness is a problem when using the OADP projection (Figure S5).
 After selecting a subset of individuals, we apply several steps of outlier detection, one for outlier variants that capture long-range LD variation (automatic), and one for detecting outlier samples (semi-automatic and visual).
@@
 In contrast, long-range LD structure is essentially capturing the variation inside one long-range LD region (so localised in the genome), so that we expect the loadings to be very large in that region only (one peak).
-Therefore, PCs capturing LD structure can be identified by looking for peaks in PC loadings (e.g.\ PC17-20 in figure \ref{fig:UKBB-loadings}). 
+Therefore, PCs capturing LD structure can be identified by looking for peaks in PC loadings (e.g.\ PC17-20 in figure S10). 
 To identify which PCs capture population structure, and which ones are probably just noise, one should also look at PC scores (colored by ancestry if possible).
@@
 As in many applications, we believe a compromise between signal and noise should be preferred.
-Therefore, we recommend using only PCs that show structure (e.g.\ PC1-16 in figure \ref{fig:UKBB-scores}) and excluding PCs that do not seem to capture any population structure (e.g.\ PC17-20 in figure \ref{fig:UKBB-scores}).
+Therefore, we recommend using only PCs that show structure (e.g.\ PC1-16 in figure S9) and excluding PCs that do not seem to capture any population structure (e.g.\ PC17-20 in figure S9).
 
@@
 
-%\clearpage
-\vspace*{5em}
+\section*{Funding}
 
-\section*{Software and code availability}
-
-All code used for this paper is available at \url{https://github.com/privefl/paper4-bedpca/tree/master/code}.
-R packages bigsnpr and bigutilsr can be installed from either CRAN or GitHub (see \url{https://github.com/privefl/bigsnpr}).
-A tutorial on the steps to perform PCA on 1000G data is available at \url{https://privefl.github.io/bigsnpr/articles/bedpca.html}.
+F.P., J.M.\ and B.V.\ are supported by the Danish National Research Foundation (Niels Bohr Professorship to J.M.), and also acknowledge the Lundbeck Foundation Initiative for Integrative Psychiatric Research, iPSYCH (R248-2017-2003).
 
@@
 
-F.P., J.M.\ and B.V.\ are supported by the Danish National Research Foundation (Niels Bohr Professorship to J.M.), and also acknowledge the Lundbeck Foundation Initiative for Integrative Psychiatric Research, iPSYCH (R248-2017-2003).
+\vspace*{1em}
 
-\section*{Declaration of Interests}
-
-Michael Blum is now an employee of OWKIN France.
+\noindent\textit{Conflict of Interest}: Michael Blum is now an employee of OWKIN France.
 The other authors declare no competing interests.
@@
 
-\clearpage
+%\clearpage
 
-\bibliographystyle{natbib}
-\bibliography{refs}
+%\bibliographystyle{natbib}
+%\bibliography{refs}
+
+\begin{thebibliography}{}
+
+	\bibitem[{1000 Genomes Project Consortium} {\em et~al.}(2015){1000 Genomes
+		Project Consortium} {\em et~al.}]{10002015global}
+	{1000 Genomes Project Consortium} {\em et~al.} (2015).
+	\newblock A global reference for human genetic variation.
+	\newblock {\em Nature\/}, {\bf 526}(7571), 68.
+
+	\bibitem[Abdellaoui {\em et~al.}(2013)Abdellaoui, Hottenga, De~Knijff, Nivard,
+	Xiao, Scheet, Brooks, Ehli, Hu, Davies, {\em
+		et~al.}]{abdellaoui2013population}
+	Abdellaoui, A., Hottenga, J.-J., De~Knijff, P., Nivard, M.~G., Xiao, X.,
+	Scheet, P., Brooks, A., Ehli, E.~A., Hu, Y., Davies, G.~E., {\em et~al.}
+	(2013).
+	\newblock Population structure, migration, and diversifying selection in the
+	netherlands.
+	\newblock {\em European journal of human genetics\/}, {\bf 21}(11), 1277.
+
+	\bibitem[Abraham {\em et~al.}(2017)Abraham, Qiu, and
+	Inouye]{abraham2017flashpca2}
+	Abraham, G., Qiu, Y., and Inouye, M. (2017).
+	\newblock {FlashPCA2}: principal component analysis of biobank-scale genotype
+	datasets.
+	\newblock {\em Bioinformatics\/}.
+
+	\bibitem[Agrawal {\em et~al.}(2019)Agrawal, Chiu, Le, Halperin, and
+	Sankararaman]{agrawal2019scalable}
+	Agrawal, A., Chiu, A.~M., Le, M., Halperin, E., and Sankararaman, S. (2019).
+	\newblock Scalable probabilistic pca for large-scale genetic variation data.
+	\newblock {\em bioRxiv\/}, page 729202.
+
+	\bibitem[Bellenguez {\em et~al.}(2011)Bellenguez, Strange, Freeman, {Wellcome
+		Trust Case Control Consortium}, Donnelly, and Spencer]{bellenguez2011robust}
+	Bellenguez, C., Strange, A., Freeman, C., {Wellcome Trust Case Control
+		Consortium}, Donnelly, P., and Spencer, C.~C. (2011).
+	\newblock A robust clustering algorithm for identifying problematic samples in
+	genome-wide association studies.
+	\newblock {\em Bioinformatics\/}, {\bf 28}(1), 134--135.
+
+	\bibitem[Bose {\em et~al.}(2019)Bose, Kalantzis, Kontopoulou, Elkady, Paschou,
+	and Drineas]{bose2019terapca}
+	Bose, A., Kalantzis, V., Kontopoulou, E.-M., Elkady, M., Paschou, P., and
+	Drineas, P. (2019).
+	\newblock {TeraPCA}: a fast and scalable software package to study genetic
+	variation in tera-scale genotypes.
+	\newblock {\em Bioinformatics\/}.
+
+	\bibitem[Brand(2003)Brand]{brand2003fast}
+	Brand, M. (2003).
+	\newblock Fast online svd revisions for lightweight recommender systems.
+	\newblock In {\em Proceedings of the 2003 {SIAM} international conference on
+		data mining\/}, pages 37--46. SIAM.
+
+	\bibitem[Brys {\em et~al.}(2004)Brys, Hubert, and Struyf]{brys2004robust}
+	Brys, G., Hubert, M., and Struyf, A. (2004).
+	\newblock A robust measure of skewness.
+	\newblock {\em Journal of Computational and Graphical Statistics\/}, {\bf
+		13}(4), 996--1017.
+
+	\bibitem[Bycroft {\em et~al.}(2018)Bycroft, Freeman, Petkova, Band, Elliott,
+	Sharp, Motyer, Vukcevic, Delaneau, O'Connell, {\em
+		et~al.}]{bycroft2017genome}
+	Bycroft, C., Freeman, C., Petkova, D., Band, G., Elliott, L.~T., Sharp, K.,
+	Motyer, A., Vukcevic, D., Delaneau, O., O'Connell, J., {\em et~al.} (2018).
+	\newblock The {UK} biobank resource with deep phenotyping and genomic data.
+	\newblock {\em Nature\/}, {\bf 562}(7726), 203.
+
+	\bibitem[Chang {\em et~al.}(2015)Chang, Chow, Tellier, Vattikuti, Purcell, and
+	Lee]{chang2015second}
+	Chang, C.~C., Chow, C.~C., Tellier, L.~C., Vattikuti, S., Purcell, S.~M., and
+	Lee, J.~J. (2015).
+	\newblock Second-generation {PLINK}: rising to the challenge of larger and
+	richer datasets.
+	\newblock {\em Gigascience\/}, {\bf 4}(1), 7.
+
+	\bibitem[Decamps {\em et~al.}(2020)Decamps, Priv{\'e}, Bacher, Jost, Waguet,
+	Houseman, Lurie, Lutsik, Milosavljevic, Scherer, {\em
+		et~al.}]{decamps2019guidelines}
+	Decamps, C., Priv{\'e}, F., Bacher, R., Jost, D., Waguet, A., Houseman, E.~A.,
+	Lurie, E., Lutsik, P., Milosavljevic, A., Scherer, M., {\em et~al.} (2020).
+	\newblock Guidelines for cell-type heterogeneity quantification based on a
+	comparative analysis of reference-free dna methylation deconvolution
+	software.
+	\newblock {\em BMC bioinformatics\/}, {\bf 21}(1), 16.
+
+	\bibitem[Dey and Lee(2019)Dey and Lee]{dey2019asymptotic}
+	Dey, R. and Lee, S. (2019).
+	\newblock Asymptotic properties of principal component analysis and
+	shrinkage-bias adjustment under the generalized spiked population model.
+	\newblock {\em Journal of Multivariate Analysis\/}, {\bf 173}, 145--164.
+
+	\bibitem[Elseberg {\em et~al.}(2012)Elseberg, Magnenat, Siegwart, and
+	N{\"u}chter]{elseberg2012comparison}
+	Elseberg, J., Magnenat, S., Siegwart, R., and N{\"u}chter, A. (2012).
+	\newblock Comparison of nearest-neighbor-search strategies and implementations
+	for efficient shape registration.
+	\newblock {\em Journal of Software Engineering for Robotics\/}, {\bf 3}(1),
+	2--12.
+
+	\bibitem[Galinsky {\em et~al.}(2016)Galinsky, Bhatia, Loh, Georgiev, Mukherjee,
+	Patterson, and Price]{galinsky2016fast}
+	Galinsky, K.~J., Bhatia, G., Loh, P.-R., Georgiev, S., Mukherjee, S.,
+	Patterson, N.~J., and Price, A.~L. (2016).
+	\newblock Fast principal-component analysis reveals convergent evolution of
+	{ADH1B} in europe and east asia.
+	\newblock {\em The American Journal of Human Genetics\/}, {\bf 98}(3),
+	456--472.
+
+	\bibitem[Gnanadesikan and Kettenring(1972)Gnanadesikan and
+	Kettenring]{gnanadesikan1972robust}
+	Gnanadesikan, R. and Kettenring, J.~R. (1972).
+	\newblock Robust estimates, residuals, and outlier detection with multiresponse
+	data.
+	\newblock {\em Biometrics\/}, pages 81--124.
+
+	\bibitem[Hubert and Vandervieren(2008)Hubert and
+	Vandervieren]{hubert2008adjusted}
+	Hubert, M. and Vandervieren, E. (2008).
+	\newblock An adjusted boxplot for skewed distributions.
+	\newblock {\em Computational statistics \& data analysis\/}, {\bf 52}(12),
+	5186--5201.
+
+	\bibitem[{International HapMap 3 Consortium} {\em et~al.}(2010){International
+		HapMap 3 Consortium} {\em et~al.}]{international2010integrating}
+	{International HapMap 3 Consortium} {\em et~al.} (2010).
+	\newblock Integrating common and rare genetic variation in diverse human
+	populations.
+	\newblock {\em Nature\/}, {\bf 467}(7311), 52.
+
+	\bibitem[Kriegel {\em et~al.}(2009)Kriegel, Kr{\"o}ger, Schubert, and
+	Zimek]{kriegel2009loop}
+	Kriegel, H.-P., Kr{\"o}ger, P., Schubert, E., and Zimek, A. (2009).
+	\newblock {LoOP}: local outlier probabilities.
+	\newblock In {\em Proceedings of the 18th {ACM} conference on Information and
+		knowledge management\/}, pages 1649--1652. ACM.
+
+	\bibitem[Lee {\em et~al.}(2010)Lee, Zou, and Wright]{lee2010convergence}
+	Lee, S., Zou, F., and Wright, F.~A. (2010).
+	\newblock Convergence and prediction of principal component scores in
+	high-dimensional settings.
+	\newblock {\em Annals of statistics\/}, {\bf 38}(6), 3605.
+
+	\bibitem[Lehoucq and Sorensen(1996)Lehoucq and Sorensen]{Lehoucq1996}
+	Lehoucq, R.~B. and Sorensen, D.~C. (1996).
+	\newblock {Deflation Techniques for an Implicitly Restarted Arnoldi Iteration}.
+	\newblock {\em SIAM Journal on Matrix Analysis and Applications\/}, {\bf
+		17}(4), 789--821.
+
+	\bibitem[Loh {\em et~al.}(2015a)Loh, Bhatia, Gusev, Finucane, Bulik-Sullivan,
+	Pollack, de~Candia, Lee, Wray, Kendler, {\em et~al.}]{loh2015contrasting}
+	Loh, P.-R., Bhatia, G., Gusev, A., Finucane, H.~K., Bulik-Sullivan, B.~K.,
+	Pollack, S.~J., de~Candia, T.~R., Lee, S.~H., Wray, N.~R., Kendler, K.~S.,
+	{\em et~al.} (2015a).
+	\newblock Contrasting genetic architectures of schizophrenia and other complex
+	diseases using fast variance-components analysis.
+	\newblock {\em Nature genetics\/}, {\bf 47}(12), 1385.
+
+	\bibitem[Loh {\em et~al.}(2015b)Loh, Tucker, Bulik-Sullivan, Vilhjalmsson,
+	Finucane, Salem, Chasman, Ridker, Neale, Berger, {\em
+		et~al.}]{loh2015efficient}
+	Loh, P.-R., Tucker, G., Bulik-Sullivan, B.~K., Vilhjalmsson, B.~J., Finucane,
+	H.~K., Salem, R.~M., Chasman, D.~I., Ridker, P.~M., Neale, B.~M., Berger, B.,
+	{\em et~al.} (2015b).
+	\newblock Efficient bayesian mixed-model analysis increases association power
+	in large cohorts.
+	\newblock {\em Nature genetics\/}, {\bf 47}(3), 284.
+
+	\bibitem[Luu {\em et~al.}(2017)Luu, Bazin, and Blum]{luu2017pcadapt}
+	Luu, K., Bazin, E., and Blum, M. G.~B. (2017).
+	\newblock pcadapt: an {R} package to perform genome scans for selection based
+	on principal component analysis.
+	\newblock {\em Molecular ecology resources\/}, {\bf 17}(1), 67--77.
+
+	\bibitem[Manichaikul {\em et~al.}(2010)Manichaikul, Mychaleckyj, Rich, Daly,
+	Sale, and Chen]{manichaikul2010robust}
+	Manichaikul, A., Mychaleckyj, J.~C., Rich, S.~S., Daly, K., Sale, M., and Chen,
+	W.-M. (2010).
+	\newblock Robust relationship inference in genome-wide association studies.
+	\newblock {\em Bioinformatics\/}, {\bf 26}(22), 2867--2873.
+
+	\bibitem[Maronna and Zamar(2002)Maronna and Zamar]{maronna2002robust}
+	Maronna, R.~A. and Zamar, R.~H. (2002).
+	\newblock Robust estimates of location and dispersion for high-dimensional
+	datasets.
+	\newblock {\em Technometrics\/}, {\bf 44}(4), 307--317.
+
+	\bibitem[McVean(2009)McVean]{mcvean2009genealogical}
+	McVean, G. (2009).
+	\newblock A genealogical interpretation of principal components analysis.
+	\newblock {\em PLoS genetics\/}, {\bf 5}(10), e1000686.
+
+	\bibitem[Meyer(2019)Meyer]{meyer2019genotype}
+	Meyer, H. (2019).
+	\newblock Processing 1000 genomes reference data for ancestry estimation.
+
+	\bibitem[Novembre and Stephens(2008)Novembre and
+	Stephens]{novembre2008interpreting}
+	Novembre, J. and Stephens, M. (2008).
+	\newblock Interpreting principal component analyses of spatial population
+	genetic variation.
+	\newblock {\em Nature genetics\/}, {\bf 40}(5), 646.
+
+	\bibitem[Patterson {\em et~al.}(2006)Patterson, Price, and
+	Reich]{patterson2006population}
+	Patterson, N., Price, A.~L., and Reich, D. (2006).
+	\newblock Population structure and eigenanalysis.
+	\newblock {\em PLoS genetics\/}, {\bf 2}(12), e190.
+
+	\bibitem[Pickrell {\em et~al.}(2010)Pickrell, Marioni, Pai, Degner, Engelhardt,
+	Nkadori, Veyrieras, Stephens, Gilad, and
+	Pritchard]{pickrell2010understanding}
+	Pickrell, J.~K., Marioni, J.~C., Pai, A.~A., Degner, J.~F., Engelhardt, B.~E.,
+	Nkadori, E., Veyrieras, J.-B., Stephens, M., Gilad, Y., and Pritchard, J.~K.
+	(2010).
+	\newblock Understanding mechanisms underlying human gene expression variation
+	with rna sequencing.
+	\newblock {\em Nature\/}, {\bf 464}(7289), 768.
+
+	\bibitem[Price {\em et~al.}(2006)Price, Patterson, Plenge, Weinblatt, Shadick,
+	and Reich]{price2006principal}
+	Price, A.~L., Patterson, N.~J., Plenge, R.~M., Weinblatt, M.~E., Shadick,
+	N.~A., and Reich, D. (2006).
+	\newblock Principal components analysis corrects for stratification in
+	genome-wide association studies.
+	\newblock {\em Nature genetics\/}, {\bf 38}(8), 904.
+
+	\bibitem[Price {\em et~al.}(2008)Price, Weale, Patterson, Myers, Need, Shianna,
+	Ge, Rotter, Torres, Taylor, {\em et~al.}]{price2008long}
+	Price, A.~L., Weale, M.~E., Patterson, N., Myers, S.~R., Need, A.~C., Shianna,
+	K.~V., Ge, D., Rotter, J.~I., Torres, E., Taylor, K.~D., {\em et~al.} (2008).
+	\newblock Long-range {LD} can confound genome scans in admixed populations.
+	\newblock {\em The American Journal of Human Genetics\/}, {\bf 83}(1),
+	132--135.
+
+	\bibitem[Price {\em et~al.}(2010)Price, Zaitlen, Reich, and
+	Patterson]{price2010new}
+	Price, A.~L., Zaitlen, N.~A., Reich, D., and Patterson, N. (2010).
+	\newblock New approaches to population stratification in genome-wide
+	association studies.
+	\newblock {\em Nature Reviews Genetics\/}, {\bf 11}(7), 459.
+
+	\bibitem[Priv{\'e} {\em et~al.}(2018)Priv{\'e}, Aschard, Ziyatdinov, and
+	Blum]{prive2017efficient}
+	Priv{\'e}, F., Aschard, H., Ziyatdinov, A., and Blum, M. G.~B. (2018).
+	\newblock Efficient analysis of large-scale genome-wide data with two {R}
+	packages: bigstatsr and bigsnpr.
+	\newblock {\em Bioinformatics\/}, {\bf 34}(16), 2781--2787.
+
+	\bibitem[Priv{\'e} {\em et~al.}(2019)Priv{\'e}, Aschard, and
+	Blum]{prive2019efficient}
+	Priv{\'e}, F., Aschard, H., and Blum, M. G.~B. (2019).
+	\newblock Efficient implementation of penalized regression for genetic risk
+	prediction.
+	\newblock {\em Genetics\/}, {\bf 212}(1), 65--74.
+
+	\bibitem[Priv{\'e} {\em et~al.}(2020)Priv{\'e}, Luu, Vilhj{\'a}lmsson, and
+	Blum]{prive2020performing}
+	Priv{\'e}, F., Luu, K., Vilhj{\'a}lmsson, B.~J., and Blum, M. G.~B. (2020).
+	\newblock Performing highly efficient genome scans for local adaptation with
+	{R} package pcadapt version 4.
+	\newblock {\em Molecular Biology and Evolution\/}.
+
+	\bibitem[Tukey(1977)Tukey]{tukey77}
+	Tukey, J.~W. (1977).
+	\newblock {\em Exploratory Data Analysis\/}.
+	\newblock Addison-Wesley.
+
+	\bibitem[Wang {\em et~al.}(2015)Wang, Zhan, Liang, Abecasis, and
+	Lin]{wang2015improved}
+	Wang, C., Zhan, X., Liang, L., Abecasis, G.~R., and Lin, X. (2015).
+	\newblock Improved ancestry estimation for both genotyping and sequencing data
+	using projection procrustes analysis and genotype imputation.
+	\newblock {\em The American Journal of Human Genetics\/}, {\bf 96}(6),
+	926--937.
+
+	\bibitem[Wu {\em et~al.}(2017)Wu, Romero, and Stathopoulos]{wu2017primme}
+	Wu, L., Romero, E., and Stathopoulos, A. (2017).
+	\newblock Primme\_svds: A high-performance preconditioned svd solver for
+	accurate large-scale computations.
+	\newblock {\em {SIAM} Journal on Scientific Computing\/}, {\bf 39}(5),
+	S248--S271.
+
+	\bibitem[Yang {\em et~al.}(2010)Yang, Benyamin, McEvoy, Gordon, Henders,
+	Nyholt, Madden, Heath, Martin, Montgomery, {\em et~al.}]{yang2010common}
+	Yang, J., Benyamin, B., McEvoy, B.~P., Gordon, S., Henders, A.~K., Nyholt,
+	D.~R., Madden, P.~A., Heath, A.~C., Martin, N.~G., Montgomery, G.~W., {\em
+		et~al.} (2010).
+	\newblock Common snps explain a large proportion of the heritability for human
+	height.
+	\newblock {\em Nature genetics\/}, {\bf 42}(7), 565.
+
+	\bibitem[Yohai and Zamar(1988)Yohai and Zamar]{yohai1988high}
+	Yohai, V.~J. and Zamar, R.~H. (1988).
+	\newblock High breakdown-point estimates of regression by means of the
+	minimization of an efficient scale.
+	\newblock {\em Journal of the American statistical association\/}, {\bf
+		83}(402), 406--413.
+
+	\bibitem[Zhang {\em et~al.}(2020)Zhang, Dey, and Lee]{zhang2019fast}
+	Zhang, D., Dey, R., and Lee, S. (2020).
+	\newblock {Fast and robust ancestry prediction using principal component
+		analysis}.
+	\newblock {\em Bioinformatics\/}.
+	\newblock btaa152.
+
+	\bibitem[Zou {\em et~al.}(2010)Zou, Lee, Knowles, and
+	Wright]{zou2010quantification}
+	Zou, F., Lee, S., Knowles, M.~R., and Wright, F.~A. (2010).
+	\newblock Quantification of population structure using correlated snps by
+	shrinkage principal components.
+	\newblock {\em Human heredity\/}, {\bf 70}(1), 9--22.
+
+\end{thebibliography}
 
